### 2015-06-10 writeup/parkinsons.tex
Add writeup latex draft for data preprocessing

\documentclass[letter,12pt]{article}
\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{tablefootnote}
\pagestyle{fancy}
\lhead{Jesse Mu}
\rhead{CSCI339 Term Project}
\graphicspath{ {../figures/} }

\begin{document}

\title{Cluster Analysis: Identifying Parkinson's Disease Subtypes}
\date{Wednesday, June 10}
\author{Jesse Mu}
\maketitle

\section{Preprocessing}

\subsection{Dataset Description}
951 subjects, 145 metrics, collected 15-4-2012. From Pablo Martinez Mart\'in.
170 subjects with missing values (brought down to 781); these were removed
automatically, even if the missing values were not included in the selected
features below. This will need to be changed later on, by keeping those
removed that still have all selceted features and perhaps with some
compensation for missing values.

\subsection{Selected Features}

Combination of non-motor scale (NMS) symptoms and standard motor symptoms.

\begin{table}[h]
  \centering
  \begin{tabular}{l|l|l|l}
    Name & Type & Format & Description \\
    \hline
    nms\_d1 & byte & \%8.0g & cardiovascular \\
    nms\_d2 & byte & \%8.0g & sleep/fatigue \\
    nms\_d3 & byte & \%8.0g & mood/cognition \\
    nms\_d4 & byte & \%8.0g & percep/hallucinations \\
    nms\_d5 & byte & \%8.0g & attention/memory \\
    nms\_d6 & byte & \%8.0g & gastrointestinal \\
    nms\_d7 & byte & \%8.0g & urinary \\
    nms\_d8 & byte & \%8.0g & sexual function \\
    nms\_d9 & byte & \%8.0g & miscellaneous \\
    tremor & float & \%9.0g & tremor \\
    bradykin & float & \%9.0g & bradykinesia\tablefootnote{Impaired ability to
    adjust the body's position.} \\
    rigidity & float & \%9.0g & rigidity \\
    axial & float & \%9.0g & axial\tablefootnote{Issues affecting the middle of
    the body.} \\
    pigd & float & \%9.0g & postural instability and gait difficulty \\
  \end{tabular}
  \caption{Selected Features and Details}
  \label{tab:selcted_features}
\end{table}

\begin{table}[h]
  \centering
  \begin{tabular}{l|l|l|l}
  Name  &       $\mu$ & $\sigma$ & min-max \\
         \hline
nms\_d1&   1.76&  3.32&   0-24 \\
nms\_d2&   8.71&  8.76&   0-48 \\
nms\_d3&   8.70& 11.83&   0-60 \\
nms\_d4&   1.65&  3.94&   0-33 \\
nms\_d5&   5.22&  7.44&   0-36 \\
nms\_d6&   5.67&  6.92&   0-36 \\
nms\_d7&   8.02&  9.09&   0-36 \\
nms\_d8&   3.57&  5.97&   0-24 \\
nms\_d9&   6.99&  7.74&   0-48 \\
tremor&   2.59&  2.63&   0-12 \\
bradykin& 2.49&  1.39&   0-6 \\
rigidity& 2.34&  1.36&   0-6 \\
axial&    3.28&  2.75&   0-12 \\
pigd&     3.36&  2.77&   0-12 \\
  \end{tabular}
  \caption{Descriptive Statistics}
  \label{tab:descriptive_statistics}
\end{table}

\subsection{Dimensionality Reduction: PCA}

May not be useful? If we're trying to identify \emph{clinically} relevant
features, merging them may not be a good idea.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\linewidth]{pca-eigenvalues.png}
  \caption{Scree test: eigenvalues by factor}
  \label{fig:pca-eigenvalues}
\end{figure}

Figure~\ref{fig:pca-eigenvalues} shows scree test elbow occurs around 2 or 3. Also, eigenvalues 1 and $2 > 1$, while 3
is around .9

\section{$k$-means}
\subsection{Identifying optimal number of clusters}

\begin{figure}[h]
  \centering
  \includegraphics[width=0.8\linewidth]{kmeans-wss-error.png}
  \caption{Scree test: WSS error by cluster size}
  \label{fig:kmeans-wss-error}
\end{figure}

Figure~\ref{fig:kmeans-wss-error} shows no optimal elbow in scree test! Maybe 2-3?

% TODO: More optimal cluster determination (PAMK, NbClust)

\subsection{Decision Tree classifier based on clusters}

% Pruned, non-pruned tree

\section{Biclustering}

\section{Subspace clustering}

% TODO: Wat.

\section{Bayesian Networks}

\end{document}

### 2015-06-10 R/preprocessing.R

@@ -2,6 +2,7 @@
 library(plyr)
 # For PCA
 library(FactoMineR)
+library(psych)
 
 # ==== GLOBAL CONSTANTS ====
 DB_FILE = "../data/DATABASE_NMS Burden levels_15-4-2012.csv"
@@ -26,7 +27,7 @@ MOTOR.SYMPTOMS = c("tremor",
                    "pigd")
 # Change this variable to change symptoms
 SYMPTOMS.TO.USE = ALL.SYMPTOMS
-  
+
 # ==== UTILITY FUNCTIONS ====
 splitdf <- function(dataframe, seed=NULL, trainfrac=0.7) {
   if (trainfrac<=0 | trainfrac>=1) stop("Training fraction must be between 0 and 1, not inclusive")
@@ -48,7 +49,12 @@ raw.omitted <- na.omit(raw)
 # raw with only select nms and motor symptoms
 raw.filtered <- raw.omitted[, SYMPTOMS.TO.USE]
 
-# Standardized 0 mean, 1 stddev
+# ==== DESCRIPTIVE STATISTICS (BEFORE STANDARDIZATION) ====
+# TODO - change raw.filtered to raw.omitted.filtered, or something like that
+raw.omitted.stats <- describe(raw.omitted)
+raw.filtered.stats <- describe(raw.filtered)
+
+# ==== STANDARDIZATION ====
 raw.filtered <- as.data.frame(scale(raw.filtered))
 
 # ==== PCA (Note: not helpful for identifying specific factors) ====
@@ -57,5 +63,8 @@ pca <- PCA(raw.filtered)
 # pca$eig
 # pca$var$coord
 # head(pca$ind$coord)
-# plot(x = 1:length(rownames(pca$eig)), y = (pca$eig$eigenvalue), type="b")
+plot(x = 1:length(rownames(pca$eig)), y = (pca$eig$eigenvalue), type="b",
+     xlab="Factor", ylab="Eigenvalue", xaxp=c(0, 14, 14))
 # Elbow appears around 3 factors
+# Reset par for the rest of the script
+par(mfrow = c(1, 1))

### 2015-06-10 R/kmeans-dtree.R
add nbclust/pamk cluster estimation

@@ -2,6 +2,8 @@
 library(cluster)
 library(rpart)
 library(plyr)
+library(fpc)
+library(NbClust)
 
 # ==== LOAD DATA ====
 source('./preprocessing.R')
@@ -14,9 +16,21 @@ for (i in 2:15) {
   wss[i] <- sum(kmeans(raw.filtered, i)$withinss)
 }
 
-plot(1:15, wss, type="b", xlab="Number of Clusters", ylab="Within groups sum of squares")
+plot(1:15, wss, type="b",
+     xlab="Number of Clusters", ylab="Within groups sum of squares")
 
-# ==== TODO: OTHER WAYS TO FIND OPTIMAL K ====
+# ==== NBCLUST ESTIMATION FOR OPTIMAL K (30 metrics) ====
+# This is taking a long time
+# nb <- NbClust(raw.filtered, distance = "euclidean",
+#               min.nc=2, max.nc=15, method = "kmeans",
+#               index = "alllong", alphaBeale = 0.1)
+# hist(nb$Best.nc[1,], breaks = max(na.omit(nb$Best.nc[1,])))
+
+# ==== PAM ESTIMATION FOR OPTIMAL K ====
+# Estimates 2
+pamk.best <- pamk(raw.filtered)
+cat("number of clusters estimated by optimum average silhouette width:", pamk.best$nc, "\n")
+plot(pam(raw.filtered, pamk.best$nc), main="Optimal clustering")
 
 # ==== INITIAL KMEANS CLUSTERING ====
 splits = splitdf(raw.filtered)

### 2015-06-11 writeup/parkinsons.tex
add extra cluster estimation methods

@@
-\documentclass[letter,12pt]{article}
+\documentclass[letterpaper,12pt]{article}
 \usepackage{fancyhdr}
@@
 \usepackage{tablefootnote}
-\pagestyle{fancy}
-\lhead{Jesse Mu}
-\rhead{CSCI339 Term Project}
+% \pagestyle{fancy}
+% \lhead{Jesse Mu}
+% \rhead{CSCI339 Term Project}
 \graphicspath{ {../figures/} }
@@
 
-\begin{table}[h]
+\begin{table}[ht]
   \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
   \centering
@@
 
-\begin{figure}[h]
+\subsubsection{WSS Error Scree Test}
+
+\begin{figure}[ht]
   \centering
@@
 
-% TODO: More optimal cluster determination (PAMK, NbClust)
+\subsubsection{Gap Statistic}
+
+Optimal cluster is the local maximum of the gap statistic, but it appears to be
+consistently increasing in Figure~\ref{fig:gap-statistic}.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{gap-statistic.png}
+  \caption{Gap statistic by cluster size}
+  \label{fig:gap-statistic}
+\end{figure}
+
+\subsubsection{Average Silhouette Width}
+
+Figure~\ref{fig:asw} shows average silhouette width as being consistently under
+0.25 for all clusters, implying the data is not well structured.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{asw.png}
+  \caption{Average silhouette width by cluster size}
+  \label{fig:asw}
+\end{figure}
+
+% \subsubsection{\texttt{NbClust} package}
 
@@
 
-% Pruned, non-pruned tree
+% TODO: Use compressed function to plot decision trees based on clusters with
+% 2, 3, 4 groups.
+% See if there are any common factors here!
 

### 2015-06-11 R/kmeans-dtree.R

@@ -19,6 +19,20 @@ for (i in 2:15) {
 plot(1:15, wss, type="b",
      xlab="Number of Clusters", ylab="Within groups sum of squares")
 
+# ==== TODO: BIC ====
+# See http://stackoverflow.com/questions/15376075/cluster-analysis-in-r-determine-the-optimal-number-of-clusters
+library(mclust)
+# NOW - does this have to do with EM/hierarchical models. not cluster-based?
+# Run the function to see how many clusters
+# it finds to be optimal, set it to search for
+# at least 1 model and up 20.
+# Traditionally this takes a while, so it's uncommented
+# d_clust <- Mclust(as.matrix(raw.filtered), G=1:20)
+# m.best <- dim(d_clust$z)[2]
+# cat("model-based optimal number of clusters:", m.best, "\n")
+# # 3 clusters
+# plot(d_clust)
+
 # ==== NBCLUST ESTIMATION FOR OPTIMAL K (30 metrics) ====
 # This is taking a long time
 # nb <- NbClust(raw.filtered, distance = "euclidean",
@@ -28,9 +42,15 @@ plot(1:15, wss, type="b",
 
 # ==== PAM ESTIMATION FOR OPTIMAL K ====
 # Estimates 2
-pamk.best <- pamk(raw.filtered)
-cat("number of clusters estimated by optimum average silhouette width:", pamk.best$nc, "\n")
-plot(pam(raw.filtered, pamk.best$nc), main="Optimal clustering")
+pam_sils <- c()
+for (i in 1:15) {
+  pam_sils <- c(pam_sils, pam(raw.filtered, k=i)$silinfo$avg.width)
+}
+plot(x=1:14, y=pam_sils, xlab="Clusters", ylab="Average Silhouette Width", type="b")
+
+# ==== GAP STATISTIC ESTIMATION ====
+gaps <- clusGap(raw.filtered, kmeans, 14, B = 100)
+plot(x=1:14, y=gaps$Tab[, "gap"], xlab="Clusters", ylab="Gap Statitsic", type="b")
 
 # ==== INITIAL KMEANS CLUSTERING ====
 splits = splitdf(raw.filtered)

### 2015-06-11 R/kmeans-dtree.R
add kmeans compressed function for iteration
to be used later to obtain statistics and decision tree plots of
varying numbers of k, since k hasn't been determined quite yet!

@@ -60,7 +60,8 @@ str(splits)
 splits$trainset = raw.filtered
 
 # FIXME: How are we identifying k = 4? (Arbitrary right now)
-cl <- kmeans(splits$trainset, 4, nstart = 25)
+K = 3
+cl <- kmeans(splits$trainset, K, nstart = 25)
 
 # Add cluster label to original data, rename
 trainset.labeled <- cbind(splits$trainset, cl$cluster)
@@ -93,6 +94,43 @@ pruned.t <- prune(t, cp = cp)
 plot(pruned.t, uniform=TRUE, main="Decision Tree (Pruned)")
 text(pruned.t, use.n=TRUE, all=TRUE, cex=.8, pos=1)
 # stopping point - pruned tree
+
+# ==== KMEANS CLUSTERING COMPRESSED FUNCTION ====
+# Use for iteration!
+kmeans_dtree <- function(data, k, save = FALSE) {
+  cl <- kmeans(splits$trainset, K, nstart = 25)
+  labeled_data <- cbind(data, cl$cluster)
+  # Add cluster label to original data, rename
+  labeled_data <- rename(labeled_data, c("cl$cluster"="cluster"))
+  # Convert to factor
+  labeled_data$cluster <- as.factor(labeled_data$cluster)
+  t <- rpart(cluster ~ ., trainset.labeled)
+  # Print error statistics
+  printcp(t)
+  # Find resubstitution rate
+  pred.t <- table(predict(t, type="class"), trainset.labeled$cluster)
+  resub <- 1-sum(diag(pred.t))/sum(pred.t)
+
+  # Prune, find minimum cp
+  cp <- t$cptable[which.min(t$cptable[,"xerror"]),"CP"]
+  pruned.t <- prune(t, cp = cp)
+
+  if (save) {
+    # TODO: Find a way to save the plots!!
+  }
+  plot(t, uniform=TRUE, main="Decision Tree")
+  text(t, use.n=TRUE, all=TRUE, cex=.8)
+
+  plot(pruned.t, uniform=TRUE, main="Decision Tree (Pruned)")
+  text(pruned.t, use.n=TRUE, all=TRUE, cex=.8)
+
+  list(
+    "data" = labeled_data,
+    "unpruned.tree" = t,
+    "pruned.tree" = pruned.t,
+    "resub.error" = resub
+  )
+}
 
 # ==== NOTES ====
 # Note - with clustering, indeed the main feature (root of the tree) doesn't carry much

### 2015-06-12 R/kmeans-dtree.R
Add functionality for saving dtrees TODO: Convert other figures into pngs (better for pdflatex!)

@@ -1,6 +1,7 @@
 # ==== LOAD LIBRARIES ====
 library(cluster)
 library(rpart)
+library(rpart.plot)
 library(plyr)
 library(fpc)
 library(NbClust)
@@ -60,8 +61,8 @@ str(splits)
 splits$trainset = raw.filtered
 
 # FIXME: How are we identifying k = 4? (Arbitrary right now)
-K = 3
-cl <- kmeans(splits$trainset, K, nstart = 25)
+# Arbitrary 4 choice - variable is later
+cl <- kmeans(splits$trainset, 4, nstart = 25)
 
 # Add cluster label to original data, rename
 trainset.labeled <- cbind(splits$trainset, cl$cluster)
@@ -98,38 +99,83 @@ text(pruned.t, use.n=TRUE, all=TRUE, cex=.8, pos=1)
 # ==== KMEANS CLUSTERING COMPRESSED FUNCTION ====
 # Use for iteration!
 kmeans_dtree <- function(data, k, save = FALSE) {
-  cl <- kmeans(splits$trainset, K, nstart = 25)
+  cl <- kmeans(splits$trainset, k, nstart = 25)
   labeled_data <- cbind(data, cl$cluster)
   # Add cluster label to original data, rename
   labeled_data <- rename(labeled_data, c("cl$cluster"="cluster"))
   # Convert to factor
   labeled_data$cluster <- as.factor(labeled_data$cluster)
   t <- rpart(cluster ~ ., trainset.labeled)
-  # Print error statistics
-  printcp(t)
-  # Find resubstitution rate
-  pred.t <- table(predict(t, type="class"), trainset.labeled$cluster)
-  resub <- 1-sum(diag(pred.t))/sum(pred.t)
 
-  # Prune, find minimum cp
+  # Find 10-fold CV error rate for UNPRUNED tree
+  # Old resub rate method
+  # pred.t <- table(predict(t, type="class"), trainset.labeled$cluster)
+  # resub <- 1-sum(diag(pred.t))/sum(pred.t)
+
+  # From printcp source, this is how root node error is calculated
+  # cat("Root node error: ", format(t$frame$dev[1L], digits = digits),
+  #     "/", t$frame$n[1L], " = ", format(t$frame$dev[1L]/t$frame$n[1L],
+  #                                     digits = digits), "\n\n", sep = "")
[…]

### 2015-06-12 R/kmeans-dtree.R
option to print global tree stats

@@ -178,6 +178,17 @@ for (i in 2:4) {
   trees[[istr]] <-  kmeans_dtree(raw.filtered, i, save = TRUE)
 }
 
+# ==== PRINT GLOBAL TREE STATS ====
+for (i in 2:4) {
+  istr <- paste("clusters", i, sep="")
+  t <- trees[[istr]]$pruned.tree
+  cat("CLUSTERS: ", i, "\n", sep="")
+  cat("================================\n")
+  cat("Root node error: ", t$frame$dev[1L] / t$frame$n[1L], "\n", sep="")
+  cat("Complexity Parameter: ", tail(t$cptable[, "CP"], n=1), "\n", sep="")
+  cat("10-fold CV error: ", trees[[istr]]$xv.error.pruned, "\n", sep="")
+}
+
 # ==== NOTES ====
 # Note - with clustering, indeed the main feature (root of the tree) doesn't carry much
 # information

### 2015-06-12 R/kmeans-dtree.R
Fixed incorrect dtree classification bug
Would constantly classify on trainset.labeled wihch was a global
variable defined outside of the compressed function, rather than the
renamed labeled_data

@@ -105,7 +105,7 @@ kmeans_dtree <- function(data, k, save = FALSE) {
   labeled_data <- rename(labeled_data, c("cl$cluster"="cluster"))
   # Convert to factor
   labeled_data$cluster <- as.factor(labeled_data$cluster)
-  t <- rpart(cluster ~ ., trainset.labeled)
+  t <- rpart(cluster ~ ., labeled_data)
 
   # Find 10-fold CV error rate for UNPRUNED tree
   # Old resub rate method

### 2015-06-12 writeup/parkinsons.tex
add dtree statistics tables and figures

@@
 \usepackage{graphicx}
+% Random packages from
+% http://tex.stackexchange.com/questions/50070/landscape-figure-in-latex
+% Necessary for sideways pictures
+\usepackage{wrapfig}
+\usepackage{lscape}
+\usepackage{rotating}
+\usepackage{epstopdf}
 \usepackage{tablefootnote}
@@
 features below. This will need to be changed later on, by keeping those
-removed that still have all selceted features and perhaps with some
+removed that still have all selected features and perhaps with some
 compensation for missing values.
@@
   \caption{Selected Features and Details}
-  \label{tab:selcted_features}
+  \label{tab:selected-features}
 \end{table}
@@
   \caption{Descriptive Statistics}
-  \label{tab:descriptive_statistics}
+  \label{tab:descriptive-statistics}
 \end{table}
@@
 
-% TODO: Use compressed function to plot decision trees based on clusters with
-% 2, 3, 4 groups.
-% See if there are any common factors here!
+% CLUSTERS: 2
+% ================================
+% Root node error: 0.2573624
+% Complexity Parameter: 0.01741294
+% 10-fold CV error: 0.1241997
+% CLUSTERS: 3
+% ================================
+% Root node error: 0.4571063
+% Complexity Parameter: 0.0140056
+% 10-fold CV error: 0.1766965
+% CLUSTERS: 4
+% ================================
+% Root node error: 0.5236876
+% Complexity Parameter: 0.01222494
+% 10-fold CV error: 0.2535211
+\begin{table}[ht]
+  \centering
+  \begin{tabular}{l|l|l|l|l|l}
+    K & CP\footnote{Complexity Parameter} & CV Xerror\footnote{10-fold cross
+    validation} & Root Feature &
+    Root Error & Figure \\
+    \hline
+    2 & 0.017 & 0.124 & & 0.257 & Figure~\ref{fig:kmeans-dtree-2} \\
+    3 & 0.014 & 0.177 & & 0.457 & Figure~\ref{fig:kmeans-dtree-3} \\
+    4 & 0.012 & 0.254 & & 0.524 & Figure~\ref{fig:kmeans-dtree-4} \\
+  \end{tabular}
+  \caption{$k$-kmeans decision trees statistics}
+  \label{tab:k-means-dtrees}
+\end{table}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-2.pdf}
+  \caption{Decision Tree from $k$-means clustering, 2 clusters}
+  \label{fig:kmeans-dtree-2}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-3.pdf}
+  \caption{Decision Tree from $k$-means clustering, 3 clusters}
+  \label{fig:kmeans-dtree-3}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-4.pdf}
+  \caption{Decision Tree from $k$-means clustering, 4 clusters}
+  \label{fig:kmeans-dtree-4}
+\end{figure}
 

### 2015-06-12 writeup/parkinsons.tex
tablefootnotes in table 3

@@
   \begin{tabular}{l|l|l|l|l|l}
-    K & CP\footnote{Complexity Parameter} & CV Xerror\footnote{10-fold cross
+    K & CP\tablefootnote{Complexity Parameter} & CV Xerror\tablefootnote{10-fold cross
     validation} & Root Feature &

### 2015-06-12 R/kmeans-dtree.R
add clustering information and reproducibility

@@ -98,7 +98,9 @@ text(pruned.t, use.n=TRUE, all=TRUE, cex=.8, pos=1)
 
 # ==== KMEANS CLUSTERING COMPRESSED FUNCTION ====
 # Use for iteration!
-kmeans_dtree <- function(data, k, save = FALSE) {
+kmeans_dtree <- function(data, k, save = FALSE, seed = 911) {
+  # Reproducibility!
+  set.seed(seed)
   cl <- kmeans(splits$trainset, k, nstart = 25)
   labeled_data <- cbind(data, cl$cluster)
   # Add cluster label to original data, rename
@@ -162,6 +164,7 @@ kmeans_dtree <- function(data, k, save = FALSE) {
 
   list(
     "data" = labeled_data,
+    "clustering" = cl,
     "unpruned.tree" = t,
     # These are the same, you idiot.
     "xv.error.unpruned" = xv.error.unpruned,
@@ -184,11 +187,20 @@ for (i in 2:4) {
   t <- trees[[istr]]$pruned.tree
   cat("CLUSTERS: ", i, "\n", sep="")
   cat("================================\n")
-  cat("Root node error: ", t$frame$dev[1L] / t$frame$n[1L], "\n", sep="")
   cat("Complexity Parameter: ", tail(t$cptable[, "CP"], n=1), "\n", sep="")
   cat("10-fold CV error: ", trees[[istr]]$xv.error.pruned, "\n", sep="")
+  cat("Root node error: ", t$frame$dev[1L] / t$frame$n[1L], "\n", sep="")
 }
 
+# ==== PRINT CLUSTER STATS ====
+for (i in 2:4) {
+  istr <- paste("clusters", i, sep="")
+  cl <- trees[[istr]]$clustering
+  cat("CLUSTERS: ", i, "\n", sep="")
+  cat("================================\n")
+  cat("WithinSS:", cl$withinss, "\n", sep=" ")
+  cat("Sizes:", cl$size, "\n", sep=" ")
+}
 # ==== NOTES ====
 # Note - with clustering, indeed the main feature (root of the tree) doesn't carry much
 # information

### 2015-06-12 writeup/parkinsons.tex
add updated cluster stats for 911 seed

--- 2015-06-12 writeup/parkinsons.tex
@@
 \usepackage{tablefootnote}
+% for word wrap verbatim
+\usepackage{listings}
+\lstset{
+   breaklines=true,
+   basicstyle=\ttfamily}
 % \pagestyle{fancy}
@@
 
-\subsection{Decision Tree classifier based on clusters}
-
+\subsection{Cluster statistics}
 % CLUSTERS: 2
 % ================================
-% Root node error: 0.2573624
-% Complexity Parameter: 0.01741294
-% 10-fold CV error: 0.1241997
+% Sizes: 201 580
+% WithinSS: 4248.585 4132.434
+% Sum WithinSS: 8381.019
 % CLUSTERS: 3
 % ================================
-% Root node error: 0.4571063
-% Complexity Parameter: 0.0140056
-% 10-fold CV error: 0.1766965
+% Sizes: 420 231 130
+% WithinSS: 2618.368 1973.82 3076.542
+% Sum WithinSS: 7668.73
 % CLUSTERS: 4
 % ================================
+% Sizes: 61 372 145 203
+% WithinSS: 1481.25 1845.389 2147.988 1609.555
+% Sum WithinSS: 7084.183
+\begin{table}[ht]
+  \centering
+  \begin{tabular}{l|l|l|l}
+    $k$ & $n$ & Within SS & sum(Within SS) \\
+    \hline
+    2 & 201/580 & 4248.585/4132.434 & 8381.019 \\
+    3 & 420/231/130 & 2618.368/1973.82/3076.542 &  7668.73 \\
+    4 & 61/372/145/203 & 1481.25/1845.389/2147.988/1609.555 & 7084.183
+  \end{tabular}
+  \caption{Cluster statistics}
+  \label{tab:cluster-statistics}
+\end{table}
+
+\subsection{Centers}
+Omitted; too much information.
+\iffalse
+\begin{lstlisting}
+CLUSTERS: 2
+================================
+> 1   nms_d1    nms_d2    nms_d3    nms_d4    nms_d5    nms_d6    nms_d7    nms_d8    nms_d9    tremor  bradykin
+0.7328282 0.9345720 0.9810287 0.8195052 0.7908599 0.8551764 0.8493069 0.6376458 0.6289463 0.2008254 0.8019272
+ rigidity     axial      pigd
+0.6840038 1.0834910 1.0634082
+> 2    nms_d1     nms_d2     nms_d3     nms_d4     nms_d5     nms_d6     nms_d7     nms_d8     nms_d9     tremor
+-0.2539629 -0.3238776 -0.3399772 -0.2840009 -0.2740739 -0.2963629 -0.2943288 -0.2209772 -0.2179624 -0.0695964
+  bradykin   rigidity      axial       pigd
+-0.2779093 -0.2370427 -0.3754857 -0.3685259
+CLUSTERS: 3
+================================
+> 1    nms_d1     nms_d2     nms_d3     nms_d4     nms_d5     nms_d6     nms_d7     nms_d8     nms_d9     tremor
+-0.2699345 -0.3571672 -0.3574942 -0.2776501 -0.2579928 -0.3084614 -0.3030016 -0.2270260 -0.1867338 -0.2531402
+  bradykin   rigidity      axial       pigd
+-0.6091393 -0.5542033 -0.5792769 -0.5775312
+> 2     nms_d1      nms_d2      nms_d3      nms_d4      nms_d5      nms_d6      nms_d7      nms_d8      nms_d9
+-0.13369057 -0.07938936 -0.05832302 -0.22742096 -0.18981647 -0.02540265 -0.15964421 -0.08973885 -0.16993231
+     tremor    bradykin    rigidity       axial        pigd
+ 0.39256552  0.69210577  0.63956741  0.43974428  0.44762184
+> 3   nms_d1    nms_d2    nms_d3    nms_d4    nms_d5    nms_d6    nms_d7    nms_d8    nms_d9    tremor  bradykin
+1.1096539 1.2949935 1.2586166 1.3011330 1.1708044 1.0417061 1.2626038 0.8929277 0.9052504 0.1202787 0.7381697
+ rigidity     axial      pigd
+0.6540408 1.0901181 1.0704805
+CLUSTERS: 4
+================================
+> 1   nms_d1    nms_d2    nms_d3    nms_d4    nms_d5    nms_d6    nms_d7    nms_d8    nms_d9    tremor  bradykin
+1.5558981 1.4133002 1.1184738 1.9877193 1.2550406 1.5834288 1.4194860 0.7793719 0.8480830 0.4984324 1.5155436
+ rigidity     axial      pigd
+1.5055810 1.9495563 1.9242566
+> 2    nms_d1     nms_d2     nms_d3     nms_d4     nms_d5     nms_d6     nms_d7     nms_d8     nms_d9     tremor
+-0.3206187 -0.4993585 -0.4670797 -0.3084269 -0.3380326 -0.3885648 -0.3868053 -0.2817080 -0.3620072 -0.2526726
+  bradykin   rigidity      axial       pigd
+-0.5841212 -0.5460419 -0.5968519 -0.5927227
+> 3     nms_d1      nms_d2      nms_d3      nms_d4      nms_d5      nms_d6      nms_d7      nms_d8      nms_d9
+ 0.35964469  0.82182933  0.92114183  0.30981044  0.75178256  0.42320932  0.66431905  0.63701383  0.85893238
+     tremor    bradykin    rigidity       axial        pigd
+-0.34050774 -0.15173201 -0.20269121  0.04852427  0.04032785
+> 4     nms_d1      nms_d2      nms_d3      nms_d4      nms_d5      nms_d6      nms_d7      nms_d8      nms_d9
+-0.13688715 -0.09662665 -0.13812235 -0.25339210 -0.29466914 -0.06605132 -0.19223309 -0.17297198 -0.20498319
+     tremor    bradykin    rigidity       axial        pigd
+ 0.55647025  0.72337965  0.69299204  0.47325105  0.47914113
+\end{lstlisting}
+\fi
+
+\subsection{Decision tree classifier based on clusters}
+% seed = 911
+% CLUSTERS: 2
+% ================================
+% Complexity Parameter: 0.03482587
+% 10-fold CV error: 0.134443
+% Root node error: 0.2573624
+% CLUSTERS: 3
+% ================================
+% Complexity Parameter: 0.01
+% 10-fold CV error: 0.1946223
+% Root node error: 0.4622279
+% CLUSTERS: 4
+% ================================
+% Complexity Parameter: 0.01
+% 10-fold CV error: 0.2483995
 % Root node error: 0.5236876
-% Complexity Parameter: 0.01222494
-% 10-fold CV error: 0.2535211
 \begin{table}[ht]
@@
   \begin{tabular}{l|l|l|l|l|l}
-    K & CP\footnote{Complexity Parameter} & CV Xerror\footnote{10-fold cross
+    $k$ & CP\tablefootnote{Complexity Parameter} & CV Xerror\tablefootnote{10-fold cross
     validation} & Root Feature &
@@
     \hline
-    2 & 0.017 & 0.124 & & 0.257 & Figure~\ref{fig:kmeans-dtree-2} \\
-    3 & 0.014 & 0.177 & & 0.457 & Figure~\ref{fig:kmeans-dtree-3} \\
-    4 & 0.012 & 0.254 & & 0.524 & Figure~\ref{fig:kmeans-dtree-4} \\
+    2 & 0.0348 & 0.134 & axial $\geq$ 0.44 & 0.257 & Figure~\ref{fig:kmeans-dtree-2} \\
+    3 & 0.0100 & 0.194 & bradykin $<$ 0.0041 & 0.462 & Figure~\ref{fig:kmeans-dtree-3} \\
+    4 & 0.0100 & 0.248 & bradykin $<$ 0.0041 & 0.523 & Figure~\ref{fig:kmeans-dtree-4} \\
   \end{tabular}

### 2015-06-12 R/kmeans-dtree.R
add print centers functionality

@@ -198,9 +198,24 @@ for (i in 2:4) {
   cl <- trees[[istr]]$clustering
   cat("CLUSTERS: ", i, "\n", sep="")
   cat("================================\n")
-  cat("WithinSS:", cl$withinss, "\n", sep=" ")
   cat("Sizes:", cl$size, "\n", sep=" ")
+  cat("WithinSS:", cl$withinss, "\n", sep=" ")
+  cat("Sum WithinSS:", sum(cl$withinss), "\n", sep=" ")
 }
+
+# ==== PRINT CENTERS ====
+for (i in 2:4) {
+  istr <- paste("clusters", i, sep="")
+  cl <- trees[[istr]]$clustering
+  cat("CLUSTERS: ", i, "\n", sep="")
+  cat("================================\n")
+  for (j in 1:i) {
+    center <- cl$centers[j, ]
+    cat("> ", j, sep="")
+    print(center)
+  }
+}
+
 # ==== NOTES ====
 # Note - with clustering, indeed the main feature (root of the tree) doesn't carry much
 # information

### 2015-06-24 R/kmeans-dtree.R
add cluster to data, add cluster summaries

@@ -216,6 +216,47 @@ for (i in 2:4) {
   }
 }
 
+# ==== BIND CLUSTER TO ORIGINAL DATA ====
+labeled <- vector(mode = "list", length = 3)
+labeled <- lapply(1:3, function(i) cbind(raw.omitted, trees[[paste("clusters", i+1, sep="")]]$clustering$cluster))
+lnames <- c("clusters2", "clusters3", "clusters4")
+names(labeled) <- lnames
+# I use 145 indexing because I'm not sure if tail() copies
+for (name in lnames) {
+  names(labeled[[name]])[145] <- "cluster"
+}
+
+# ==== COMBINE LABELS WITH ORIGINAL DATA ====
+clusters <- lapply(2:4, function(i) {
+  name <- paste("clusters", i, sep="")
+  current <- labeled[[name]]
+  separated <- lapply(1:i, function(c) {
+    current[current$cluster == c, ]
+  })
+})
+names(clusters) <- lnames
+
+for (i in 2:4) {
+  names(clusters[[paste("clusters", i, sep="")]]) <- 1:i
+}
+
+# ==== SUMMARY OF CLUSTER STATISTICS ====
+# TODO: Find similarity among the variables "For the interpreter"
+
+interpreted <- c("age", "sex", "pdonset", "durat_pd", "cisitot")
+for (i in 2:4) {
+  name <- paste("clusters", i, sep="")
+  current <- clusters[[name]]
+  for (c in 1:i) {
+    cname <- paste(c)
+    # WE ONLY CARE ABOUT THE INTERPRETED
+    interpreted <- current[[cname]][, interpreted]
+    cat("k = ", i, ", ", "CLUSTER ", cname, '\n', sep="")
+    # Select a couple of statistics
+    write.csv(round(describe(interpreted)[, c('mean', 'sd', 'min', 'max')], 2))
+  }
+}
+
 # ==== NOTES ====
 # Note - with clustering, indeed the main feature (root of the tree) doesn't carry much
 # information

### 2015-06-24 writeup/parkinsons.tex
update pdf to include new data

@@
 \usepackage{bm}
+\usepackage{numprint}
 \usepackage[margin=1in]{geometry}
@@
 \graphicspath{ {../figures/} }
+% Or this, if run from main folder
+% \graphicspath{ {./figures/} }
 
@@
 
+\subsection{Summary Statistics based on Clusters}
+
+\begin{lstlisting}
+k = 2, cluster 1
+var,mean,sd,min,max
+age,68.38,9.4,42,89
+sex,0.43,0.5,0,1
+pdonset,57.65,10.81,32,87
+durat_pd,10.73,7.14,0,40
+cisitot,13.07,4.55,2,24
+k = 2, cluster 2
+var,mean,sd,min,max
+age,63.43,9.55,37,89
+sex,0.36,0.48,0,1
+pdonset,56.44,10.62,28,89
+durat_pd,6.99,4.94,0,28
+cisitot,6.64,3.42,0,16
+
+k = 3, cluster 1
+var,mean,sd,min,max
+age,62.69,9.43,37,89
+sex,0.39,0.49,0,1
+pdonset,56.04,10.41,28,89
+durat_pd,6.65,4.66,0,26
+cisitot,5.76,3.15,0,15
+k = 3, cluster 2
+var,mean,sd,min,max
+age,66.17,9.74,38,86
+sex,0.32,0.47,0,1
+pdonset,57.78,11.19,32,84
+durat_pd,8.39,5.79,0,31
+cisitot,9.95,3.5,2,22
+k = 3, cluster 3
+var,mean,sd,min,max
+age,68.62,9.2,50,89
+sex,0.46,0.5,0,1
+pdonset,57.24,10.46,35,87
+durat_pd,11.38,7.52,0,40
+cisitot,13.52,5.01,2,24
+
+k = 4, cluster 1
+var,mean,sd,min,max
+age,71.9,8.1,54,89
+sex,0.46,0.5,0,1
+pdonset,58.28,10.27,35,87
+durat_pd,13.62,7.76,0,40
+cisitot,16.72,4.14,4,24
+k = 4, cluster 2
+var,mean,sd,min,max
+age,62.65,9.63,37,89
+sex,0.36,0.48,0,1
+pdonset,56.13,10.47,28,89
+durat_pd,6.52,4.66,0,25
+cisitot,5.58,3.12,0,15
+k = 4, cluster 3
+var,mean,sd,min,max
+age,64.79,9.3,44,86
+sex,0.47,0.5,0,1
+pdonset,56.2,10.71,35,85
+durat_pd,8.59,6.16,0,27
+cisitot,9.23,3.88,2,19
+k = 4, cluster 4
+var,mean,sd,min,max
+age,66.27,9.47,40,86
+sex,0.33,0.47,0,1
+pdonset,57.84,11.06,32,84
+durat_pd,8.43,5.64,0,31
+cisitot,10.06,3.49,3,22
+\end{lstlisting}
 \section{Biclustering}
+
+Used BCBimax clustering algorithm. Clusters seem quite sparse.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{biclust-16.pdf}
+  \caption{Biclustering $N = 16$}
+  \label{fig:biclust-16}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{bubbleplot-16.pdf}
+  \caption{Bubbleplot $N = 16$}
+  \label{fig:bubbleplot-16}
+\end{figure}
 

### 2015-06-24 R/biclustering.R
didn't know Rstudio's comment block format

@@ -1,24 +1,24 @@
-# ==== LIBRARIES ====
+# LIBRARIES ====
 # http://cran.r-project.org/web/packages/biclust/biclust.pdf
 library(biclust)  # Quite a recent package (May 12, 2015)
 
-# ==== LOAD DATA ====
+# LOAD DATA ====
 source('./preprocessing.R')
 
-# ==== CONSTANTS ====
+# CONSTANTS ====
 # Number of biclusters
 # Note - N can't be more than 16 for par, will need to fix that later
 # (multiple plots/pagination)
 N <- 16
 
-# ==== BICLUSTERING ====
+# BICLUSTERING ====
 # Bimax algorithm
 bcl <- biclust(as.matrix(raw.filtered), method=BCBimax(), minr=2, minc=2, number=N)
 
-# ==== BUBBLEPLOT (?) ====
+# BUBBLEPLOT (?) ====
 bubbleplot(raw.filtered, bcl)
 
-# ==== CLUSTER HEATMAPS ====
+# CLUSTER HEATMAPS ====
 
 # Optimal grid organization
 par(mfrow = n2mfrow(N))

### 2015-06-24 R/kmeans-dtree.R

@@ -1,4 +1,4 @@
-# ==== LOAD LIBRARIES ====
+# LOAD LIBRARIES ====
 library(cluster)
 library(rpart)
 library(rpart.plot)
@@ -6,10 +6,10 @@ library(plyr)
 library(fpc)
 library(NbClust)
 
-# ==== LOAD DATA ====
+# LOAD DATA ====
 source('./preprocessing.R')
 
-# ==== VISUALIZE WSS ERROR TO FIND OPTIMAL K ====
+# VISUALIZE WSS ERROR TO FIND OPTIMAL K ====
 # NOTE: Doesn't work well, there isn't any elbow
 
 wss <- (nrow(raw.filtered)-1)*sum(apply(raw.filtered, 2, var))
@@ -20,7 +20,7 @@ for (i in 2:15) {
 plot(1:15, wss, type="b",
      xlab="Number of Clusters", ylab="Within groups sum of squares")
 
-# ==== TODO: BIC ====
+# TODO: BIC ====
 # See http://stackoverflow.com/questions/15376075/cluster-analysis-in-r-determine-the-optimal-number-of-clusters
 library(mclust)
 # NOW - does this have to do with EM/hierarchical models. not cluster-based?
@@ -34,14 +34,14 @@ library(mclust)
 # # 3 clusters
 # plot(d_clust)
 
-# ==== NBCLUST ESTIMATION FOR OPTIMAL K (30 metrics) ====
+# NBCLUST ESTIMATION FOR OPTIMAL K (30 metrics) ====
 # This is taking a long time
 # nb <- NbClust(raw.filtered, distance = "euclidean",
 #               min.nc=2, max.nc=15, method = "kmeans",
 #               index = "alllong", alphaBeale = 0.1)
 # hist(nb$Best.nc[1,], breaks = max(na.omit(nb$Best.nc[1,])))
 
-# ==== PAM ESTIMATION FOR OPTIMAL K ====
+# PAM ESTIMATION FOR OPTIMAL K ====
 # Estimates 2
 pam_sils <- c()
 for (i in 1:15) {
@@ -49,11 +49,11 @@ for (i in 1:15) {
 }
 plot(x=1:14, y=pam_sils, xlab="Clusters", ylab="Average Silhouette Width", type="b")
 
-# ==== GAP STATISTIC ESTIMATION ====
+# GAP STATISTIC ESTIMATION ====
[…]

### 2015-06-24 R/preprocessing.R

@@ -1,10 +1,10 @@
-# ==== GLOBAL LIBRARIES ====
+# GLOBAL LIBRARIES ====
 library(plyr)
 # For PCA
 library(FactoMineR)
 library(psych)
 
-# ==== GLOBAL CONSTANTS ====
+# GLOBAL CONSTANTS ====
 DB_FILE = "../data/DATABASE_NMS Burden levels_15-4-2012.csv"
 ALL.SYMPTOMS = c("nms_d1",
                  "nms_d2",
@@ -28,7 +28,7 @@ MOTOR.SYMPTOMS = c("tremor",
 # Change this variable to change symptoms
 SYMPTOMS.TO.USE = ALL.SYMPTOMS
 
-# ==== UTILITY FUNCTIONS ====
+# UTILITY FUNCTIONS ====
 splitdf <- function(dataframe, seed=NULL, trainfrac=0.7) {
   if (trainfrac<=0 | trainfrac>=1) stop("Training fraction must be between 0 and 1, not inclusive")
   if (!is.null(seed)) set.seed(seed)
@@ -39,7 +39,7 @@ splitdf <- function(dataframe, seed=NULL, trainfrac=0.7) {
   list(trainset=trainset,testset=testset)
 }
 
-# ==== IMPORT DATA, PREPROCESSING ====
+# IMPORT DATA, PREPROCESSING ====
 raw <- read.csv(DB_FILE)
 # Study is irrelevant
 raw$study <- NULL
@@ -49,15 +49,15 @@ raw.omitted <- na.omit(raw)
 # raw with only select nms and motor symptoms
 raw.filtered <- raw.omitted[, SYMPTOMS.TO.USE]
 
-# ==== DESCRIPTIVE STATISTICS (BEFORE STANDARDIZATION) ====
+# DESCRIPTIVE STATISTICS (BEFORE STANDARDIZATION) ====
 # TODO - change raw.filtered to raw.omitted.filtered, or something like that
 raw.omitted.stats <- describe(raw.omitted)
 raw.filtered.stats <- describe(raw.filtered)
 
-# ==== STANDARDIZATION ====
+# STANDARDIZATION ====
 raw.filtered <- as.data.frame(scale(raw.filtered))
 
-# ==== PCA (Note: not helpful for identifying specific factors) ====
+# PCA (Note: not helpful for identifying specific factors) ====
 par(mfrow = c(1, 2))
 pca <- PCA(raw.filtered)
 # pca$eig

### 2015-06-24 R/kmeans-dtree.R
fix variable name conflict, default no save dtrees

@@ -5,6 +5,14 @@ library(rpart.plot)
 library(plyr)
 library(fpc)
 library(NbClust)
+# library for TODO: BIC
+library(mclust)
+
+# CONSTANTS ====
+# Remember - PDFs can vary even if the (seeded) clusters don't
+# So this should be false unless I've changed something about the
+# kmeans analysis
+SAVE.DTREES = FALSE
 
 # LOAD DATA ====
 source('./preprocessing.R')
@@ -22,7 +30,6 @@ plot(1:15, wss, type="b",
 
 # TODO: BIC ====
 # See http://stackoverflow.com/questions/15376075/cluster-analysis-in-r-determine-the-optimal-number-of-clusters
-library(mclust)
 # NOW - does this have to do with EM/hierarchical models. not cluster-based?
 # Run the function to see how many clusters
 # it finds to be optimal, set it to search for
@@ -178,7 +185,8 @@ trees <- vector(mode = "list", length = 3)
 names(trees) <- c("clusters2", "clusters3", "clusters4")
 for (i in 2:4) {
   istr <- paste("clusters", i, sep="")
-  trees[[istr]] <-  kmeans_dtree(raw.filtered, i, save = TRUE)
+  # Remember - PDFs can vary even if the (seeded) clusters don't
+  trees[[istr]] <-  kmeans_dtree(raw.filtered, i, save = SAVE.DTREES)
 }
 
 # PRINT GLOBAL TREE STATS ====
@@ -250,10 +258,10 @@ for (i in 2:4) {
   for (c in 1:i) {
     cname <- paste(c)
     # WE ONLY CARE ABOUT THE INTERPRETED
-    interpreted <- current[[cname]][, interpreted]
+    current.filtered <- current[[cname]][, interpreted]
     cat("k = ", i, ", ", "CLUSTER ", cname, '\n', sep="")
     # Select a couple of statistics
-    write.csv(round(describe(interpreted)[, c('mean', 'sd', 'min', 'max')], 2))
+    write.csv(round(describe(current.filtered)[, c('mean', 'sd', 'min', 'max')], 2))
   }
 }
 

### 2015-06-24 R/kmeans-dtree.R
automatically save kmeans summaries

@@ -250,19 +250,37 @@ for (i in 2:4) {
 
 # SUMMARY OF CLUSTER STATISTICS ====
 # TODO: Find similarity among the variables "For the interpreter"
+# Rounding constant
+PRECISION = 3
 
 interpreted <- c("age", "sex", "pdonset", "durat_pd", "cisitot")
 for (i in 2:4) {
   name <- paste("clusters", i, sep="")
   current <- clusters[[name]]
+  # Begin collecting clusters of csv
+  clusters.csv <- c()
   for (c in 1:i) {
     cname <- paste(c)
     # WE ONLY CARE ABOUT THE INTERPRETED
     current.filtered <- current[[cname]][, interpreted]
+    # Print out stuff first
+    # NOTE: this format differs from what is written to csv (csv has cluster column)
     cat("k = ", i, ", ", "CLUSTER ", cname, '\n', sep="")
+    filtered.description <- round(describe(current.filtered)[, c('mean', 'sd', 'min', 'max')],
+                                  PRECISION)
     # Select a couple of statistics
-    write.csv(round(describe(current.filtered)[, c('mean', 'sd', 'min', 'max')], 2))
+    write.csv(filtered.description)
+    # Attach varaible rownames, number of cluster (c)
+    filtered.description <- cbind(variable=rownames(filtered.description), filtered.description)
+    filtered.description <- cbind(cluster=rep(c, length(interpreted)), filtered.description)
+    # Then add to csv list
+    clusters.csv <- rbind(clusters.csv, filtered.description)
   }
+  # Write cluster summaries as csv to data folder
+  # NOTE: This (should) stays the same because of the seeding, unlike the PDFs,
+  # so we don't need a boolean flag here
+  write.csv(clusters.csv, file = paste("../data/kmeans-summaries-", i, ".csv", sep=""),
+            row.names = FALSE)
 }
 
 # NOTES ====

### 2015-06-24 R/kmeans-dtree.R
add todo comment to remind me of #1

@@ -13,6 +13,7 @@ library(mclust)
 # So this should be false unless I've changed something about the
 # kmeans analysis
 SAVE.DTREES = FALSE
+# TODO: Make k = 2, 3, 4 modifiable via constant
 
 # LOAD DATA ====
 source('./preprocessing.R')

### 2015-06-24 writeup/parkinsons.tex
Add bar plots of interpreted cluster vars

--- 2015-06-12 writeup/parkinsons.tex
@@
 
-\subsection{Decision Tree classifier based on clusters}
-
+\subsection{Cluster statistics}
 % CLUSTERS: 2
 % ================================
-% Root node error: 0.2573624
-% Complexity Parameter: 0.01741294
-% 10-fold CV error: 0.1241997
+% Sizes: 201 580
+% WithinSS: 4248.585 4132.434
+% Sum WithinSS: 8381.019
 % CLUSTERS: 3
 % ================================
-% Root node error: 0.4571063
-% Complexity Parameter: 0.0140056
-% 10-fold CV error: 0.1766965
+% Sizes: 420 231 130
+% WithinSS: 2618.368 1973.82 3076.542
+% Sum WithinSS: 7668.73
 % CLUSTERS: 4
 % ================================
+% Sizes: 61 372 145 203
+% WithinSS: 1481.25 1845.389 2147.988 1609.555
+% Sum WithinSS: 7084.183
+\begin{table}[ht]
+  \centering
+  \begin{tabular}{l|l|l|l}
+    $k$ & $n$ & Within SS & sum(Within SS) \\
+    \hline
+    2 & 201/580 & 4248.585/4132.434 & 8381.019 \\
+    3 & 420/231/130 & 2618.368/1973.82/3076.542 &  7668.73 \\
+    4 & 61/372/145/203 & 1481.25/1845.389/2147.988/1609.555 & 7084.183
+  \end{tabular}
+  \caption{Cluster statistics}
+  \label{tab:cluster-statistics}
+\end{table}
+
+\subsection{Decision trees based on clusters}
+% seed = 911
+% CLUSTERS: 2
+% ================================
+% Complexity Parameter: 0.03482587
+% 10-fold CV error: 0.134443
+% Root node error: 0.2573624
+% CLUSTERS: 3
+% ================================
+% Complexity Parameter: 0.01
+% 10-fold CV error: 0.1946223
+% Root node error: 0.4622279
+% CLUSTERS: 4
+% ================================
+% Complexity Parameter: 0.01
+% 10-fold CV error: 0.2483995
 % Root node error: 0.5236876
-% Complexity Parameter: 0.01222494
-% 10-fold CV error: 0.2535211
 \begin{table}[ht]
@@
   \begin{tabular}{l|l|l|l|l|l}
-    K & CP\tablefootnote{Complexity Parameter} & CV Xerror\tablefootnote{10-fold cross
+    $k$ & CP\tablefootnote{Complexity Parameter} & CV Xerror\tablefootnote{10-fold cross
     validation} & Root Feature &
@@
 
+\subsection{Interpretation of Clusters}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\linewidth]{kmeans-summaries-2.pdf}
+  \caption{Cluster Summaries, $k = 2$}
+  \label{fig:kmeans-summaries-2}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\linewidth]{kmeans-summaries-3.pdf}
+  \caption{Cluster Summaries, $k = 3$}
+  \label{fig:kmeans-summaries-3}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=\linewidth]{kmeans-summaries-4.pdf}
+  \caption{Cluster Summaries, $k = 4$}
+  \label{fig:kmeans-summaries-4}
+\end{figure}
+
+\end{lstlisting}
 \section{Biclustering}
+
+Used BCBimax clustering algorithm. Clusters seem quite sparse.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{biclust-16.pdf}
+  \caption{Biclustering $N = 16$}
+  \label{fig:biclust-16}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{bubbleplot-16.pdf}
+  \caption{Bubbleplot $N = 16$}
+  \label{fig:bubbleplot-16}
+\end{figure}
 

### 2015-06-24 R/kmeans-dtree.R

@@ -7,6 +7,7 @@ library(fpc)
 library(NbClust)
 # library for TODO: BIC
 library(mclust)
+library(ggplot2)
 
 # CONSTANTS ====
 # Remember - PDFs can vary even if the (seeded) clusters don't
@@ -182,6 +183,7 @@ kmeans_dtree <- function(data, k, save = FALSE, seed = 911) {
 }
 
 # KMEANS DTREE OBTAINING PLOTS ====
+# FIXME: This can be vectorized with lapply
 trees <- vector(mode = "list", length = 3)
 names(trees) <- c("clusters2", "clusters3", "clusters4")
 for (i in 2:4) {
@@ -282,6 +284,32 @@ for (i in 2:4) {
   # so we don't need a boolean flag here
   write.csv(clusters.csv, file = paste("../data/kmeans-summaries-", i, ".csv", sep=""),
             row.names = FALSE)
+}
+
+# PLOT CLUSTER RESULTS ====
+# TODO: I may just need to plot everything, not just variables for the interpreter
+# Then ideally (but not necessary) I'd want to find some way to differentiate the variables
+# "For the interpreter" and the clustered variables and serialize that into the csv
+summaries <- lapply(2:4, function(i) {
+  raw <- read.csv(paste("../data/kmeans-summaries-", i, ".csv", sep=""),
+                  stringsAsFactors = TRUE)
+  # Turn numeric clusters into factors
+  raw$cluster = factor(raw$cluster)
+  raw
+})
+# FIXME: By using standard numeric list names, I can't use summaries$2 or summaries[[2]],
+# only summaries$"2" or summaries[["2"]]
+# seems confusing and not standardized with the rest of the script
+names(summaries) <- 2:4 # NOTE: These numbers become factors (strings)!
+for (k in c("2", "3", "4")) {
+  summary <- summaries[[k]]
+  p <- ggplot(summary, aes(x = cluster, y = mean, fill=cluster)) +
+    guides(fill = FALSE) +
+    facet_wrap( ~ variable, scales = "free") +
+    geom_bar(stat = "identity", position = position_dodge()) +
+    geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), position = position_dodge(), width = 0.2)
+  print(p)
+  ggsave(paste("../figures/kmeans-summaries-", k, ".pdf", sep=""))
 }
 
 # NOTES ====

### 2015-06-24 R/kmeans-dtree.R
don't need to preallocate vector before lapply

@@ -228,7 +228,6 @@ for (i in 2:4) {
 }
 
 # BIND CLUSTER TO ORIGINAL DATA ====
-labeled <- vector(mode = "list", length = 3)
 labeled <- lapply(1:3, function(i) cbind(raw.omitted, trees[[paste("clusters", i+1, sep="")]]$clustering$cluster))
 lnames <- c("clusters2", "clusters3", "clusters4")
 names(labeled) <- lnames

### 2015-06-24 R/kmeans-dtree.R
Expand interpreted variables to ca vars
Because it's worth looking at general trends, obviously

@@ -14,6 +14,7 @@ library(ggplot2)
 # So this should be false unless I've changed something about the
 # kmeans analysis
 SAVE.DTREES = FALSE
+SAVE.BARPLOTS = TRUE
 # TODO: Make k = 2, 3, 4 modifiable via constant
 
 # LOAD DATA ====
@@ -255,7 +256,14 @@ for (i in 2:4) {
 # Rounding constant
 PRECISION = 3
 
-interpreted <- c("age", "sex", "pdonset", "durat_pd", "cisitot")
+# FIXME: The results from shortened list of interpreted variables below, which I've saved in
+# the data folder (for now), are not automatically saved to the data or figures folders. If
+# they're needed, we need to save them again.
+# interpreted <- c("age", "sex", "pdonset", "durat_pd", "cisitot")
+interpreted <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
+                 "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
+                 "nms_d6", "nms_d7", "nms_d8", "nms_d9",
+                 "tremor", "bradykin", "rigidity", "axial", "pigd")
 for (i in 2:4) {
   name <- paste("clusters", i, sep="")
   current <- clusters[[name]]
@@ -264,6 +272,7 @@ for (i in 2:4) {
   for (c in 1:i) {
     cname <- paste(c)
     # WE ONLY CARE ABOUT THE INTERPRETED
+    # FALSE - now we're going to just include everything
     current.filtered <- current[[cname]][, interpreted]
     # Print out stuff first
     # NOTE: this format differs from what is written to csv (csv has cluster column)
@@ -308,7 +317,9 @@ for (k in c("2", "3", "4")) {
     geom_bar(stat = "identity", position = position_dodge()) +
     geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), position = position_dodge(), width = 0.2)
   print(p)
-  ggsave(paste("../figures/kmeans-summaries-", k, ".pdf", sep=""))
+  if (SAVE.BARPLOTS) {
+    ggsave(paste("../figures/kmeans-summaries-", k, ".pdf", sep=""))
+  }
 }
 
 # NOTES ====

### 2015-06-24 writeup/parkinsons.tex
add brief description of summary plots
k = 4!!!!

@@
 
+\subsubsection{Cluster summaries}
+
+Available in
+Figures~\ref{fig:kmeans-summaries-2},~\ref{fig:kmeans-summaries-3},
+and~\ref{fig:kmeans-summaries-4}. Error bar is standard error.
+
+\subsubsection{Interpretation}
+
+$k = 2$ seems too basic. Cluster is organized solely by severity - all
+symptoms, including motor and nonmotor, are higher in severity in cluster 1,
+and lower in cluster 2. Quite consistently, groups in cluster 1 are generally
+of slightly higher age and pd duration.
+
+$k = 3$ seems like a further development of $k = 2$, where clusters are simply
+organized by linearly increasing severity.
+
+$k = 4$ is where it gets interesting.
+
 \begin{figure}[ht]

### 2015-06-24 R/kmeans-dtree.R
use standard error in error bars, not sd

@@ -277,7 +277,7 @@ for (i in 2:4) {
     # Print out stuff first
     # NOTE: this format differs from what is written to csv (csv has cluster column)
     cat("k = ", i, ", ", "CLUSTER ", cname, '\n', sep="")
-    filtered.description <- round(describe(current.filtered)[, c('mean', 'sd', 'min', 'max')],
+    filtered.description <- round(describe(current.filtered)[, c('mean', 'sd', 'min', 'max', 'se')],
                                   PRECISION)
     # Select a couple of statistics
     write.csv(filtered.description)
@@ -315,7 +315,8 @@ for (k in c("2", "3", "4")) {
     guides(fill = FALSE) +
     facet_wrap( ~ variable, scales = "free") +
     geom_bar(stat = "identity", position = position_dodge()) +
-    geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), position = position_dodge(), width = 0.2)
+    geom_errorbar(aes(ymin = mean - se, ymax = mean + se),
+                  position = position_dodge(), width = 0.2)
   print(p)
   if (SAVE.BARPLOTS) {
     ggsave(paste("../figures/kmeans-summaries-", k, ".pdf", sep=""))

### 2015-06-24 R/kmeans-dtree.R
save raw binded cluster data for box plots

@@ -10,11 +10,17 @@ library(mclust)
 library(ggplot2)
 
 # CONSTANTS ====
+# INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot")
+INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
+                 "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
+                 "nms_d6", "nms_d7", "nms_d8", "nms_d9",
+                 "tremor", "bradykin", "rigidity", "axial", "pigd")
+
 # Remember - PDFs can vary even if the (seeded) clusters don't
 # So this should be false unless I've changed something about the
 # kmeans analysis
 SAVE.DTREES = FALSE
-SAVE.BARPLOTS = TRUE
+SAVE.BARPLOTS = FALSE
 # TODO: Make k = 2, 3, 4 modifiable via constant
 
 # LOAD DATA ====
@@ -251,7 +257,9 @@ for (i in 2:4) {
   names(clusters[[paste("clusters", i, sep="")]]) <- 1:i
 }
 
-# SUMMARY OF CLUSTER STATISTICS ====
+# COLLECTED CLUSTERS AND SUMMARY OF CLUSTER STATISTICS ====
+# Get both the summary of the statistics and collect the raw clusters
+# in the form clusters$clustersN$[[N]] for box plotting later on
 # TODO: Find similarity among the variables "For the interpreter"
 # Rounding constant
 PRECISION = 3
@@ -260,20 +268,23 @@ PRECISION = 3
 # the data folder (for now), are not automatically saved to the data or figures folders. If
 # they're needed, we need to save them again.
 # interpreted <- c("age", "sex", "pdonset", "durat_pd", "cisitot")
-interpreted <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
-                 "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
-                 "nms_d6", "nms_d7", "nms_d8", "nms_d9",
-                 "tremor", "bradykin", "rigidity", "axial", "pigd")
+# NOTE: This is now set as a constant up above
+
+# This is a global collection of clusters to be used
+clusters.raw <- vector(mode = "list", length = 3)
+names(clusters.raw) <- 2:4
+
 for (i in 2:4) {
   name <- paste("clusters", i, sep="")
   current <- clusters[[name]]
[…]

### 2015-06-24 R/kmeans-dtree.R
don't use c as a variable name you dork

@@ -280,8 +280,8 @@ for (i in 2:4) {
   # Begin collecting clusters of csv
   summaries.csv <- c()
   clusters.csv <- c()
-  for (c in 1:i) {
-    cname <- paste(c)
+  for (clus in 1:i) {
+    cname <- paste(clus)
     # WE ONLY CARE ABOUT THE INTERPRETED
     # FALSE - now we're going to just include everything
     current.filtered <- current[[cname]][, INTERPRETED]
@@ -294,12 +294,12 @@ for (i in 2:4) {
     write.csv(filtered.description)
     # Attach variable rownames, number of cluster (c)
     filtered.description <- cbind(variable=rownames(filtered.description), filtered.description)
-    filtered.description <- cbind(cluster=rep(c, length(INTERPRETED)), filtered.description)
+    filtered.description <- cbind(cluster=rep(clus, length(INTERPRETED)), filtered.description)
     # Then add to csv list
     summaries.csv <- rbind(summaries.csv, filtered.description)
 
     # Attach cluster id to all elements in the cluster
-    filtered.with_clusters <- cbind(cluster=rep(c, nrow(current.filtered)), current.filtered)
+    filtered.with_clusters <- cbind(cluster=rep(clus, nrow(current.filtered)), current.filtered)
     clusters.csv <- rbind(clusters.csv, filtered.with_clusters)
   }
   # Write cluster summaries as csv to data folder

### 2015-06-24 R/kmeans-dtree.R
draw box plots instead of bar plots

@@ -8,6 +8,7 @@ library(NbClust)
 # library for TODO: BIC
 library(mclust)
 library(ggplot2)
+library(tidyr)
 
 # CONSTANTS ====
 # INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot")
@@ -272,7 +273,7 @@ PRECISION = 3
 
 # This is a global collection of clusters to be used
 clusters.raw <- vector(mode = "list", length = 3)
-names(clusters.raw) <- 2:4
+names(clusters.raw) <- c("2", "3", "4")
 
 for (i in 2:4) {
   name <- paste("clusters", i, sep="")
@@ -309,32 +310,27 @@ for (i in 2:4) {
             row.names = FALSE)
   write.csv(clusters.csv, file = paste("../data/kmeans-raw-", i, ".csv", sep=""),
             row.names = FALSE)
-  clusters.raw[[i]] <- clusters.csv
+  clusters.raw[[paste(i)]] <- clusters.csv
 }
 
+# CLUSTERS RAW WIDE -> LONG ====
+clusters.raw.long <- vector(mode = "list", length = 3)
+names(clusters.raw.long) <- c("2", "3", "4")
+for (i in c("2", "3", "4")) {
+  clusters.raw.long[[i]] <- gather(clusters.raw[[i]], variable, measurement, age:pigd)
+}
+
+
 # PLOT CLUSTER RESULTS ====
-# TODO: I may just need to plot everything, not just variables for the interpreter
-# Then ideally (but not necessary) I'd want to find some way to differentiate the variables
-# "For the interpreter" and the clustered variables and serialize that into the csv
-summaries <- lapply(2:4, function(i) {
-  raw <- read.csv(paste("../data/kmeans-summaries-", i, ".csv", sep=""),
-                  stringsAsFactors = TRUE)
-  # Turn numeric clusters into factors
-  raw$cluster = factor(raw$cluster)
-  raw
-})
 # FIXME: By using standard numeric list names, I can't use summaries$2 or summaries[[2]],
 # only summaries$"2" or summaries[["2"]]
 # seems confusing and not standardized with the rest of the script
-names(summaries) <- 2:4 # NOTE: These numbers become factors (strings)!
-for (k in c("2", "3", "4")) {
-  summary <- summaries[[k]]
-  p <- ggplot(summary, aes(x = cluster, y = mean, fill=cluster)) +
+for (i in c("2", "3", "4")) {
[…]

### 2015-06-24 R/kmeans-dtree.R
fix up boxplot saving, save, rerun pdflatex

@@ -21,7 +21,7 @@ INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
 # So this should be false unless I've changed something about the
 # kmeans analysis
 SAVE.DTREES = FALSE
-SAVE.BARPLOTS = FALSE
+SAVE.BOXPLOTS = FALSE
 # TODO: Make k = 2, 3, 4 modifiable via constant
 
 # LOAD DATA ====
@@ -332,8 +332,8 @@ for (i in c("2", "3", "4")) {
     guides(fill = FALSE) +
     facet_wrap( ~ variable, scales = "free")
   print(p)
-  if (SAVE.BARPLOTS) {
-    ggsave(paste("../figures/kmeans-summaries-", k, ".pdf", sep=""))
+  if (SAVE.BOXPLOTS) {
+    ggsave(paste("../figures/kmeans-summaries-", i, ".pdf", sep=""))
   }
 }
 

### 2015-06-25 R/kmeans-dtree.R
Add unscaled decision trees
NOTE: Weird happenings re: the k = 4 pruned tree. Seems like in the
previous commit (#a0250ee) k = 4 was actually an unpruned tree.
Rerunning the script with SAVE.DTREES = TRUE and pdflatex fixes this.
I'm going to push a rerun of pdflatex soon, which then can be compared
to the previous version - not sure what happened.

@@ -115,16 +115,20 @@ text(pruned.t, use.n=TRUE, all=TRUE, cex=.8, pos=1)
 
 # KMEANS CLUSTERING COMPRESSED FUNCTION ====
 # Use for iteration!
-kmeans_dtree <- function(data, k, save = FALSE, seed = 911) {
+kmeans.dtree <- function(data, data.unscaled, k, save = FALSE, seed = 911) {
   # Reproducibility!
   set.seed(seed)
   cl <- kmeans(splits$trainset, k, nstart = 25)
-  labeled_data <- cbind(data, cl$cluster)
-  # Add cluster label to original data, rename
-  labeled_data <- rename(labeled_data, c("cl$cluster"="cluster"))
+  labeled.data <- cbind(data, cluster=cl$cluster)
   # Convert to factor
-  labeled_data$cluster <- as.factor(labeled_data$cluster)
-  t <- rpart(cluster ~ ., labeled_data)
+  labeled.data$cluster <- as.factor(labeled.data$cluster)
+
+  # Do the same thing with unscaled data to compare
+  labeled.data.unscaled <- cbind(data.unscaled, cluster=cl$cluster)
+  labeled.data.unscaled$cluster <- as.factor(labeled.data.unscaled$cluster)
+
+  t.unscaled <- rpart(cluster ~ ., labeled.data.unscaled)
+  t <- rpart(cluster ~ ., labeled.data)
 
   # Find 10-fold CV error rate for UNPRUNED tree
   # Old resub rate method
@@ -153,6 +157,10 @@ kmeans_dtree <- function(data, k, save = FALSE, seed = 911) {
   xerror.min.pruned <- tail(pruned.t$cptable[, "xerror"], n=1)
   xv.error.pruned <- xerror.min.pruned * root.node.error.pruned
 
+  # Do the same with the unscaled tree
+  cp.unscaled <- t.unscaled$cptable[which.min(t.unscaled$cptable[,"xerror"]),"CP"]
+  pruned.t.unscaled <- prune(t.unscaled, cp = cp)
+
   # Plot and save trees
   # Also look at fancyRpartPlot()
   # fancyRpartPlot(tree.2)
@@ -179,8 +187,16 @@ kmeans_dtree <- function(data, k, save = FALSE, seed = 911) {
     dev.off()
   }
 
+  # Plot the unscaled tree
+  prp(pruned.t.unscaled, extra = 1,
[…]

### 2015-06-25 R/preprocessing.R

@@ -55,6 +55,7 @@ raw.omitted.stats <- describe(raw.omitted)
 raw.filtered.stats <- describe(raw.filtered)
 
 # STANDARDIZATION ====
+raw.filtered.unscaled <- raw.filtered
 raw.filtered <- as.data.frame(scale(raw.filtered))
 
 # PCA (Note: not helpful for identifying specific factors) ====

### 2015-06-25 R/kmeans-dtree.R
Reorder factors in k = 4 in increasing cisitot
...for simplicity of interpretation in the graphs.
The issue with this, of course, is that the numbers are backwards, but
that can be fixed later.

@@ -341,6 +341,11 @@ for (i in c("2", "3", "4")) {
 # FIXME: By using standard numeric list names, I can't use summaries$2 or summaries[[2]],
 # only summaries$"2" or summaries[["2"]]
 # seems confusing and not standardized with the rest of the script
+
+# Reorder clusters to c(2, 3, 4, 1) for k = 4 to simplify interpretation
+clusters.raw.long[["4"]]$cluster <- factor(clusters.raw.long[["4"]]$cluster,
+                                              levels = c(2, 3, 4, 1))
+
 for (i in c("2", "3", "4")) {
   clus <- clusters.raw.long[[i]]
   p <- ggplot(clus, aes(x = factor(cluster), y = measurement, fill = factor(cluster))) +

### 2015-06-25 R/kmeans-dtree.R
apply factor reordering to all k

@@ -336,15 +336,20 @@ for (i in c("2", "3", "4")) {
   clusters.raw.long[[i]] <- gather(clusters.raw[[i]], variable, measurement, age:pigd)
 }
 
+# REORDER FACTORS BY INCREASING CISITOT ====
+for (i in c("2", "3", "4")) {
+  cisitot <- clusters.raw.long[[i]][clusters.raw.long[[i]]$variable == "cisitot", ] 
+  cisitot.means <- sapply(factor(1:as.integer(i)), function(i) {
+    mean(cisitot[cisitot$cluster == i, ]$measurement)
+  })
+  clusters.raw.long[[i]]$cluster <- factor(clusters.raw.long[[i]]$cluster,
+                                           levels = order(cisitot.means))
+}
 
 # PLOT CLUSTER RESULTS ====
 # FIXME: By using standard numeric list names, I can't use summaries$2 or summaries[[2]],
 # only summaries$"2" or summaries[["2"]]
 # seems confusing and not standardized with the rest of the script
-
-# Reorder clusters to c(2, 3, 4, 1) for k = 4 to simplify interpretation
-clusters.raw.long[["4"]]$cluster <- factor(clusters.raw.long[["4"]]$cluster,
-                                              levels = c(2, 3, 4, 1))
 
 for (i in c("2", "3", "4")) {
   clus <- clusters.raw.long[[i]]

### 2015-06-25 writeup/parkinsons.tex
fix tex syntax error
Switch to using unscaled decision trees
These are more important for sensible interpretation
Add silhouette plots to writeup
Note - silhouette plots look good but continue to observe that the
average silhouette width (done by PAM, which should be comparable since
I can't find a lazy method for kmeans) is very weak, indicating little
structure?

@@
 
+\subsection{Silhouette plots}
+
+Available in
+Figures~\ref{fig:kmeans-silhouette-2},~\ref{fig:kmeans-silhouette-3},
+and~\ref{fig:kmeans-silhouette-4}. Note: constructed with standardized
+$z$-score data.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{kmeans-silhouette-2.pdf}
+  \caption{$k$-means cluster silhouette plot, $k = 2$}
+  \label{fig:kmeans-silhouette-2}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{kmeans-silhouette-3.pdf}
+  \caption{$k$-means cluster silhouette plot, $k = 3$}
+  \label{fig:kmeans-silhouette-3}
+\end{figure}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{kmeans-silhouette-4.pdf}
+  \caption{$k$-means cluster silhouette plot, $k = 4$}
+  \label{fig:kmeans-silhouette-4}
+\end{figure}
+
 \subsection{Decision trees based on clusters}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-2.pdf}
+  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-unscaled-2.pdf}
   \caption{Decision Tree from $k$-means clustering, 2 clusters}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-3.pdf}
+  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-unscaled-3.pdf}
   \caption{Decision Tree from $k$-means clustering, 3 clusters}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-4.pdf}
+  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-unscaled-4.pdf}
   \caption{Decision Tree from $k$-means clustering, 4 clusters}
@@
 
-\end{lstlisting}
 \section{Biclustering}

### 2015-06-25 R/kmeans-dtree.R

@@ -119,6 +119,14 @@ kmeans.dtree <- function(data, data.unscaled, k, save = FALSE, seed = 911) {
   # Reproducibility!
   set.seed(seed)
   cl <- kmeans(splits$trainset, k, nstart = 25)
+  # Visualize clustering with clusplot
+  clusplot(raw.filtered, cl$cluster, main = paste("Silhouette plot k =", 4))
+  if (save) {
+    # FIXME: This save boolean flag is SAVE.DTREES outside of this function
+    # which is slightly ambiguous (these are silhouette plots!)
+    dev.copy(pdf, paste('../figures/kmeans-silhouette-', k, '.pdf', sep=''))
+    dev.off()
+  }
   labeled.data <- cbind(data, cluster=cl$cluster)
   # Convert to factor
   labeled.data$cluster <- as.factor(labeled.data$cluster)

### 2015-06-25 writeup/parkinsons.tex
Initial affinity propagation
Next step - interpreting the results of the 8 clusters and seeing if
they mesh with the 4 k-means clusters

@@
 
+\section{Affinity Propagation}
+
+\subsection{Clustering}
+
+AP with input preferences minimized ($q = 0$) resulted in 8 clusters.
+With the standard median input preferences ($q = 0.5$), algorithm failed to
+converge with default parameters. Even setting damping factor to 0.98, maximum
+iterations to 10000, and convergence iterations to 1000 failed to converge.
+Might need to try a longer term work.
+
+\emph{However}, given that input preferences control how many clusters are
+found, I don't think it's very useful to have some dozen clusters running
+around.
+
+\subsubsection{Silhouette Plots}
+
+Silhouette plot in Figure~\ref{fig:ap-silhouette} looks pretty weak, really.
+Tons of overlap between the clusters.
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{ap-silhouette.pdf}
+  \caption{AP silhouette plot, $k = 8$}
+  \label{fig:ap-silhouette}
+\end{figure}
+
+\subsection{Interpretation}
+
+Boxplot summary of clusters available in Figure~\ref{fig:ap-summaries}.
+\textbf{Discussion forthcoming.}
+
+\begin{figure}[ht]
+  \centering
+  \includegraphics[width=0.8\linewidth]{ap-summaries.pdf}
+  \caption{AP Boxplot Summaries}
+  \label{fig:ap-summaries}
+\end{figure}
+
 \section{Biclustering}

### 2015-06-25 R/affinity-propagation.R

@@ -1,0 +1,82 @@
+# LIBRARIES ====
+library(apcluster)
+library(tidyr)
+
+# PREPROCESSING ====
+source('./preprocessing.R')
+
+# CONSTANTS ====
+# TODO: Because this is used by this and kmeans-dtree
+# it should probably go in as a constant in preprocessing.R
+INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
+                 "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
+                 "nms_d6", "nms_d7", "nms_d8", "nms_d9",
+                 "tremor", "bradykin", "rigidity", "axial", "pigd")
+# save.plots for now since I don't know what else to plot!
+SAVE.PLOTS <- TRUE
+
+# CLUSTERING ====
+# Lambda - dampening factor - usually 0.9, higher takes longer to converge
+# but lets it converge in cases where it doesn't
+# q = 0 - smallest input preference, lowest number of clusters
+raw.filtered.apclus <- apcluster(negDistMat(r=2), raw.filtered,
+                                 details = TRUE, q = 0, lam = 0.98,
+                                 maxits=10000, convits=1000)
+nclust <- length(raw.filtered.apclus@clusters)
+apclusters <- raw.filtered.apclus@clusters
+cat("Found ", nclust, " clusters\n", sep="")
+
+# BIND TO ORIGINAL DATA ====
+
+unscaled.apclus <- raw.omitted[, INTERPRETED]
+unscaled.apclus$cluster <- 0
+for (i in 1:nclust) {
+  unscaled.apclus[apclusters[[i]], ]$cluster <- i
+}
+
+# WIDE -> LONG ====
+unscaled.apclus.long <- gather(unscaled.apclus, variable, measurement, age:pigd)
+
+# FACTOR CLUSTERS, ORDER BY BY CISITOT ====
+# FIXME: Is it bad that the clusters are only factored here rather than
+# in the above code block?
+cisitot <- unscaled.apclus.long[unscaled.apclus.long$variable == "cisitot", ]
+cisitot.means <- sapply(factor(1:nclust), function(i) {
+  mean(cisitot[cisitot$cluster == i, ]$measurement)
+})
+unscaled.apclus.long$cluster <- factor(unscaled.apclus.long$cluster,
+                                       levels = order(cisitot.means))
+
[…]

### 2015-06-26 R/kmeans-dtree.R
use nstart = 25 for wss scree plot

@@ -32,7 +32,7 @@ source('./preprocessing.R')
 
 wss <- (nrow(raw.filtered)-1)*sum(apply(raw.filtered, 2, var))
 for (i in 2:15) {
-  wss[i] <- sum(kmeans(raw.filtered, i)$withinss)
+  wss[i] <- sum(kmeans(raw.filtered, i, nstart = 25)$withinss)
 }
 
 plot(1:15, wss, type="b",

### 2015-06-26 R/affinity-propagation.R
add apcluster size table

@@ -73,6 +73,12 @@ if (SAVE.PLOTS) {
   dev.off()
 }
 
+# GET SIZES OF CLUSTERS ====
+cat("cluster,size\n")
+for (i in 1:nclust) {
+  cat(i, ",", length(apclusters[[i]]), "\n", sep="")
+}
+
 # Visualization
 # heatmap(raw.filtered.apclus)
 # plot(raw.filtered.apclus, raw.filtered)

### 2015-06-26 writeup/parkinsons.tex
add apcluster size table
Use full \linewidth for graphics
...since there's no reason to use 0.8

@@
   \centering
-  \includegraphics[width=0.8\linewidth]{pca-eigenvalues.png}
+  \includegraphics[width=\linewidth]{pca-eigenvalues.png}
   \caption{Scree test: eigenvalues by factor}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{kmeans-wss-error.png}
+  \includegraphics[width=\linewidth]{kmeans-wss-error.png}
   \caption{Scree test: WSS error by cluster size}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{gap-statistic.png}
+  \includegraphics[width=\linewidth]{gap-statistic.png}
   \caption{Gap statistic by cluster size}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{asw.png}
+  \includegraphics[width=\linewidth]{asw.png}
   \caption{Average silhouette width by cluster size}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{kmeans-silhouette-2.pdf}
+  \includegraphics[width=\linewidth]{kmeans-silhouette-2.pdf}
   \caption{$k$-means cluster silhouette plot, $k = 2$}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{kmeans-silhouette-3.pdf}
+  \includegraphics[width=\linewidth]{kmeans-silhouette-3.pdf}
   \caption{$k$-means cluster silhouette plot, $k = 3$}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{kmeans-silhouette-4.pdf}
+  \includegraphics[width=\linewidth]{kmeans-silhouette-4.pdf}
   \caption{$k$-means cluster silhouette plot, $k = 4$}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-unscaled-2.pdf}
+  \includegraphics[width=\linewidth]{dtree-kmeans-pruned-unscaled-2.pdf}
   \caption{Decision Tree from $k$-means clustering, 2 clusters}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-unscaled-3.pdf}
+  \includegraphics[width=\linewidth]{dtree-kmeans-pruned-unscaled-3.pdf}
   \caption{Decision Tree from $k$-means clustering, 3 clusters}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{dtree-kmeans-pruned-unscaled-4.pdf}
+  \includegraphics[width=\linewidth]{dtree-kmeans-pruned-unscaled-4.pdf}
   \caption{Decision Tree from $k$-means clustering, 4 clusters}
@@
 
+Package \texttt{apcluster} was used. Distance matrix was the negative euclidean
+squared distance ($r = 2$).
+
 AP with input preferences minimized ($q = 0$) resulted in 8 clusters.
@@
 iterations to 10000, and convergence iterations to 1000 failed to converge.
-Might need to try a longer term work.
+Might need to try a longer run.
 
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{ap-silhouette.pdf}
+  \includegraphics[width=\linewidth]{ap-silhouette.pdf}
   \caption{AP silhouette plot, $k = 8$}
@@
 
+Cluster sizes are available in Table~\ref{tab:ap-cluster-sizes}
+
+\begin{table}[ht]
+  \centering
+  \begin{tabular}{l|l}
+  Cluster & Size \\
+  \hline
+  1 & 54 \\
+  2 & 68 \\
+  3 & 25 \\
+  4 & 259 \\
+  5 & 102 \\
+  6 & 68 \\
+  7 & 185 \\
+  8 & 20 \\
+  \end{tabular}
+  \caption{AP Cluster Sizes}
+  \label{tab:ap-cluster-sizes}
+\end{table}
+
 Boxplot summary of clusters available in Figure~\ref{fig:ap-summaries}.
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{ap-summaries.pdf}
+  \includegraphics[width=\linewidth]{ap-summaries.pdf}
   \caption{AP Boxplot Summaries}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{biclust-16.pdf}
+  \includegraphics[width=\linewidth]{biclust-16.pdf}
   \caption{Biclustering $N = 16$}
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{bubbleplot-16.pdf}
+  \includegraphics[width=\linewidth]{bubbleplot-16.pdf}
   \caption{Bubbleplot $N = 16$}

### 2015-06-26 writeup/parkinsons.tex
Add hierarchical clustering, change ht to h
LaTeX warns for h but I feel like there's no problem...it's just an
indication that the figure must be placed on the [t]op of the next page

Right?

@@
 
-\begin{table}[ht]
+\begin{table}[h]
   \centering
@@
 
-\begin{figure}[ht]
+\begin{figure}[h]
   \centering
@@
 
+\clearpage
 \section{$k$-means}
@@
 % Sum WithinSS: 7084.183
-\begin{table}[ht]
+\begin{table}[h]
   \centering
@@
 % Root node error: 0.5236876
-\begin{table}[ht]
+\begin{table}[h]
   \centering
@@
 
+\clearpage
 \section{Affinity Propagation}
@@
 
+\clearpage
+\section{Hierarchical Clustering}
+
+\subsection{Clustering}
+
+Four dissimilarity methods were used with a euclidean distance matrix.
+Dendrograms available in Figure~\ref{fig:hc-dendrograms}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{hc-dendrograms.pdf}
+  \caption{Dendrograms}
+  \label{fig:hc-dendrograms}
+\end{figure}
+
+\subsection{Cutting Trees}
+\begin{table}[h]
+  \centering
+  \begin{tabular}{l|l|l|l}
+    Method & Condition & n & Figure \\
+    \hline
+    Complete & $k = 4$ & 4 (665/74/35/7) & ~\ref{fig:hc-summaries-complete-k4} \\
+    Complete & \texttt{dynamicTreeCut}\tablefootnote{Package \texttt{dynamicTreeCut} in R (Langfelder P,
+  Zhang B, Horvath S (2007)). Hybrid method, minimum cluster selection
+parameters} & 11 (7/270/138/83/79/49/46/39/35/35) &
+    ~\ref{fig:hc-summaries-complete-dynamic} \\
+    Ward & $k = 4$ & 4 (294/236/120/131) & ~\ref{fig:hc-summaries-ward-D-k4} \\
+    Ward & $h = 60$ & 6 (97/236/120/197/91/40) &
+    ~\ref{fig:hc-summaries-ward-D-h60} \\
+  \end{tabular}
+  \caption{Clusters from Tree Cutting}
+  \label{tab:tree-cutting}
+\end{table}
+
+\subsection{Interpretation}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{hc-summaries-complete-k4.pdf}
+  \caption{Using maximum (complete linkage) dissimilarity, cutting tree for $k = 4$}
+  \label{fig:hc-summaries-complete-k4}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{hc-summaries-complete-dynamic.pdf}
+  \caption{Using maximum (complete linkage) dissimilarity, cutting tree
+    dynamically}
+  \label{fig:hc-summaries-complete-dynamic}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{hc-summaries-ward-D-k4.pdf}
+  \caption{Using Ward (1963) dissimilarity, cutting tree for $k = 4$}
+  \label{fig:hc-summaries-ward-D-k4}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{hc-summaries-ward-D-h60.pdf}
+  \caption{Using Ward (1963) dissimilarity, cutting tree at $h = 60$}
+  \label{fig:hc-summaries-ward-D-h60}
+\end{figure}
+
 \subsection{Interpretation}

### 2015-06-26 R/hierarchical.R

@@ -1,0 +1,122 @@
+# LIBRARIES ====
+library(gclus)
+library(ggplot2)
+library(tidyr)
+library(dynamicTreeCut)
+
+# PREPROCESSING ====
+source('./preprocessing.R')
+
+# CONSTANTS ====
+# TODO: Because this is used by this and kmeans-dtree
+# it should probably go in as a constant in preprocessing.R
+INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
+                 "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
+                 "nms_d6", "nms_d7", "nms_d8", "nms_d9",
+                 "tremor", "bradykin", "rigidity", "axial", "pigd")
+SAVE.PLOTS = TRUE
+
+# COMPUTE DISTANCE MATRIX (EUCLIDEAN) ====
+hclust.dist <- dist(raw.filtered)
+
+# HCLUST - MAX, AVERAGE, WARD, WARD.D2 ====
+# Look at all 4 clustering methods side by side
+par(mfrow=n2mfrow(4))
+
+# Maximum (Complete Linkage)
+hcl.complete <- hclust(hclust.dist)
+plot(hcl.complete)
+
+# Average
+hcl.average <- hclust(hclust.dist, method = "average")
+plot(hcl.average)
+
+# Ward
+hcl.ward.D <- hclust(hclust.dist, method = "ward.D")
+plot(hcl.ward.D)
+
+# Ward D2 (squared distance)
+hcl.ward.D2 <- hclust(hclust.dist, method = "ward.D2")
+plot(hcl.ward.D2)
+
+if (SAVE.PLOTS) {
+  dev.copy(pdf, "../figures/hc-dendrograms.pdf")
+  dev.off()
+}
+
+# Reset
+par(mfrow=c(1, 1))
+
+# CUT TREES ====
+# Let's pick Max for now
+
+# 11 results in 4 clusters
+HEIGHT = 11
+K = 4
+
+# Complete linkage cuts
+# K = 4 cut
+hcl.complete.mem <- cutree(hcl.complete, k = K)
+
+# fixed height cut
+# hcl.complete.mem <- cutree(hcl.complete, h = HEIGHT)
+# dynamicTreeCut
+# Be advised the default minimum cluster size is 20
+# Hybrid (bottom up)
+# K = 11
+# hcl.complete.mem <- cutreeDynamic(hcl.complete, method = "hybrid",
+#                                   distM = as.matrix(hclust.dist),
[…]

### 2015-06-26 writeup/parkinsons.tex
Only remove NAs from interpreted features
This makes the number of data points rise from 781 to 901 participants.

Will need to rerun all figures and recalculate statistics, so I pushed
a tag up. Get ready!

@@
 \subsection{Dataset Description}
-951 subjects, 145 metrics, collected 15-4-2012. From Pablo Martinez Mart\'in.
-170 subjects with missing values (brought down to 781); these were removed
-automatically, even if the missing values were not included in the selected
-features below. This will need to be changed later on, by keeping those
-removed that still have all selected features and perhaps with some
-compensation for missing values.
+951 subjects, 145 metrics, collected 15-4-2012 from Pablo Martinez \'in. Only
+19 features used for clustering and/or interpretation.  50 subjects with
+missing values of the features to be used in clustering (brought down to 901).
+Imputation may be a good idea later on.
 

### 2015-06-26 R/preprocessing.R

@@ -25,6 +25,10 @@ MOTOR.SYMPTOMS = c("tremor",
                    "rigidity",
                    "axial",
                    "pigd")
+INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
+                 "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
+                 "nms_d6", "nms_d7", "nms_d8", "nms_d9",
+                 "tremor", "bradykin", "rigidity", "axial", "pigd")
 # Change this variable to change symptoms
 SYMPTOMS.TO.USE = ALL.SYMPTOMS
 
@@ -44,10 +48,12 @@ raw <- read.csv(DB_FILE)
 # Study is irrelevant
 raw$study <- NULL
 # Get rid of NAs (later may use some kind of missing value compensation method
-raw.omitted <- na.omit(raw)
+# This only brings it down to 901, rather than ~700 something!
+raw.omitted <- raw[, INTERPRETED]
+raw.omitted <- na.omit(raw.omitted)
 
 # raw with only select nms and motor symptoms
-raw.filtered <- raw.omitted[, SYMPTOMS.TO.USE]
+raw.filtered <- raw.omitted
 
 # DESCRIPTIVE STATISTICS (BEFORE STANDARDIZATION) ====
 # TODO - change raw.filtered to raw.omitted.filtered, or something like that

### 2015-06-26 R/preprocessing.R
rerun preprocessing graphs, pca

@@ -5,8 +5,9 @@ library(FactoMineR)
 library(psych)
 
 # GLOBAL CONSTANTS ====
-DB_FILE = "../data/DATABASE_NMS Burden levels_15-4-2012.csv"
-ALL.SYMPTOMS = c("nms_d1",
+SAVE.PREPROCESSING.PLOTS <- FALSE
+DB_FILE <- "../data/DATABASE_NMS Burden levels_15-4-2012.csv"
+ALL.SYMPTOMS <- c("nms_d1",
                  "nms_d2",
                  "nms_d3",
                  "nms_d4",
@@ -20,7 +21,7 @@ ALL.SYMPTOMS = c("nms_d1",
                  "rigidity",
                  "axial",
                  "pigd")
-MOTOR.SYMPTOMS = c("tremor",
+MOTOR.SYMPTOMS <- c("tremor",
                    "bradykin",
                    "rigidity",
                    "axial",
@@ -30,7 +31,7 @@ INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
                  "nms_d6", "nms_d7", "nms_d8", "nms_d9",
                  "tremor", "bradykin", "rigidity", "axial", "pigd")
 # Change this variable to change symptoms
-SYMPTOMS.TO.USE = ALL.SYMPTOMS
+SYMPTOMS.TO.USE <- ALL.SYMPTOMS
 
 # UTILITY FUNCTIONS ====
 splitdf <- function(dataframe, seed=NULL, trainfrac=0.7) {
@@ -67,11 +68,20 @@ raw.filtered <- as.data.frame(scale(raw.filtered))
 # PCA (Note: not helpful for identifying specific factors) ====
 par(mfrow = c(1, 2))
 pca <- PCA(raw.filtered)
+if (SAVE.PREPROCESSING.PLOTS) {
+  dev.copy(pdf, "../figures/pca.pdf")
+  dev.off()
+}
 # pca$eig
 # pca$var$coord
 # head(pca$ind$coord)
+par(mfrow = c(1, 1))
 plot(x = 1:length(rownames(pca$eig)), y = (pca$eig$eigenvalue), type="b",
-     xlab="Factor", ylab="Eigenvalue", xaxp=c(0, 14, 14))
+     xlab="Factor", ylab="Eigenvalue", xaxp=c(0, 19, 19))
+if (SAVE.PREPROCESSING.PLOTS) {
[…]

### 2015-06-26 R/kmeans-dtree.R
rerun determining k graphs, png -> pdf

@@ -22,6 +22,7 @@ INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
 # kmeans analysis
 SAVE.DTREES = FALSE
 SAVE.BOXPLOTS = FALSE
+SAVE.EXPLORE.PLOTS <- FALSE
 # TODO: Make k = 2, 3, 4 modifiable via constant
 
 # LOAD DATA ====
@@ -37,6 +38,10 @@ for (i in 2:15) {
 
 plot(1:15, wss, type="b",
      xlab="Number of Clusters", ylab="Within groups sum of squares")
+if (SAVE.EXPLORE.PLOTS) {
+  dev.copy(pdf, "../figures/kmeans-wss-error.pdf")
+  dev.off()
+}
 
 # TODO: BIC ====
 # See http://stackoverflow.com/questions/15376075/cluster-analysis-in-r-determine-the-optimal-number-of-clusters
@@ -65,10 +70,18 @@ for (i in 1:15) {
   pam_sils <- c(pam_sils, pam(raw.filtered, k=i)$silinfo$avg.width)
 }
 plot(x=1:14, y=pam_sils, xlab="Clusters", ylab="Average Silhouette Width", type="b")
+if (SAVE.EXPLORE.PLOTS) {
+  dev.copy(pdf, "../figures/asw.pdf")
+  dev.off()
+}
 
 # GAP STATISTIC ESTIMATION ====
 gaps <- clusGap(raw.filtered, kmeans, 14, B = 100)
 plot(x=1:14, y=gaps$Tab[, "gap"], xlab="Clusters", ylab="Gap Statitsic", type="b")
+if (SAVE.EXPLORE.PLOTS) {
+  dev.copy(pdf, "../figures/gap-statistic.pdf")
+  dev.off()
+}
 
 # INITIAL KMEANS CLUSTERING ====
 splits = splitdf(raw.filtered)

### 2015-06-26 R/kmeans-dtree.R
rerun kmeans dtrees, figures

@@ -20,9 +20,10 @@ INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
 # Remember - PDFs can vary even if the (seeded) clusters don't
 # So this should be false unless I've changed something about the
 # kmeans analysis
-SAVE.DTREES = FALSE
-SAVE.BOXPLOTS = FALSE
+# explore plots is the determining clusters plots
 SAVE.EXPLORE.PLOTS <- FALSE
+SAVE.DTREES <- FALSE
+SAVE.BOXPLOTS <- FALSE
 # TODO: Make k = 2, 3, 4 modifiable via constant
 
 # LOAD DATA ====
@@ -278,7 +279,8 @@ lnames <- c("clusters2", "clusters3", "clusters4")
 names(labeled) <- lnames
 # I use 145 indexing because I'm not sure if tail() copies
 for (name in lnames) {
-  names(labeled[[name]])[145] <- "cluster"
+  # A convoluted way to get the last element - tail was throwing an error
+  names(labeled[[name]])[length(names(labeled[[name]]))] <- "cluster"
 }
 
 # COMBINE LABELS WITH ORIGINAL DATA ====

### 2015-06-26 writeup/parkinsons.tex
rerun preprocessing graphs, pca
rerun determining k graphs, png -> pdf
update kmeans statistics after rerun

@@
          \hline
-nms\_d1&   1.76&  3.32&   0-24 \\
-nms\_d2&   8.71&  8.76&   0-48 \\
-nms\_d3&   8.70& 11.83&   0-60 \\
-nms\_d4&   1.65&  3.94&   0-33 \\
-nms\_d5&   5.22&  7.44&   0-36 \\
-nms\_d6&   5.67&  6.92&   0-36 \\
-nms\_d7&   8.02&  9.09&   0-36 \\
-nms\_d8&   3.57&  5.97&   0-24 \\
-nms\_d9&   6.99&  7.74&   0-48 \\
-tremor&   2.59&  2.63&   0-12 \\
-bradykin& 2.49&  1.39&   0-6 \\
-rigidity& 2.34&  1.36&   0-6 \\
-axial&    3.28&  2.75&   0-12 \\
-pigd&     3.36&  2.77&   0-12 \\
+nms\_d1&   1.73&  3.35&   0-24 \\
+nms\_d2&   8.75&  8.70&   0-48 \\
+nms\_d3&   8.68& 11.55&   0-60 \\
+nms\_d4&   1.64&  3.86&   0-33 \\
+nms\_d5&   5.42&  7.43&   0-36 \\
+nms\_d6&   5.53&  6.79&   0-36 \\
+nms\_d7&   8.08&  8.94&   0-36 \\
+nms\_d8&   3.52&  5.97&   0-24 \\
+nms\_d9&   7.13&  7.79&   0-48 \\
+tremor&   2.59&  2.58&   0-12 \\
+bradykin& 2.40&  1.41&   0-6 \\
+rigidity& 2.24&  1.36&   0-6 \\
+axial&    3.25&  2.68&   0-12 \\
+pigd&     3.31&  2.71&   0-12 \\
   \end{tabular}
@@
 May not be useful? If we're trying to identify \emph{clinically} relevant
-features, merging them may not be a good idea.
+features, merging them may not be a good idea. Regardless, Figure~\ref{fig:pca}
+shows results of preliminary PCA.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{pca-eigenvalues.png}
+  \includegraphics[width=\linewidth]{pca.pdf}
+  \caption{PCA Analysis}
+  \label{fig:pca}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{pca-eigenvalues.pdf}
   \caption{Scree test: eigenvalues by factor}
@@
 
-Figure~\ref{fig:pca-eigenvalues} shows scree test elbow occurs around 2 or 3. Also, eigenvalues 1 and $2 > 1$, while 3
-is around .9
+Figure~\ref{fig:pca-eigenvalues} shows scree test elbow occurs around 2 or 2 or
+.4 Also, eigenvalues $1-5 > 1$.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{kmeans-wss-error.png}
+  \includegraphics[width=\linewidth]{kmeans-wss-error.pdf}
   \caption{Scree test: WSS error by cluster size}
@@
   \centering
-  \includegraphics[width=\linewidth]{gap-statistic.png}
+  \includegraphics[width=\linewidth]{gap-statistic.pdf}
   \caption{Gap statistic by cluster size}
@@
   \centering
-  \includegraphics[width=\linewidth]{asw.png}
+  \includegraphics[width=\linewidth]{asw.pdf}
   \caption{Average silhouette width by cluster size}
@@
 % ================================
-% Sizes: 201 580
-% WithinSS: 4248.585 4132.434
-% Sum WithinSS: 8381.019
+% Sizes: 229 672
+% WithinSS: 6117.986 7695.411
+% Sum WithinSS: 13813.4
 % CLUSTERS: 3
 % ================================
-% Sizes: 420 231 130
-% WithinSS: 2618.368 1973.82 3076.542
-% Sum WithinSS: 7668.73
+% Sizes: 333 134 434
+% WithinSS: 4669.037 4009.387 4153.892
+% Sum WithinSS: 12832.32
 % CLUSTERS: 4
 % ================================
-% Sizes: 61 372 145 203
-% WithinSS: 1481.25 1845.389 2147.988 1609.555
-% Sum WithinSS: 7084.183
+% Sizes: 79 394 275 153
+% WithinSS: 2366.585 3356.709 3454.142 2879.508
+% Sum WithinSS: 12056.94
 \begin{table}[h]
@@
     \hline
-    2 & 201/580 & 4248.585/4132.434 & 8381.019 \\
-    3 & 420/231/130 & 2618.368/1973.82/3076.542 &  7668.73 \\
-    4 & 61/372/145/203 & 1481.25/1845.389/2147.988/1609.555 & 7084.183
+    2 & 229/672 & 6118/7695 & 13813 \\
+    3 & 333/134/434 & 4669/40009/4154 & 12832 \\
+    4 & 79/394/275/153 & 2367/3357/3454/2880  & 12057 \\
   \end{tabular}
@@
 % ================================
-% Complexity Parameter: 0.03482587
-% 10-fold CV error: 0.134443
-% Root node error: 0.2573624
+% Complexity Parameter: 0.02183406
+% 10-fold CV error: 0.1132075
+% Root node error: 0.254162
 % CLUSTERS: 3
 % ================================
-% Complexity Parameter: 0.01
-% 10-fold CV error: 0.1946223
-% Root node error: 0.4622279
+% Complexity Parameter: 0.01070664
+% 10-fold CV error: 0.190899
+% Root node error: 0.518313
 % CLUSTERS: 4
@@
 % Complexity Parameter: 0.01
-% 10-fold CV error: 0.2483995
-% Root node error: 0.5236876
+% 10-fold CV error: 0.2552719
+% Root node error: 0.5627081
 \begin{table}[h]
@@
     \hline
-    2 & 0.0348 & 0.134 & axial $\geq$ 0.44 & 0.257 & Figure~\ref{fig:kmeans-dtree-2} \\
-    3 & 0.0100 & 0.194 & bradykin $<$ 0.0041 & 0.462 & Figure~\ref{fig:kmeans-dtree-3} \\
-    4 & 0.0100 & 0.248 & bradykin $<$ 0.0041 & 0.523 & Figure~\ref{fig:kmeans-dtree-4} \\
+    2 & 0.0218 & 0.113 & axial $\geq$ 4.5 & 0.254 & Figure~\ref{fig:kmeans-dtree-2} \\
+    3 & 0.0107 & 0.191 & pigd $\geq$ 2.5 & 0.518 & Figure~\ref{fig:kmeans-dtree-3} \\
+    4 & 0.0100 & 0.255 & pigd $<$ 2.5 & 0.563 & Figure~\ref{fig:kmeans-dtree-4} \\
   \end{tabular}

### 2015-06-26 R/biclustering.R
rerun biclustering plots

@@ -10,13 +10,22 @@ source('./preprocessing.R')
 # Note - N can't be more than 16 for par, will need to fix that later
 # (multiple plots/pagination)
 N <- 16
+SAVE.PLOTS <- FALSE
 
 # BICLUSTERING ====
 # Bimax algorithm
 bcl <- biclust(as.matrix(raw.filtered), method=BCBimax(), minr=2, minc=2, number=N)
+if (SAVE.PLOTS) {
+  dev.copy(pdf, paste("../figures/biclust-", N, ".pdf", sep=""))
+  dev.off()
+}
 
 # BUBBLEPLOT (?) ====
 bubbleplot(raw.filtered, bcl)
+if (SAVE.PLOTS) {
+  dev.copy(pdf, paste("../figures/biclust-bubbleplot-", N, ".pdf", sep=""))
+  dev.off()
+}
 
 # CLUSTER HEATMAPS ====
 
@@ -25,3 +34,7 @@ par(mfrow = n2mfrow(N))
 for (i in 1:N) {
   drawHeatmap(raw.filtered, bicResult=bcl, number=i)
 }
+# if (SAVE.PLOTS) {
+#   dev.copy(pdf, paste("../figures/biclust-heatmap-", N, ".pdf", sep=""))
+#   dev.off()
+# }

### 2015-06-26 R/affinity-propagation.R
update ap statistics to match rerun

@@ -13,7 +13,7 @@ INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
                  "nms_d6", "nms_d7", "nms_d8", "nms_d9",
                  "tremor", "bradykin", "rigidity", "axial", "pigd")
 # save.plots for now since I don't know what else to plot!
-SAVE.PLOTS <- TRUE
+SAVE.PLOTS <- FALSE
 
 # CLUSTERING ====
 # Lambda - dampening factor - usually 0.9, higher takes longer to converge

### 2015-06-26 writeup/parkinsons.tex
rerun biclustering plots
update ap statistics to match rerun
rerun plots, data for hc

@@
     \hline
-    Complete & $k = 4$ & 4 (665/74/35/7) & ~\ref{fig:hc-summaries-complete-k4} \\
+    Complete & $k = 4$ & 4 (790/81/18/12) & ~\ref{fig:hc-summaries-complete-k4} \\
     Complete & \texttt{dynamicTreeCut}\tablefootnote{Package \texttt{dynamicTreeCut} in R (Langfelder P,
   Zhang B, Horvath S (2007)). Hybrid method, minimum cluster selection
-parameters} & 11 (7/270/138/83/79/49/46/39/35/35) &
+parameters} & 13 (255/99/77/64/62/58/56/46/44/41/37/32/30) &
     ~\ref{fig:hc-summaries-complete-dynamic} \\
-    Ward & $k = 4$ & 4 (294/236/120/131) & ~\ref{fig:hc-summaries-ward-D-k4} \\
-    Ward & $h = 60$ & 6 (97/236/120/197/91/40) &
-    ~\ref{fig:hc-summaries-ward-D-h60} \\
+    Ward & $k = 4$ & 4 (200/237/263/201) & ~\ref{fig:hc-summaries-ward-D-k4} \\
+    Ward & $h = 100$ & 3 (437/263/201) &
+    ~\ref{fig:hc-summaries-ward-D-h100} \\
   \end{tabular}
@@
   \centering
-  \includegraphics[width=\linewidth]{hc-summaries-ward-D-h60.pdf}
-  \caption{Using Ward (1963) dissimilarity, cutting tree at $h = 60$}
-  \label{fig:hc-summaries-ward-D-h60}
+  \includegraphics[width=\linewidth]{hc-summaries-ward-D-h100.pdf}
+  \caption{Using Ward (1963) dissimilarity, cutting tree at $h = 100$}
+  \label{fig:hc-summaries-ward-D-h100}
 \end{figure}
@@
   \hline
-  1 & 54 \\
-  2 & 68 \\
-  3 & 25 \\
-  4 & 259 \\
-  5 & 102 \\
-  6 & 68 \\
-  7 & 185 \\
-  8 & 20 \\
+  1 & 63 \\
+  2 & 53 \\
+  3 & 85 \\
+  4 & 122 \\
+  5 & 48 \\
+  6 & 126 \\
+  7 & 123 \\
+  8 & 102 \\
+  9 & 166 \\
+  10 & 13 \\
   \end{tabular}
@@
   \centering
-  \includegraphics[width=\linewidth]{bubbleplot-16.pdf}
+  \includegraphics[width=\linewidth]{biclust-bubbleplot-16.pdf}
   \caption{Bubbleplot $N = 16$}

### 2015-06-26 R/hierarchical.R

@@ -14,7 +14,7 @@ INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
                  "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
                  "nms_d6", "nms_d7", "nms_d8", "nms_d9",
                  "tremor", "bradykin", "rigidity", "axial", "pigd")
-SAVE.PLOTS = TRUE
+SAVE.PLOTS <- FALSE
 
 # COMPUTE DISTANCE MATRIX (EUCLIDEAN) ====
 hclust.dist <- dist(raw.filtered)
@@ -51,7 +51,7 @@ par(mfrow=c(1, 1))
 # Let's pick Max for now
 
 # 11 results in 4 clusters
-HEIGHT = 11
+HEIGHT = 100
 K = 4
 
 # Complete linkage cuts
@@ -73,8 +73,8 @@ hcl.complete.mem <- cutree(hcl.complete, k = K)
 #                                   deepSplit = FALSE)
 
 # Ward (1963) linkage cuts
-# hcl.complete.mem <- cutree(hcl.ward.D, h = 60)
 # hcl.complete.mem <- cutree(hcl.ward.D, k = K)
+# hcl.complete.mem <- cutree(hcl.ward.D, h = HEIGHT)
 
 # Then print out the number of clusters and frequencies
 nclust <- length(unique(hcl.complete.mem))
@@ -94,8 +94,8 @@ labeled.complete.long <- gather(labeled.complete, variable, measurement, age:pig
 
 # ORDER BY INCREASING CISITOT ====
 # We need to shift by 1 if using dynamic, since dynamicTreeCut starts at 0.
-DYNAMIC = FALSE
-# DYNAMIC = TRUE
+DYNAMIC <- FALSE
+# DYNAMIC <- TRUE
 cisitot <- labeled.complete.long[labeled.complete.long$variable == "cisitot", ] 
 anyNA(cisitot)
 cisitot.means <- sapply(1:nclust, function(i) {
@@ -116,7 +116,7 @@ print(p)
 
 if (SAVE.PLOTS) {
   ggsave("../figures/hc-summaries-complete-k4.pdf")
-  # ggsave("../figures/hc-summaries-ward-D-k4.pdf")
-  # ggsave("../figures/hc-summaries-ward-D-h60.pdf")
   # ggsave("../figures/hc-summaries-complete-dynamic.pdf")
[…]

### 2015-06-26 R/biclustering.R
draw the correct bcl heatmap

@@ -15,10 +15,6 @@ SAVE.PLOTS <- FALSE
 # BICLUSTERING ====
 # Bimax algorithm
 bcl <- biclust(as.matrix(raw.filtered), method=BCBimax(), minr=2, minc=2, number=N)
-if (SAVE.PLOTS) {
-  dev.copy(pdf, paste("../figures/biclust-", N, ".pdf", sep=""))
-  dev.off()
-}
 
 # BUBBLEPLOT (?) ====
 bubbleplot(raw.filtered, bcl)
@@ -34,7 +30,7 @@ par(mfrow = n2mfrow(N))
 for (i in 1:N) {
   drawHeatmap(raw.filtered, bicResult=bcl, number=i)
 }
-# if (SAVE.PLOTS) {
-#   dev.copy(pdf, paste("../figures/biclust-heatmap-", N, ".pdf", sep=""))
-#   dev.off()
-# }
+if (SAVE.PLOTS) {
+  dev.copy(pdf, paste("../figures/biclust-heatmaps-", N, ".pdf", sep=""))
+  dev.off()
+}

### 2015-06-26 R/kmeans-dtree.R
change k in silhouette plot titles

@@ -134,7 +134,7 @@ kmeans.dtree <- function(data, data.unscaled, k, save = FALSE, seed = 911) {
   set.seed(seed)
   cl <- kmeans(splits$trainset, k, nstart = 25)
   # Visualize clustering with clusplot
-  clusplot(raw.filtered, cl$cluster, main = paste("Silhouette plot k =", 4))
+  clusplot(raw.filtered, cl$cluster, main = paste("Silhouette plot k =", k))
   if (save) {
     # FIXME: This save boolean flag is SAVE.DTREES outside of this function
     # which is slightly ambiguous (these are silhouette plots!)

### 2015-07-15 R/bnets.R
bayesian nets draft

@@ -1,5 +1,22 @@
 # Libraries ====
 library(bnlearn)
-library(gRain)
+library(gRain) # Note - depends on biocLite graph and RBGL packages
+
+# Preprocessing ====
+source('./preprocessing.R')
+# Need stuff from kmeans dtree - so run that
+
+clus4.numclus <- clus4.wide
+# Convert this to numeric
+clus4.numclus$cluster <- as.numeric(clus4.wide$cluster)
+# I don't know why this is required
+clus4.df <- as.data.frame(as.matrix(clus4.numclus))
+head(clus4.df)
 
 # Learning a bayesian network ====
+clus4.iamb <- iamb(as.data.frame(as.matrix(clus4.numclus)))
+# Tons of warnings
+plot(clus4.iamb)
+mb(clus4.iamb, 'cluster')
+plot(hc(clus4.df))
+mb(hc(clus4.df), 'cluster')

### 2015-07-15 writeup/parkinsons.tex
draw the correct bcl heatmap
fix name typo
add feature ranking, start difference tests

@@
 \subsection{Dataset Description}
-951 subjects, 145 metrics, collected 15-4-2012 from Pablo Martinez \'in. Only
+951 subjects, 145 metrics, collected 15-4-2012 from Pablo Martinez Mart\'in. Only
 19 features used for clustering and/or interpretation.  50 subjects with
@@
 
+\subsubsection{Statistical Significance Tests, $k = 4$}
+Using one-way ANOVA for multiple means, we reject the null
+hypothesis that the means are the same with $p < 0.05$ for every variable
+\emph{except} \texttt{pdonset}.
+
+\subsubsection{Ranked Features}
+
+\begin{table}[h]
+  \centering
+  \caption{Features ranked by information gain}
+  \label{tab:info_gain}
+  \begin{tabular}{l|l}
+    variable & information gain \\
+    \hline
+    axial      & 0.20640691 \\
+    cisitot      & 0.20008571 \\
+    pigd      & 0.18193982 \\
+    nms\_d2      & 0.13178572 \\
+    nms\_d9      & 0.12116024 \\
+    bradykin      & 0.11966097 \\
+    nms\_d3      & 0.09421859 \\
+    rigidity      & 0.09260628 \\
+    nms\_d5      & 0.07579997 \\
+    nms\_d4      & 0.07438784 \\
+    nms\_d6      & 0.06620599 \\
+    nms\_d7      & 0.05574956 \\
+    nms\_d1      & 0.05509838 \\
+    tremor      & 0.04140473 \\
+    nms\_d8      & 0.03786173 \\
+    durat\_pd      & 0.02794420 \\
+    age      & 0.00000000 \\
+    sex      & 0.00000000 \\
+    pdonset      & 0.00000000 \\
+  \end{tabular}
+\end{table}
+
 \clearpage
@@
   \centering
-  \includegraphics[width=\linewidth]{biclust-16.pdf}
-  \caption{Biclustering $N = 16$}
-  \label{fig:biclust-16}
+  \includegraphics[width=\linewidth]{biclust-heatmaps-16.pdf}
+  \caption{Biclustering heatmaps $N = 16$}
+  \label{fig:biclust-heatmaps-16}
 \end{figure}

### 2015-07-15 R/kmeans-dtree.R

@@ -419,9 +419,13 @@ print(p)
 # Significant differneces between population means ====
 # Try axial (most demonstrative feature)
 clus4.wide <- clusters.raw[["4"]]
+# Why isn't this already a factor? Really confused
+# Because it wasn't set in clusters.raw - if this is a bad thing lmk
+clus4.wide$cluster <- as.factor(clus4.wide$cluster)
 # Assuming 1st column is cluster (which it should be)
 oneways <- lapply(colnames(clus4.wide[, -1]), function(col) {
-  oneway.test(substitute(i ~ cluster, list(i = as.name(col))), data = clus4.wide)
+  fm <- substitute(i ~ cluster, list(i = as.name(col)))
+  oneway.test(fm, clus4.wide)
 })
 for (test in oneways) {
   if (test$p.value < 0.05) {
@@ -431,6 +435,39 @@ for (test in oneways) {
     cat(test$data.name, '\n')
   }
 }
+
+# Tukey's HSD test ====
+
+tukeys <- lapply(colnames(clus4.wide[, -1]), function(col) {
+  # Doesn't work the oneway way for some reason!
+  fm <- as.formula(paste(col, '~ cluster'))
+  TukeyHSD(aov(fm, clus4.wide))$cluster
+})
+names(tukeys) <- colnames(clus4.wide[, -1])
+
+for (var in names(tukeys)) {
+  test <- tukeys[[var]]
+  # Check for nonsignificant, since there are more significant
+  sigs <- test[test[, "p adj"] > 0.05, ]
+  if (!identical(logical(0), as.logical(sigs))) { 
+    # Super hacky to figure out if null matrix without type error
+    cat(var, ' insignificant differences', ':', '\n', sep='')
+    if (class(sigs) == 'numeric') {  # If returned just a single vector, can't do anything
+      # Print em all, don't know how to get around this
+      print(test)
+    }
+    print(sigs)
+  } else {
+    cat(var, 'nothing', '\n')
+  }
+}
+
+# Rank by most informative features ====
+features.ranked <- information.gain(cluster ~ ., clus4.wide)
+# Add to a separate column because rownames are annoying
+features.ranked$variable <- rownames(features.ranked)
+rownames(features.ranked) <- NULL
+features.ranked[with(features.ranked, order(-attr_importance)), c('variable', 'attr_importance')]
 
 # NOTES ====
 # Note - with clustering, indeed the main feature (root of the tree) doesn't carry much

### 2015-07-16 R/biclustering.R
add exploration, improved biclusts, separate clus4.wide

@@ -16,10 +16,36 @@ raw.nopdonset.unscaled$pdonset <- NULL
 N <- 4
 SAVE.PLOTS <- FALSE
 
+# Trim this ====
+nmses <- sapply(1:9, function(i) paste('nms_d', i, sep=''))
+extras <- c('age', 'durat_pd', 'cisitot')
+raw.nms <- raw.nopdonset[, c(nmses, extras)]
+raw.nms.unscaled <- raw.nopdonset.unscaled[, c(nmses, extras)]
+raw.motor <- raw.nopdonset[, c('rigidity', 'axial', 'bradykin', 'tremor', 'pigd')]
+raw.motor.unscaled <- raw.nopdonset.unscaled[, c('rigidity', 'axial', 'bradykin', 'tremor', 'pigd')]
+
 # BICLUSTERING ====
 # Bimax algorithm
-bcl <- biclust(as.matrix(raw.nopdonset), method=BCPlaid(), iter.startup = 50, iter.layer = 50)
-bcls <- bicluster(raw.nopdonset.unscaled, bcl)
+set.seed(0)
+bcl <- biclust(as.matrix(raw.nms.unscaled), method=BCCC())
+bcls <- bicluster(raw.nms.unscaled, bcl)
+for (i in 1:100) {
+  cat(i, '\n')
+  varname <- paste('Bicluster', i, sep='')
+  bc <- bcls[[varname]]
+  plot(bc)
+  x <- readline()
+  corrplot(cor(bc), method = 'number')
+  x <- readline()
+  if (x == 'q') break
+}
+plot(bcls$Bicluster7)
+drawHeatmap(raw.nms.unscaled, bicResult = bcl, number = 3)
+for (bc in bcls) {
+  cat('dim:', dim(bc), '\n')
+}
+
+# TODO: CORRPLOT
 
 # Transform pdonset to number of years w/ parkinsons?
 # Then we want to look at some kind of comparison between NONMOTOR...

### 2015-07-16 R/exploration.R

@@ -1,0 +1,66 @@
+# Libraries ====
+library(corrplot)
+library(ggplot2)
+
+# Need clus4.wide from kmeans.dtree (the one with the class modified)
+
+# Random exploration ====
+c1 <- clus4.wide[clus4.wide$class == 1, ]
+c2 <- clus4.wide[clus4.wide$class == 2, ]
+c3 <- clus4.wide[clus4.wide$class == 3, ]
+c4 <- clus4.wide[clus4.wide$class == 4, ]
+
+# Plot w/o cluster
+par(mfrow=c(2, 2))
+corrplot(cor(c2[, -20]), method = 'pie', title = 'Correlation 2', diag = F, type = 'lower')
+corrplot(cor(c4[, -20]), method = 'pie', title = 'Correlation 4', diag = F, type = 'lower')
+corrplot(cor(c3[, -20]), method = 'pie', title = 'Correlation 3', diag = F, type = 'lower')
+corrplot(cor(c1[, -20]), method = 'pie', title = 'Correlation 1', diag = F, type = 'lower')
+par(mfrow=c(1, 1))
+
+par(mfrow=c(2, 2))
+
+# Check out relationship between bradykinesia and cisitot -
+# Correlation in c1 but not c2, c3, c4
+clus4.wide$class <- factor(clus4.wide$class, levels = c(2, 4, 3, 1))
+ggplot(clus4.wide, aes(x = bradykin, y = cisitot)) +
+  facet_wrap( ~ class, scales = 'free') +
+  geom_point() +
+  stat_smooth(method = 'lm') +
+  theme_bw()
+
+# Same for rigidity - more severe PD has larger dynamics of
+# brady/rigidity expression?
+ggplot(clus4.wide, aes(x = rigidity, y = cisitot)) +
+  facet_wrap( ~ class, scales = 'free') +
+  geom_point() +
+  stat_smooth(method = 'lm') +
+  theme_bw()
+
+# Check chisq independence of sex and nms_d8 for c1 (severe group)
+# Being male or female affects sexual function reports, in general
+# (Assuming male here)
+chisq.test(c2$rigidity, c2$cisitot, simulate.p.value = T)
+chisq.test(c4$pdonset, c4$age, simulate.p.value = T)
+cor.test(c4$bradykin, c4$durat_pd)
+
+
+# Bayes nets ===
[…]

### 2015-07-16 R/kmeans-dtree.R

@@ -418,14 +418,14 @@ print(p)
 
 # Significant differneces between population means ====
 # Try axial (most demonstrative feature)
-clus4.wide <- clusters.raw[["4"]]
+clus4.wide.st <- clusters.raw[["4"]]
 # Why isn't this already a factor? Really confused
 # Because it wasn't set in clusters.raw - if this is a bad thing lmk
-clus4.wide$cluster <- as.factor(clus4.wide$cluster)
+clus4.wide.st$cluster <- as.factor(clus4.wide.st$cluster)
 # Assuming 1st column is cluster (which it should be)
-oneways <- lapply(colnames(clus4.wide[, -1]), function(col) {
+oneways <- lapply(colnames(clus4.wide.st[, -1]), function(col) {
   fm <- substitute(i ~ cluster, list(i = as.name(col)))
-  oneway.test(fm, clus4.wide)
+  oneway.test(fm, clus4.wide.st)
 })
 for (test in oneways) {
   if (test$p.value < 0.05) {
@@ -438,12 +438,12 @@ for (test in oneways) {
 
 # Tukey's HSD test ====
 
-tukeys <- lapply(colnames(clus4.wide[, -1]), function(col) {
+tukeys <- lapply(colnames(clus4.wide.st[, -1]), function(col) {
   # Doesn't work the oneway way for some reason!
   fm <- as.formula(paste(col, '~ cluster'))
-  TukeyHSD(aov(fm, clus4.wide))$cluster
+  TukeyHSD(aov(fm, clus4.wide.st))$cluster
 })
-names(tukeys) <- colnames(clus4.wide[, -1])
+names(tukeys) <- colnames(clus4.wide.st[, -1])
 
 for (var in names(tukeys)) {
   test <- tukeys[[var]]

### 2015-07-16 writeup/abbreviated/abbreviated.tex
save figures for an abbreviated writeup

--- 2015-07-15 writeup/parkinsons.tex
@@
 % \rhead{CSCI339 Term Project}
-\graphicspath{ {../figures/} }
-% Or this, if run from main folder
-% \graphicspath{ {./figures/} }
+% One more .. since we're in the abbreviated subdir
+\graphicspath{ {../../figures/abbreviated/} {../../figures/} }
 
@@
 \title{Cluster Analysis: Identifying Parkinson's Disease Subtypes}
-\date{Wednesday, June 10}
-\author{Jesse Mu}
 \maketitle
@@
 
-\subsection{Dimensionality Reduction: PCA}
+\section{$k$-means}
 
-May not be useful? If we're trying to identify \emph{clinically} relevant
-features, merging them may not be a good idea. Regardless, Figure~\ref{fig:pca}
-shows results of preliminary PCA.
+$k$-means clustering with $k = 4$ was tried. $k = 2, 3$ provided models that
+were too simplistic. $k = 5$ did not provide any new information, but rather
+just fragmented existing groups.
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{pca.pdf}
-  \caption{PCA Analysis}
-  \label{fig:pca}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{pca-eigenvalues.pdf}
-  \caption{Scree test: eigenvalues by factor}
-  \label{fig:pca-eigenvalues}
-\end{figure}
-
-Figure~\ref{fig:pca-eigenvalues} shows scree test elbow occurs around 2 or 2 or
-.4 Also, eigenvalues $1-5 > 1$.
-
-\clearpage
-\section{$k$-means}
-\subsection{Identifying optimal number of clusters}
-
-\subsubsection{WSS Error Scree Test}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{kmeans-wss-error.pdf}
-  \caption{Scree test: WSS error by cluster size}
-  \label{fig:kmeans-wss-error}
-\end{figure}
-
-Figure~\ref{fig:kmeans-wss-error} shows no optimal elbow in scree test! Maybe 2-3?
-
-\subsubsection{Gap Statistic}
-
-Optimal cluster is the local maximum of the gap statistic, but it appears to be
-consistently increasing in Figure~\ref{fig:gap-statistic}.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{gap-statistic.pdf}
-  \caption{Gap statistic by cluster size}
-  \label{fig:gap-statistic}
-\end{figure}
-
-\subsubsection{Average Silhouette Width}
-
-Figure~\ref{fig:asw} shows average silhouette width as being consistently under
-0.25 for all clusters, implying the data is not well structured.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{asw.pdf}
-  \caption{Average silhouette width by cluster size}
-  \label{fig:asw}
-\end{figure}
-
-% \subsubsection{\texttt{NbClust} package}
-
-\subsection{Cluster statistics}
-% CLUSTERS: 2
-% ================================
-% Sizes: 229 672
-% WithinSS: 6117.986 7695.411
-% Sum WithinSS: 13813.4
-% CLUSTERS: 3
-% ================================
-% Sizes: 333 134 434
-% WithinSS: 4669.037 4009.387 4153.892
-% Sum WithinSS: 12832.32
-% CLUSTERS: 4
-% ================================
-% Sizes: 79 394 275 153
-% WithinSS: 2366.585 3356.709 3454.142 2879.508
-% Sum WithinSS: 12056.94
 \begin{table}[h]
   \centering
-  \begin{tabular}{l|l|l|l}
-    $k$ & $n$ & Within SS & sum(Within SS) \\
+  \caption{Cluster statistics}
+  \label{tab:cluster_stats}
+  \begin{tabular}{l|l}
+    Cluster & $n$ \\
     \hline
-    2 & 229/672 & 6118/7695 & 13813 \\
-    3 & 333/134/434 & 4669/40009/4154 & 12832 \\
-    4 & 79/394/275/153 & 2367/3357/3454/2880  & 12057 \\
+    1 & 79 \\
+    2 & 394 \\
+    3 & 275 \\
+    4 & 153 \\
   \end{tabular}
-  \caption{Cluster statistics}
-  \label{tab:cluster-statistics}
 \end{table}
 
-\subsection{Silhouette plots}
-
-Available in
-Figures~\ref{fig:kmeans-silhouette-2},~\ref{fig:kmeans-silhouette-3},
-and~\ref{fig:kmeans-silhouette-4}. Note: constructed with standardized
-$z$-score data.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{kmeans-silhouette-2.pdf}
-  \caption{$k$-means cluster silhouette plot, $k = 2$}
-  \label{fig:kmeans-silhouette-2}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{kmeans-silhouette-3.pdf}
-  \caption{$k$-means cluster silhouette plot, $k = 3$}
-  \label{fig:kmeans-silhouette-3}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{kmeans-silhouette-4.pdf}
-  \caption{$k$-means cluster silhouette plot, $k = 4$}
-  \label{fig:kmeans-silhouette-4}
-\end{figure}
-
-\subsection{Decision trees based on clusters}
-% seed = 911
-% CLUSTERS: 2
-% ================================
-% Complexity Parameter: 0.02183406
-% 10-fold CV error: 0.1132075
-% Root node error: 0.254162
-% CLUSTERS: 3
-% ================================
-% Complexity Parameter: 0.01070664
-% 10-fold CV error: 0.190899
-% Root node error: 0.518313
-% CLUSTERS: 4
-% ================================
-% Complexity Parameter: 0.01
-% 10-fold CV error: 0.2552719
-% Root node error: 0.5627081
+\subsection{Decision tree}
 \begin{table}[h]
@@
     \hline
-    2 & 0.0218 & 0.113 & axial $\geq$ 4.5 & 0.254 & Figure~\ref{fig:kmeans-dtree-2} \\
-    3 & 0.0107 & 0.191 & pigd $\geq$ 2.5 & 0.518 & Figure~\ref{fig:kmeans-dtree-3} \\
     4 & 0.0100 & 0.255 & pigd $<$ 2.5 & 0.563 & Figure~\ref{fig:kmeans-dtree-4} \\
@@
 \end{table}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{dtree-kmeans-pruned-unscaled-2.pdf}
-  \caption{Decision Tree from $k$-means clustering, 2 clusters}
-  \label{fig:kmeans-dtree-2}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{dtree-kmeans-pruned-unscaled-3.pdf}
-  \caption{Decision Tree from $k$-means clustering, 3 clusters}
-  \label{fig:kmeans-dtree-3}
-\end{figure}
 
@@
 
-Available in
-Figures~\ref{fig:kmeans-summaries-2},~\ref{fig:kmeans-summaries-3},
-and~\ref{fig:kmeans-summaries-4}. Error bar is standard error.
-
-\subsubsection{Interpretation}
-
-$k = 2$ seems too basic. Cluster is organized solely by severity - all
-symptoms, including motor and nonmotor, are higher in severity in cluster 1,
-and lower in cluster 2. Quite consistently, groups in cluster 1 are generally
-of slightly higher age and pd duration.
-
-$k = 3$ seems like a further development of $k = 2$, where clusters are simply
-organized by linearly increasing severity.
-
-$k = 4$ is where it gets interesting.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{kmeans-summaries-2.pdf}
-  \caption{Cluster Summaries, $k = 2$}
-  \label{fig:kmeans-summaries-2}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{kmeans-summaries-3.pdf}
-  \caption{Cluster Summaries, $k = 3$}
-  \label{fig:kmeans-summaries-3}
-\end{figure}
+Available in Figure~\ref{fig:kmeans-summaries-4}. Error bar is standard error.
 
@@
 
+\subsubsection{Interpretation}
+
 \subsubsection{Statistical Significance Tests, $k = 4$}
@@
 
-\subsubsection{Ranked Features}
+Post-hoc analysis using Tukey's HSD:
+\begin{verbatim}
+age insignificant differences:
+         diff         lwr        upr        p adj
+4-3 -2.271990  -4.7160927  0.1721117 7.920343e-02
+sex insignificant differences:
+           diff        lwr        upr     p adj
+2-1 -0.09275204 -0.2454183 0.05991417 0.4000134
+3-1 -0.14660529 -0.3046921 0.01148156 0.0803067
+4-1  0.04757177 -0.1240051 0.21914864 0.8917054
+3-2 -0.05385325 -0.1511666 0.04346014 0.4843788
+pdonset insignificant differences:
+           diff       lwr      upr     p adj
+2-1 -0.90162565 -4.302019 2.498768 0.9037828
+3-1 -0.05040276 -3.571532 3.470727 0.9999820
+4-1 -0.53727145 -4.358869 3.284326 0.9837825
+3-2  0.85122289 -1.316276 3.018722 0.7431486
+4-2  0.36435420 -2.263251 2.991959 0.9844187
+4-3 -0.48686869 -3.268952 2.295214 0.9695444
+nms_d1 insignificant differences:
+          diff         lwr       upr        p adj
+3-2  0.5565021 -0.01586385  1.128868 6.018674e-02
+nms_d3 insignificant differences:
+          diff        lwr        upr       p adj
+4-1  -1.192024  -4.493510   2.109461 0.789133802
+nms_d4 insignificant differences:
+          diff        lwr       upr        p adj
+3-2  0.2951915 -0.3318339  0.922217 6.195406e-01
+nms_d5 insignificant differences:
+           diff         lwr       upr        p adj
+3-2   0.7841071  -0.4814825  2.049697 0.3821565990
+nms_d8 insignificant differences:
+          diff       lwr      upr     p adj
+4-1 -0.9010507 -2.846119 1.044017 0.6318563
+3-2  0.7255838 -0.377602 1.828770 0.3280035
+nms_d9 insignificant differences:
+          diff         lwr       upr       p adj
+4-1  0.7048896  -1.6510530  3.060832 0.867951568
+tremor insignificant differences:
+          diff       lwr        upr      p adj
+3-1 -0.2662831 -1.064520 0.53195417 0.82613187
+4-2 -0.5667198 -1.162396 0.02895624 0.06895758
+rigidity insignificant differences:
+          diff         lwr        upr        p adj
+4-2  0.2310142 -0.03536748  0.4973959 1.153952e-01
+\end{verbatim}
+
+\subsubsection{Ranked Features by Information Gain}
 
@@
 
-\clearpage
-\section{Affinity Propagation}
+\subsubsection{Correlation Plots}
 
-\subsection{Clustering}
-
-Package \texttt{apcluster} was used. Distance matrix was the negative euclidean
-squared distance ($r = 2$).
-
-AP with input preferences minimized ($q = 0$) resulted in 8 clusters.
-With the standard median input preferences ($q = 0.5$), algorithm failed to
-converge with default parameters. Even setting damping factor to 0.98, maximum
-iterations to 10000, and convergence iterations to 1000 failed to converge.
-Might need to try a longer run.
-
-\emph{However}, given that input preferences control how many clusters are
-found, I don't think it's very useful to have some dozen clusters running
-around.
-
-\subsubsection{Silhouette Plots}
-
-Silhouette plot in Figure~\ref{fig:ap-silhouette} looks pretty weak, really.
-Tons of overlap between the clusters.
+Figure~\ref{fig:corrplots}.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{ap-silhouette.pdf}
-  \caption{AP silhouette plot, $k = 8$}
-  \label{fig:ap-silhouette}
+  \includegraphics[width=\linewidth]{corrplots.pdf}
+  \caption{Correlation plots}
+  \label{fig:corrplots}
 \end{figure}
 
-\clearpage
-\section{Hierarchical Clustering}
+\subsubsection{Bradykinesia and rigidity}
 
-\subsection{Clustering}
-
-Four dissimilarity methods were used with a euclidean distance matrix.
-Dendrograms available in Figure~\ref{fig:hc-dendrograms}
+Figures~\ref{fig:bradykin-cisitot} and~\ref{fig:rigidity-cisitot}
 
@@
   \centering
-  \includegraphics[width=\linewidth]{hc-dendrograms.pdf}
-  \caption{Dendrograms}
-  \label{fig:hc-dendrograms}
-\end{figure}
-
-\subsection{Cutting Trees}
-\begin{table}[h]
-  \centering
-  \begin{tabular}{l|l|l|l}
-    Method & Condition & n & Figure \\
-    \hline
-    Complete & $k = 4$ & 4 (790/81/18/12) & ~\ref{fig:hc-summaries-complete-k4} \\
-    Complete & \texttt{dynamicTreeCut}\tablefootnote{Package \texttt{dynamicTreeCut} in R (Langfelder P,
-  Zhang B, Horvath S (2007)). Hybrid method, minimum cluster selection
-parameters} & 13 (255/99/77/64/62/58/56/46/44/41/37/32/30) &
-    ~\ref{fig:hc-summaries-complete-dynamic} \\
-    Ward & $k = 4$ & 4 (200/237/263/201) & ~\ref{fig:hc-summaries-ward-D-k4} \\
-    Ward & $h = 100$ & 3 (437/263/201) &
-    ~\ref{fig:hc-summaries-ward-D-h100} \\
-  \end{tabular}
-  \caption{Clusters from Tree Cutting}
-  \label{tab:tree-cutting}
-\end{table}
-
-\subsection{Interpretation}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{hc-summaries-complete-k4.pdf}
-  \caption{Using maximum (complete linkage) dissimilarity, cutting tree for $k = 4$}
-  \label{fig:hc-summaries-complete-k4}
+  \includegraphics[width=\linewidth]{bradykin-cisitot.pdf}
+  \caption{Relationship between bradykinesia and cisitot}
+  \label{fig:bradykin-cisitot}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\linewidth]{hc-summaries-complete-dynamic.pdf}
-  \caption{Using maximum (complete linkage) dissimilarity, cutting tree
-    dynamically}
-  \label{fig:hc-summaries-complete-dynamic}
+  \includegraphics[width=\linewidth]{bradykin-cisitot-condensed.pdf}
+  \caption{Relationship between bradykinesia and cisitot (condensed)}
+  \label{fig:bradykin-cisitot-condensed}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\linewidth]{hc-summaries-ward-D-k4.pdf}
-  \caption{Using Ward (1963) dissimilarity, cutting tree for $k = 4$}
-  \label{fig:hc-summaries-ward-D-k4}
+  \includegraphics[width=\linewidth]{bradykin-cisitot-all.pdf}
+  \caption{Relationship between bradykinesia and cisitot (entire data)}
+  \label{fig:bradykin-cisitot-all}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\linewidth]{hc-summaries-ward-D-h100.pdf}
-  \caption{Using Ward (1963) dissimilarity, cutting tree at $h = 100$}
-  \label{fig:hc-summaries-ward-D-h100}
+  \includegraphics[width=\linewidth]{rigidity-cisitot.pdf}
+  \caption{Similar relationship between rigidity and cisitot}
+  \label{fig:rigidity-cisitot}
 \end{figure}
 
-\subsection{Interpretation}
+\section{Other Work}
 
-Cluster sizes are available in Table~\ref{tab:ap-cluster-sizes}
+\subsection{Bayesian Networks}
 
-\begin{table}[h]
-  \centering
-  \begin{tabular}{l|l}
-  Cluster & Size \\
-  \hline
-  1 & 63 \\
-  2 & 53 \\
-  3 & 85 \\
-  4 & 122 \\
-  5 & 48 \\
-  6 & 126 \\
-  7 & 123 \\
-  8 & 102 \\
-  9 & 166 \\
-  10 & 13 \\
-  \end{tabular}
-  \caption{AP Cluster Sizes}
-  \label{tab:ap-cluster-sizes}
-\end{table}
-
-Boxplot summary of clusters available in Figure~\ref{fig:ap-summaries}.
-\textbf{Discussion forthcoming.}
+In Figure~\ref{fig:bnets}. Structure is too sparse, need to discretize or use
+some kind of regularization (e.g. a lasso)
 
@@
   \centering
-  \includegraphics[width=\linewidth]{ap-summaries.pdf}
-  \caption{AP Boxplot Summaries}
-  \label{fig:ap-summaries}
+  \includegraphics[width=0.8\linewidth]{bnets.pdf}
+  \caption{Bayesian Networks}
+  \label{fig:bnets}
 \end{figure}
 
-\section{Biclustering}
-
-Used BCBimax clustering algorithm. Clusters seem quite sparse.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{biclust-heatmaps-16.pdf}
-  \caption{Biclustering heatmaps $N = 16$}
-  \label{fig:biclust-heatmaps-16}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{biclust-bubbleplot-16.pdf}
-  \caption{Bubbleplot $N = 16$}
-  \label{fig:bubbleplot-16}
-\end{figure}
-
-\section{Subspace clustering}
-
-% TODO: Wat.
-
-\section{Bayesian Networks}
-
 \end{document}

### 2015-07-16 R/exploration.R
add more exploration code, ARFF writing

@@ -1,3 +1,6 @@
+# NOTE: This is just bits and pieces of sandbox code.
+# Not really meant to be sourced.
+
 # Libraries ====
 library(corrplot)
 library(ggplot2)
@@ -64,3 +67,50 @@ plot(hc(c2.df[, -20]))
 plot(hc(c4.df[, -20]))
 plot(hc(c3.df[, -20]))
 plot(hc(c1.df[, -20]))
+
+# Markov blanket feature selection
+parkinsons.chr <- parkinsons
+parkinsons$class <- as.factor(as.character(parkinsons$class))
+# Discretize first
+pd.discrete <- discretize(parkinsons, method = "quantile")
+iamb(parkinsons)
+tan_cl(parkinsons, class = "class")
+
+# More correlation (from asdm) ====
+pd.cor <- cor(parkinsons[, 1:19])
+corrplot(pd.cor, method = 'ellipse', type = 'lower', diag = F)
+pd.cor['pigd', 'axial']
+pd.cor['pigd', 'cisitot']
+pd.cor['axial', 'cisitot']
+pd.cor['pdonset', 'age']
+pd.cor['bradykin', 'rigidity']
+
+pd.cor['nms_d2', 'nms_d3']
+
+pd <- parkinsons
+pd[, 1:19] <- scale(pd[, 1:19])
+pd.iamb <- iamb(pd)
+plot(pd.iamb)
+pd.tan_cl <- tan_cl(pd, class = "class")
+pd.hc <- hc(pd)
+pd.cpt <- cptable(pd)
+as.grain(pd.cpt)
+pd.whitelist <- data.frame(
+  from = c('axial', 'pigd', 'tremor', 'bradykin', 'rigidity'),
+  to = c('cisitot', 'cisitot', 'cisitot', 'cisitot', 'cisitot')
+)
+pd.hc.wl <- hc(pd, whitelist = pd.whitelist)
+pd.hc <- si.hiton.pc(pd)
+plot(pd.hc)
+pd.hc.mb <- mb(pd.hc, 'class')
+pd.hc.mb
+length(mb(pd.hc.mb, 'class'))
+
+pd.woclass <- pd[, 1:19]
+pd.woclass.hc <- hc(pd[, 1:19])
+plot(pd.woclass.hc)
+mb(pd.woclass.hc, 'cisitot')
+
+# PCA ====
+pc <- princomp(pd[, 1:19])
+plot(pc)

### 2015-07-16 R/kmeans-dtree.R

@@ -361,7 +361,7 @@ for (i in c("2", "3", "4")) {
 
 # REORDER FACTORS BY INCREASING CISITOT ====
 for (i in c("2", "3", "4")) {
-  cisitot <- clusters.raw.long[[i]][clusters.raw.long[[i]]$variable == "cisitot", ] 
+  cisitot <- clusters.raw.long[[i]][clusters.raw.long[[i]]$variable == "cisitot", ]
   cisitot.means <- sapply(factor(1:as.integer(i)), function(i) {
     mean(cisitot[cisitot$cluster == i, ]$measurement)
   })
@@ -449,7 +449,7 @@ for (var in names(tukeys)) {
   test <- tukeys[[var]]
   # Check for nonsignificant, since there are more significant
   sigs <- test[test[, "p adj"] > 0.05, ]
-  if (!identical(logical(0), as.logical(sigs))) { 
+  if (!identical(logical(0), as.logical(sigs))) {
     # Super hacky to figure out if null matrix without type error
     cat(var, ' insignificant differences', ':', '\n', sep='')
     if (class(sigs) == 'numeric') {  # If returned just a single vector, can't do anything
@@ -476,6 +476,18 @@ features.ranked[with(features.ranked, order(-attr_importance)), c('variable', 'a
 # 90/214/56/421
 # 89/421/57/214
 # 214/421/57/89
+
 # but nms_d3 < 17.5 seems to be a main classifier here, eeven though it's not very
 # practically informative
 # also nms_d7
+
+# WRITE TO ARFF FOR WEKA ====
+clus4.wide <- clusters.raw[["4"]][, c(2:20, 1)]
+clus4.wide <- rename(clus4.wide, c("cluster" = "class"))
+clus4.wide$class <- as.factor(clus4.wide$class)
+head(clus4.wide)
+# Rename to parkinsons because it makes more sense
+parkinsons <- clus4.wide
+if (assertthat::are_equal(ncol(clusters.raw[["4"]]), ncol(parkinsons))) {
+  foreign::write.arff(parkinsons, './parkinsons-k4.arff')
+}

### 2015-07-22 R/bnets.R
Some preliminary bnets for cluster 3

@@ -20,3 +20,11 @@ plot(clus4.iamb)
 mb(clus4.iamb, 'cluster')
 plot(hc(clus4.df))
 mb(hc(clus4.df), 'cluster')
+
+# Learning only in c3 ====
+# Standardize to force gaussian vars
+c3.woclus <- c3[, -1]
+c3.std <- as.data.frame(scale(c3.woclus))
+c3.hc <- hc(c3.std)
+plot(c3.hc)
+mb(c3.hc, "cisitot")

### 2015-07-22 R/exploration.R

@@ -13,6 +13,14 @@ c2 <- clus4.wide[clus4.wide$class == 2, ]
 c3 <- clus4.wide[clus4.wide$class == 3, ]
 c4 <- clus4.wide[clus4.wide$class == 4, ]
 
+# OR alternative cluster usage if the full kmeans-dtree script hasn't been run
+if (FALSE) {
+  c1 <- clus4.wide[clus4.wide$cluster == 1, ]
+  c2 <- clus4.wide[clus4.wide$cluster == 2, ]
+  c3 <- clus4.wide[clus4.wide$cluster == 3, ]
+  c4 <- clus4.wide[clus4.wide$cluster == 4, ]
+}
+
 # Plot w/o cluster
 par(mfrow=c(2, 2))
 corrplot(cor(c2[, -20]), method = 'pie', title = 'Correlation 2', diag = F, type = 'lower')

### 2015-07-22 R/affinity-propagation.R
Remove pigd mentions, update indices, backup

@@ -11,7 +11,7 @@ source('./preprocessing.R')
 INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
                  "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
                  "nms_d6", "nms_d7", "nms_d8", "nms_d9",
-                 "tremor", "bradykin", "rigidity", "axial", "pigd")
+                 "tremor", "bradykin", "rigidity", "axial")
 # save.plots for now since I don't know what else to plot!
 SAVE.PLOTS <- FALSE
 
@@ -35,7 +35,7 @@ for (i in 1:nclust) {
 }
 
 # WIDE -> LONG ====
-unscaled.apclus.long <- gather(unscaled.apclus, variable, measurement, age:pigd)
+unscaled.apclus.long <- gather(unscaled.apclus, variable, measurement, age:axial)
 
 # FACTOR CLUSTERS, ORDER BY BY CISITOT ====
 # FIXME: Is it bad that the clusters are only factored here rather than

### 2015-07-22 R/biclustering.R

@@ -21,8 +21,8 @@ nmses <- sapply(1:9, function(i) paste('nms_d', i, sep=''))
 extras <- c('age', 'durat_pd', 'cisitot')
 raw.nms <- raw.nopdonset[, c(nmses, extras)]
 raw.nms.unscaled <- raw.nopdonset.unscaled[, c(nmses, extras)]
-raw.motor <- raw.nopdonset[, c('rigidity', 'axial', 'bradykin', 'tremor', 'pigd')]
-raw.motor.unscaled <- raw.nopdonset.unscaled[, c('rigidity', 'axial', 'bradykin', 'tremor', 'pigd')]
+raw.motor <- raw.nopdonset[, c('rigidity', 'axial', 'bradykin', 'tremor')]
+raw.motor.unscaled <- raw.nopdonset.unscaled[, c('rigidity', 'axial', 'bradykin', 'tremor')]
 
 # BICLUSTERING ====
 # Bimax algorithm

### 2015-07-22 R/exploration.R

@@ -23,10 +23,10 @@ if (FALSE) {
 
 # Plot w/o cluster
 par(mfrow=c(2, 2))
-corrplot(cor(c2[, -20]), method = 'pie', title = 'Correlation 2', diag = F, type = 'lower')
-corrplot(cor(c4[, -20]), method = 'pie', title = 'Correlation 4', diag = F, type = 'lower')
-corrplot(cor(c3[, -20]), method = 'pie', title = 'Correlation 3', diag = F, type = 'lower')
-corrplot(cor(c1[, -20]), method = 'pie', title = 'Correlation 1', diag = F, type = 'lower')
+corrplot(cor(c2[, -19]), method = 'pie', title = 'Correlation 2', diag = F, type = 'lower')
+corrplot(cor(c4[, -19]), method = 'pie', title = 'Correlation 4', diag = F, type = 'lower')
+corrplot(cor(c3[, -19]), method = 'pie', title = 'Correlation 3', diag = F, type = 'lower')
+corrplot(cor(c1[, -19]), method = 'pie', title = 'Correlation 1', diag = F, type = 'lower')
 par(mfrow=c(1, 1))
 
 par(mfrow=c(2, 2))
@@ -64,17 +64,17 @@ c3.df <- as.data.frame(as.matrix(c3))
 c1.df <- as.data.frame(as.matrix(c1))
 
 # iamb w/o class
-plot(iamb(c2.df[, -20]), main = '2')
-plot(iamb(c4.df[, -20]), main = '4')
-plot(iamb(c3.df[, -20]), main = '3')
-plot(iamb(c1.df[, -20]), main = '1')
+plot(iamb(c2.df[, -19]), main = '2')
+plot(iamb(c4.df[, -19]), main = '4')
+plot(iamb(c3.df[, -19]), main = '3')
+plot(iamb(c1.df[, -19]), main = '1')
 
 # hc w/o class
 # Too sparse - need more data
-plot(hc(c2.df[, -20]))
-plot(hc(c4.df[, -20]))
-plot(hc(c3.df[, -20]))
-plot(hc(c1.df[, -20]))
+plot(hc(c2.df[, -19]))
+plot(hc(c4.df[, -19]))
+plot(hc(c3.df[, -19]))
+plot(hc(c1.df[, -19]))
 
[…]

### 2015-07-22 R/hierarchical.R

@@ -13,7 +13,7 @@ source('./preprocessing.R')
 INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
                  "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
                  "nms_d6", "nms_d7", "nms_d8", "nms_d9",
-                 "tremor", "bradykin", "rigidity", "axial", "pigd")
+                 "tremor", "bradykin", "rigidity", "axial")
 SAVE.PLOTS <- FALSE
 
 # COMPUTE DISTANCE MATRIX (EUCLIDEAN) ====
@@ -90,13 +90,13 @@ print(count(hcl.complete.mem))
 labeled.complete <- cbind(cluster=hcl.complete.mem, raw.omitted[, INTERPRETED])
 
 # WIDE -> LONG ====
-labeled.complete.long <- gather(labeled.complete, variable, measurement, age:pigd)
+labeled.complete.long <- gather(labeled.complete, variable, measurement, age:axial)
 
 # ORDER BY INCREASING CISITOT ====
 # We need to shift by 1 if using dynamic, since dynamicTreeCut starts at 0.
 DYNAMIC <- FALSE
 # DYNAMIC <- TRUE
-cisitot <- labeled.complete.long[labeled.complete.long$variable == "cisitot", ] 
+cisitot <- labeled.complete.long[labeled.complete.long$variable == "cisitot", ]
 anyNA(cisitot)
 cisitot.means <- sapply(1:nclust, function(i) {
   mean(cisitot[cisitot$cluster == i - DYNAMIC, ]$measurement)

### 2015-07-22 R/kmeans-dtree.R

@@ -15,7 +15,7 @@ library(tidyr)
 INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
                  "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
                  "nms_d6", "nms_d7", "nms_d8", "nms_d9",
-                 "tremor", "bradykin", "rigidity", "axial", "pigd")
+                 "tremor", "bradykin", "rigidity", "axial")
 
 # Remember - PDFs can vary even if the (seeded) clusters don't
 # So this should be false unless I've changed something about the
@@ -356,7 +356,7 @@ for (i in 2:4) {
 # CLUSTERS RAW WIDE -> LONG ====
 # NOTE: This lapply conversion hasn't been tested fully yet!
 clusters.raw.long <- lapply(c("2", "3", "4"), function(i) {
-  gather(clusters.raw[[i]], variable, measurement, age:pigd)
+  gather(clusters.raw[[i]], variable, measurement, age:axial)
 })
 names(clusters.raw.long) <- c("2", "3", "4")
 
@@ -483,7 +483,8 @@ features.ranked[with(features.ranked, order(-attr_importance)), c('variable', 'a
 # also nms_d7
 
 # WRITE TO ARFF FOR WEKA ====
-clus4.wide <- clusters.raw[["4"]][, c(2:20, 1)]
+clus4.wide <- clusters.raw[["4"]][, c(2:19, 1)]
+# NOTE: Here is where the cluster/class renaming happens (.arff)
 clus4.wide <- rename(clus4.wide, c("cluster" = "class"))
 clus4.wide$class <- as.factor(clus4.wide$class)
 head(clus4.wide)

### 2015-07-22 R/nms.R

@@ -29,7 +29,7 @@ names(c4.labeled) <- 2:4
 
 # Convert wide -> long ====
 c4.labeled.long <- lapply(c("2", "3", "4"), function(i) {
-  gather(c4.labeled[[i]], variable, measurement, age:pigd)
+  gather(c4.labeled[[i]], variable, measurement, age:axial)
 })
 names(c4.labeled.long) <- c("2", "3", "4")
 

### 2015-07-22 R/preprocessing.R

@@ -19,17 +19,15 @@ ALL.SYMPTOMS <- c("nms_d1",
                  "tremor",
                  "bradykin",
                  "rigidity",
-                 "axial",
-                 "pigd")
+                 "axial")
 MOTOR.SYMPTOMS <- c("tremor",
                    "bradykin",
                    "rigidity",
-                   "axial",
-                   "pigd")
+                   "axial")
 INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
                  "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
                  "nms_d6", "nms_d7", "nms_d8", "nms_d9",
-                 "tremor", "bradykin", "rigidity", "axial", "pigd")
+                 "tremor", "bradykin", "rigidity", "axial")
 # Change this variable to change symptoms
 SYMPTOMS.TO.USE <- ALL.SYMPTOMS
 

### 2015-07-22 R/kmeans-dtree.R
Rerun preprocessing and kmeans-dtree w/o pigd
Tons of changes, hard to keep track of. Will need to validate these
later.

@@ -9,6 +9,7 @@ library(NbClust)
 library(mclust)
 library(ggplot2)
 library(tidyr)
+library(FSelector)
 
 # CONSTANTS ====
 # INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot")
@@ -33,6 +34,8 @@ source('./preprocessing.R')
 # VISUALIZE WSS ERROR TO FIND OPTIMAL K ====
 # NOTE: Doesn't work well, there isn't any elbow
 
+# Just in case it's still defined from below
+var <- NULL
 wss <- (nrow(raw.filtered)-1)*sum(apply(raw.filtered, 2, var))
 for (i in 2:15) {
   wss[i] <- sum(kmeans(raw.filtered, i, nstart = 25)$withinss)
@@ -130,7 +133,7 @@ text(pruned.t, use.n=TRUE, all=TRUE, cex=.8, pos=1)
 
 # KMEANS CLUSTERING COMPRESSED FUNCTION ====
 # Use for iteration!
-kmeans.dtree <- function(data, data.unscaled, k, save = FALSE, seed = 911) {
+kmeans.dtree <- function(data, data.unscaled, k, save = FALSE, seed = 0) {
   # Reproducibility!
   set.seed(seed)
   cl <- kmeans(splits$trainset, k, nstart = 25)
@@ -430,9 +433,9 @@ oneways <- lapply(colnames(clus4.wide.st[, -1]), function(col) {
 })
 for (test in oneways) {
   if (test$p.value < 0.05) {
-    cat('insig\n')
+    cat('sig\n')
   } else {
-    cat('SIG:\n')
+    cat('INSIG:\n')
     cat(test$data.name, '\n')
   }
 }
@@ -446,29 +449,30 @@ tukeys <- lapply(colnames(clus4.wide.st[, -1]), function(col) {
 })
 names(tukeys) <- colnames(clus4.wide.st[, -1])
 
-for (var in names(tukeys)) {
-  test <- tukeys[[var]]
+for (v in names(tukeys)) {
+  test <- tukeys[[v]]
   # Check for nonsignificant, since there are more significant
   sigs <- test[test[, "p adj"] > 0.05, ]
   if (!identical(logical(0), as.logical(sigs))) {
     # Super hacky to figure out if null matrix without type error
-    cat(var, ' insignificant differences', ':', '\n', sep='')
+    cat(v, ' insignificant differences', ':', '\n', sep='')
[…]

### 2015-07-22 R/preprocessing.R

@@ -52,7 +52,7 @@ raw.omitted <- raw[, INTERPRETED]
 raw.omitted <- na.omit(raw.omitted)
 
 # raw with only select nms and motor symptoms
-raw.filtered <- raw.omitted
+raw.filtered <- raw.omitted[, SYMPTOMS.TO.USE]
 
 # DESCRIPTIVE STATISTICS (BEFORE STANDARDIZATION) ====
 # TODO - change raw.filtered to raw.omitted.filtered, or something like that

### 2015-07-22 writeup/abbreviated/abbreviated.tex
Update abbr. writeup with new data, corrplot

@@
 Combination of non-motor scale (NMS) symptoms and standard motor symptoms.
+Note: PIGD was deleted after 2015-07-16 meeting.
 
@@
     the body.} \\
-    pigd & float & \%9.0g & postural instability and gait difficulty \\
   \end{tabular}
@@
 axial&    3.25&  2.68&   0-12 \\
-pigd&     3.31&  2.71&   0-12 \\
   \end{tabular}
@@
     \hline
-    1 & 79 \\
-    2 & 394 \\
-    3 & 275 \\
-    4 & 153 \\
+    1 & 189 \\
+    2 & 88 \\
+    3 & 221 \\
+    4 & 406 \\
   \end{tabular}
@@
 \subsection{Decision tree}
-\begin{table}[h]
-  \centering
-  \begin{tabular}{l|l|l|l|l|l}
-    $k$ & CP\tablefootnote{Complexity Parameter} & CV Xerror\tablefootnote{10-fold cross
-    validation} & Root Feature &
-    Root Error & Figure \\
-    \hline
-    4 & 0.0100 & 0.255 & pigd $<$ 2.5 & 0.563 & Figure~\ref{fig:kmeans-dtree-4} \\
-  \end{tabular}
-  \caption{$k$-kmeans decision trees statistics}
-  \label{tab:k-means-dtrees}
-\end{table}
+% \begin{table}[h]
+  % \centering
+  % \begin{tabular}{l|l|l|l|l|l}
+    % $k$ & CP\tablefootnote{Complexity Parameter} & CV Xerror\tablefootnote{10-fold cross
+    % validation} & Root Feature &
+    % Root Error & Figure \\
+    % \hline
+    % 4 & 0.0100 & 0.255 & pigd $<$ 2.5 & 0.563 & Figure~\ref{fig:kmeans-dtree-4} \\
+  % \end{tabular}
+  % \caption{$k$-kmeans decision trees statistics}
+  % \label{tab:k-means-dtrees}
+% \end{table}
 
@@
           diff       lwr       upr     p adj
-4-3 -2.271990  -4.7160927  0.1721117 7.920343e-02
+3-1  0.9947808 -1.458184 3.4477455 0.7236845
+4-1 -1.2838898 -3.464063 0.8962832 0.4284274
 sex insignificant differences:
            diff        lwr        upr      p adj
-2-1 -0.09275204 -0.2454183 0.05991417 0.4000134
-3-1 -0.14660529 -0.3046921 0.01148156 0.0803067
-4-1  0.04757177 -0.1240051 0.21914864 0.8917054
-3-2 -0.05385325 -0.1511666 0.04346014 0.4843788
+2-1 -0.05044493 -0.2106093 0.10971941 0.84945412
+4-1 -0.09897829 -0.2082638 0.01030726 0.09181043
+3-2 -0.12633690 -0.2827741 0.03010026 0.16087872
+4-2 -0.04853336 -0.1944676 0.09740091 0.82744866
+4-3  0.07780354 -0.0259428 0.18154987 0.21607772
 pdonset insignificant differences:
           diff        lwr      upr     p adj
-2-1 -0.90162565 -4.302019 2.498768 0.9037828
-3-1 -0.05040276 -3.571532 3.470727 0.9999820
-4-1 -0.53727145 -4.358869 3.284326 0.9837825
-3-2  0.85122289 -1.316276 3.018722 0.7431486
-4-2  0.36435420 -2.263251 2.991959 0.9844187
-4-3 -0.48686869 -3.268952 2.295214 0.9695444
+2-1  2.9315777 -0.6232172 6.486373 0.1466742
+3-1  1.7136632 -1.0153886 4.442715 0.3699280
+4-1  0.7453932 -1.6801637 3.170950 0.8585776
+3-2 -1.2179144 -4.6899860 2.254157 0.8033301
+4-2 -2.1861845 -5.4251477 1.052779 0.3049434
+4-3 -0.9682701 -3.2708860 1.334346 0.7004488
+durat_pd insignificant differences:
+          diff       lwr        upr        p adj
+3-1 -0.7188824 -2.140915  0.7031499 5.624040e-01
+cisitot insignificant differences:
+          diff         lwr       upr        p adj
+3-1  0.4942421  -0.4388731  1.427357 5.228228e-01
 nms_d1 insignificant differences:
           diff        lwr        upr        p adj
-3-2  0.5565021 -0.01586385  1.128868 6.018674e-02
-nms_d3 insignificant differences:
-          diff        lwr        upr       p adj
-4-1  -1.192024  -4.493510   2.109461 0.789133802
+4-3 -0.3798787 -0.9604053  0.2006478 3.325894e-01
 nms_d4 insignificant differences:
           diff        lwr        upr        p adj
-3-2  0.2951915 -0.3318339  0.922217 6.195406e-01
+4-3 -0.3362459 -0.9972012  0.3247094 5.571480e-01
 nms_d5 insignificant differences:
            diff        lwr         upr        p adj
-3-2   0.7841071  -0.4814825  2.049697 0.3821565990
+4-3  -0.4117201  -1.743630   0.9201902 8.564409e-01
 nms_d8 insignificant differences:
           diff        lwr        upr        p adj
-4-1 -0.9010507 -2.846119 1.044017 0.6318563
-3-2  0.7255838 -0.377602 1.828770 0.3280035
+4-3 -0.9953302 -2.1560641  0.1654036 1.220509e-01
 nms_d9 insignificant differences:
           diff       lwr        upr      p adj
-4-1  0.7048896  -1.6510530  3.060832 0.867951568
+2-1  0.8708514 -1.297413 3.03911557 0.72966265
+4-3 -1.3221920 -2.726684 0.08229967 0.07350641
 tremor insignificant differences:
           diff         lwr        upr        p adj
-3-1 -0.2662831 -1.064520 0.53195417 0.82613187
-4-2 -0.5667198 -1.162396 0.02895624 0.06895758
-rigidity insignificant differences:
-          diff         lwr        upr        p adj
-4-2  0.2310142 -0.03536748  0.4973959 1.153952e-01
+4-1  0.3346105 -0.19270261  0.8619236 3.603863e-01
 \end{verbatim}
@@
     \hline
-    axial      & 0.20640691 \\
-    cisitot      & 0.20008571 \\
-    pigd      & 0.18193982 \\
-    nms\_d2      & 0.13178572 \\
-    nms\_d9      & 0.12116024 \\
-    bradykin      & 0.11966097 \\
-    nms\_d3      & 0.09421859 \\
-    rigidity      & 0.09260628 \\
-    nms\_d5      & 0.07579997 \\
-    nms\_d4      & 0.07438784 \\
-    nms\_d6      & 0.06620599 \\
-    nms\_d7      & 0.05574956 \\
-    nms\_d1      & 0.05509838 \\
-    tremor      & 0.04140473 \\
-    nms\_d8      & 0.03786173 \\
-    durat\_pd      & 0.02794420 \\
-    age      & 0.00000000 \\
+    bradykin      & 0.31574672 \\
+    rigidity      & 0.29560018 \\
+    nms\_d2      & 0.24218407 \\
+    cisitot      & 0.22920103 \\
+    axial      & 0.22780750 \\
+    nms\_d3      & 0.20480570 \\
+    nms\_d9      & 0.15782743 \\
+    nms\_d7      & 0.15290569 \\
+    nms\_d5      & 0.14454931 \\
+    nms\_d6      & 0.14025139 \\
+    nms\_d1      & 0.13212756 \\
+    tremor      & 0.10937168 \\
+    nms\_d4      & 0.10710526 \\
+    nms\_d8      & 0.10005480 \\
+    durat\_pd      & 0.02876190 \\
+    age      & 0.02346158 \\
     sex      & 0.00000000 \\
@@
 
-\subsubsection{Bradykinesia and rigidity}
+\subsubsection{One vs all decision trees}
 
-Figures~\ref{fig:bradykin-cisitot} and~\ref{fig:rigidity-cisitot}
+Figures~\ref{fig:1va},~\ref{fig:2va},~\ref{fig:3va},~\ref{fig:4va}.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{bradykin-cisitot.pdf}
-  \caption{Relationship between bradykinesia and cisitot}
-  \label{fig:bradykin-cisitot}
+  \includegraphics[width=\linewidth]{dtree-1va-pruned.pdf}
+  \caption{Cluster 1 vs all}
+  \label{fig:1va}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\linewidth]{bradykin-cisitot-condensed.pdf}
-  \caption{Relationship between bradykinesia and cisitot (condensed)}
-  \label{fig:bradykin-cisitot-condensed}
+  \includegraphics[width=\linewidth]{dtree-2va-pruned.pdf}
+  \caption{Cluster 2 vs all}
+  \label{fig:2va}
+\end{figure}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{dtree-3va-pruned.pdf}
+  \caption{Cluster 3 vs all}
+  \label{fig:3va}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\linewidth]{bradykin-cisitot-all.pdf}
-  \caption{Relationship between bradykinesia and cisitot (entire data)}
-  \label{fig:bradykin-cisitot-all}
+  \includegraphics[width=\linewidth]{dtree-4va-pruned.pdf}
+  \caption{Cluster 4 vs all}
+  \label{fig:4va}
 \end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{rigidity-cisitot.pdf}
-  \caption{Similar relationship between rigidity and cisitot}
-  \label{fig:rigidity-cisitot}
-\end{figure}
-
 \section{Other Work}

### 2015-07-22 R/exploration.R

@@ -23,10 +23,10 @@ if (FALSE) {
 
 # Plot w/o cluster
 par(mfrow=c(2, 2))
-corrplot(cor(c2[, -19]), method = 'pie', title = 'Correlation 2', diag = F, type = 'lower')
-corrplot(cor(c4[, -19]), method = 'pie', title = 'Correlation 4', diag = F, type = 'lower')
-corrplot(cor(c3[, -19]), method = 'pie', title = 'Correlation 3', diag = F, type = 'lower')
-corrplot(cor(c1[, -19]), method = 'pie', title = 'Correlation 1', diag = F, type = 'lower')
+corrplot(cor(c4[, -19]), method = 'ellipse', title = 'Correlation 4', diag = F, type = 'lower')
+corrplot(cor(c1[, -19]), method = 'ellipse', title = 'Correlation 1', diag = F, type = 'lower')
+corrplot(cor(c3[, -19]), method = 'ellipse', title = 'Correlation 3', diag = F, type = 'lower')
+corrplot(cor(c2[, -19]), method = 'ellipse', title = 'Correlation 2', diag = F, type = 'lower')
 par(mfrow=c(1, 1))
 
 par(mfrow=c(2, 2))
@@ -48,6 +48,7 @@ ggplot(clus4.wide, aes(x = rigidity, y = cisitot)) +
   stat_smooth(method = 'lm') +
   theme_bw()
 
+# NOTE: This hsan't been updated for new group names!
 # Check chisq independence of sex and nms_d8 for c1 (severe group)
 # Being male or female affects sexual function reports, in general
 # (Assuming male here)

### 2015-07-22 R/bnets.R
Add Bayes nets for nms-dominated cluster 1

@@ -2,6 +2,9 @@
 library(bnlearn)
 library(gRain) # Note - depends on biocLite graph and RBGL packages
 
+# Constants ====
+SAVE.BNETS <- F
+
 # Preprocessing ====
 source('./preprocessing.R')
 # Need stuff from kmeans dtree - so run that
@@ -22,9 +25,34 @@ plot(hc(clus4.df))
 mb(hc(clus4.df), 'cluster')
 
 # Learning only in c3 ====
+# These vars are set in exploration.R
 # Standardize to force gaussian vars
-c3.woclus <- c3[, -1]
-c3.std <- as.data.frame(scale(c3.woclus))
-c3.hc <- hc(c3.std)
-plot(c3.hc)
-mb(c3.hc, "cisitot")
+c1.woclus <- c1
+c1.woclus$class <- NULL
+c1.std <- as.data.frame(scale(c1.woclus))
+# Hill climbing ====
+c1.hc <- bnlearn::hc(c1.std)
+plot(c1.hc)
+title('Hill-Climbing Algorithm on Cluster 1')
+if (SAVE.BNETS) {
+  dev.copy(pdf, '../figures/clus1-bnet-hc.pdf')
+  dev.off()
+}
+# Grow shrink ====
+c1.gs <- bnlearn::gs(c1.std)
+plot(c1.gs)
+title('Grow-Shrink Algorithm on Cluster 1')
+if (SAVE.BNETS) {
+  dev.copy(pdf, '../figures/clus1-bnet-gs.pdf')
+  dev.off()
+}
+# Max-min hill-climbing ====
+c1.gs <- bnlearn::mmhc(c1.std)
+plot(c1.mmhc)
+title('Min-Max Hill Climbing Algorithm on Cluster 1')
+if (SAVE.BNETS) {
+  dev.copy(pdf, '../figures/clus1-bnet-mmhc.pdf')
+  dev.off()
+}
+
+mb(c1.hc, "cisitot")

### 2015-07-22 R/bnets.R
Switch nms from clus4 to 1, add summaries

@@ -47,7 +47,7 @@ if (SAVE.BNETS) {
   dev.off()
 }
 # Max-min hill-climbing ====
-c1.gs <- bnlearn::mmhc(c1.std)
+c1.mmhc <- bnlearn::mmhc(c1.std)
 plot(c1.mmhc)
 title('Min-Max Hill Climbing Algorithm on Cluster 1')
 if (SAVE.BNETS) {

### 2015-07-22 R/nms.R

@@ -9,51 +9,51 @@ library(ggplot2)
 # source('./preprocessing.R')
 
 # Constants ====
-SAVE.C4PLOTS <- FALSE
+SAVE.C1PLOTS <- FALSE
 
 
 # Do kmeans clustering on subgroup ====
-# c4 is what we're interested in, that's set in kmeans dtree
+# c1 is what we're interested in, that's set in kmeans dtree
 set.seed(0)
-c4.woclass <- c4
-c4.woclass$class <- NULL
-c4.woclass.scaled <- as.data.frame(scale(c4.woclass))
-c4.clus <- lapply(2:4, function(i) kmeans(c4.woclass.scaled, i, nstart = 25))
-names(c4.clus) <- 2:4
+c1.woclass <- c1
+c1.woclass$class <- NULL
+c1.woclass.scaled <- as.data.frame(scale(c1.woclass))
+c1.clus <- lapply(2:4, function(i) kmeans(c1.woclass.scaled, i, nstart = 25))
+names(c1.clus) <- 2:4
 
 # Bind cluster to orig data
-c4.labeled <- lapply(c("2", "3", "4"), function(i) {
-  cbind(c4.woclass, cluster = c4.clus[[i]]$cluster)
+c1.labeled <- lapply(c("2", "3", "4"), function(i) {
+  cbind(c1.woclass, cluster = c1.clus[[i]]$cluster)
 })
-names(c4.labeled) <- 2:4
+names(c1.labeled) <- 2:4
 
 # Convert wide -> long ====
-c4.labeled.long <- lapply(c("2", "3", "4"), function(i) {
-  gather(c4.labeled[[i]], variable, measurement, age:axial)
+c1.labeled.long <- lapply(c("2", "3", "4"), function(i) {
+  gather(c1.labeled[[i]], variable, measurement, age:axial)
 })
-names(c4.labeled.long) <- c("2", "3", "4")
+names(c1.labeled.long) <- c("2", "3", "4")
 
 # REORDER FACTORS BY INCREASING CISITOT ====
 for (i in c("2", "3", "4")) {
-  c4.cisitot <- c4.labeled.long[[i]][c4.labeled.long[[i]]$variable == "cisitot", ]
-  c4.cisitot.means <- sapply(factor(1:as.integer(i)), function(i) {
-    mean(c4.cisitot[c4.cisitot$cluster == i, ]$measurement)
+  c1.cisitot <- c1.labeled.long[[i]][c1.labeled.long[[i]]$variable == "cisitot", ]
[…]

### 2015-07-30 writeup/abbreviated/abbreviated.tex
Add Bayes nets for nms-dominated cluster 1
Add c1 summares to abbreviated
Add pamk/ASW k estimation, start writeup table

@@
 
-$k$-means clustering with $k = 4$ was tried. $k = 2, 3$ provided models that
-were too simplistic. $k = 5$ did not provide any new information, but rather
-just fragmented existing groups.
+$k$-means clustering with $k = 4$ was tried. Statistics for determining the
+optimal number of clusters were used, but were inconclusive: results in
+Figure~\ref{tab:numclus}. This probably indicates that the data is not very
+well clustered. $k = 2, 3$ provided models that were too simplistic. $k = 5$
+did not provide any new information, but rather just fragmented existing
+groups.
 
+\begin{table}[h]
+  \centering
+  \caption{caption}
+  \label{tab:numclus}
+  \begin{tabular}{l|l}
+    Criterion & Optimal $k$ \\
+    \hline
+    Minimum ASW & Inconclusive  \\
+    % TODO: Finish this
+  \end{tabular}
+\end{table}
 \begin{table}[h]
@@
 
-\subsection{Bayesian Networks}
+\subsection{Bayesian Networks on Cluster 1}
 
-In Figure~\ref{fig:bnets}. Structure is too sparse, need to discretize or use
-some kind of regularization (e.g. a lasso)
+Figures~\ref{fig:bnet-gs},~\ref{fig:bnet-hc}, and~\ref{fig:bnet-mmhc} show various
+learning algorithms.
 
@@
   \centering
-  \includegraphics[width=0.8\linewidth]{bnets.pdf}
-  \caption{Bayesian Networks}
-  \label{fig:bnets}
+  \includegraphics[width=\linewidth]{clus1-bnet-gs.pdf}
+  \caption{Grow Shrink Algorithm}
+  \label{fig:bnet-gs}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{clus1-bnet-hc.pdf}
+  \caption{Hill Climbing Algorithm}
+  \label{fig:bnet-hc}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{clus1-bnet-mmhc.pdf}
+  \caption{Min-Max Hill Climbing Algorithm}
+  \label{fig:bnet-mmhc}
+\end{figure}
+
+\subsection{$k$-means on Cluster 1}
+
+Figures~\ref{fig:c1-summaries-2},~\ref{fig:c1-summaries-3},
+and~\ref{fig:c1-summaries-4}.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{c1-summaries-2.pdf}
+  \caption{$k = 2$}
+  \label{fig:c1-summaries-2}
+\end{figure}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{c1-summaries-3.pdf}
+  \caption{$k = 3$}
+  \label{fig:c1-summaries-3}
+\end{figure}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{c1-summaries-4.pdf}
+  \caption{$k = 4$}
+  \label{fig:c1-summaries-4}
 \end{figure}

### 2015-07-30 R/kmeans-dtree.R

@@ -81,6 +81,11 @@ if (SAVE.EXPLORE.PLOTS) {
   dev.off()
 }
 
+library(fpc)
+pamk.best <- pamk(raw.filtered)
+cat("number of clusters estimated by optimum average silhouette width:", pamk.best$nc, "\n")
+plot(pam(raw.filtered, pamk.best$nc))
+
 # GAP STATISTIC ESTIMATION ====
 gaps <- clusGap(raw.filtered, kmeans, 14, B = 100)
 plot(x=1:14, y=gaps$Tab[, "gap"], xlab="Clusters", ylab="Gap Statitsic", type="b")

### 2015-10-09 R/kmeans-dtree.R
add extra processing, nonworking box colors

@@ -10,6 +10,8 @@ library(mclust)
 library(ggplot2)
 library(tidyr)
 library(FSelector)
+library(apclust)
+library(RColorBrewer) # For colors for decision tree plots
 
 # CONSTANTS ====
 # INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot")
@@ -56,11 +58,11 @@ if (SAVE.EXPLORE.PLOTS) {
 # it finds to be optimal, set it to search for
 # at least 1 model and up 20.
 # Traditionally this takes a while, so it's uncommented
-# d_clust <- Mclust(as.matrix(raw.filtered), G=1:20)
-# m.best <- dim(d_clust$z)[2]
-# cat("model-based optimal number of clusters:", m.best, "\n")
-# # 3 clusters
-# plot(d_clust)
+d_clust <- Mclust(as.matrix(raw.filtered), G=1:20)
+m.best <- dim(d_clust$z)[2]
+cat("model-based optimal number of clusters:", m.best, "\n")
+# 3 clusters
+plot(d_clust)
 
 # NBCLUST ESTIMATION FOR OPTIMAL K (30 metrics) ====
 # This is taking a long time
@@ -93,6 +95,18 @@ if (SAVE.EXPLORE.PLOTS) {
   dev.copy(pdf, "../figures/gap-statistic.pdf")
   dev.off()
 }
+
+# Affinity propagation ====
+# Auto commented out because this takes a while
+# library(apcluster)
+# raw.filtered.apclus <- apcluster(negDistMat(r=2), raw.filtered,
+#           details = TRUE, q = 0, lam = 0.98,
+#           maxits=10000, convits=1000)
+# cat("affinity propogation optimal number of clusters:",
+#     length(raw.filtered.apclus@clusters), "\n")
+# # 4
+# heatmap(raw.filtered.apclus)
+# plot(raw.filtered.apclus, raw.filtered)
 
 # INITIAL KMEANS CLUSTERING ====
 splits = splitdf(raw.filtered)
@@ -199,7 +213,8 @@ kmeans.dtree <- function(data, data.unscaled, k, save = FALSE, seed = 0) {
   # plot(t, uniform=TRUE, main="Decision Tree")
   # text(t, use.n=TRUE, all=TRUE, cex=.8)
   # ?prp for extra keyword args - this one displays class proportions
[…]

### 2015-10-09 R/nms.R

@@ -80,3 +80,6 @@ cat("affinity propogation optimal number of clusters:", length(d.apclus@clusters
 # 4
 heatmap(d.apclus)
 plot(d.apclus, c1.woclass)
+
+
+# write.csv(c1.woclass, '~/Downloads/nms-unscaled.csv')

### 2015-10-09 R/nms.R
determine clusters for nms subcluster

@@ -11,6 +11,87 @@ library(ggplot2)
 # Constants ====
 SAVE.C1PLOTS <- FALSE
 
+# VISUALIZE WSS ERROR TO FIND OPTIMAL K ====
+# NOTE: Doesn't work well, there isn't any elbow
+
+# Just in case it's still defined from below
+var <- NULL
+wss <- (nrow(c1.woclass)-1)*sum(apply(c1.woclass, 2, var))
+for (i in 2:15) {
+  wss[i] <- sum(kmeans(c1.woclass, i, nstart = 25)$withinss)
+}
+
+plot(1:15, wss, type="b",
+     xlab="Number of Clusters", ylab="Within groups sum of squares")
+if (SAVE.EXPLORE.PLOTS) {
+  dev.copy(pdf, "../figures/kmeans-wss-error.pdf")
+  dev.off()
+}
+
+# TODO: BIC ====
+# See http://stackoverflow.com/questions/15376075/cluster-analysis-in-r-determine-the-optimal-number-of-clusters
+# NOW - does this have to do with EM/hierarchical models. not cluster-based?
+# Run the function to see how many clusters
+# it finds to be optimal, set it to search for
+# at least 1 model and up 20.
+# Traditionally this takes a while, so it's uncommented
+d_clust <- Mclust(as.matrix(c1.woclass), G=1:20)
+m.best <- dim(d_clust$z)[2]
+cat("model-based optimal number of clusters:", m.best, "\n")
+# 3 clusters
+plot(d_clust)
+
+# NBCLUST ESTIMATION FOR OPTIMAL K (30 metrics) ====
+# This is taking a long time
+# nb <- NbClust(c1.woclass, distance = "euclidean",
+#               min.nc=2, max.nc=15, method = "kmeans",
+#               index = "alllong", alphaBeale = 0.1)
+# hist(nb$Best.nc[1,], breaks = max(na.omit(nb$Best.nc[1,])))
+
+# PAM ESTIMATION FOR OPTIMAL K ====
+# Estimates 2
+pam_sils <- c()
+for (i in 1:15) {
+  pam_sils <- c(pam_sils, pam(c1.woclass, k=i)$silinfo$avg.width)
+}
+plot(x=1:14, y=pam_sils, xlab="Clusters", ylab="Average Silhouette Width", type="b")
+if (SAVE.EXPLORE.PLOTS) {
+  dev.copy(pdf, "../figures/asw.pdf")
+  dev.off()
+}
[…]

### 2015-10-09 writeup/abbreviated/abbreviated.tex
update writeup

--- 2015-07-22 writeup/abbreviated/abbreviated.tex
@@
 \title{Cluster Analysis: Identifying Parkinson's Disease Subtypes}
+\author{Jesse Mu}
 \maketitle
@@
 missing values of the features to be used in clustering (brought down to 901).
-Imputation may be a good idea later on.
+It was decided to not impute the data.
 
@@
 Combination of non-motor scale (NMS) symptoms and standard motor symptoms.
-Note: PIGD was deleted after 2015-07-16 meeting.
+PIGD was deleted after 2015-07-16 meeting.
 
@@
   \centering
-  \begin{tabular}{l|l|l|l}
-    Name & Type & Format & Description \\
+  \begin{tabular}{l|l|l}
+    Name & Type & Description \\
     \hline
-    nms\_d1 & byte & \%8.0g & cardiovascular \\
-    nms\_d2 & byte & \%8.0g & sleep/fatigue \\
-    nms\_d3 & byte & \%8.0g & mood/cognition \\
-    nms\_d4 & byte & \%8.0g & percep/hallucinations \\
-    nms\_d5 & byte & \%8.0g & attention/memory \\
-    nms\_d6 & byte & \%8.0g & gastrointestinal \\
-    nms\_d7 & byte & \%8.0g & urinary \\
-    nms\_d8 & byte & \%8.0g & sexual function \\
-    nms\_d9 & byte & \%8.0g & miscellaneous \\
-    tremor & float & \%9.0g & tremor \\
-    bradykin & float & \%9.0g & bradykinesia\tablefootnote{Impaired ability to
+    nms\_d1 & byte & cardiovascular \\
+    nms\_d2 & byte & sleep/fatigue \\
+    nms\_d3 & byte & mood/cognition \\
+    nms\_d4 & byte & percep/hallucinations \\
+    nms\_d5 & byte & attention/memory \\
+    nms\_d6 & byte & gastrointestinal \\
+    nms\_d7 & byte & urinary \\
+    nms\_d8 & byte & sexual function \\
+    nms\_d9 & byte & miscellaneous \\
+    tremor & float & tremor \\
+    bradykin & float & bradykinesia\tablefootnote{Impaired ability to
     adjust the body's position.} \\
-    rigidity & float & \%9.0g & rigidity \\
-    axial & float & \%9.0g & axial\tablefootnote{Issues affecting the middle of
+    rigidity & float & rigidity \\
+    axial & float & axial\tablefootnote{Issues affecting the middle of
     the body.} \\
@@
 
-\section{$k$-means}
+\section{Clustering}
 
-$k$-means clustering with $k = 4$ was tried. $k = 2, 3$ provided models that
-were too simplistic. $k = 5$ did not provide any new information, but rather
-just fragmented existing groups.
-
+$k$-means clustering with $k = 4$ was tried. Statistics for determining the
+optimal number of clusters were used, but were inconclusive: results in
+Figure~\ref{tab:numclus}. This probably indicates that the data is not very
+well clustered. $k = 2, 3$ provided models that were too simplistic. $k = 5$
+did not provide any new information, but rather just fragmented existing
+groups.
 \begin{table}[h]
   \centering
-  \caption{Cluster statistics}
-  \label{tab:cluster_stats}
   \begin{tabular}{l|l}
-    Cluster & $n$ \\
+    Criterion & Optimal $k$ \\
     \hline
-    1 & 189 \\
-    2 & 88 \\
-    3 & 221 \\
-    4 & 406 \\
+    Minimum ASW & 2 \\
+    BIC & 18 \\
+    SSE Scree Plot & Inconclusive \\
+    Gap Statistic & 4, 13? \\
+    Affinity Propagation\tablefootnote{$\lambda = 0.98$, q = 0, maxits = 1000,
+    convits = 100} & 8 \\
+
   \end{tabular}
+  \caption{Results of various techniques for determining $k$}
+  \label{tab:numclus}
 \end{table}
@@
 % \end{table}
+
+Decision tree for $k = 4$ created via recursive partitioning is available in
+Figure~\ref{fig:kmeans-dtree-4}. More discussion about the decision tree is
+located in Section~\ref{feature-importance}.
 
@@
 \subsubsection{Interpretation}
+$k$-means clustering ($k = 4$) found four clusters. They are numbered (somewhat
+confusingly) according to the automatically assigned numbers of the cluster
+analysis, and not by any intuitive pattern. With a brief description, they are:
+
+\begin{itemize}
+  \item[4.] $(n = 406)$ Mildly affected in all domains.
+  \item[1.] $(n = 189)$ Severely affected in nonmotor domains; mildly
+    affected in motor domains.
+  \item[3.] $(n = 221)$ Severely affected in motor domains; mildly
+    affected in nonmotor domains.
+  \item[2.] $(n = 88)$ Severely affected in all domains.
+\end{itemize}
+
+These cluster results are identical to the four clusters found in van Rooden et
+al \cite{vanrooden10}, which was done with a separate dataset using a different
+modeling method (expectation-maximization), supporting these subtype
+classifications. Unlike van Rooden, mean disease durations differences do exist
+between subtypes 4 (mild) and 2 (severe), likely due to further development of
+the disease, although the differences between 1 and 3 (nonmotor/motor
+predominated) subtypes are insignificant, suggesting different developmental
+paths of the disease.
 
@@
 hypothesis that the means are the same with $p < 0.05$ for every variable
-\emph{except} \texttt{pdonset}.
+\emph{except} pdonset.
 
-Post-hoc analysis using Tukey's HSD:
-\begin{verbatim}
-age insignificant differences:
-          diff       lwr       upr     p adj
-3-1  0.9947808 -1.458184 3.4477455 0.7236845
-4-1 -1.2838898 -3.464063 0.8962832 0.4284274
-sex insignificant differences:
-           diff        lwr        upr      p adj
-2-1 -0.05044493 -0.2106093 0.10971941 0.84945412
-4-1 -0.09897829 -0.2082638 0.01030726 0.09181043
-3-2 -0.12633690 -0.2827741 0.03010026 0.16087872
-4-2 -0.04853336 -0.1944676 0.09740091 0.82744866
-4-3  0.07780354 -0.0259428 0.18154987 0.21607772
-pdonset insignificant differences:
-          diff        lwr      upr     p adj
-2-1  2.9315777 -0.6232172 6.486373 0.1466742
-3-1  1.7136632 -1.0153886 4.442715 0.3699280
-4-1  0.7453932 -1.6801637 3.170950 0.8585776
-3-2 -1.2179144 -4.6899860 2.254157 0.8033301
-4-2 -2.1861845 -5.4251477 1.052779 0.3049434
-4-3 -0.9682701 -3.2708860 1.334346 0.7004488
-durat_pd insignificant differences:
-          diff       lwr        upr        p adj
-3-1 -0.7188824 -2.140915  0.7031499 5.624040e-01
-cisitot insignificant differences:
-          diff         lwr       upr        p adj
-3-1  0.4942421  -0.4388731  1.427357 5.228228e-01
-nms_d1 insignificant differences:
-          diff        lwr        upr        p adj
-4-3 -0.3798787 -0.9604053  0.2006478 3.325894e-01
-nms_d4 insignificant differences:
-          diff        lwr        upr        p adj
-4-3 -0.3362459 -0.9972012  0.3247094 5.571480e-01
-nms_d5 insignificant differences:
-           diff        lwr         upr        p adj
-4-3  -0.4117201  -1.743630   0.9201902 8.564409e-01
-nms_d8 insignificant differences:
-          diff        lwr        upr        p adj
-4-3 -0.9953302 -2.1560641  0.1654036 1.220509e-01
-nms_d9 insignificant differences:
-          diff       lwr        upr      p adj
-2-1  0.8708514 -1.297413 3.03911557 0.72966265
-4-3 -1.3221920 -2.726684 0.08229967 0.07350641
-tremor insignificant differences:
-          diff         lwr        upr        p adj
-4-1  0.3346105 -0.19270261  0.8619236 3.603863e-01
-\end{verbatim}
-
-\subsubsection{Ranked Features by Information Gain}
+Post-hoc analysis using Tukey's HSD to examine statistically significant
+differences between individual means is available in Table~\ref{tab:tukeyhsd}.
+Only statistically insignificant relations are provided; all other relations
+are significant with $p < 0.05$.
 
@@
   \centering
+  \begin{tabular}{l|l|l}
+  Variable & Cluster Relation & $p$ \\
+  \hline
+  age & 3-1 & 0.724 \\
+      & 4-1 & 0.428 \\
+  \hline
+  sex & 2-1 & 0.849 \\
+      & 4-1 & 0.092 \\
+      & 3-2 & 0.161 \\
+      & 4-2 & 0.827 \\
+      & 4-3 & 0.216 \\
+  \hline
+  pdonset & 2-1 & 0.147 \\
+          & 3-1 & 0.370 \\
+          & 4-1 & 0.859 \\
+          & 3-2 & 0.803 \\
+          & 4-2 & 0.305 \\
+          & 4-3 & 0.700 \\
+  \hline
+  durat\_pd & 3-1 & 0.562 \\
+  \hline
+  cisitot & 3-1 & 0.523 \\
+  \hline
+  nms\_d1 & 4-3 & 0.333 \\
+  \hline
+  nms\_d4 & 4-3 & 0.557 \\
+  \hline
+  nms\_d5 & 4-3 & 0.856 \\
+  \hline
+  nms\_d8 & 4-3 & 0.122 \\
+  \hline
+  nms\_d9 & 2-1 & 0.730 \\
+         & 4-3 & 0.074 \\
+  \hline
+  tremor & 4-1 & 0.360 \\
+  \end{tabular}
+  \caption{Tukey's HSD Insignificant Differences}
+  \label{tab:tukeyhsd}
+\end{table}
+
+\subsubsection{Feature importance}
+\label{feature-importance}
+
+Features ranked by information gain with respect to cluster are available in
+Table~\ref{tab:info_gain}. Also, in the 4-cluster decision tree in
+Figure~\ref{fig:kmeans-dtree-4}, features are ranked implicitly by importance
+in determining clusters. We see, quite naturally, that standard measures of
+motor symptoms rank very highly (1, 2, 4, 5) in information gain \emph{except}
+termor (12). Similarly, bradykinesia (1) is used as the root node of the
+4-cluster decision tree, although other motor symptoms are used further down
+the tree, since immediately successive motor symptom decision nodes would, due
+to their determination of clusters, be redundant.
+
+The most informative nonmotor
+symptoms are nms\_d2 (sleep/fatigue) at 2, along with nms\_d3
+(mood/cognition). As discussed later in Section~\ref{sub:ova} these features
+become critical in one-versus-all decision trees for distinguishing various
+subtypes. The importance of these nonmotor symptoms confirms the longitudinal
+study by Fereshtehnejad et al. \cite{fereshtehnejad15} who cites a 3-cluster PD
+subtype identification based primarily on nonmotor symptoms including cognitive
+impairment, rapid eye movement sleep disorder (RBD), anxiety, and depression,
+conditions that align closely with nms\_d2 and nms\_d3 as tested in this
+dataset. More analysis needs to be done on whether there are parallels between
+Fereshtehnejad's 3-cluster longitudinal study and the clusters found in both
+this investigation and van Rooden.
+
+Interestingly, demographic information, including durat\_pd, age, sex, and
+pdonset, plays almost no role in the determination of these clusters. That the
+time of onset of PD or sex nulls important, clinically-relevant questions about
+the demographic sources of these different subtypes.
+
+\begin{table}[h]
+  \centering
+  \begin{tabular}{l|l|l}
+    rank & variable & information gain \\
+    \hline
+    1 & bradykin      & 0.31574672 \\
+    2 & rigidity      & 0.29560018 \\
+    3 & nms\_d2      & 0.24218407 \\
+    4 & cisitot      & 0.22920103 \\
+    5 & axial      & 0.22780750 \\
+    6 & nms\_d3      & 0.20480570 \\
+    7 & nms\_d9      & 0.15782743 \\
+    8 & nms\_d7      & 0.15290569 \\
+    9 & nms\_d5      & 0.14454931 \\
+    10 & nms\_d6      & 0.14025139 \\
+    11 & nms\_d1      & 0.13212756 \\
+    12 & tremor      & 0.10937168 \\
+    13 & nms\_d4      & 0.10710526 \\
+    14 & nms\_d8      & 0.10005480 \\
+    15 & durat\_pd      & 0.02876190 \\
+    16 & age      & 0.02346158 \\
+    17 & sex      & 0.00000000 \\
+    18 & pdonset      & 0.00000000 \\
+  \end{tabular}
   \caption{Features ranked by information gain}
   \label{tab:info_gain}
-  \begin{tabular}{l|l}
-    variable & information gain \\
-    \hline
-    bradykin      & 0.31574672 \\
-    rigidity      & 0.29560018 \\
-    nms\_d2      & 0.24218407 \\
-    cisitot      & 0.22920103 \\
-    axial      & 0.22780750 \\
-    nms\_d3      & 0.20480570 \\
-    nms\_d9      & 0.15782743 \\
-    nms\_d7      & 0.15290569 \\
-    nms\_d5      & 0.14454931 \\
-    nms\_d6      & 0.14025139 \\
-    nms\_d1      & 0.13212756 \\
-    tremor      & 0.10937168 \\
-    nms\_d4      & 0.10710526 \\
-    nms\_d8      & 0.10005480 \\
-    durat\_pd      & 0.02876190 \\
-    age      & 0.02346158 \\
-    sex      & 0.00000000 \\
-    pdonset      & 0.00000000 \\
-  \end{tabular}
 \end{table}
@@
 
-Figure~\ref{fig:corrplots}.
+The interplay between specific symptoms in each of the four clusters was
+examined in Figure~\ref{fig:corrplots}, but nothing too interesting was found. There is, perhaps somewhat
+interestingly, a somewhat higher correlation between bradykinesia, rigidity,
+and nms\_d6 (gastrointestinal), but it was not judged to be significant.
 
@@
 
-\subsubsection{One vs all decision trees}
+\section{Nonmotor-predominant subtype analysis}
+\subsection{$k$-means sub-subdivision on Cluster 1}
 
-Figures~\ref{fig:1va},~\ref{fig:2va},~\ref{fig:3va},~\ref{fig:4va}.
+In an attempt to understand further the properties of the nonmotor-dominated
+subtypes, $k$-means analysis was run again on specifically this subtype to
+examine any possible patterns.
+
+ The same $k$-determining tests were run on subtype 1 and are displayed
+in Table~\ref{tab:numclus-nms}.
+
+\begin{table}[h]
+  \centering
+  \begin{tabular}{l|l}
+    Criterion & Optimal $k$ \\
+    \hline
+    Minimum ASW & 2 \\
+    BIC & 1 (?) \\
+    SSE Scree Plot & Inconclusive \\
+    Gap Statistic & 3 \\
+    Affinity Propagation\tablefootnote{$\lambda = 0.98$, q = 0, maxits = 1000,
+    convits = 100} & 5 \\
+  \end{tabular}
+  \caption{Results of various techniques for determining $k$, applied to
+  subtype 1}
+  \label{tab:numclus-nms}
+\end{table}
+
+Boxplots for $k$-means run for $k = 2, 3, 4$ can
+be seen in Figures~\ref{fig:c1-summaries-2},~\ref{fig:c1-summaries-3},
+and~\ref{fig:c1-summaries-4}.
+
+\subsection{Interpretation}
+\label{sub:nms-interp}
+
+An interesting set of subtleties occurs when $k = 2$ and 3. When $k = 2$, the
+two groups
+are generally somewhat similar, except nms\_d2 and nms\_d3 move in opposite
+directions, i.e.\ sub-subtype 2 generally has higher nms\_d2 scores but lower
+nms\_d3 scores. In addition, sub-subtype 2 has higher axial scores. This
+difference can be once again observed in $k = 3$, where subtype 3 is shown to
+have higher nms\_d2 scores, lower nms\_d3 scores, and higher axial scores. This
+may indicate a trend in nonmotor-dominated PD for nms\_d2 and nms\_d3 to live
+on opposite ends of a spectrum, i.e.\ patients especially severe in one may not
+be as severe in the other. This difference can also been seen in
+Section~\ref{sub:ova}, when OVA decision trees are discussed for the the
+nonmotor-dominated group.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{c1-summaries-2.pdf}
+  \caption{Clustering on nonmotor group: $k = 2$}
+  \label{fig:c1-summaries-2}
+\end{figure}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{c1-summaries-3.pdf}
+  \caption{Clustering on nonmotor group: $k = 3$}
+  \label{fig:c1-summaries-3}
+\end{figure}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{c1-summaries-4.pdf}
+  \caption{Clustering on nonmotor group: $k = 4$}
+  \label{fig:c1-summaries-4}
+\end{figure}
+
+\section{Modeling}
+
+One further step of this investigation was to produce accurate, practical
+models that could be used in a clinical setting to predict the subtype of PD
+based on previous clustering results. Cluster assignments were treated as
+labels in a supervised classification problem in an attempt to produce useful
+models.
+
+\subsection{One-versus-all decision trees}
+\label{sub:ova}
+
+While the decision tree in Figure~\ref{fig:kmeans-dtree-4} is useful, it could
+be considered overly complicated. Additionally, a model is not necessarily
+needed to make simpler diagnoses such as classifying a patient as mildly
+affected (subtype 4) or severely affected (subtype 2). One-versus-all (OVA)
+decision trees were thus considered, in order to isolate the classification
+problem and look at possible distinguishing characteristics of individual
+subtypes. These OVA decision trees for all 4 subtypes are located in
+Figures~\ref{fig:4va},~\ref{fig:1va},~\ref{fig:3va}, and~\ref{fig:2va}.
+
+\subsubsection{4 (mild)}
+This tree is very odd. The root node considers nms\_d9 $< 7.5$ (miscellaneous)
+and classifies by far the biggest majority of negative examples when this
+symptom is milder, even though the subgroup 4 is the mild subtype. After
+intuitively then classifying on the mildness of rigidity the tree then proceeds
+to classify once again negative examples based on mildness of nms\_d2 (sleep)
+and positive examples solely based on more severe manifestations of nms\_d2,
+despite the means of these nonmotor symptoms being quite similar according to
+the boxplots.  However, accuracy on the final node is quite poor (61
+misclassifications) which could be an indicator that the cluster is not well
+defined.
+
+\subsubsection{1 (nonmotor-predominant)}
+Quite characteristically, this OVA decision tree clusters first on the severity
+of bradykinesia. When mild (i.e.\ $< 2.5$), all of the subsequent decision
+nodes ask whether the patient has correspondingly severe forms of primarily
+nms\_d3 (mood/cognition), then nms\_d7 (urinary) and nms\_d9 (miscellaneous).
+
+When bradykinesia (and all motor symptoms) are relatively severe, however,
+there is still a small subset of patients who classify into subtype 1 by having
+severe nms\_d2 (sleep) symptoms. While I am not sure if the sample size is
+large enough here (9/34), I propose this could be one of the ``subtypes'' of
+the nonmotor-predominant subtype, i.e.\ a group that, when accompanied with
+high motor symptoms, generally will have high nonmotor symptoms manifest
+primarily in the form of sleep disorders. This aligns with the findings
+previously mentioned in Section~\ref{sub:nms-interp} a disjunct between
+symptoms nms\_d2 and nms\_d3. As briefly mentioned in that section, axial motor
+symptom severity also plays a role. In this tree, mildness of axial symptoms is
+the main classifier of 280 negative examples when identifying
+nonmotor-predominant
+subtypes, indicating it is an important distinguishing factor of this unique
+group of 34 nonmotor-predominant PD patients.
+
+More discussion in the
+conclusion.
+
+\subsubsection{3 (motor-predominant)}
+
+This tree classifies overwhelmingly on severity of bradykinesia, with 476
+negative examples when bradykinesia is less than 2.5. The resulting tree is
+quite complex, but generally, nodes check again for severity of motor
+symptoms (tremor is the next node) and end up classifying positive examples
+based on both mildness of nonmotor symptoms and severity of motor symptoms. For
+example, in the furthest right branch, once nms\_d2 (as we know, an important
+feature) is established to be relatively mild (< 12), the test for subtype 3
+involves two more nodes verifying the severity of rigidty and tremor. Similar
+behavior can be found in the tree branch testing nms\_d7 and nms\_d6, then
+rigidity and axial.
+
+\subsubsection{2 (severe)}
+
+The OVA tree for severely affected in all areas is predictable, testing
+entirely on whether or not symptoms (both motor and nonmotor) are relatively
+severe. Ppositive nodes always appear to the right (no) of less-than checks.
+Interestingly, however, nms\_d4 (percep/hallucinations), previously not of note, is used twice as the
+root node of a tree and again further down. As the boxplot display in
+Figure~\ref{fig:kmeans-dtree-4} shows, nms\_d4 is perhaps the most
+distinguishing symptom of subtype 4 against nonmotor-predominant subtype 1 in
+particular, as subtype 1 is relatively mild, in contrast to relatively
+comparable levels of severity for other nonmotor symptoms. This shows that
+issues with
+perception and hallucinations generally occur in only the most severe cases of
+PD, and are relatively rare when a patient exhibits a nonmotor-predominant form
+of PD.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{dtree-4va-pruned.pdf}
+  \caption{Cluster 4 (mild) vs all}
+  \label{fig:4va}
+\end{figure}
 
@@
   \includegraphics[width=\linewidth]{dtree-1va-pruned.pdf}
-  \caption{Cluster 1 vs all}
+  \caption{Cluster 1 (nonmotor-dominated) vs all}
   \label{fig:1va}
@@
   \centering
-  \includegraphics[width=\linewidth]{dtree-2va-pruned.pdf}
-  \caption{Cluster 2 vs all}
-  \label{fig:2va}
-\end{figure}
-\begin{figure}[h]
-  \centering
   \includegraphics[width=\linewidth]{dtree-3va-pruned.pdf}
-  \caption{Cluster 3 vs all}
+  \caption{Cluster 3 (motor-dominated) vs all}
   \label{fig:3va}
@@
   \centering
-  \includegraphics[width=\linewidth]{dtree-4va-pruned.pdf}
-  \caption{Cluster 4 vs all}
-  \label{fig:4va}
+  \includegraphics[width=\linewidth]{dtree-2va-pruned.pdf}
+  \caption{Cluster 2 (severe) vs all}
+  \label{fig:2va}
 \end{figure}
-\section{Other Work}
+
+\section{Conclusion}
+
+$k$-means clustering on this Parkinsons' Disease data set reveals clusters that
+confirm previous computationally-based findings in the field
+\cite{vanrooden10}, mainly concerning the identification of four subtypes of
+Parkinson's disease: mild, nonmotor-predominant, motor-predominant, and severe.
+The most important nonmotor symptoms in determining these clusters were nms\_d2
+(sleep) and nms\_d3 (mood/cognition), which echo findings of Fereshtehnejad's
+longitudinal study \cite{fereshtehnejad15}. More work
+
+Nonmotor symptoms nms\_d2 and nms\_d3 became critical not only in
+classification trees distinguishing between the various symptoms but in the
+nonmotor-predominant subgroup itself, where both standard $k$-means analysis
+and decision tree branches show two possible trends in the
+manifestation of nonmotor-predominant PD:
+
+\begin{itemize}
+  \item[1.] Axial and sleep-severe PD
+  \item[2.] mood/cognition-severe PD
+\end{itemize}
+
+I am, however, not sure if these groups hae enough members to warrant a
+subtype, or whether this could just be chance or noise.
+
+It remains to be seen whether these classification models, especially the
+one-vs-all decision trees, are useful in clinical practice.
 
@@
 
-In Figure~\ref{fig:bnets}. Structure is too sparse, need to discretize or use
-some kind of regularization (e.g. a lasso)
+\subsubsection{On all data}
 
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=0.8\linewidth]{bnets.pdf}
-  \caption{Bayesian Networks}
-  \label{fig:bnets}
-\end{figure}
+I decided to discretize the data into three uniform-width groups based on the
+scales of each symptom. In other words, each symptom was discretized into a
+mild, moderate, and severe bin. Continuous data was unreliable on my computer,
+and updating intricately connected nodes like nms\_d2 resulted in slowdowns and
+crashes on my computer.
+
+Two bayesian network algorithms were tried: the default Bayesian score-search
+algorithm and the PC conditional independence tests algorithm. I couldn't find
+the exact name of the Bayesian search implementation, but it was the default
+method used by GeNIe. GeNIe files will be attached electronically.
+
+I assume these models are to be looked at by Dr. Mart\'in. I have not done too
+much investigation myself, as I'm not exactly sure what I'm looking for).
+
+\subsubsection{On nms-dominated data}
+I tried to construct Bayesian networks based on the nms-dominated subtype, but
+the data was too sparse to create a very informative network, even when leaving
+the information continuous. However, I'm not sure this is necessary. If it is,
+I can work on this problem more.
+
+\begin{thebibliography}{9}
+
+\bibitem{fereshtehnejad15}
+Fereshtehnejad et al (June 15, 2015). New Clinical Subtypes of Parkinson
+Disease and Their Longitudinal Progression
+
+\bibitem{vanrooden10}
+van Rooden et al (2010). The Identification of Parkinson's Disease Subtypes
+Using Cluster Analysis: A Systematic Review
+
+\end{thebibliography}
 

### 2015-11-23 writeup/abbreviated/abbreviated.tex
Update writeup with additional pots

@@
-\documentclass[letterpaper,12pt]{article}
+\documentclass[10pt]{article}
 \usepackage{fancyhdr}
@@
 \usepackage{listings}
+\usepackage{caption}
+\usepackage{subfig}
 \lstset{
@@
 % One more .. since we're in the abbreviated subdir
-\graphicspath{ {../../figures/abbreviated/} {../../figures/} }
+\graphicspath{ {../../figures/} }
+\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}
 
@@
 missing values of the features to be used in clustering (brought down to 901).
-It was decided to not impute the data.
+It was decided to not impute the data. Data was scaled to $\mu = 0, \sigma =
+1$ during clustering and modeling, then unscaled for visualization.
 
@@
     SSE Scree Plot & Inconclusive \\
-    Gap Statistic & 4, 13? \\
-    Affinity Propagation\tablefootnote{$\lambda = 0.98$, q = 0, maxits = 1000,
-    convits = 100} & 8 \\
+    Gap Statistic & 4 \\
+    Affinity Propagation & 8 \\%\tablefootnote{$\lambda = 0.98$, q = 0, maxits = 1000, %convits = 100} & 8 \\
 
@@
 
-\begin{figure}[h]
+\begin{figure*}[h]
   \centering
@@
   \label{fig:kmeans-summaries-4}
-\end{figure}
+\end{figure*}
 
 \subsubsection{Interpretation}
-$k$-means clustering ($k = 4$) found four clusters. They are numbered (somewhat
-confusingly) according to the automatically assigned numbers of the cluster
-analysis, and not by any intuitive pattern. With a brief description, they are:
+$k$-means clustering ($k = 4$) found four clusters. With a brief description,
+they are:
 
+% TODO: Figure out where the cluster reordering appears in other graphs
+% (kmeans, correlation) and redo those!!!
+% AND THE REST OF THE PAPER
+% TODO: LEVELS play a big factor - when I factor by reordering levels\dots
 \begin{itemize}
-  \item[4.] $(n = 406)$ Mildly affected in all domains.
-  \item[1.] $(n = 189)$ Severely affected in nonmotor domains; mildly
+  \item[1.] $(n = 406)$ Mildly affected in all domains.
+  \item[2.] $(n = 189)$ Severely affected in nonmotor domains; mildly
     affected in motor domains.
@@
     affected in nonmotor domains.
-  \item[2.] $(n = 88)$ Severely affected in all domains.
+  \item[4.] $(n = 88)$ Severely affected in all domains.
 \end{itemize}
 
-These cluster results are identical to the four clusters found in van Rooden et
-al \cite{vanrooden10}, which was done with a separate dataset using a different
-modeling method (expectation-maximization), supporting these subtype
-classifications. Unlike van Rooden, mean disease durations differences do exist
-between subtypes 4 (mild) and 2 (severe), likely due to further development of
-the disease, although the differences between 1 and 3 (nonmotor/motor
-predominated) subtypes are insignificant, suggesting different developmental
-paths of the disease.
-
 \subsubsection{Statistical Significance Tests, $k = 4$}
-Using one-way ANOVA for multiple means, we reject the null
-hypothesis that the means are the same with $p < 0.05$ for every variable
-\emph{except} pdonset.
+For each variable $i$ and cluster means $\mu_i^1, \mu_i^2, \mu_i^3, \mu_i^4$,
+we use one-way ANOVA for multiple means and reject the null hypothesis that
+$\mu_i^1 = \mu_i^2 = \mu_i^3 = \mu_i^4$ with $p < 0.05$ for every variable
+except pdonset.
 
@@
 differences between individual means is available in Table~\ref{tab:tukeyhsd}.
-Only statistically insignificant relations are provided; all other relations
-are significant with $p < 0.05$.
+For brevity, only statistically insignificant relations are provided; all other
+relations are significant with $p < 0.05$.
 
@@
   \hline
-  age & 3-1 & 0.724 \\
-      & 4-1 & 0.428 \\
+age & 2-1 & 0.428 \\
+    & 3-2 & 0.724 \\
     \hline
-  sex & 2-1 & 0.849 \\
-      & 4-1 & 0.092 \\
-      & 3-2 & 0.161 \\
-      & 4-2 & 0.827 \\
-      & 4-3 & 0.216 \\
+sex & 2-1 & 0.0918 \\
+    & 3-1 & 0.216 \\
+    & 4-1 & 0.827 \\
+    & 4-2 & 0.849 \\
+    & 4-3 & 0.161 \\
     \hline
-  pdonset & 2-1 & 0.147 \\
-          & 3-1 & 0.370 \\
-          & 4-1 & 0.859 \\
-          & 3-2 & 0.803 \\
-          & 4-2 & 0.305 \\
-          & 4-3 & 0.700 \\
+pdonset & 2-1 & 0.859 \\
+        & 3-1 & 0.700 \\
+        & 4-1 & 0.305 \\
+        & 3-2 & 0.370 \\
+        & 4-2 & 0.147 \\
+        & 4-3 & 0.803 \\
 \hline
-  durat\_pd & 3-1 & 0.562 \\
+durat\_pd & 3-2 & 0.562 \\
 \hline
-  cisitot & 3-1 & 0.523 \\
+cisitot & 3-2 & 0.522 \\
 \hline
-  nms\_d1 & 4-3 & 0.333 \\
+nms\_d1 & 3-1 & 0.333 \\
 \hline
-  nms\_d4 & 4-3 & 0.557 \\
+nms\_d4 & 3-1 & 0.557 \\
 \hline
-  nms\_d5 & 4-3 & 0.856 \\
+nms\_d5 & 3-1 & 0.856 \\
 \hline
-  nms\_d8 & 4-3 & 0.122 \\
+nms\_d8 & 3-1 & 0.122 \\
 \hline
-  nms\_d9 & 2-1 & 0.730 \\
-         & 4-3 & 0.074 \\
+nms\_d9 & 3-1 & 0.0735 \\
+       & 4-2 & 0.730 \\
 \hline
-  tremor & 4-1 & 0.360 \\
+tremor & 2-1 & 0.360 \\
   \end{tabular}
@@
 in determining clusters. We see, quite naturally, that standard measures of
-motor symptoms rank very highly (1, 2, 4, 5) in information gain \emph{except}
-termor (12). Similarly, bradykinesia (1) is used as the root node of the
+motor symptoms rank very highly (ranks 1, 2, 4, 5) in information gain \emph{except}
+tremor (12). Similarly, bradykinesia (1) is used as the root node of the
 4-cluster decision tree, although other motor symptoms are used further down
@@
 pdonset, plays almost no role in the determination of these clusters. That the
-time of onset of PD or sex nulls important, clinically-relevant questions about
-the demographic sources of these different subtypes.
+time of onset of PD or sex is largely irrelevant provides an important negative answer to
+clinically-relevant questions about the demographic sources of these different
+subtypes.
 
@@
   \centering
-  \begin{tabular}{l|l|l}
+  \begin{tabular}{r|l|l}
     rank & variable & information gain \\
     \hline
-    1 & bradykin      & 0.31574672 \\
-    2 & rigidity      & 0.29560018 \\
-    3 & nms\_d2      & 0.24218407 \\
-    4 & cisitot      & 0.22920103 \\
-    5 & axial      & 0.22780750 \\
-    6 & nms\_d3      & 0.20480570 \\
-    7 & nms\_d9      & 0.15782743 \\
-    8 & nms\_d7      & 0.15290569 \\
-    9 & nms\_d5      & 0.14454931 \\
-    10 & nms\_d6      & 0.14025139 \\
-    11 & nms\_d1      & 0.13212756 \\
-    12 & tremor      & 0.10937168 \\
-    13 & nms\_d4      & 0.10710526 \\
-    14 & nms\_d8      & 0.10005480 \\
-    15 & durat\_pd      & 0.02876190 \\
-    16 & age      & 0.02346158 \\
-    17 & sex      & 0.00000000 \\
-    18 & pdonset      & 0.00000000 \\
+    1 & bradykin      & 0.316 \\
+    2 & rigidity      & 0.296 \\
+    3 & nms\_d2      & 0.242 \\
+    4 & cisitot      & 0.229 \\
+    5 & axial      & 0.228 \\
+    6 & nms\_d3      & 0.205 \\
+    7 & nms\_d9      & 0.158 \\
+    8 & nms\_d7      & 0.153 \\
+    9 & nms\_d5      & 0.145 \\
+    10 & nms\_d6      & 0.140 \\
+    11 & nms\_d1      & 0.132 \\
+    12 & tremor      & 0.109 \\
+    13 & nms\_d4      & 0.107 \\
+    14 & nms\_d8      & 0.100 \\
+    15 & durat\_pd      & 0.0288 \\
+    16 & age      & 0.0235 \\
+    17 & sex      & 0.000 \\
+    18 & pdonset      & 0.000 \\
   \end{tabular}
@@
 The interplay between specific symptoms in each of the four clusters was
-examined in Figure~\ref{fig:corrplots}, but nothing too interesting was found. There is, perhaps somewhat
-interestingly, a somewhat higher correlation between bradykinesia, rigidity,
-and nms\_d6 (gastrointestinal), but it was not judged to be significant.
+examined in Figure~\ref{fig:corrplots}. There are two points of note. The first
+is that there is a higher correlation in cluster 4 (severe) between overall
+severity (cisitot) and bradykinesia and rigidity, illustrated in
+Figure~\ref{fig:corrxy-br}.
+Second, there exists a somewhat higher correlation between
+bradykinesia, rigidity, and nms\_d6 (gastrointestinal) in cluster 4,
+illustrated in Figure~\ref{fig:corrxy-d6}. These differences are statistically
+significant; correlation tests are located in Table~\ref{tab:cortests}. I am
+unsure of the significance or proper interpretation of these results.
+
+\begin{table*}[h]
+  \centering
+  \begin{tabular}{l|l|l|l}
+ Cluster & Variables & 95\% CI & $p$ \\
+    \hline
+1 & bradykin, cisitot&[-0.0225, 0.171]&0.131 \\
+  & rigidity, cisitot&[-0.000406, 0.192]&0.0510 \\
+  & bradykin, nms\_d6&[-0.0634, 0.131]&0.493 \\
+  & rigidity, nms\_d6&[-0.101, 0.0932]&0.934 \\
+  \hline
+2 & bradykin, cisitot&[0.0786, 0.351]&$0.00248^{(**)}$ \\
+  & rigidity, cisitot&[-0.215, 0.069]&0.310 \\
+  & bradykin, nms\_d6&[-0.152, 0.133]&0.897 \\
+  & rigidity, nms\_d6&[-0.123, 0.163]&0.781 \\
+  \hline
+3 & bradykin, cisitot&[0.0995, 0.350]&$0.000620^{(***)}$ \\
+  & rigidity, cisitot&[0.0687, 0.322]&$0.00298^{(**)}$ \\
+  & bradykin, nms\_d6&[0.0350, 0.292]&$0.0134^{(*)}$ \\
+  & rigidity, nms\_d6&[-0.0846, 0.179]&0.478 \\
+  \hline
+  4 & bradykin, cisitot&[0.454, 0.724]&${3.97\e{-10}}^{(***)}$ \\
+& rigidity, cisitot&[0.375, 0.675]&${4.99\e{-08}}^{(***)}$ \\
+  & bradykin, nms\_d6&[0.297, 0.624]&${2.60\e{-06}}^{(***)}$ \\
+  & rigidity, nms\_d6&[0.278, 0.611]&${6.43\e{-06}}^{(***)}$ \\
+  \end{tabular}
+  \caption{Correlation tests. (*) $p < 0.05$, (**) $p < 0.01$, (***), $p < 0.001$}
+  \label{tab:cortests}
+\end{table*}
 
@@
   \centering
-  \includegraphics[width=\linewidth]{corrplots.pdf}
+  \subfloat[][]{%
+  \includegraphics[clip,width=0.8\linewidth]{bradykin-v-cisitot.pdf}
+  \label{bradyv}
+}
+
+\subfloat[][]{%
+  \includegraphics[clip,width=0.8\linewidth]{rigidity-v-cisitot.pdf}
+  \label{rigidv}
+}
+  \caption{Relationship between \protect\subref{bradyv} bradykinesia,
+  \protect\subref{rigidv} rigidity and overall severity (cisitot). Shaded band
+is 95\% confidence interval.}
+  \label{fig:corrxy-br}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \subfloat[][]{%
+  \includegraphics[clip,width=0.8\linewidth]{bradykin-v-nms_d6.pdf}
+  \label{brady6}
+}
+
+\subfloat[][]{%
+  \includegraphics[clip,width=0.8\linewidth]{rigidity-v-nms_d6.pdf}
+  \label{rigid6}
+}
+  \caption{Relationship between \protect\subref{brady6} bradykinesia,
+  \protect\subref{rigid6} rigidity and nms\_d6 (gastrointestinal). Shaded band
+is 95\% confidence interval.}
+  \label{fig:corrxy-d6}
+\end{figure}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.68\linewidth]{corrplots.pdf}
   \caption{Correlation plots}
@@
 \section{Nonmotor-predominant subtype analysis}
-\subsection{$k$-means sub-subdivision on Cluster 1}
+\subsection{$k$-means sub-subdivision on Cluster 2}
+% TODO: Synthesize changed subtype 2 analysis with discussion HERE and in
+% conclusion. Make sure to reread and ensure that all plots have been updated
+% (in particular, the 4 cluster decision tree?). Incorporate some stuff from
+% standard progression of motor diseases paper and the new nms subtype paper.
 
@@
 
- The same $k$-determining tests were run on subtype 1 and are displayed
+The same $k$-determining tests were run on subtype 2 and are displayed
 in Table~\ref{tab:numclus-nms}.
@@
   \caption{Results of various techniques for determining $k$, applied to
-  subtype 1}
+  subtype 2}
   \label{tab:numclus-nms}
@@
 Boxplots for $k$-means run for $k = 2, 3, 4$ can
-be seen in Figures~\ref{fig:c1-summaries-2},~\ref{fig:c1-summaries-3},
-and~\ref{fig:c1-summaries-4}.
+be seen in Figures~\ref{fig:c2-summaries-2},~\ref{fig:c2-summaries-3},
+and~\ref{fig:c2-summaries-4}. Clusters are ordered by increasing cisitot.
 
@@
 An interesting set of subtleties occurs when $k = 2$ and 3. When $k = 2$, the
-two groups
-are generally somewhat similar, except nms\_d2 and nms\_d3 move in opposite
-directions, i.e.\ sub-subtype 2 generally has higher nms\_d2 scores but lower
-nms\_d3 scores. In addition, sub-subtype 2 has higher axial scores. This
-difference can be once again observed in $k = 3$, where subtype 3 is shown to
-have higher nms\_d2 scores, lower nms\_d3 scores, and higher axial scores. This
-may indicate a trend in nonmotor-dominated PD for nms\_d2 and nms\_d3 to live
-on opposite ends of a spectrum, i.e.\ patients especially severe in one may not
-be as severe in the other. This difference can also been seen in
-Section~\ref{sub:ova}, when OVA decision trees are discussed for the the
-nonmotor-dominated group.
+two groups are divided generally by PD severity (see cisitot and especially
+axial). The specific symptoms of the two groups follow this trend, except
+nms\_d3 and tremor, which are actually decreasing, and other symptoms like
+rigidity, nms\_d4, and nms\_d9, which are more indeterminate.
+
+When $k = 3$, the symptoms that continue show a non-monotonically increasing
+trend are nms\_d2, tremor, and rigidity scores, where patients in the 3rd
+subtype exhibit lower severities. nms\_d4 and nms\_d9 differences turn out to be
+not as pronounced.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{c1-summaries-2.pdf}
+  \includegraphics[width=\linewidth]{c2-summaries-2.pdf}
   \caption{Clustering on nonmotor group: $k = 2$}
-  \label{fig:c1-summaries-2}
+  \label{fig:c2-summaries-2}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\linewidth]{c1-summaries-3.pdf}
+  \includegraphics[width=\linewidth]{c2-summaries-3.pdf}
   \caption{Clustering on nonmotor group: $k = 3$}
-  \label{fig:c1-summaries-3}
+  \label{fig:c2-summaries-3}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\linewidth]{c1-summaries-4.pdf}
+  \includegraphics[width=\linewidth]{c2-summaries-4.pdf}
   \caption{Clustering on nonmotor group: $k = 4$}
-  \label{fig:c1-summaries-4}
+  \label{fig:c2-summaries-4}
 \end{figure}
 
-\section{Modeling}
+\section{Further modeling}
 
@@
 models that could be used in a clinical setting to predict the subtype of PD
-based on previous clustering results. Cluster assignments were treated as
-labels in a supervised classification problem in an attempt to produce useful
+based on previous clustering results. Cluster assignments obtained from
+previous $k$-means investigation were treated as labels in a supervised
+classification problem in an attempt to produce useful and easily interpretable
 models.
@@
 needed to make simpler diagnoses such as classifying a patient as mildly
-affected (subtype 4) or severely affected (subtype 2). One-versus-all (OVA)
+affected (subtype 1) or severely affected (subtype 4). One-versus-all (OVA)
 decision trees were thus considered, in order to isolate the classification
@@
 subtypes. These OVA decision trees for all 4 subtypes are located in
-Figures~\ref{fig:4va},~\ref{fig:1va},~\ref{fig:3va}, and~\ref{fig:2va}.
+Figures~\ref{fig:1va},~\ref{fig:2va},~\ref{fig:3va}, and~\ref{fig:4va}. Trees
+are pruned by selecting the version of tree with the minimum 10-fold
+cross-validated error.
 
-\subsubsection{4 (mild)}
-This tree is very odd. The root node considers nms\_d9 $< 7.5$ (miscellaneous)
-and classifies by far the biggest majority of negative examples when this
-symptom is milder, even though the subgroup 4 is the mild subtype. After
-intuitively then classifying on the mildness of rigidity the tree then proceeds
-to classify once again negative examples based on mildness of nms\_d2 (sleep)
-and positive examples solely based on more severe manifestations of nms\_d2,
-despite the means of these nonmotor symptoms being quite similar according to
-the boxplots.  However, accuracy on the final node is quite poor (61
-misclassifications) which could be an indicator that the cluster is not well
-defined.
+\subsubsection{1 (mild)}
+The tree for the mild subtype classifies mainly based on negative responses to
+nodes asking whether the patient has a relatively severe manifestation of a
+symptom. The majority of examples are classified by following the bradykinesia
+$<$ 2.5, which subsequently tests the severity of several nonmotor symptoms.
+Most of subtype 1 patients that score relatively mildly on these scales are
+classified this way. There are also small populations of patients who 1) score
+% http://www.ncbi.nlm.nih.gov/pubmed/22233888
+higher on bradykinesia but lower with axial, tremor, rigidity, and nms\_d2 and
+2) score higher in nms\_d2 (sleep) but lower with nms\_d7 (urinary).
+% This tree is very odd. The root node considers nms\_d9 $< 7.5$  and nms\_d3 (miscellaneous)
+% and classifies by far the biggest majority of negative examples when this
+% symptom is milder, even though the subgroup 1 is the mild subtype. After
+% intuitively then classifying on the mildness of rigidity (rigidity >= 3.5) the tree then proceeds
+% to classify once again negative examples mostly based on mildness of nms\_d2
+% (sleep)
+% and positive examples based on more severe manifestations of nms\_d2,
+% despite the means of these nonmotor symptoms being quite similar according to
+% the boxplots.  However, accuracy on the final nodes are quite poor, which could
+% be an indicator that the cluster is not well defined.
 
-\subsubsection{1 (nonmotor-predominant)}
-Quite characteristically, this OVA decision tree clusters first on the severity
-of bradykinesia. When mild (i.e.\ $< 2.5$), all of the subsequent decision
-nodes ask whether the patient has correspondingly severe forms of primarily
-nms\_d3 (mood/cognition), then nms\_d7 (urinary) and nms\_d9 (miscellaneous).
+\subsubsection{2 (nonmotor-predominant)}
+The decision tree for the nonmotor-predominant subtype is quite simple.
+Interestingly, although nms\_d9 (miscellaneous) is not the most important
+nonmotor symptom, since the information gain is less than nms\_d2 and nms\_d3
+and it does not appear very high in the 4-class decision tree, it is used as
+the root node of this decision tree, classifying over half of the negative
+examples based on whether the subject has a low severity of miscellaneous
+symptoms (nms\_d9 $<$ 7.5)\footnote{Recall that 0 $<$ nms\_d9 $<$ 48.}. This could be an indication that nonmotor-predominant PD patients do
+indeed have a wide manifestation and variety of nonmotor symptoms. After
+classifying on nms\_d9, the tree then classifies negatiave examples as having
+rigidity $\geq$ 3.5, an example of how subtype 2 patients have relatively low motor
+symptoms. Finally, the tree classifies on the nonmotor symptom with the most
+information gain, nms\_d2, where patients $\geq$ 7.5 are classified as falling into
+subtype 2.
 
-When bradykinesia (and all motor symptoms) are relatively severe, however,
-there is still a small subset of patients who classify into subtype 1 by having
-severe nms\_d2 (sleep) symptoms. While I am not sure if the sample size is
-large enough here (9/34), I propose this could be one of the ``subtypes'' of
-the nonmotor-predominant subtype, i.e.\ a group that, when accompanied with
-high motor symptoms, generally will have high nonmotor symptoms manifest
-primarily in the form of sleep disorders. This aligns with the findings
-previously mentioned in Section~\ref{sub:nms-interp} a disjunct between
-symptoms nms\_d2 and nms\_d3. As briefly mentioned in that section, axial motor
-symptom severity also plays a role. In this tree, mildness of axial symptoms is
-the main classifier of 280 negative examples when identifying
-nonmotor-predominant
-subtypes, indicating it is an important distinguishing factor of this unique
-group of 34 nonmotor-predominant PD patients.
+% Quite characteristically, this OVA decision tree clusters first on the severity
+% of bradykinesia. When mild (i.e.\ bradykin $< 2.5$), all of the subsequent decision
+% nodes ask whether the patient has correspondingly severe forms of primarily
+% nms\_d3 (mood/cognition), then nms\_d7 (urinary) and nms\_d9 (miscellaneous).
 
-More discussion in the
-conclusion.
+% When bradykinesia (and all motor symptoms) are relatively severe, however,
+% there is still a small subset of patients who classify into subtype 2 by having
+% severe nms\_d2 (sleep) symptoms. While I am not sure if the sample size is
+% large enough here (9/34), I propose this could be one of the ``subtypes'' of
+% the nonmotor-predominant subtype, i.e.\ a group that, when accompanied with
+% high motor symptoms, generally will have high nonmotor symptoms manifest
+% primarily in the form of sleep disorders. This aligns with the findings
+% previously mentioned in Section~\ref{sub:nms-interp} a disjunct between
+% symptoms nms\_d2 and nms\_d3. As briefly mentioned in that section, axial motor
+% symptom severity also plays a role. In this tree, mildness of axial symptoms is
+% the main classifier of 280 negative examples when identifying
+% nonmotor-predominant
+% subtypes, indicating it is an important distinguishing factor of this unique
+% group of 34 nonmotor-predominant PD patients.
+
+% More discussion on this decision tree, synthesized with the other results of
+% this research, is in the conclusion.
 
@@
 example, in the furthest right branch, once nms\_d2 (as we know, an important
-feature) is established to be relatively mild (< 12), the test for subtype 3
-involves two more nodes verifying the severity of rigidty and tremor. Similar
-behavior can be found in the tree branch testing nms\_d7 and nms\_d6, then
-rigidity and axial.
+feature) is established to be relatively mild ($<$ 12), the test for subtype 3
+involves several more nodes verifying the severity of rigidty, tremor, and
+axial, and the mildness of nms\_d7 (urinary).
 
-\subsubsection{2 (severe)}
+\subsubsection{4 (severe)}
 
-The OVA tree for severely affected in all areas is predictable, testing
+The OVA tree for patients severely affected in all areas is predictable, testing
 entirely on whether or not symptoms (both motor and nonmotor) are relatively
-severe. Ppositive nodes always appear to the right (no) of less-than checks.
+severe. Positive nodes always appear to the right (no) of less-than checks.
 Interestingly, however, nms\_d4 (percep/hallucinations), previously not of note, is used twice as the
 root node of a tree and again further down. As the boxplot display in
-Figure~\ref{fig:kmeans-dtree-4} shows, nms\_d4 is perhaps the most
-distinguishing symptom of subtype 4 against nonmotor-predominant subtype 1 in
-particular, as subtype 1 is relatively mild, in contrast to relatively
-comparable levels of severity for other nonmotor symptoms. This shows that
-issues with
-perception and hallucinations generally occur in only the most severe cases of
-PD, and are relatively rare when a patient exhibits a nonmotor-predominant form
-of PD.
+Figure~\ref{fig:kmeans-summaries-4} shows, nms\_d4 is perhaps the most
+distinguishing symptom of subtype 4 against nonmotor-predominant
+subtype 2 in particular, as subtype 2 has relatively mild percep/hallucination
+symptoms, in contrast to the comparable levels of severity for other nonmotor
+symptoms in both groups. This shows that issues with perception and
+hallucinations generally occur in only the most severe cases of PD, and are
+relatively rare when a patient exhibits a nonmotor-predominant form of PD.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{dtree-4va-pruned.pdf}
-  \caption{Cluster 4 (mild) vs all}
-  \label{fig:4va}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{dtree-1va-pruned.pdf}
-  \caption{Cluster 1 (nonmotor-dominated) vs all}
+  \includegraphics[width=0.6\linewidth]{dtree-1va-pruned.pdf}
+  \caption{Cluster 1 (mild) vs all}
   \label{fig:1va}
@@
   \centering
-  \includegraphics[width=\linewidth]{dtree-3va-pruned.pdf}
+  \includegraphics[width=0.6\linewidth]{dtree-2va-pruned.pdf}
+  \caption{Cluster 2 (nonmotor-dominated) vs all}
+  \label{fig:2va}
+\end{figure}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.6\linewidth]{dtree-3va-pruned.pdf}
   \caption{Cluster 3 (motor-dominated) vs all}
   \label{fig:3va}
+\end{figure}
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.6\linewidth]{dtree-4va-pruned.pdf}
+  \caption{Cluster 4 (severe) vs all}
+  \label{fig:4va}
+\end{figure}
+
+\subsection{Different angles of exploration: 2 and 4, 2 and 3 vs rest}
+
+There are many more interesting questions to be asked when examining the
+relationship between these clusters. One thing that may be helpful in
+understanding the relationship between the clusters is exploring different
+groupings of clusters for decision trees. The trees in
+Figure~\ref{fig:dtree-2and4va-pruned} and~\ref{fig:dtree-2and3va-pruned} are
+preliminary examples of this kind
+of exploration.
+
+\subsubsection{2 and 4 vs rest}
+In this tree, the node classifying examples as subtype 4 is
+localized to the furthest right branch. Predictably, examples in this node have
+scored relatively higher in rigidity ($\geq$ 3.5). Interestingly, a classification
+decision that is replicated in the 4 versus all decision tree is the decision
+to use nms\_d7 (urinary) as a node, where subtype 4 is classified as having
+relatively high nms\_d7 components ($\geq$ 12). Indeed, as shown in
+Figure~\ref{fig:kmeans-summaries-4}, the mean of nms\_d7 severity is similar to
+nms\_d4 in that it is especially higher in cluster 4 than in cluster 2.
+
+\subsubsection{2 and 3 vs rest}
+In this tree, classification of patients in subtype 3 is primarily dependent on
+asserting bradykin $\geq$ 2.5 then splitting on tremor $<$ 3.5 is considered.
+Interestingly, the two nonmotor symptoms that differentiate cluster 3 are
+nms\_d5 (attention/memory) and nms\_d7 (urinary). Additionally, nms\_d2, a
+quite important symptom, does not appear in the classification tree for this
+task.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=0.6\linewidth]{dtree-2and4va-pruned.pdf}
+  \caption{Clusters 2 (nms) and 4 (severe) vs rest (1 and 3)}
+  \label{fig:dtree-2and4va-pruned}
 \end{figure}
@@
   \centering
-  \includegraphics[width=\linewidth]{dtree-2va-pruned.pdf}
-  \caption{Cluster 2 (severe) vs all}
-  \label{fig:2va}
+  \includegraphics[width=0.6\linewidth]{dtree-2and3va-pruned.pdf}
+  \caption{Clusters 2 (nms) and 3 (motor) vs rest (1 and 4)}
+  \label{fig:dtree-2and3va-pruned}
 \end{figure}
-
-\section{Conclusion}
-
-$k$-means clustering on this Parkinsons' Disease data set reveals clusters that
-confirm previous computationally-based findings in the field
-\cite{vanrooden10}, mainly concerning the identification of four subtypes of
-Parkinson's disease: mild, nonmotor-predominant, motor-predominant, and severe.
-The most important nonmotor symptoms in determining these clusters were nms\_d2
-(sleep) and nms\_d3 (mood/cognition), which echo findings of Fereshtehnejad's
-longitudinal study \cite{fereshtehnejad15}. More work
-
-Nonmotor symptoms nms\_d2 and nms\_d3 became critical not only in
-classification trees distinguishing between the various symptoms but in the
-nonmotor-predominant subgroup itself, where both standard $k$-means analysis
-and decision tree branches show two possible trends in the
-manifestation of nonmotor-predominant PD:
-
-\begin{itemize}
-  \item[1.] Axial and sleep-severe PD
-  \item[2.] mood/cognition-severe PD
-\end{itemize}
-
-I am, however, not sure if these groups hae enough members to warrant a
-subtype, or whether this could just be chance or noise.
-
-It remains to be seen whether these classification models, especially the
-one-vs-all decision trees, are useful in clinical practice.
 
@@
 I assume these models are to be looked at by Dr. Mart\'in. I have not done too
-much investigation myself, as I'm not exactly sure what I'm looking for).
+much investigation myself, as I'm not exactly sure what I'm looking for.
 
@@
 
+\section{Preliminary Conclusions}
+
+\subsection{Overall clustering}
+$k$-means clustering on this Parkinsons' Disease data set reveals clusters that
+confirm previous computationally-based findings in the field, mainly van Rooden
+et al. \cite{vanrooden10} and the identification of four subtypes of
+Parkinson's disease: mild, nonmotor-predominant, motor-predominant, and severe.
+van Rooden's work was done with a separate dataset using a different
+modeling method (expectation-maximization), and this investigation
+independently confirms these subtype classifications. Unlike van Rooden, mean
+disease durations differences do exist between subtypes 1 (mild) and 4
+(severe), likely due to further development of the disease, although the
+differences between 2 and 3 (nonmotor/motor predominated) subtypes are
+insignificant (Table~\ref{tab:tukeyhsd}), suggesting different developmental paths of the disease.
+
+Overall, little information was found in pdonset,
+durat\_pd, or current age, according to Tables~\ref{tab:tukeyhsd}
+and~\ref{tab:info_gain}.  Mean ages were similar for subgroups 1, 2, and 3 ($p >
+0.05$, but different for the severe subtype 4, which makes sense given that
+patients in 4 also have longer disease durations. Specifically, clusters 1 and
+4 seem to be phenotypically quite similar, except at different stages of
+disease progression, given cluster 4's higher age and durat\_pd scores.
+
+However, clusters 2 and 3 clearly show different disease progression, one in
+the motor direction, and one in the nonmotor. Both groups have similar age,
+pdonset, and durat\_pd scores, but differ wildly in symptomatic expression.
+Cluster 2 is dominated by a high prevalence of nonmotor symptoms, such as
+nms\_d2, nms\_d3, nms\_d7, and nms\_d9. Cluster 3, however, is dominated by a
+high prevalence of motor symptoms, while most motor symptoms are similar to the
+mild cluster 1. Of note is that the tremor population mean
+is the highest cluster mean, even higher than the severe subtype 4. This
+motor-dominant cluster may thus overlap with Ma's tremor dominant/slow
+progression cluster \cite{ma15}.
+
+Generally, given stable pdonset scores and predictably increasing durat\_pd
+scores for clusters 1 and 4, Ma et al's rapid disease
+progression/late onset and tremor dominant/slow progression clusters
+\cite{ma15} were mostly not found in this dataset, save for the tremor-dominant
+motor cluster.
+
+The most important nonmotor symptoms in determining these clusters were nms\_d2
+(sleep) and nms\_d3 (mood/cognition), which echo findings of Fereshtehnejad's
+longitudinal study \cite{fereshtehnejad15} and are similar to Sauerbier's
+identification of sleep dominant and cognitive dominant clinical NMS subtypes
+\cite{sauerbier15}. Compared to Erro et al. \cite{erro13}, nonmotor/motor
+dominant subtypes were indeed found, but an additional subgroup with relatively
+severe levels of both motor and nonmotor symptoms were found. Erro's benign
+subtype groups possibly overlap with the mild cluster 1 found in this
+investigation.
+
+\subsection{Nonmotor subtype: clustering and modeling}
+Nonmotor symptoms nms\_d2 and nms\_d3 became critical not only in
+classification trees distinguishing between the various symptoms but in the
+nonmotor-predominant subgroup itself. In $k$-means subdivision of the
+nonmotor-dominant subtype where $k = 2$ and $k = 3$, opposite trends were
+confirmed with nms\_d2 and nms\_d3 symptoms. Similarly, in the 2 and 4 vs rest
+decision tree (Figure~\ref{fig:dtree-2and4va-pruned}), nms\_d2 and nms\_d3 nodes were
+used to differentiate various categories of nomotor-dominant patients.
+When $k = 3$, the subtype with the highest nms\_d2 scores and lowest
+nms\_d3 scores had by far the highest axial scores, nms\_d6 (gastrointestinal)
+scores, and nms\_d7 (urinary) scores. Thus subtype 3 of the nonmotor-dominated
+group could include patients falling into the cognitive/depression-dominant or
+autonomic dominant subtypes.
+
+Despite the variety in symptomatic expression in this nonmotor group, what
+seems most consistent is the presence of nms\_d9 (miscellaneous) nonmotor
+symptoms, as it is used as the root node of the 2 vs all decision tree
+(Figure~\ref{fig:2va}) and the 2 and 4 vs rest decision tree
+(Figure~\ref{fig:dtree-2and4va-pruned}).
+
+It remains to be seen whether these classification models, especially the
+one-vs-all decision trees, are useful in clinical practice.
+
 \begin{thebibliography}{9}
 
+\bibitem{erro13}
+Erro et al (2013). The Heterogeneity of Early Parkinson’s Disease: A Cluster
+Analysis on Newly Diagnosed Untreated Patients
+
 \bibitem{fereshtehnejad15}
-Fereshtehnejad et al (June 15, 2015). New Clinical Subtypes of Parkinson
+Fereshtehnejad et al (2015). New Clinical Subtypes of Parkinson
 Disease and Their Longitudinal Progression
+
+\bibitem{ma15}
+Ma et al (2015). Heterogeneity among patients with Parkinson's disease: Cluster
+analysis and genetic association
+
+\bibitem{sauerbier15}
+Sauerbier et al (2015). Non motor subtypes and Parkinson's disease.
 

### 2015-11-23 R/kmeans-dtree.R

@@ -10,8 +10,10 @@ library(mclust)
 library(ggplot2)
 library(tidyr)
 library(FSelector)
-library(apclust)
+# library(apclust)
 library(RColorBrewer) # For colors for decision tree plots
+my.palette = brewer.pal(8, "Pastel1")
+
 
 # CONSTANTS ====
 # INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot")
@@ -25,15 +27,34 @@ INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
 # kmeans analysis
 # explore plots is the determining clusters plots
 SAVE.EXPLORE.PLOTS <- FALSE
-SAVE.DTREES <- FALSE
+SAVE.DTREES <- TRUE
 SAVE.OVA.DTREES <- FALSE
 SAVE.BOXPLOTS <- FALSE
 SAVE.ARFF <- FALSE
+SAVE.2VA.DTREES <- FALSE
 # TODO: Make k = 2, 3, 4 modifiable via constant
 
 # LOAD DATA ====
 source('./preprocessing.R')
 
+rename.clusters <- function(cl) {
+  # Does passing in the argument not pass by reference?
+  # I'm just putting this (ugly) code everywhere else
+  for (i in 1:length(cl$cluster)) {
+    curr.cluster <- cl$cluster[[i]]
+
+    if (curr.cluster == 1) { # nms
+      cl$cluster[[i]] <- 2
+    } else if (curr.cluster == 2) { # all severe
+      cl$cluster[[i]] <- 4
+    } else if (curr.cluster == 3) { # motor
+      cl$cluster[[i]] <- 3
+    } else if (curr.cluster == 4) { # mild
+      cl$cluster[[i]] <- 1
+    }
+
+  }
+}
 # VISUALIZE WSS ERROR TO FIND OPTIMAL K ====
 # NOTE: Doesn't work well, there isn't any elbow
 
@@ -58,11 +79,11 @@ if (SAVE.EXPLORE.PLOTS) {
 # it finds to be optimal, set it to search for
 # at least 1 model and up 20.
 # Traditionally this takes a while, so it's uncommented
-d_clust <- Mclust(as.matrix(raw.filtered), G=1:20)
-m.best <- dim(d_clust$z)[2]
-cat("model-based optimal number of clusters:", m.best, "\n")
+# d_clust <- Mclust(as.matrix(raw.filtered), G=1:20)
[…]

### 2015-11-23 R/nms.R

@@ -1,24 +1,29 @@
 # This contains further exploration of the nms-dominated subgroup
+# added 20151022: it's c2 now
 
 # Libraries ====
 library(plyr)
 library(ggplot2)
+library(mclust)
+library(cluster)
+library(tidyr)
 
 # Preprocessing ====
 # At this point not necessary since I'm doing ad hoc exploration
 # source('./preprocessing.R')
+green.palette = c("#E6FAE6", "#80E680", "#00CD00", "#006600")
 
 # Constants ====
-SAVE.C1PLOTS <- FALSE
+SAVE.C2PLOTS <- TRUE
 
 # VISUALIZE WSS ERROR TO FIND OPTIMAL K ====
 # NOTE: Doesn't work well, there isn't any elbow
 
 # Just in case it's still defined from below
 var <- NULL
-wss <- (nrow(c1.woclass)-1)*sum(apply(c1.woclass, 2, var))
+wss <- (nrow(c2.woclass)-1)*sum(apply(c2.woclass, 2, var))
 for (i in 2:15) {
-  wss[i] <- sum(kmeans(c1.woclass, i, nstart = 25)$withinss)
+  wss[i] <- sum(kmeans(c2.woclass, i, nstart = 25)$withinss)
 }
 
 plot(1:15, wss, type="b",
@@ -35,15 +40,15 @@ if (SAVE.EXPLORE.PLOTS) {
 # it finds to be optimal, set it to search for
 # at least 1 model and up 20.
 # Traditionally this takes a while, so it's uncommented
-d_clust <- Mclust(as.matrix(c1.woclass), G=1:20)
+d_clust <- Mclust(as.matrix(c2.woclass), G=1:20)
 m.best <- dim(d_clust$z)[2]
 cat("model-based optimal number of clusters:", m.best, "\n")
 # 3 clusters
-plot(d_clust)
+# plot(d_clust) #, this is interactive use 0 to quit
 
 # NBCLUST ESTIMATION FOR OPTIMAL K (30 metrics) ====
 # This is taking a long time
-# nb <- NbClust(c1.woclass, distance = "euclidean",
+# nb <- NbClust(c2.woclass, distance = "euclidean",
 #               min.nc=2, max.nc=15, method = "kmeans",
 #               index = "alllong", alphaBeale = 0.1)
 # hist(nb$Best.nc[1,], breaks = max(na.omit(nb$Best.nc[1,])))
@@ -52,7 +57,7 @@ plot(d_clust)
[…]

### 2016-03-03 R/nms30.R
More work on clustering with nms30

@@ -1,0 +1,366 @@
+# Performing column clustering on only nms30. like nms-part2.R but graduated
+# Depends on everything.wide data frame used in longitudinal.R
+# NMS.30, NMS.30.NAMES from preprocessing
+library(plyr)
+library(FactoMineR)
+library(rpart)
+library(rpart.plot)
+library(nFactors)
+library(gplots) # For heatmaps
+# Better dendrograms
+library(dendextend)
+library(dynamicTreeCut)
+# Trying to get some color here
+library(colorspace)
+library(sparcl)
+library(cluster)
+library(fpc)
+library(mclust)
+library(RColorBrewer)
+library(fmsb)
+library(proxy)
+library(reshape2)
+library(ggplot2)
+library(tidyr)
+library(DMwR)
+gg_color_hue <- function(n) {
+  hues = seq(15, 375, length=n+1)
+  hcl(h=hues, l=65, c=100)[1:n]
+}
+# library(apcluster)
+
+SAVE.NMS30.PLOTS <- TRUE
+
+# Preprocess ====
+NMS.30.MAP <- setNames(NMS.30.NAMES, NMS.30)
+
+nms30 <- everything.wide[, NMS.30]
+nms30 <- rename(nms30, NMS.30.MAP)
+# Standardize
+nms30.s <- as.data.frame(scale(nms30))
+
+# With motor
+nms30m <- nms30
+nms30m[, MOTOR.SYMPTOMS] <- everything.wide[, MOTOR.SYMPTOMS]
+
+nms30m.scale <- scale(nms30m)
+nms30m.s <- as.data.frame(nms30m.scale)
+
+# Transposes (right now, don't need)
+# nms30.t <- as.data.frame(t(nms30))
+# nms30.st <- as.data.frame(t(nms30.s))
+
+# Dimensionality on this is ridiculous. Not sure it makes sense.
+
+# Factor analysis ====
+# FactoMineR PCA
+pca <- PCA(nms30.s)
+# Obviously depression symptoms are on their own
+# Plot variance explained
+plot(pca$eig$`cumulative percentage of variance`,
+     main="Cumulative variance explained",
+     ylab="% variance",
+     xlab="Component"
+     )
+
+# Look at PCA contributions
+# HELP: Not sure how to interpret PCA
[…]

### 2016-03-03 R/preprocessing.R

@@ -24,6 +24,38 @@ MOTOR.SYMPTOMS <- c("tremor",
                    "bradykin",
                    "rigidity",
                    "axial")
+NMS.30.NAMES <- c(
+  'd1-1-lightheaded',
+  'd1-2-fainting',
+  'd2-3-drowsiness',
+  'd2-4-fatigue',
+  'd2-5-insomnia',
+  'd2-6-rls',
+  'd3-7-loss_interest',
+  'd3-8-loss_activities',
+  'd3-9-anxiety',
+  'd3-10-depression',
+  'd3-11-flat_affect',
+  'd3-12-loss_pleasure',
+  'd4-13-hallucination',
+  'd4-14-delusion',
+  'd4-15-diplopia',
+  'd5-16-loss_concentration',
+  'd5-17-forget_explicit',
+  'd5-18-forget_implicit',
+  'd6-19-drooling',
+  'd6-20-swallowing',
+  'd6-21-constipation',
+  'd7-22-urinary_urgency',
+  'd7-23-urinary_frequency',
+  'd7-24-nocturia',
+  'd8-25-sex_drive',
+  'd8-26-sex_dysfunction',
+  'd9-27-unexplained_pain',
+  'd9-28-gust_olfact',
+  'd9-29-weight_change',
+  'd9-30-sweating'
+)
 NMS.30 <- c(
   "nms1", "nms2", "nms3", "nms4", "nms5", "nms6",
   "nms7", "nms8", "nms9", "nms10", "nms11", "nms12",
@@ -31,6 +63,8 @@ NMS.30 <- c(
   "nms19", "nms20", "nms21", "nms22", "nms23", "nms24",
   "nms25", "nms26", "nms27", "nms28", "nms29", "nms30"
 )
+ALL.BUT.NMS <- c('age', 'sex', 'pdonset', 'durat_pd', 'cisitot',
+                 'tremor', 'bradykin', 'rigidity', 'axial')
 INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",
                  "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5",
                  "nms_d6", "nms_d7", "nms_d8", "nms_d9",

### 2016-03-03 writeup/abbreviated/abbreviated.tex
Almost done with second version of abbrv

@@
 \usepackage{numprint}
-\usepackage[margin=1in]{geometry}
+\usepackage{float}
+\usepackage[margin=0.75in]{geometry}
 \usepackage{graphicx}
@@
 \usepackage{caption}
+\usepackage{multicol}
 \usepackage{subfig}
@@
    basicstyle=\ttfamily}
-% \pagestyle{fancy}
-% \lhead{Jesse Mu}
-% \rhead{CSCI339 Term Project}
-% One more .. since we're in the abbreviated subdir
-\graphicspath{ {../../figures/} }
+% Include longitudinal
+\graphicspath{ {../../figures/} {../../figures/longitudinal/} }
 \providecommand{\e}[1]{\ensuremath{\times 10^{#1}}}
@@
 \maketitle
+
+\tableofcontents
+\newpage
+
+\begin{multicols}{2}
+
+\section{New Content}
+
+Section~\ref{sec:longitudinal} on longitudinal analysis; section~\ref{sec:nms30} on
 
@@
 
-\begin{table}[h]
+\begin{table}[H]
   \centering
@@
 $k$-means clustering with $k = 4$ was tried. Statistics for determining the
-optimal number of clusters were used, but were inconclusive: results in
+optimal number of clusters were used, but were generally inconclusive: results in
 Figure~\ref{tab:numclus}. This probably indicates that the data is not very
@@
 groups.
-\begin{table}[h]
+\begin{table}[H]
   \centering
@@
     Affinity Propagation & 8 \\%\tablefootnote{$\lambda = 0.98$, q = 0, maxits = 1000, %convits = 100} & 8 \\
-
+    clValid stability measures & 4 \\
   \end{tabular}
@@
 \subsection{Decision tree}
-% \begin{table}[h]
+% \begin{table}[H]
   % \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
@@
 
-\begin{figure*}[h]
+\begin{figure*}[t]
   \centering
@@
 
-\begin{table*}[h]
+\begin{table*}[t]
   \centering
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
   \subfloat[][]{%
-  \includegraphics[clip,width=0.8\linewidth]{bradykin-v-cisitot.pdf}
+  \includegraphics[clip,width=\linewidth]{bradykin-v-cisitot.pdf}
   \label{bradyv}
@@
 \subfloat[][]{%
-  \includegraphics[clip,width=0.8\linewidth]{rigidity-v-cisitot.pdf}
+  \includegraphics[clip,width=\linewidth]{rigidity-v-cisitot.pdf}
   \label{rigidv}
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
   \subfloat[][]{%
-  \includegraphics[clip,width=0.8\linewidth]{bradykin-v-nms_d6.pdf}
+  \includegraphics[clip,width=\linewidth]{bradykin-v-nms_d6.pdf}
   \label{brady6}
@@
 \subfloat[][]{%
-  \includegraphics[clip,width=0.8\linewidth]{rigidity-v-nms_d6.pdf}
+  \includegraphics[clip,width=\linewidth]{rigidity-v-nms_d6.pdf}
   \label{rigid6}
@@
 \end{figure}
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
-  \includegraphics[width=0.68\linewidth]{corrplots.pdf}
+  \includegraphics[width=\linewidth]{corrplots.pdf}
   \caption{Correlation plots}
@@
 
-\begin{figure}[h]
+\begin{figure*}[p]
   \centering
-  \includegraphics[width=\linewidth]{c2-summaries-2.pdf}
+  \includegraphics[width=0.65\linewidth]{c2-summaries-2.pdf}
   \caption{Clustering on nonmotor group: $k = 2$}
   \label{fig:c2-summaries-2}
-\end{figure}
-\begin{figure}[h]
+\end{figure*}
+\begin{figure*}[p]
   \centering
-  \includegraphics[width=\linewidth]{c2-summaries-3.pdf}
+  \includegraphics[width=0.65\linewidth]{c2-summaries-3.pdf}
   \caption{Clustering on nonmotor group: $k = 3$}
   \label{fig:c2-summaries-3}
-\end{figure}
-\begin{figure}[h]
+\end{figure*}
+\begin{figure*}[p]
   \centering
-  \includegraphics[width=\linewidth]{c2-summaries-4.pdf}
+  \includegraphics[width=0.65\linewidth]{c2-summaries-4.pdf}
   \caption{Clustering on nonmotor group: $k = 4$}
   \label{fig:c2-summaries-4}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
-  \includegraphics[width=0.6\linewidth]{dtree-1va-pruned.pdf}
+  \includegraphics[width=0.75\linewidth]{dtree-1va-pruned.pdf}
   \caption{Cluster 1 (mild) vs all}
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
-  \includegraphics[width=0.6\linewidth]{dtree-2va-pruned.pdf}
+  \includegraphics[width=0.75\linewidth]{dtree-2va-pruned.pdf}
   \caption{Cluster 2 (nonmotor-dominated) vs all}
@@
 \end{figure}
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
-  \includegraphics[width=0.6\linewidth]{dtree-3va-pruned.pdf}
+  \includegraphics[width=0.75\linewidth]{dtree-3va-pruned.pdf}
   \caption{Cluster 3 (motor-dominated) vs all}
@@
 \end{figure}
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
-  \includegraphics[width=0.6\linewidth]{dtree-4va-pruned.pdf}
+  \includegraphics[width=0.75\linewidth]{dtree-4va-pruned.pdf}
   \caption{Cluster 4 (severe) vs all}
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
-  \includegraphics[width=0.6\linewidth]{dtree-2and4va-pruned.pdf}
+  \includegraphics[width=0.75\linewidth]{dtree-2and4va-pruned.pdf}
   \caption{Clusters 2 (nms) and 4 (severe) vs rest (1 and 3)}
@@
 
-\begin{figure}[h]
+\begin{figure}[H]
   \centering
-  \includegraphics[width=0.6\linewidth]{dtree-2and3va-pruned.pdf}
+  \includegraphics[width=0.75\linewidth]{dtree-2and3va-pruned.pdf}
   \caption{Clusters 2 (nms) and 3 (motor) vs rest (1 and 4)}
@@
 
+\section{Longitudinal Analysis}
+\label{sec:longitudinal}
+
+I wanted to observe the progression of nonmotor symptoms of Parkinson's disease akin to work done
+by various studies (e.g.\ \cite{onsetpd,vu12,zahodne12}). The data was a little bit too noisy when
+I tried to examine correlations, so I decided to average the values into bins for
+\texttt{durat\_pd}: 0-2 years, 2-4 years, 4-6 years, etc. First, the correlation based on PD
+duration for each symptom is displayed in Figure~\ref{fig:longcorr}.
+
+\begin{quote}
+  Note: I have not included all of the graphs I have, just those I found interesting. If you would
+  like all of the graphs, I can send those to you in a separate email.
+\end{quote}
+
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=0.9\linewidth]{pd-durat-cor.pdf}
+  \caption{Correlation of variables according to PD Duration}
+  \label{fig:longcorr}
+\end{figure*}
+
+Most symptoms have at least a minor positive correlation. Interestingly, tremor, nms9 (anxiety)
+and nms10 (depression) do not have a strong correlation with PD duration.
+% This could give evidence that whether or not someone has high tremor is a distinct subtype?
+
+\subsection{Anxiety}
+I plotted the mean nms\_9 (anxiety) score for each durat\_pd bin in
+Figures~\ref{fig:nms9-long} and ~\ref{fig:nms9-multi}
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=0.8\linewidth]{nms9-durat-counts.pdf}
+  \caption{Mean anxiety (nms\_9) score by PD duration, with number of patients in each group at
+  bottom.}
+  \label{fig:nms9-long}
+\end{figure*}
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=0.8\linewidth]{nms9-multi-durat-counts.pdf}
+  \caption{Mean anxiety (nms\_9) score per subtype, with number of patients in each group at
+  bottom.}
+  \label{fig:nms9-multi}
+\end{figure*}
+
+\subsection{Depression}
+I plotted the mean nms\_10 (depression) score for each durat\_pd bin in
+Figures~\ref{fig:nms10-long} and ~\ref{fig:nms10-multi}.
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=0.8\linewidth]{nms10-durat-counts.pdf}
+  \caption{Mean depression (nms\_10) score by PD duration, with number of patients in each group at
+  bottom.}
+  \label{fig:nms10-long}
+\end{figure*}
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=0.8\linewidth]{nms10-multi-durat-counts.pdf}
+  \caption{Mean depression (nms\_10) score per subtype, with number of patients in each group at
+  bottom.}
+  \label{fig:nms10-multi}
+\end{figure*}
+
+\subsection{Cisitot}
+
+Note that cisitot has the highest correlation of all of the PD symptoms. I plotted cisitot per
+subtype in Figure~\ref{fig:cisitot-multi}. Interestingly, subtypes 2, 3, and 4 generally start from
+the same mean motor symptom score, suggesting different paths of disease progression.
+% same global mean motor symptom score. % TODO: This is important.
+
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=0.8\linewidth]{cisitot-multi-durat-counts.pdf}
+  \caption{Mean cisitot score per subtype, with number of patients in each group at
+  bottom.}
+  \label{fig:cisitot-multi}
+\end{figure*}
+
+\subsection{Tremor}
+
+I plotted tremor symptoms in Figure~\ref{fig:tremor-multi}. Notice how in this group, subtype 3
+(motor-dom) is higher than all groups towards the beginning of disease progression. This supports
+the idea that Subtype 3 is Ma's \cite{ma15} tremor-dominant cluster.
+
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=0.8\linewidth]{tremor-multi-durat-counts.pdf}
+  \caption{Mean tremor score per subtype, with number of patients in each group at
+  bottom.}
+  \label{fig:tremor-multi}
+\end{figure*}
+
+% TODO: Maybe later
+% \subsection{By \texttt{pdonset}}
+
+\section{Extended Nonmotor Symptoms}
+\label{sec:nms30}
+
+\subsection{Adding nonmotor symptoms to original $k$-means clustering}
+
+First, I used the clustering with the original 9 nms symptoms, but plotted the original 30 symptoms
+instead. The reason for this was to treat the clustering of the 9 symptoms as a form of
+dimensionality reduction, since we know each symptom of nms\{1-30\} is colinear with several other
+symptoms, and they form the broader nms\_d\{1-9\} symptoms. The results are in
+Figure~\ref{fig:d9to30}
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=\linewidth]{d9to30.png}
+  \caption{All symptoms (nms\{1-30\}) plotted with the original clustering results of the the 9
+broader symptoms. Interesting parts of the figure are are labeled in red.}
+  \label{fig:d9to30}
+\end{figure*}
+
+\subsection{\emph{New} clustering with 30 symptoms}
+
+\subsubsection{Principal Component Analysis}
+
+I ran PCA on the dataset with solely the 30 nonmotor symptoms. A plot of various metrics for
+determining the optimal number of eigenvectors is in Figure~\ref{fig:nms30only-eigs}. Most metrics agree
+Nthat the optimal number of eigenvalues is around 6, but notice how one component has by far the
+highest eigenvalue; this component can be interpreted as a general ``nonmotor severity'' measure,
+indicating that much of the general nonmotor symptomatic expression falls along the same dimension.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{nms30only-eigs.pdf}
+  \caption{Eigenvalues and other metrics for Principal Component Analysis applied to the NMS30 data.}
+  \label{fig:nms30only-eigs}
+\end{figure}
+
+This principal component can be shown more clearly in the correlation plot of each symptom with the
+first 5 components located in Figure~\ref{fig:nms30only-eigscorr}. In the first component, all of the
+symptoms have at least a minor positive correlation. The remaining components are composed
+primarily of dimensionality reduction in domain 3 (mood/cognition), domain 5 (attention/memory),
+domain 7 (urinary), and domain 9 (misc). It is once again demonstrated that mood/cognition symptoms
+are very important parts of this dataset.
+
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=\linewidth]{nms30only-eigscorr.pdf}
+  \caption{Correlation of each nonmotor symptom with the first 6 components}
+  \label{fig:nms30only-eigscorr}
+\end{figure*}
+
+\subsubsection{Symptoms clustering}
+
+\paragraph{Nonmotor only}
+
+I used hierarchical clustering to cluster the 30 symptoms of the dataset using complete linkage.
+Results are in Figure~\ref{fig:nms30only-colhc}. Naturally, symptoms that belong to the same domain are
+clustered together. But the hierarchical clustering also shows how some symptomatic domains are
+closely related; for example, d6 (gastro), d7 (urinary), and d8(sexual). Additionally,
+miscellaneous symptoms are quite distinct from the other symptoms.
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=\linewidth]{nms30only-colhc.pdf} \caption{Hierarchical clustering of symptoms.
+    Dendrogram colored with 5 clusters. Naming is: d\{1-9\}-\{1-30\}-description}
+  \label{fig:nms30only-colhc}
+\end{figure*}
+
+\paragraph{With motor symptoms}
+
+In Figure~\ref{fig:nms30m-colhc}, the results are similar to the original hierarchical
+clustering; but notice that motor symptoms are clustered together \emph{except} for tremor. In
+fact, tremor is the most dissimilar symptom of them all!
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=\linewidth]{nms30m-colhc.pdf} \caption{Hierarchical clustering of symptoms
+  including motor symptoms.  Dendrogram colored with 4 clusters.}
+  \label{fig:nms30m-colhc}
+\end{figure*}
+
+\subsubsection{$k$-means}
+
+More so than previously, metrics for identifying the number of clusters proved inconclusive. Most
+metrics suggested 2 clusters, which isn't a useful clustering. I decided to stick with $k = 4$ to
+compare the results of this new clustering with the previous run. All 30 NMS symptoms were included
+as well as the four motor symptoms axial, rigidity, bradykinesia, and tremor.
+In this clustering, 4 slightly different subtypes are identified:
+
+\begin{enumerate}
+  \item ($n = 509$) Mildly affected in all domains.
+  \item ($n = 97$) Higher than average nonmotor symptoms, but severely affected in depressive
+    symptoms especially.
+  \item ($n = 249$) Averagely affected in all domains.
+  \item ($n = 49$) Severely affected in all domains.
+\end{enumerate}
+
+Because the number of symptoms is high, I don't provide the boxplots I have previously provided;
+instead, I provide a heatmap of $z$-scores in Figure~\ref{fig:nms30m-k4-heatmap-improved} and a
+table of results in Table~\ref{tab:nms30_k4}.
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=0.6\linewidth]{nms30m-k4-heatmap-improved.png}
+  \caption{Heatmap for $k$-means clustering, $k = 4$, with the 30 nonmotor symptoms. Notice that
+  subtype 2 is a ``Depression-Dominant'' subtype.} \label{fig:nms30m-k4-heatmap-improved}
+\end{figure*}
+
+\begin{table*}[b]
+  \centering
+  \caption{Mean symptom scores for the centers of the $k$-means clustering where $k = 4$. In
+  this case, a Depression-Dominant subtype has been identified.}
+  \label{tab:nms30_k4}
+  \begin{tabular}{l|l|l|l|l}
+cluster & 1 & 2 & 3 & 4 \\
+\hline
+d1-1-lightheaded & 2.08 & 2.43 & 0.55 & 5.18 \\
+d1-2-fainting & 0.17 & 0.53 & 0.13 & 2.45 \\
+d2-3-drowsiness & 2.51 & 2.77 & 0.97 & 7 \\
+d2-4-fatigue & 4.84 & 6.29 & 1.3 & 7.31 \\
+d2-5-insomnia & 3.16 & 5.46 & 1.2 & 4.22 \\
+d2-6-rls & 1.88 & 2.31 & 0.66 & 3.88 \\
+d3-7-loss\_interest & 1.04 & \textbf{6.08} & 0.41 & 4.59 \\
+d3-8-loss\_activities & 1.7 & \textbf{7.14} & 0.7 & 5.37 \\
+d3-9-anxiety & 1.97 & \textbf{6.63} & 0.91 & 4.57 \\
+d3-10-depression & 2.35 & \textbf{7.81} & 0.81 & 4.29 \\
+d3-11-flat\_affect & 0.91 & \textbf{4.95} & 0.39 & 2.71 \\
+d3-12-loss\_pleasure & 1.09 & \textbf{6.82} & 0.36 & 3.43 \\
+d4-13-hallucination & 0.58 & 0.82 & 0.2 & 4.63 \\
+d4-14-delusion & 0.16 & 1.49 & 0.11 & 3.14 \\
+d4-15-diplopia & 0.56 & 1.01 & 0.16 & 4.39 \\
+d5-16-loss\_concentration & 2.47 & 4.02 & 0.91 & 6.55 \\
+d5-17-forget\_explicit & 2.18 & 3.77 & 0.86 & 6.69 \\
+d5-18-forget\_implicit & 1.76 & 3.12 & 0.68 & 6.65 \\
+d6-19-drooling & 3 & 2.99 & 0.67 & 6.1 \\
+d6-20-swallowing & 1.74 & 1.24 & 0.37 & 4.53 \\
+d6-21-constipation & 3.31 & 3.63 & 1.65 & 6.96 \\
+d7-22-urinary\_urgency & 3.84 & 3.49 & 0.99 & 7.8 \\
+d7-23-urinary\_frequency & 3.79 & 3.98 & 0.94 & 6.51 \\
+d7-24-nocturia & 5.13 & 4.99 & 1.67 & 7.65 \\
+d8-25-sex\_drive & 2.34 & 4.46 & 0.76 & 5.16 \\
+d8-26-sex\_dysfunction & 2.39 & 3.21 & 0.79 & 4.61 \\
+d9-27-unexplained\_pain & 2.98 & 2.48 & 0.7 & 4.24 \\
+d9-28-gust\_olfact & 3.18 & 3.57 & 1.49 & 4.73 \\
+d9-29-weight\_change & 1.86 & 2.54 & 0.88 & 3.92 \\
+d9-30-sweating & 2.77 & 2.2 & 0.68 & 3.18 \\
+\hline
+tremor & 2.59 & 2.12 & 2.53 & 4.1 \\
+bradykin & 2.83 & 2.76 & 1.98 & 3.86 \\
+rigidity & 2.61 & 2.42 & 1.87 & 3.61 \\
+axial & 4.22 & 4.4 & 2.17 & 7.24 \\
+  \end{tabular}
+\end{table*}
+
+\subsubsection{Model-based expectation-maximization}
+
+Because the dimensionality is quite high in this case; $k$-means may not have provided particularly
+accurate results; thus, I tried using a Gaussian mixture model-based clustering package. Various
+models were tried, with Bayesian Information Criterion (BIC) plotted in
+Figure~\ref{fig:nms30m-bic}. The best
+fitting model(s) that maximized the Bayesian Information Criterion include the diagonal, varying
+volume, equal shape Gaussian mixture model (VEI) at 11 clusters and the ellipsoidal, equal shape
+model (VEV) with 3 clusters. However, the VEV model wasn't interesting; it simply partitioned
+individuals into low, medium, and high overall severity.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{nms30m-bic.pdf}
+  \caption{Bayesian Information Criterion for various Gaussian mixture models, plotted against the
+  number of clusters in each model. The best model found is the diagonal, varying volume, equal
+shape Gaussian mixture model (VEI) at 11 clusters; notice also that the ellipsoidal, equal shape
+model (VEV) is able to achieve a relatively high BIC with only 3 clusters. The VEV model, however,
+is not particularly informative.}
+  \label{fig:nms30m-bic}
+\end{figure}
+
+The VEI model, with 11 clusters, is more interesting. There are many clusters in the data that are
+expected, including a few generally mild clusters and a few generally severe clusters.  But, by
+isolating more ``interesting'' clusters, we begin to see smaller, more specific proportions of PD
+patients. These clusters seem to clarify and expand upon the results found in previous analyses.
+The clusters, summarized, are:
+
+\begin{itemize}
+  \item[1-3.] ($n = 41,\; 77,\; 150$) Mild in all domains. Cluster 1 is especially mild in motor
+    symptoms (see cisitot).
+  \item[4.] ($n = 38$) Insomnia-dominant.
+  \item[5.] ($n = 43$) Motor-dominant.
+  \item[6.] ($n = 163$) Average.
+  \item[7.] ($n = 54$) Urinary-dominant.
+  \item[8.] ($n = 177$) Above-average in all domains.
+  \item[9.] ($n = 67$) Depression-dominant.
+  \item[10.] ($n = 81$) Nonmotor-dominant.
+  \item[11.] ($n = 13$) Severe in all domains.
+% 1	2  3	4   5  6    7  8    9   10  11
+% 41  77 150  38  43 163  54 177  67  81  13
+\end{itemize}
+
+A heatmap of all 11 clusters found is located in Figure~\ref{fig:nms30m-mclust-heatmap-improved}.
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=0.7\linewidth]{nms30m-mclust-heatmap-improved.png}
+  \caption{Heatmap of clusters found using the VEI Gaussian Mixture Model.}
+  \label{fig:nms30m-mclust-heatmap-improved}
+\end{figure*}
+
+\subsection{Decision Trees}
+
+I created decision trees for the $k$-means $k = 4$ clustering, located in
+Figure~\ref{fig:nms30m-dtree-all}
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=\linewidth]{nms30m-dtree-all.pdf}
+  \caption{Pruned decision tree for one vs all, clustered on nms\{1-30\}}
+  \label{fig:nms30m-dtree-all}
+\end{figure*}
+
 \section{Preliminary Conclusions}
@@
 \subsection{Overall clustering}
-$k$-means clustering on this Parkinsons' Disease data set reveals clusters that
-confirm previous computationally-based findings in the field, mainly van Rooden
-et al. \cite{vanrooden10} and the identification of four subtypes of
-Parkinson's disease: mild, nonmotor-predominant, motor-predominant, and severe.
-van Rooden's work was done with a separate dataset using a different
-modeling method (expectation-maximization), and this investigation
-independently confirms these subtype classifications. Unlike van Rooden, mean
-disease durations differences do exist between subtypes 1 (mild) and 4
-(severe), likely due to further development of the disease, although the
-differences between 2 and 3 (nonmotor/motor predominated) subtypes are
-insignificant (Table~\ref{tab:tukeyhsd}), suggesting different developmental paths of the disease.
+$k$-means clustering on this Parkinsons' Disease data set reveals clusters that confirm previous
+findings in the field, mainly van Rooden et al. \cite{vanrooden10} and the identification of four
+subtypes of Parkinson's disease: mild, nonmotor-predominant, motor-predominant, and severe.  van
+Rooden's work was done with a separate dataset using a different modeling method
+(expectation-maximization), and this investigation independently confirms these subtype
+classifications. Unlike van Rooden, mean disease durations differences do exist between subtypes 1
+(mild) and 4 (severe), likely due to further development of the disease, although the differences
+between 2 and 3 (nonmotor/motor predominated) subtypes are insignificant
+(Table~\ref{tab:tukeyhsd}), suggesting different developmental paths of the disease.
 
@@
 
-It remains to be seen whether these classification models, especially the
-one-vs-all decision trees, are useful in clinical practice.
+It remains to be seen whether these classification models, especially the one-vs-all decision
+trees, are useful in clinical practice.
 
-\begin{thebibliography}{9}
+\bibliographystyle{Science}
+\bibliography{abbreviated}
 
-\bibitem{erro13}
-Erro et al (2013). The Heterogeneity of Early Parkinson’s Disease: A Cluster
-Analysis on Newly Diagnosed Untreated Patients
-
-\bibitem{fereshtehnejad15}
-Fereshtehnejad et al (2015). New Clinical Subtypes of Parkinson
-Disease and Their Longitudinal Progression
-
-\bibitem{ma15}
-Ma et al (2015). Heterogeneity among patients with Parkinson's disease: Cluster
-analysis and genetic association
-
-\bibitem{sauerbier15}
-Sauerbier et al (2015). Non motor subtypes and Parkinson's disease.
-
-\bibitem{vanrooden10}
-van Rooden et al (2010). The Identification of Parkinson's Disease Subtypes
-Using Cluster Analysis: A Systematic Review
-
-\end{thebibliography}
-
+\end{multicols}
 \end{document}

### 2016-05-02 writeup/manuscript/manuscript.tex
Begin drafting manuscript

% PREAMBLE ====
\documentclass[preprint,3p]{elsarticle} % TODO: What model?

% Tables
\usepackage{booktabs}
\usepackage{threeparttable}
\setlength{\tabcolsep}{10pt}

% Graphics
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf,.png}

% References
\usepackage{numcompress}
\bibliographystyle{model3-num-names}

% DOCUMENT ====
\begin{document}
% FRONTMATTER ====
\begin{frontmatter}
\title{Parkinson's disease cluster analysis}

\author[c3]{Pablo Martinez-Martin}
\ead{pmartinez@isciii.es}
\author[bc]{Jesse Mu}
\ead{muj@bc.edu}
\author[cig]{Concha Bielza}
\ead{mcbielza@fi.upm.es}
\author[cig]{Pedro Larra\~naga}
\ead{plarranaga@fi.upm.es}
\address[c3]{Area of Applied Epidemiology, National Centre of Epidemiology and CIBERNED, Carlos III Institute of Health, Madrid, Spain}
\address[bc]{Department of Computer Science, Boston College, Chestnut Hill, Massachusetts, USA}
\address[cig]{Computational Intelligence Group, Polytechnic University of Madrid, Madrid, Spain}

\begin{abstract}
\paragraph{Background}
\paragraph{Methods}
\paragraph{Findings}
\paragraph{Interpretation}
\paragraph{Funding}
\end{abstract}

\journal{The Lancet Neurology}

\end{frontmatter}

% BODY ====

\section{Introduction}
\section{Methods}

\section{Results}

$k$-means clustering was used.

\subsection{Clustering on nonmotor domains}

$k$-means clustering on the nine nonmotor domains and the four motor symptoms found four clusters,
available in Table~\ref{tab:nmd}. Using one-way ANOVA, symptom means for all clusters were found to
be different with $p < 0.05$, and the differences among pairwise clusters were tested post-hoc
using Tukey's range test. On average, cluster 1 ($n = 406$) patients were mildly affected in all
domains, cluster 2 ($n = 189$) patients were severely affected in nonmotor domains but mildly
affected in motor domains (Nonmotor-Dom), cluster 3 ($n = 221$) patients were severely affected in
motor domains but mildly affected in nonmotor domains (Motor-Dom), and cluster 4 ($n = 88$)
patients were severely affected in all domains.
Additionally, statistics for variables not used in the clustering were explored in
Table~\ref{tab:nmd_extra}. Like before, one-way ANOVA was used to test equality of means across all
groups, and all variables were found to be different except PD onset. Tukey's range test was used
again to compare individual differences in means.

\begin{table}[t]
  \centering
  \caption{Summaries of clustering on nonmotor domains. Bold indicates the highest score for that
  symptom out of the four clusters.}
  \label{tab:nmd}
  \begin{threeparttable}
  \begin{tabular}{l r r r r}
    \toprule
    Cluster & 1 & 2 & 3 & 4 \\
    $n$ & 406 (44.9\%) & 189 (20.9\%) & 221 (24.4\%) & 88 (9.7\%) \\
    \midrule
    \textbf{Nonmotor} (Domains) & & & & \\
    Cardiovascular (0--24) & 0.7 (1.6)\tnote{24} & 2.3 (2.9)\tnote{134} & 1.1 (2.0)\tnote{24} &
    \textbf{7.0} (6.0)\tnote{234} \\
    Sleep/fatigue (0--48) & 4.3 (4.8)\tnote{234} & 14.9 (8.2)\tnote{134} & 6.9 (6.3)\tnote{124}
    & \textbf{21.0} (10.0)\tnote{123} \\
    Mood/cognition (0--60) & 3.1 (4.5)\tnote{234} & 16.1 (14.0)\tnote{134} & 6.6
    (8.1)\tnote{124} & \textbf{23.4} (14.3)\tnote{123} \\
    Perception/hallucination (0--36) & 0.5 (1.7)\tnote{24} & 1.9 (3.5)\tnote{134} & 0.8
    (2.0)\tnote{24} & \textbf{8.6} (6.9)\tnote{123} \\
    Attention/memory (0--36) & 2.8 (4.2)\tnote{24} & 8.9 (8.1)\tnote{134} & 3.2 (4.4)\tnote{24} &
    \textbf{15.6} (11.1)\tnote{123} \\
    Gastrointestinal (0--36) & 2.8 (3.9)\tnote{234} & 8.8 (7.1)\tnote{134} & 4.1
    (4.7)\tnote{124} & \textbf{14.6} (9.6)\tnote{123} \\
    Urinary (0--36) & 4.5 (5.9)\tnote{234} & 12.4 (9.7)\tnote{134} & 6.1 (6.4)\tnote{124} &
    \textbf{20.3} (9.9)\tnote{123} \\
    Sexual function (0--24) & 1.5 (3.1)\tnote{24} & 6.4 (7.4)\tnote{134} & 2.5 (4.1)\tnote{24} &
    \textbf{9.0} (9.7)\tnote{123} \\
    Miscellaneous (0--48) & 3.9 (4.7)\tnote{24} & 13.2 (8.6)\tnote{13} & 5.2 (5.8)\tnote{24} &
    \textbf{14.0} (9.5)\tnote{13} \\
    \midrule
    \textbf{Motor} & & & & \\
    Tremor & 1.9 (1.8)\tnote{34} & 1.6 (1.9)\tnote{34} & \textbf{4.3} (2.8)\tnote{124} & 3.5
    (3.8)\tnote{123} \\
    Bradykinesia & 1.5 (0.9)\tnote{234} & 2.1 (1.1)\tnote{134} & 3.6 (0.9)\tnote{124} &
    \textbf{4.0} (1.5)\tnote{123} \\
    Rigidity & 1.5 (0.9)\tnote{234} & 1.8 (1.1)\tnote{134} & 3.3 (0.9)\tnote{124} &
    \textbf{3.8} (1.4)\tnote{123} \\
    Axial & 1.8 (1.6)\tnote{234} & 3.3 (2.1)\tnote{134} & 4.3 (2.4)\tnote{124} &
    \textbf{7.4} (2.9)\tnote{123} \\
    \bottomrule
  \end{tabular}
  \begin{tablenotes}
    \small
    \item[1] Significant difference with cluster 1 ($p < 0.05$)
    \item[2] Significant difference with cluster 2 ($p < 0.05$)
    \item[3] Significant difference with cluster 3 ($p < 0.05$)
    \item[4] Significant difference with cluster 4 ($p < 0.05$)
    \item[\textdagger] Another footnote.
  \end{tablenotes}
  \end{threeparttable}
\end{table}

\begin{table}[t]
  \centering
  \caption{Cluster statistics for variables not used in the analysis}
  \label{tab:nmd_extra}
  \begin{threeparttable}
  \begin{tabular}{l r r r r}
    \toprule
    & 1 & 2 & 3 & 4 \\
    \midrule
      Age & 62.7 (9.7) & 64 (9) & 65 (10.2) & 70.5 (8.9) \\ 
      Sex & 0.4 (0.5) & 0.5 (0.5) & 0.3 (0.5) & 0.4 (0.5) \\ 
      PD\_Onset & 56 (10.6) & 55.2 (10.1) & 56.9 (11.1) & 58.2 (11.4) \\ 
      PD\_Duration & 6.7 (4.8) & 8.7 (5.7) & 8 (5.5) & 12.3 (8.1) \\ 
      CISI\_Total & 5.7 (3.2) & 9.2 (3.9) & 9.7 (3.5) & 14.8 (5.1) \\ 
      ldopa & NA (NA) & NA (NA) & NA (NA) & NA (NA) \\ 
      Surgery & 0 (0.2) & 0.1 (0.2) & 0 (0.2) & 0 (0.1) \\ 
    \bottomrule
  \end{tabular}
  \begin{tablenotes}
    \small
    \item[1] Significant difference with cluster 1 ($p < 0.05$)
    \item[2] Significant difference with cluster 2 ($p < 0.05$)
    \item[3] Significant difference with cluster 3 ($p < 0.05$)
    \item[4] Significant difference with cluster 4 ($p < 0.05$)
    \item[\textdagger] Another footnote.
  \end{tablenotes}
  \end{threeparttable}
\end{table}
\subsection{Clustering on nonmotor symptoms}
% From 931 to 904 patients

\section{Discussion}

$k$-means clustering on this Parkinsons' Disease data set reveals clusters that confirm previous
findings in the field, mainly van Rooden et al. \cite{vanrooden10} and the identification of four
subtypes of Parkinson's disease: mild, nonmotor-predominant, motor-predominant, and severe.  van
Rooden's work was done with a separate dataset using a different modeling method
(expectation-maximization), and this investigation independently confirms these subtype
classifications. Unlike van Rooden, mean disease durations differences do exist between subtypes 1
(mild) and 4 (severe), likely due to further development of the disease, although the differences
between 2 and 3 (nonmotor/motor predominated) subtypes are insignificant
(Table~\ref{tab:tukeyhsd}), suggesting different developmental paths of the disease.

Overall, little information was found in pdonset,
durat\_pd, or current age, according to Tables~\ref{tab:tukeyhsd}
and~\ref{tab:info_gain}.  Mean ages were similar for subgroups 1, 2, and 3 ($p >
0.05$), but different for the severe subtype 4, which makes sense given that
patients in 4 also have longer disease durations. Specifically, clusters 1 and
4 seem to be phenotypically quite similar, except at different stages of
disease progression, given cluster 4's higher age and durat\_pd scores.

However, clusters 2 and 3 clearly show different disease progression, one in
the motor direction, and one in the nonmotor. Both groups have similar age,
pdonset, and durat\_pd scores, but differ wildly in symptomatic expression.
Cluster 2 is dominated by a high prevalence of nonmotor symptoms, such as
nms\_d2, nms\_d3, nms\_d7, and nms\_d9. Cluster 3, however, is dominated by a
high prevalence of motor symptoms, while most motor symptoms are similar to the
mild cluster 1. Of note is that the tremor population mean
is the highest cluster mean, even higher than the severe subtype 4. This
motor-dominant cluster may thus overlap with Ma's tremor dominant/slow
progression cluster \cite{ma15}.

Generally, given stable pdonset scores and predictably increasing durat\_pd
scores for clusters 1 and 4, Ma et al's rapid disease
progression/late onset and tremor dominant/slow progression clusters
\cite{ma15} were mostly not found in this dataset, save for the tremor-dominant
motor cluster.

The most important nonmotor symptoms in determining these clusters were nms\_d2
(sleep) and nms\_d3 (mood/cognition), which echo findings of Fereshtehnejad's
longitudinal study \cite{fereshtehnejad15} and are similar to Sauerbier's
identification of sleep dominant and cognitive dominant clinical NMS subtypes
\cite{sauerbier15}. Compared to Erro et al. \cite{erro13}, nonmotor/motor
dominant subtypes were indeed found, but an additional subgroup with relatively
severe levels of both motor and nonmotor symptoms were found. Erro's benign
subtype groups possibly overlap with the mild cluster 1 found in this
investigation.

\subsection{Nonmotor subtype: clustering and modeling}
Nonmotor symptoms nms\_d2 and nms\_d3 became critical not only in
classification trees distinguishing between the various symptoms but in the
nonmotor-predominant subgroup itself. In $k$-means subdivision of the
nonmotor-dominant subtype where $k = 2$ and $k = 3$, opposite trends were
confirmed with nms\_d2 and nms\_d3 symptoms. Similarly, in the 2 and 4 vs rest
decision tree (Figure~\ref{fig:dtree-2and4va-pruned}), nms\_d2 and nms\_d3 nodes were
used to differentiate various categories of nomotor-dominant patients.
When $k = 3$, the subtype with the highest nms\_d2 scores and lowest
nms\_d3 scores had by far the highest axial scores, nms\_d6 (gastrointestinal)
scores, and nms\_d7 (urinary) scores. Thus subtype 3 of the nonmotor-dominated
group could include patients falling into the cognitive/depression-dominant or
autonomic dominant subtypes.

Despite the variety in symptomatic expression in this nonmotor group, what
seems most consistent is the presence of nms\_d9 (miscellaneous) nonmotor
symptoms, as it is used as the root node of the 2 vs all decision tree
(Figure~\ref{fig:2va}) and the 2 and 4 vs rest decision tree
(Figure~\ref{fig:dtree-2and4va-pruned}).

It remains to be seen whether these classification models, especially the one-vs-all decision
trees, are useful in clinical practice.

\subsection{New conclusions}
\label{ssec:newconc}

First, the longitudinal analysis gives more insight into the clusters found when clustering on
nms\_d\{1-9\}. According to Figure~\ref{fig:longcorr}, most symptoms are highly correlated, with PD
duration, but notably, mood/cognition symptoms (nms9, nms10, nms12) and tremor are not correlated
highly with PD duration. The differences in disease progression can be seen by the corresponding
graphs, Figures~\ref{fig:nms9-multi} and~\ref{fig:nms10-multi}. In both graphs, what is interesting
is that Subtype 2 (Nonmotor-Dominant) starts at higher scores for nms\_9 and nms\_10, thus
indicating that these patients' subtype is can be determined early in PD duration from depressive
symptom score. Similarly, when examining Subtype 3 (Motor-dominant) in
Figure~\ref{fig:tremor-multi}, the mean tremor score is substantially higher from PD onset.
Interestingly, Subtype 4 (Severe) generally starts at lower tremor and motor scores during disease
onset (Figure~\ref{fig:cisitot-multi}), but then rises sharply, exceeding other Subtypes. More evidence
that tremor is a unique motor symptom is located in Figure~\ref{fig:nms30m-colhc}, where it is the most
distant symptom from all other symptoms.

When examining all 30 symptoms, more evidence is given that the previously-discovered Subtype
2 (Nonmotor-Dominant) may be primarily characterized by high depressive symptoms. By graphing the
30 subtypes attached to the original clustering, as in Figure~\ref{fig:d9to30}, the mean scores of
nms8, nms9, and nms10 for Subtype 2 are substantially higher than Subtypes 1 and 3.

Indeed, PCA on the 30 nonmotor symptoms identifies the second-most prominent component as a general
mood/cognition component, and $k$-means clustering on the 30 symptoms only
(Figure~\ref{fig:nms30m-k4-heatmap-improved}) divides the 1000 patients into four slightly
different groups, a mild, average, depression-dominant, and severe group.

The Gaussian mixture model identified in Figure~\ref{fig:nms30m-mclust-heatmap-improved} fragments
the previously-discovered clusters into more groups. Here, a wide variety of specialized subtypes
of PD are displayed, including insomnia, urinary, motor, nonmotor, and depression-dominant groups,
as well as the expected mild, average, and severe subtypes. It is likely that the previous analysis
with only nms\_d\{1-9\} combined most of those specialized groups into the more general
nonmotor-dominant Subtype 2.

It's intuitive that a Depression-Dominant group emerges when clustering on nms\{1-30\}, since
domain 3 consists of 5 separate measures. Thus, any high expression of depressive symptoms is
magnified in clustering, since the symptoms are highly similar (Figure~\ref{fig:nms30m-colhc}) and
treated with equal weight. Once again reinforcing what was discovered previously, depressive
symptoms have been shown to be very important in determining subtypes of PD.

\section{Supplementary Material}
\section{References}

\bibliography{manuscript}

\end{document}

### 2016-05-02 R/kmeans-dtree.R
Improve plotting of clvalid measures

@@ -1,5 +1,7 @@
 # LOAD LIBRARIES ====
 library(cluster)
+library(reshape2)
+library(clValid)
 library(rpart)
 library(rpart.plot)
 library(corrplot)
@@ -681,12 +683,22 @@ corrplot(cor(cor.nms.raw4), method="ellipse", order="hclust", main="4")
 corrplot(cor(cor.nms.raw2), method="ellipse", order="hclust", main="2")
 
 # 2016 update - some better cluster validation ====
-clres.stability <- clValid(obj = cor.nms.raw, nClust = 2:8, clMethods = "kmeans",
+clres.stability <- clValid(obj = raw.filtered, nClust = 2:8, clMethods = "kmeans",
                  validation = "stability", maxitems=1000)
+plot(clres.stability, measure = c("APN", "AD", "ADM"))
+optimalScores(clres.stability)
 stab.measures <- as.data.frame(clres.stability@measures)
 colnames(stab.measures) <- 2:8
-clres.internal <- clValid(obj = cor.nms.raw, nClust = 2:8, clMethods = "kmeans",
+stab.measures$measure <- rownames(stab.measures)
+stab.measures.long <- melt(stab.measures, id.vars = c("measure"),
+                           variable.name = "clusters",
+                           value.name = c("value"))
+ggplot(stab.measures.long) +
+  geom_line(aes(x = clusters, y = value, group = measure, color = measure)) +
+  theme_bw()
+clres.internal <- clValid(obj = raw.filtered, nClust = 2:8, clMethods = "kmeans",
                  validation = "internal", maxitems=1000)
+plot(clres.internal)
 stab.measures
 # Stability measures look good for 4, surprisingly
 # http://bioinformatics.oxfordjournals.org/content/19/4/459.full.pdf

### 2016-05-02 R/preprocessing.R
Extra preprocessing (camera-ready names)

@@ -56,6 +56,38 @@ NMS.30.NAMES <- c(
   'd9-29-weight_change',
   'd9-30-sweating'
 )
+NMS.30.NAMES.PUB <- c(
+  'Lightheadedness',
+  'Fainting',
+  'Drowsiness',
+  'Fatigue',
+  'Insomnia',
+  'RLS',
+  'Loss_interest',
+  'Loss_activities',
+  'Anxiety',
+  'Depression',
+  'Flat_affect',
+  'Loss_pleasure',
+  'Hallucination',
+  'Delusion',
+  'Diplopia',
+  'Loss_concentration',
+  'Forget_explicit',
+  'Forget_implicit',
+  'Drooling',
+  'Swallowing',
+  'Constipation',
+  'Urinary_urgency',
+  'Urinary_frequency',
+  'Nocturia',
+  'Sex_drive',
+  'Sex_dysfunction',
+  'Unexplained_pain',
+  'Gustation/olfacttion',
+  'Weight_change',
+  'Sweating'
+)
 NMS.30 <- c(
   "nms1", "nms2", "nms3", "nms4", "nms5", "nms6",
   "nms7", "nms8", "nms9", "nms10", "nms11", "nms12",
@@ -63,6 +95,29 @@ NMS.30 <- c(
   "nms19", "nms20", "nms21", "nms22", "nms23", "nms24",
   "nms25", "nms26", "nms27", "nms28", "nms29", "nms30"
 )
+NMS.D <- c(
+  "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5", "nms_d6", "nms_d7", "nms_d8", "nms_d9"
+)
+NMS.D.NAMES <- c(
+  "Cardiovascular", "Sleep_fatigue", "Mood_cognition", "Perception_hallucination", "Attention_memory",
+  "Gastrointestinal", "Urinary", "Sexual", "Miscellaneous"
+)
+NMS.D.MAP.PUB <- setNames(NMS.D.NAMES, NMS.D)
+NMS.30.MAP.PUB <- setNames(NMS.30.NAMES.PUB, NMS.30)
+MISC.MAP <- c(
+  'age' = 'Age',
+  'sex' = 'Sex',
+  'pdonset' = 'PD_Onset',
+  'durat_pd' = 'PD_Duration',
+  'cisitot' = 'CISI_Total',
+  'tremor' = 'Tremor',
+  'bradykin' = 'Bradykinesia',
+  'rigidity' = 'Rigidity',
[…]

### 2016-05-02 R/nms30.R
Some more nms30 (??)

@@ -120,8 +120,9 @@ if (SAVE.NMS30.PLOTS) {
 
 # Motor
 ctd.m <- cutree(hm.m$colDendrogram, 4)
+ctk.m <- hm.m$colDendrogram %>% set("branches_k_color", k = 4)
 ColorDendrogram(as.hclust(hm.m$colDendrogram), y = ctd.m, labels = names(ctd.m),
-                main = "NMS + Motor Hierarchical Clustering",
+                main = "Nonmotor + Motor Hierarchical Clustering",
                 xlab = "Symptom",
                 sub = "",
                 branchlength = 5)
@@ -327,6 +328,36 @@ if (SAVE.NMS30.PLOTS) {
   dev.off()
 }
 
+# one vs all ====
+gg_color_list <- function(i) {
+  c(gg_color_hue(i)[i], '#FFFFFF')
+}
+ova.counts <- c(509, 97, 249, 49)
+for (ONE in 1:4) {
+  set.seed(0)
+  nms30.ova <- nms30.dtree
+  nms30.ova$newcluster <- NA
+  nms30.ova[nms30.ova$cluster == ONE, ]$newcluster <- T
+  nms30.ova[nms30.ova$cluster != ONE, ]$newcluster <- F
+  this.count = length(nms30.ova[nms30.ova$newcluster == T, ]$newcluster)
+  if (this.count != ova.counts[ONE]) stop("Lengths not equal")
+  nms30.ova$cluster <- factor(nms30.ova$newcluster, levels=c(T, F))
+  nms30.ova$newcluster <- NULL
+  tree.nms30.ova <- rpart(cluster ~ ., nms30.ova)
+  cp <- tree.nms30.ova$cptable[which.min(tree.nms30.ova$cptable[,"xerror"]),"CP"]
+  if (ONE == 4 || ONE == 2) {
+    pruned.tree.nms30.ova <- tree.nms30.ova
+  } else {
+    pruned.tree.nms30.ova <- prune(tree.nms30.ova, cp = cp)
+  }
+  prp(pruned.tree.nms30.ova, extra = 1, varlen=0,
+      box.col = gg_color_list(ONE)[pruned.tree.nms30.ova$frame$yval],
+      main = paste(ONE, ' vs all decision tree', sep = ''))
+  if (SAVE.NMS30.PLOTS) {
+    dev.copy(pdf, paste('../figures/nms30m-dtree-', ONE, 'va.pdf', sep=''), width=10, height=10)
+    dev.off()
+  }
+}
 
[…]

### 2016-05-02 R/cleanup.R
Script that contains code for camera-ready info

@@ -1,0 +1,58 @@
+# Cleaning up and presenting data
+# Because I need to learn Sweave at some poin
+library(doBy)
+library(xtable)
+library(reshape)
+
+nmsd.symptoms <- c(
+  NMS.D.NAMES,
+  "Tremor",
+  "Bradykinesia",
+  "Rigidity",
+  "Axial",
+  'Cluster'
+)
+
+nmsd.extra <- c(
+  'Age',
+  'Sex',
+  'PD_Onset',
+  'PD_Duration',
+  'CISI_Total',
+  'ldopa',
+  'Surgery',
+  'Cluster'
+)
+
+# Depends on valid cluster assignment, assert that the distribution is
+# table(clus4.wide$cluster)
+#   1   2   3  4
+# 406 189 221 88
+
+# ON NONMOTOR DOMAINS ====
+# Somewhat oddly, the correct clustering vector lies in trees$clusters4$clustering$cluster
+present <- rename(raw.omitted, PUB.MAP)
+present.full <- rename(raw.omitted.full, PUB.MAP)
+present$Cluster <- trees$clusters4$clustering$cluster
+present.full$Cluster <- trees$clusters4$clustering$cluster
+
+# Funcs for latex
+mean.sd <- function(data, sig = 2) {
+  paste(round(mean(data), sig), " (", round(sd(data), sig), ")", sep = "")
+}
+to.latex <- function(df, file = NULL) {
+  summary.t <- t(summaryBy(. ~ Cluster, df, FUN = function(x) mean.sd(x, sig = 1), # Only 1 decimal
+                           keep.names = TRUE))
+  # Get rid of "cluster" row
+  summary.t <- summary.t[-which(rownames(summary.t) == 'Cluster'), ]
+  xt <- xtable(summary.t,
+               sanitize.colnames.function = function(x) gsub(pattern = '\\_', replacement = '/', x = x))
+  print(xt, type = "latex", file = file, booktabs = TRUE)
+}
+
+# Redo ANOVA + bonferroni correction.
+# Apparently Tukey takes care of multiple comparisons but make sure that's not a
+# setting or grouping you need to actually make happen in R.
+
+to.latex(present[, nmsd.symptoms], "../writeup/manuscript/include/nmsd_summaries.tex")
+to.latex(present.full[, nmsd.extra], "../writeup/manuscript/include/nmsd_extra.tex")

### 2016-05-04 writeup/manuscript/manuscript.tex
More manuscript progress
gap statistic, stability measures, publication-ready hclust and boxplots

@@
 % PREAMBLE ====
-\documentclass[preprint,3p]{elsarticle} % TODO: What model?
+\documentclass[preprint,3p,twocolumn]{elsarticle} % TODO: What model?
 
-% Tables
+% Use times font
+\usepackage{pslatex}
+
+% Math
+\usepackage{amsmath}
+
+% Tables, Figures
 \usepackage{booktabs}
 \usepackage{threeparttable}
+\usepackage{float}
+\usepackage{caption}
 \setlength{\tabcolsep}{10pt}
@@
 \DeclareGraphicsExtensions{.pdf,.png}
+\graphicspath{{../../figures/}}
 
@@
 \usepackage{numcompress}
+\biboptions{super}
 \bibliographystyle{model3-num-names}
@@
 \title{Parkinson's disease cluster analysis}
-
 \author[c3]{Pablo Martinez-Martin}
@@
 \begin{abstract}
-\paragraph{Background}
-\paragraph{Methods}
-\paragraph{Findings}
-\paragraph{Interpretation}
-\paragraph{Funding}
+\emph{Background} \quad This is the background. \\
+\emph{Methods} \quad  These are methods. \\
+\emph{Findings} \quad These are findings. \\
+\emph{Interpretation} \quad This is interpretation. \\
+\emph{Funding} \quad This is funding.
 \end{abstract}
-
 \journal{The Lancet Neurology}
-
 \end{frontmatter}
@@
 \section{Introduction}
+
 \section{Methods}
+
+\subsection{Statistical analysis}
+
+Out of the 951 patients in the study, we used listwise deletion to exclude 50 patients due to
+missing measurements in the variables of interest, resulting in 901 remaining patients. All
+symptoms were scaled to $z$-scores ($\mu = 0$, $\sigma = 1$) before clustering, and unscaled
+afterwards. Analysis was conducted in R 3.2.4 (www.r-project.org).
+
+\subsubsection{Cluster analysis}
+
+$k$-means was used for cluster analysis. We performed two analyses on the
+patients in the dataset: the first on the nine aggregate nonmotor symptom
+domains and the four motor symptoms, and the second on the 30 individual
+nonmotor symptoms collected with the four motor symptoms. Complete-linkage
+hierarchical clustering on the 30 nonmotor symptoms was also performed to observe the
+categorization of symptoms.
+
+\subsubsection{Determining $k$}
+
+Various formal measures were used to determine the optimal number of
+clusters for the dataset. First, the optimal $k$ according to the Gap Statistic and the
+1-standard-error method\cite{tibshirani01gap} was $k = 4$ (Figure~\ref{fig:gap}). Second, although
+the globally-optimal clusters reported from stability measures of the clValid R package
+\cite{brock11clvalid} stability are 2, 6, and 8 (Figure~\ref{fig:stability}), there appeared to be
+consensus that $k = 4$ is a locally-optimal number of clusters which supported the gap statistic
+result.
+
+Other cluster validation methods (SSE scree plot, minimum average silhouette width) generally
+suggested 2 clusters. As a consequence, $k = 2, 3, 4$ was tried, where $k = 2, 3$ simply divided
+the data uninformatively into groups with varying levels of overall PD severity. Thus $k = 4$ was
+selected to offer a good blend of model fit, informativeness, and parsimony.
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{gap-statistic.pdf}
+  \caption{Plot of gap statistic versus number of clusters with $k$-means on 100 bootstrapped
+    samples. Error bars are standard error. Per the method described in\cite{tibshirani01gap}, 4 is the smallest $k$ such
+    that $\text{Gap}(k) \geq \text{Gap}(k + 1) - \text{se}_{k + 1}$.}
+  \label{fig:gap}
+\end{figure}
+
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{stability-measures.pdf}
+  \caption{Average proportion of non-overlap (APN), average distance (AD), average distance
+    between means (ADM), and figure of merit (FOM) stability measures\cite{datta2006methods,
+    yeung2001validating}. Statistics should be minimized; notice the local optima at $k = 4$ for
+  APN, AD, and ADM.}
+  \label{fig:stability}
+\end{figure}
+
+\subsubsection{Interpretation}
+
+For the clustering on nonmotor domains, we displayed the distribution of each symptom for the four
+clusters using boxplots, which allowed us to visualize the center and spread of each cluster. Since
+the number of variables was larger for the clustering on the individual symptoms, we visualized
+results for the second analysis with a heatmap. Finally, for the hierarchical clustering on the
+symptoms themselves, we displayed results in a dendrogram and clustered the symptoms into four
+interpretable clusters.
+
+For each symptom in both clusterings, we used one-way ANOVA to test the equality of symptom means
+across the clusters found, using Bonferroni correction for multiple testing with corrected $p <
+0.05$ considered significant. Differences among pairwise clusters were tested post-hoc using
+Tukey's range test.
 
@@
 
-$k$-means clustering was used.
-
-\subsection{Clustering on nonmotor domains}
+\subsection{Clustering with nonmotor domains}
 
 $k$-means clustering on the nine nonmotor domains and the four motor symptoms found four clusters,
-available in Table~\ref{tab:nmd}. Using one-way ANOVA, symptom means for all clusters were found to
-be different with $p < 0.05$, and the differences among pairwise clusters were tested post-hoc
-using Tukey's range test. On average, cluster 1 ($n = 406$) patients were mildly affected in all
-domains, cluster 2 ($n = 189$) patients were severely affected in nonmotor domains but mildly
-affected in motor domains (Nonmotor-Dom), cluster 3 ($n = 221$) patients were severely affected in
-motor domains but mildly affected in nonmotor domains (Motor-Dom), and cluster 4 ($n = 88$)
-patients were severely affected in all domains.
-Additionally, statistics for variables not used in the clustering were explored in
-Table~\ref{tab:nmd_extra}. Like before, one-way ANOVA was used to test equality of means across all
-groups, and all variables were found to be different except PD onset. Tukey's range test was used
-again to compare individual differences in means.
+available in Table~\ref{tab:nmd} with boxplots in Figure~\ref{fig:box}. Cluster means for all
+symptoms were found to be statistically different, with specific pairwise differences noted in the table.
 
-\begin{table}[t]
+Cluster 1 ($n = 406$) patients were mildly affected in all domains.
+
+Cluster 2 ($n = 189$) patients were severely
+affected in nonmotor domains but mildly affected in motor domains (Nonmotor-Dom), cluster 3 ($n =
+221$) patients were severely affected in motor domains but mildly affected in nonmotor domains
+(Motor-Dom), and cluster 4 ($n = 88$) patients were severely affected in all domains. Statistics
+for variables not used in the clustering were explored in Table~\ref{tab:nmd_extra}.
+
+\begin{table*}[t]
   \centering
@@
   \end{threeparttable}
-\end{table}
+\end{table*}
 
-\begin{table}[t]
+\begin{figure*}[b]
+  \centering
+  \includegraphics[width=\linewidth]{kmeans-summaries-4-pub.pdf}
+  \vspace{-7.5em}
+  \captionsetup{justification=raggedleft,
+    singlelinecheck=false
+  }
+  \caption{Boxplot summaries for clusters found \\ when clustering with nonmotor domains.}
+  \label{fig:box}
+\end{figure*}
+
+\begin{table*}[t]
   \centering
@@
       CISI\_Total & 5.7 (3.2) & 9.2 (3.9) & 9.7 (3.5) & 14.8 (5.1) \\
-      ldopa & NA (NA) & NA (NA) & NA (NA) & NA (NA) \\ 
       Surgery & 0 (0.2) & 0.1 (0.2) & 0 (0.2) & 0 (0.1) \\
@@
   \end{threeparttable}
-\end{table}
-\subsection{Clustering on nonmotor symptoms}
-% From 931 to 904 patients
+\end{table*}
+\subsection{Clustering with nonmotor symptoms}
+
+\subsection{Hierarchical clustering on symptoms}
+
+Hierarchical clustering on the 30 nonmotor symptoms and the four motor symptoms produced the
+dendrogram in Figure~\ref{fig:hc}. Predictably, symptoms in the same nonmotor domains tended to
+cluster together, with the exceptions of diplopia (domain 4), drowsiness (domain 2), and the
+symptoms in domain 9 (miscellaneous). Notably, tremor was the most dissimilar symptom, occupying a single branch at the
+top of the tree.
+
+\begin{figure*}[t]
+  \centering
+  \includegraphics[width=\linewidth]{nms30m-colhc-pub.pdf}
+  \caption{Hierarchical clustering of motor (blue) and nonmotor (black) symptoms. Symptoms are
+  labeled with their name and corresponding domain number. Dendrogram colored with 4 clusters.}
+  \label{fig:hc}
+\end{figure*}
 
@@
 $k$-means clustering on this Parkinsons' Disease data set reveals clusters that confirm previous
-findings in the field, mainly van Rooden et al. \cite{vanrooden10} and the identification of four
+findings in the field, mainly van Rooden et al.\cite{vanrooden10} and the identification of four
 subtypes of Parkinson's disease: mild, nonmotor-predominant, motor-predominant, and severe.  van
@@
 motor-dominant cluster may thus overlap with Ma's tremor dominant/slow
-progression cluster \cite{ma15}.
+progression cluster\cite{ma15}.
 
@@
 (sleep) and nms\_d3 (mood/cognition), which echo findings of Fereshtehnejad's
-longitudinal study \cite{fereshtehnejad15} and are similar to Sauerbier's
-identification of sleep dominant and cognitive dominant clinical NMS subtypes
-\cite{sauerbier15}. Compared to Erro et al. \cite{erro13}, nonmotor/motor
+longitudinal study\cite{fereshtehnejad15} and are similar to Sauerbier's
+identification of sleep dominant and cognitive dominant clinical NMS subtypes\cite{sauerbier15}.
+Compared to Erro et al.\cite{erro13}, nonmotor/motor
 dominant subtypes were indeed found, but an additional subgroup with relatively

### 2016-05-04 R/cleanup.R

@@ -56,3 +56,107 @@ to.latex <- function(df, file = NULL) {
 
 to.latex(present[, nmsd.symptoms], "../writeup/manuscript/include/nmsd_summaries.tex")
 to.latex(present.full[, nmsd.extra], "../writeup/manuscript/include/nmsd_extra.tex")
+
+# Publication-ready dendrogram ====
+labels.wo.d <- unname(sapply(labels(hm.m$colDendrogram), function(s) {
+  splitted <- strsplit(s, '-')[[1]]
+  if (length(splitted) == 1) {
+    s  # No -, leave it alone
+  } else {
+    # Get rid of that middle one
+    # Could get rid of preceding d as well
+    domain <- splitted[1]
+    symp <- splitted[3]
+    paste(domain, symp, sep = '-')
+  }
+}))
+# TODO: Capitalize map if necessary
+par(mar=c(3, 4, 0.5, 0))
+hm.m$colDendrogram %>%
+  set("labels", labels.wo.d) %>%
+  sort(type = "nodes") %>%
+  set("branches_lwd", 2) %>%
+  hang.dendrogram(hang_height = 3) %>%
+  set("branches_k_color", k = 5) %>%
+  # Here, "1" is motor, "2" is nonmotor (sorting by nodes is convenient here)
+  color_labels(col = c(rep("blue", 4), rep("black", 30))) %>%
+  plot(ylim = c(10, 45))#, xlab = "Symptom", ylab = "Height")
+
+dev.copy(pdf, "../figures/nms30m-colhc-pub.pdf", width = 15, height = 7)
+dev.off()
+
+# Pub-ready boxplot graph
+dev.off()
+clus <- clusters.raw.long[["4"]]
+# Get rid of extra
+clus.pub <- clus
+clus.pub <- clus.pub[clus.pub$variable != "sex", ]
+clus.pub$variable <- sapply(clus.pub$variable, function(s) paste("\n", PUB.MAP.N[as.character(s)][[1]], "\n", sep = ""))
+clus.pub$variable <- factor(clus.pub$variable)
+clus.pub$variable <- factor(clus.pub$variable, levels(clus.pub$variable)[c(1, 2, 3, 4, 5, 6, 7, 8, 9, 11, 12, 16, 17, 13, 10, 15, 14)])
+# Add types. Need to do after factors have been reorganized for some reason
+clus.pub$Type <- ""
+clus.pub[clus.pub$variable %in%
[…]

### 2016-05-04 R/kmeans-dtree.R

@@ -114,12 +114,16 @@ cat("number of clusters estimated by optimum average silhouette width:", pamk.be
 plot(pam(raw.filtered, pamk.best$nc))
 
 # GAP STATISTIC ESTIMATION ====
+set.seed(0)
 gaps <- clusGap(raw.filtered, kmeans, 14, B = 100)
-plot(x=1:14, y=gaps$Tab[, "gap"], xlab="Clusters", ylab="Gap Statitsic", type="b")
+par(mar = c(4.5, 4.5, 4.5, 4.5), ps = 18)
+plot(gaps)
 if (SAVE.EXPLORE.PLOTS) {
-  dev.copy(pdf, "../figures/gap-statistic.pdf")
+  dev.copy(pdf, "../figures/gap-statistic.pdf", width = 6, height = 5)
   dev.off()
 }
+print(gaps, method = "firstSEmax", SE.factor = 1)
+
 
 # Affinity propagation ====
 # Auto commented out because this takes a while
@@ -488,7 +492,7 @@ for (i in c("2", "3", "4")) {
     facet_wrap( ~ variable, scales = "free")
   print(p)
   if (SAVE.BOXPLOTS) {
-    ggsave(paste("../figures/kmeans-summaries-", i, ".pdf", sep=""))
+    ggsave(paste("../figuresekmeans-summaries-", i, ".pdf", sep=""))
   }
 }
 
@@ -685,7 +689,12 @@ corrplot(cor(cor.nms.raw2), method="ellipse", order="hclust", main="2")
 # 2016 update - some better cluster validation ====
 clres.stability <- clValid(obj = raw.filtered, nClust = 2:8, clMethods = "kmeans",
                  validation = "stability", maxitems=1000)
-plot(clres.stability, measure = c("APN", "AD", "ADM"))
+par(mfrow=c(2, 2), mar = c(4.5, 4.5, 4.5, 4.5), oma = c(0, 0, 0, 0))
+plot(clres.stability, main = "", legend = FALSE, pch = 1)
+if (SAVE.EXPLORE.PLOTS) {
+  dev.copy(pdf, "../figures/stability-measures.pdf", width = 8, height = 6)
+  dev.off()
+}
 optimalScores(clres.stability)
 stab.measures <- as.data.frame(clres.stability@measures)
 colnames(stab.measures) <- 2:8

### 2016-05-04 R/nms30.R

@@ -9,6 +9,7 @@ library(nFactors)
 library(gplots) # For heatmaps
 # Better dendrograms
 library(dendextend)
+library(dendsort)
 library(dynamicTreeCut)
 # Trying to get some color here
 library(colorspace)
@@ -94,7 +95,7 @@ for (i in 2:15) {
 
 # Large heatmap, clustering for rows and cols ====
 hm <- heatmap.2(as.matrix(nms30.s),
-                hclustfun = function(x) hclust(x, method='complete'),
+                hclustfun = function(x) hclust(x, method='average'),
                 trace='none')
 colD <- as.ggdend(hm$colDendrogram)
 rowD <- as.ggdend(hm$rowDendrogram)
@@ -106,8 +107,13 @@ hm.m <- heatmap.2(as.matrix(nms30m.s),
                   trace='none')
 colD.m <- as.ggdend(hm.m$colDendrogram)
 
+# Average is hella different
+asdf <- hclust(dist(t(nms30m.s)), method='average')
+plot(asdf)
+
 # Plot column dendrogram, visualize dynamic clusters with cutreeDynamic
 ctd <- cutree(hm$colDendrogram, 5)
+par(mfrow = c(1, 1))
 ColorDendrogram(as.hclust(hm$colDendrogram), y = ctd, labels = names(ctd),
                 main = "NMS Hierarchical Clustering",
                 xlab = "Symptom",
@@ -130,6 +136,8 @@ if (SAVE.NMS30.PLOTS) {
   dev.copy(pdf, '../figures/nms30m-colhc.pdf', width=10, height=6)
   dev.off()
 }
+
+# A slightly better plot
 
 # Individuals: Determining k ====
 set.seed(0)

### 2016-05-04 R/preprocessing.R

@@ -103,6 +103,7 @@ NMS.D.NAMES <- c(
   "Gastrointestinal", "Urinary", "Sexual", "Miscellaneous"
 )
 NMS.D.MAP.PUB <- setNames(NMS.D.NAMES, NMS.D)
+NMS.D.MAP.PUB.N <- setNames(sapply(1:length(NMS.D.NAMES), function(i) paste(i, NMS.D.NAMES[i], sep = "-")), NMS.D)
 NMS.30.MAP.PUB <- setNames(NMS.30.NAMES.PUB, NMS.30)
 MISC.MAP <- c(
   'age' = 'Age',
@@ -118,6 +119,7 @@ MISC.MAP <- c(
   'surgery' = 'Surgery'
 )
 PUB.MAP <- c(NMS.D.MAP.PUB, NMS.30.MAP.PUB, MISC.MAP)
+PUB.MAP.N <- c(NMS.D.MAP.PUB.N, NMS.30.MAP.PUB, MISC.MAP)  # This one just has n
 ALL.BUT.NMS <- c('age', 'sex', 'pdonset', 'durat_pd', 'cisitot',
                  'tremor', 'bradykin', 'rigidity', 'axial')
 INTERPRETED <- c("age", "sex", "pdonset", "durat_pd", "cisitot",

### 2016-05-08 writeup/manuscript/manuscript.tex
Pub-ready tables, more analysis of nms30

@@
 % Use times font
-\usepackage{pslatex}
+\usepackage{mathptmx}
+\usepackage{lmodern}
 
@@
 \usepackage{threeparttable}
+% For line-breaking cells
+\newcommand{\spc}[2][c]{%
+  \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
 \usepackage{float}
@@
 
-Out of the 951 patients in the study, we used listwise deletion to exclude 50 patients due to
-missing measurements in the variables of interest, resulting in 901 remaining patients. All
+Out of the 951 patients in the study, we used listwise deletion to exclude 47 patients due to
+missing measurements in the variables of interest, resulting in 904 remaining patients. All
 symptoms were scaled to $z$-scores ($\mu = 0$, $\sigma = 1$) before clustering, and unscaled
@@
 
-For each symptom in both clusterings, we used one-way ANOVA to test the equality of symptom means
-across the clusters found, using Bonferroni correction for multiple testing with corrected $p <
-0.05$ considered significant. Differences among pairwise clusters were tested post-hoc using
-Tukey's range test.
+For each symptom in both clusterings, we used one-way ANOVA and $\chi^2$ tests to test the equality
+of symptom means and proportions across the clusters found, using Bonferroni correction for
+multiple testing with corrected $p < 0.05$ considered significant. Differences among pairwise
+clusters were tested post-hoc using Tukey's range test for continuous means, or pairwise $\chi^2$
+tests with Bonferroni correction for proportions.
 
@@
 available in Table~\ref{tab:nmd} with boxplots in Figure~\ref{fig:box}. Cluster means for all
-symptoms were found to be statistically different, with specific pairwise differences noted in the table.
+symptoms were found to be statistically different $(p < 0.05 / 13 \approx 0.003)$, with specific
+pairwise differences noted in the table. Additionally, statistics for variables not used in the
+clustering were explored similarly in Table~\ref{tab:nmd_extra}; cluster means for all variables
+were found to be statistically different, except for PD onset.
 
-Cluster 1 ($n = 406$) patients were mildly affected in all domains.
+Cluster 1 ($n = 406$) patients were mildly affected in all domains. This cluster was characterized
+by relatively lower disease durations and ages.
 
-Cluster 2 ($n = 189$) patients were severely
-affected in nonmotor domains but mildly affected in motor domains (Nonmotor-Dom), cluster 3 ($n =
-221$) patients were severely affected in motor domains but mildly affected in nonmotor domains
-(Motor-Dom), and cluster 4 ($n = 88$) patients were severely affected in all domains. Statistics
-for variables not used in the clustering were explored in Table~\ref{tab:nmd_extra}.
+Cluster 2 ($n = 189$) patients were severely affected in nonmotor domains but mildly affected in
+motor domains. This cluster had an incidence of motor symptoms relatively similar to the Cluster 1
+(mild) subtype especially in tremor, but generally expressed significantly higher scores for
+nonmotor symptoms than Clusters 1 and 3, especially in the sleep/fatigue, mood/cognition and
+miscellaneous domains.
+% But also high variability
+
+Cluster 3 ($n = 221$) patients were severely affected in motor domains but mildly affected in
+nonmotor domains. Mean motor scores were greater than the means of clusters 1 and 2, but less than
+4, with the exception of tremor, which was especially high.
+
+Cluster 4 ($n = 88$) patients were severely affected in all domains, having the greatest symptom
+mean out of all four clusters with the exception of tremor. Consequently, patients in Cluster 4 had
+the longest average PD duration and oldest ages, but did not have a significantly different age of
+PD onset.
 
@@
   \centering
-  \caption{Summaries of clustering on nonmotor domains. Bold indicates the highest score for that
-  symptom out of the four clusters.}
+  \caption{Summaries of clustering on nonmotor domains. Unless otherwise specified, statistics are
+    reported as mean (sd). \textbf{Bold} indicates the highest score for that symptom out of the
+  four clusters.}
   \label{tab:nmd}
@@
     Cluster & 1 & 2 & 3 & 4 \\
-    $n$ & 406 (44.9\%) & 189 (20.9\%) & 221 (24.4\%) & 88 (9.7\%) \\
+    $n$ & 406 (45\%) & 189 (21\%) & 221 (24\%) & 88 (10\%) \\
     \midrule
@@
     \small
-    \item[1] Significant difference with cluster 1 ($p < 0.05$)
-    \item[2] Significant difference with cluster 2 ($p < 0.05$)
-    \item[3] Significant difference with cluster 3 ($p < 0.05$)
-    \item[4] Significant difference with cluster 4 ($p < 0.05$)
+    \item[1] Significant difference with cluster 1 (adjusted $p < 0.05$)
+    \item[2] Significant difference with cluster 2 (adjusted $p < 0.05$)
+    \item[3] Significant difference with cluster 3 (adjusted $p < 0.05$)
+    \item[4] Significant difference with cluster 4 (adjusted $p < 0.05$)
     \item[\textdagger] Another footnote.
@@
     \toprule
-    & 1 & 2 & 3 & 4 \\
+    Cluster & 1 & 2 & 3 & 4 \\
     \midrule
-      Age & 62.7 (9.7) & 64 (9) & 65 (10.2) & 70.5 (8.9) \\
-      Sex & 0.4 (0.5) & 0.5 (0.5) & 0.3 (0.5) & 0.4 (0.5) \\
+      Sex (\% male) & 63 & 53\tnote{3} & 71\tnote{2} & 58 \\
+      Age & 62.7 (9.7)\tnote{34} & 64 (9.0)\tnote{4} & 65 (10.2)\tnote{14} & 70.5 (8.9)\tnote{123} \\
       PD\_Onset & 56 (10.6) & 55.2 (10.1) & 56.9 (11.1) & 58.2 (11.4) \\
-      PD\_Duration & 6.7 (4.8) & 8.7 (5.7) & 8 (5.5) & 12.3 (8.1) \\
-      CISI\_Total & 5.7 (3.2) & 9.2 (3.9) & 9.7 (3.5) & 14.8 (5.1) \\
-      Surgery & 0 (0.2) & 0.1 (0.2) & 0 (0.2) & 0 (0.1) \\
+      PD\_Duration & 6.7 (4.8)\tnote{234} & 8.7 (5.7)\tnote{14} & 8 (5.5)\tnote{14} & 12.3
+      (8.1)\tnote{123} \\
+      CISI\_Total & 5.7 (3.2)\tnote{234} & 9.2 (3.9)\tnote{14} & 9.7 (3.5)\tnote{14} & 14.8
+      (5.1)\tnote{123} \\
     \bottomrule
@@
     \small
-    \item[1] Significant difference with cluster 1 ($p < 0.05$)
-    \item[2] Significant difference with cluster 2 ($p < 0.05$)
-    \item[3] Significant difference with cluster 3 ($p < 0.05$)
-    \item[4] Significant difference with cluster 4 ($p < 0.05$)
-    \item[\textdagger] Another footnote.
+    \item[1] Significant difference with cluster 1 (adjusted $p < 0.05$)
+    \item[2] Significant difference with cluster 2 (adjusted $p < 0.05$)
+    \item[3] Significant difference with cluster 3 (adjusted $p < 0.05$)
+    \item[4] Significant difference with cluster 4 (adjusted $p < 0.05$)
   \end{tablenotes}
@@
 \end{table*}
+
 \subsection{Clustering with nonmotor symptoms}
 
-\subsection{Hierarchical clustering on symptoms}
+$k$-means clustering performed on the 30 individual nonmotor symptoms found similar patterns to the
+first clustering (Table~\ref{tab:nms30}, heatmap in Figure~\ref{fig:nms30-hm}). Means of all symptoms were found to
+differ across clusters except for PD\_Onset and tremor.
 
-Hierarchical clustering on the 30 nonmotor symptoms and the four motor symptoms produced the
-dendrogram in Figure~\ref{fig:hc}. Predictably, symptoms in the same nonmotor domains tended to
-cluster together, with the exceptions of diplopia (domain 4), drowsiness (domain 2), and the
-symptoms in domain 9 (miscellaneous). Notably, tremor was the most dissimilar symptom, occupying a single branch at the
-top of the tree.
+Clusters 1 ($n = 509$) and 4 ($n = 49$), the mild and severe subtypes of the previous clustering,
+retain their characteristics here, while clusters 2 ($n = 97$) and 3 ($n = 249$) are different. 2
+is a mood/cognition-dominant cluster, with relatively higher proportion of female patients and
+additionally a relatively more severe incidence of insomnia. 3 is not characterized by any
+especially high symptoms and is thus an ``average'' group of patients.
+% Once again, CISI TOTAL...
+
+% Perhaps this table belongs in appendix?
+\begin{table*}[t]
+  \centering
+  \caption{Summaries of clustering on individual nonmotor domains. Unless otherwise specified,
+    statistics are reported as mean (sd). \textbf{Bold} indicates the highest score for that
+  symptom out of the four clusters.}
+  \label{tab:nms30}
+  \begin{threeparttable}
+  \small
+\begin{tabular}{lrllll}
+  \toprule
+    & Cluster & 1 & 2 & 3 & 4 \\
+    & $n$ & 509 (56.3\%) & 97 (10.7\%) & 249 (27.5\%) & 49 (5.4\%) \\
+  \midrule
+  1.\ Cardiovascular &
+  Lightheadedness & 0.6 (1.3)\tnote{234} & 2.4 (3.0)\tnote{14} & 2.1 (2.9)\tnote{14} & \textbf{5.2}
+  (4.2)\tnote{123} \\
+  &Fainting & 0.1 (0.7)\tnote{24} & 0.5 (1.4)\tnote{134} & 0.2 (0.6)\tnote{24} & \textbf{2.4}
+  (3.5)\tnote{123} \\
+  \midrule
+  2.\ Sleep/fatigue &
+  Drowsiness & 1.0 (1.9)\tnote{234} & 2.8 (3.3)\tnote{14} & 2.5 (3.0)\tnote{14} & \textbf{7.0}
+  (3.8)\tnote{123} \\
+  &Fatigue & 1.3 (2.0)\tnote{234} & 6.3 (4.2)\tnote{13} & 4.8 (3.9)\tnote{124} & \textbf{7.3} (3.5)\tnote{13} \\
+  &Insomnia & 1.2 (2.6)\tnote{234} & \textbf{5.5} (4.8)\tnote{13} & 3.2 (3.8)\tnote{12} & 4.2
+  (3.9)\tnote{1} \\
+  &RLS & 0.7 (1.6)\tnote{234} & 2.3 (3.5)\tnote{14} & 1.9 (3.2)\tnote{14} & \textbf{3.9}
+  (3.9)\tnote{123} \\
+  \midrule
+  3.\ Mood/cognition &
+  Loss\_interest & 0.4 (1.1)\tnote{234} & \textbf{6.1} (3.8)\tnote{134} & 1.0 (1.8)\tnote{124} & 4.6 (3.7)\tnote{123} \\
+  &Loss\_activities & 0.7 (1.6)\tnote{234} & \textbf{7.1} (4.1)\tnote{134} & 1.7 (2.4)\tnote{124} &
+  5.4 (4.0)\tnote{123} \\
+  &Anxiety & 0.9 (1.8)\tnote{234} & \textbf{6.6} (4.3)\tnote{134} & 2.0 (2.8)\tnote{124} & 4.6 (3.9)\tnote{123} \\
+  &Depression & 0.8 (1.6)\tnote{234} & \textbf{7.8} (3.7)\tnote{134} & 2.3 (3.0)\tnote{124} & 4.3 (3.7)\tnote{123} \\
+  &Flat\_affect & 0.4 (1.1)\tnote{234} & \textbf{4.9} (4.3)\tnote{134} & 0.9 (1.9)\tnote{124} & 2.7 (3.0)\tnote{123} \\
+  &Loss\_pleasure & 0.4 (1.2)\tnote{234} & \textbf{6.8} (4.1)\tnote{134} & 1.1 (2.1)\tnote{124} & 3.4 (3.3)\tnote{123} \\
+  \midrule
+  % This one is spread across multiple cells
+  4.\ Perception/ & Hallucination & 0.2 (1.0)\tnote{234} & 0.8 (1.9)\tnote{14} & 0.6 (1.6)\tnote{14} & \textbf{4.6} (3.6)\tnote{123} \\
+  hallucination &Delusion & 0.1 (0.7)\tnote{24} & 1.5 (3.1)\tnote{124} & 0.2 (1.0)\tnote{24} &
+  \textbf{3.1} (3.8)\tnote{123} \\
+  &Diplopia & 0.2 (0.7)\tnote{234} & 1.0 (2.5)\tnote{14} & 0.6 (1.4)\tnote{14} & \textbf{4.4}
+  (4.3)\tnote{123} \\
+  \midrule
+  5.\ Attention/ & Loss\_concentration & 0.9 (1.8)\tnote{234} & 4.0 (3.6)\tnote{134} & 2.5 (3.1)\tnote{124} & \textbf{6.6} (3.9)\tnote{123} \\
+  memory &Forget\_explicit & 0.9 (1.6)\tnote{234} & 3.8 (3.9)\tnote{134} & 2.2 (2.9)\tnote{124} & \textbf{6.7} (3.6)\tnote{123} \\
+  &Forget\_implicit & 0.7 (1.6)\tnote{234} & 3.1 (3.7)\tnote{134} & 1.8 (2.6)\tnote{124} & \textbf{6.7} (4.0)\tnote{123} \\
+  \midrule
+  6.\ Gastrointestinal &
+  Drooling & 0.7 (1.6)\tnote{234} & 3.0 (4.1)\tnote{14} & 3.0 (3.8)\tnote{14} & \textbf{6.1} (4.4)\tnote{123} \\
+  &Swallowing & 0.4 (1.2)\tnote{234} & 1.2 (2.1)\tnote{14} & 1.7 (2.8)\tnote{14} & \textbf{4.5} (3.9)\tnote{123} \\
+  &Constipation & 1.6 (2.9)\tnote{234} & 3.6 (4.3)\tnote{14} & 3.3 (4.1)\tnote{14} & \textbf{7.0} (5.1)\tnote{123} \\
+  \midrule
+  7.\ Urinary &
+  Urinary\_urgency & 1.0 (1.8)\tnote{234} & 3.5 (4.1)\tnote{14} & 3.8 (3.8)\tnote{14} & \textbf{7.8} (4.0)\tnote{123} \\
+  &Urinary\_frequency & 0.9 (1.9)\tnote{234} & 4.0 (4.1)\tnote{14} & 3.8 (4.0)\tnote{14} & \textbf{6.5} (4.0)\tnote{123} \\
+  &Nocturia & 1.7 (2.3)\tnote{234} & 5.0 (4.3)\tnote{14} & 5.1 (4.1)\tnote{14} & \textbf{7.7} (3.9)\tnote{123} \\
+  \midrule
+  8.\ Sexual &
+  Sex\_drive & 0.8 (1.9)\tnote{234} & 4.5 (4.9)\tnote{13} & 2.3 (3.6)\tnote{124} & \textbf{5.2} (5.4)\tnote{13} \\
+  &Sex\_dysfunction & 0.8 (2.2)\tnote{234} & 3.2 (4.7)\tnote{1} & 2.4 (3.8)\tnote{14} &
+  \textbf{4.6} (5.4)\tnote{13} \\
+  \midrule
+  9.\ Miscellaneous &
+  Unexplained\_pain & 0.7 (1.9)\tnote{234} & 2.5 (3.7)\tnote{14} & 3.0 (4.1)\tnote{14} & \textbf{4.2} (4.3)\tnote{123} \\
+  &Gustation\_olfaction & 1.5 (2.9)\tnote{234} & 3.6 (4.3)\tnote{1} & 3.2 (3.9)\tnote{14} &
+  \textbf{4.7} (4.8)\tnote{13} \\
+  &Weight\_change & 0.9 (1.7)\tnote{234} & 2.5 (3.4)\tnote{14} & 1.9 (3.0)\tnote{14} & \textbf{3.9} (4.2)\tnote{123} \\
+  &Sweating & 0.7 (1.8)\tnote{234} & 2.2 (3.5)\tnote{1} & 2.8 (4.1)\tnote{1} & \textbf{3.2}
+  (3.8)\tnote{1} \\
+  \midrule
+  Motor &
+  Tremor & 2.5 (2.3) & 2.1 (2.4) & 2.6 (2.6) & \textbf{4.1} (4.3) \\
+  &Bradykinesia & 2.0 (1.2)\tnote{234} & 2.8 (1.4)\tnote{14} & 2.8 (1.4)\tnote{14} & \textbf{3.9} (1.8)\tnote{123} \\
+  &Rigidity & 1.9 (1.2)\tnote{234} & 2.4 (1.3)\tnote{14} & 2.6 (1.4)\tnote{14} & \textbf{3.6} (1.6)\tnote{123} \\
+  &Axial & 2.2 (1.8)\tnote{234} & 4.4 (2.8)\tnote{14} & 4.2 (2.7)\tnote{14} & \textbf{7.2} (3.2)\tnote{123} \\
+  \midrule
+  Variables not & Age & 62.7 (9.9)\tnote{34} & 64.5 (9.6)\tnote{4} & 66 (9.2)\tnote{14} & 71.5
+  (8.1)\tnote{123} \\
+  used in analysis & Sex (\% Male) & 65\tnote{2} & 51\tnote{1} & 60 & 63 \\
+  & PD\_Onset & 55.9 (10.8) & 56.2 (10.4) & 56.7 (10.5) & 57.8 (11.0) \\
+  & PD\_Duration & 6.8 (4.8)\tnote{34} & 8.3 (5.3)\tnote{4} & 9.3 (6.3)\tnote{14} & 13.8
+  (8.2)\tnote{123} \\
+  & CISI\_Total & 6.3 (3.4)\tnote{234} & 10.8 (4.7)\tnote{14} & 10 (4.0)\tnote{14} & 15.1 (5.7)\tnote{123} \\
+   \bottomrule
+\end{tabular}
+  \begin{tablenotes}
+    \small
+    \item[1] Significant difference with cluster 1 (adjusted $p < 0.05$)
+    \item[2] Significant difference with cluster 2 (adjusted $p < 0.05$)
+    \item[3] Significant difference with cluster 3 (adjusted $p < 0.05$)
+    \item[4] Significant difference with cluster 4 (adjusted $p < 0.05$)
+  \end{tablenotes}
+  \end{threeparttable}
+\end{table*}
 
@@
 
+% TODO: If the outline of the paper is good enough it might be possible to have this inline
+% in the paper somewhere. If so, change figure* -> figure, make h! (force horizontal?), and switch
+% 0.8 linewidth to full linewidth
+% Also, I'm going to include this AFTER nms30m-colhc, so that colhc/boxplots fits nicely on one
+% page
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=0.7\linewidth]{nms30-hm-pub.pdf}
+  % TODO: Variables not used in analysis? Label color - black nms, blue nonmotor, grey ?
+  \caption{Heatmap of clusters found when clustering with nonmotor symptoms. Color indicates
+    cluster mean as a $z$-score for the given symptom; red is higher, green is lower.}
+  \label{fig:nms30-hm}
+\end{figure*}
+
+\subsection{Hierarchical clustering on symptoms}
+
+Hierarchical clustering on the 30 nonmotor symptoms and the four motor symptoms produced the
+dendrogram in Figure~\ref{fig:hc}. Predictably, symptoms in the same nonmotor domains tended to
+cluster together, with the exceptions of diplopia (domain 4), drowsiness (domain 2), and the
+symptoms in domain 9 (miscellaneous). Notably, tremor was the most dissimilar symptom, occupying a
+single branch at the top of the tree.
+
 \section{Discussion}
 
-$k$-means clustering on this Parkinsons' Disease data set reveals clusters that confirm previous
+$k$-means clustering on this dataset reveals clusters that confirm previous
 findings in the field, mainly van Rooden et al.\cite{vanrooden10} and the identification of four
@@
 
-It remains to be seen whether these classification models, especially the one-vs-all decision
-trees, are useful in clinical practice.
-
-\subsection{New conclusions}
-\label{ssec:newconc}
-
 First, the longitudinal analysis gives more insight into the clusters found when clustering on

### 2016-05-08 R/cleanup.R

@@ -3,6 +3,8 @@
 library(doBy)
 library(xtable)
 library(reshape)
+library(plyr)
+library(infotheo)
 
 nmsd.symptoms <- c(
   NMS.D.NAMES,
@@ -29,10 +31,36 @@ nmsd.extra <- c(
 #   1   2   3  4
 # 406 189 221 88
 
+# V-measure ====
+# always useful...
+v.measure <- function(a, b) {
+  mi <- mutinformation(a, b)
+  entropy.a <- entropy(a)
+  entropy.b <- entropy(b)
+  if (entropy.a == 0.0) {
+    homogeneity <- 1.0
+  } else {
+    homogeneity <- mi / entropy.a
+  }
+  if (entropy.b == 0.0) {
+    completeness <- 1.0
+  } else {
+    completeness <- mi / entropy.b
+  }
+  if (homogeneity + completeness == 0.0) {
+    v.measure.score <- 0.0
+  } else {
+    v.measure.score <- (2.0 * homogeneity * completeness
+                        / (homogeneity + completeness))
+  }
+  # Can also return homogeneity and completeness if wanted
+  c(homogeneity, completeness, v.measure.score)
+}
+
 # ON NONMOTOR DOMAINS ====
 # Somewhat oddly, the correct clustering vector lies in trees$clusters4$clustering$cluster
-present <- rename(raw.omitted, PUB.MAP)
-present.full <- rename(raw.omitted.full, PUB.MAP)
+present <- reshape::rename(raw.omitted, PUB.MAP)
+present.full <- reshape::rename(raw.omitted.full, PUB.MAP)
 present$Cluster <- trees$clusters4$clustering$cluster
 present.full$Cluster <- trees$clusters4$clustering$cluster
 
@@ -58,7 +86,15 @@ to.latex(present[, nmsd.symptoms], "../writeup/manuscript/include/nmsd_summaries
 to.latex(present.full[, nmsd.extra], "../writeup/manuscript/include/nmsd_extra.tex")
 
 # Publication-ready dendrogram ====
-labels.wo.d <- unname(sapply(labels(hm.m$colDendrogram), function(s) {
+remove.domain.n <- function(s) {
+  splitted <- strsplit(s, '-')[[1]]
+  if (length(splitted) == 1) {
+    s
+  } else {
+    splitted[3]
+  }
+}
[…]

### 2016-05-08 R/kmeans-dtree.R

@@ -750,6 +750,8 @@ clus4.wide.st <- clusters.raw[["4"]]
 # Why isn't this already a factor? Really confused
 # Because it wasn't set in clusters.raw - if this is a bad thing lmk
 clus4.wide.st$cluster <- as.factor(clus4.wide.st$cluster)
+# Just NMS pls
+clus4.wide.st <- clus4.wide.st[, ]
 # Assuming 1st column is cluster (which it should be)
 oneways <- lapply(colnames(clus4.wide.st[, -1]), function(col) {
   fm <- substitute(i ~ cluster, list(i = as.name(col)))

### 2016-05-08 R/nms30.R

@@ -107,7 +107,7 @@ hm.m <- heatmap.2(as.matrix(nms30m.s),
                   trace='none')
 colD.m <- as.ggdend(hm.m$colDendrogram)
 
-# Average is hella different
+# Average is different
 asdf <- hclust(dist(t(nms30m.s)), method='average')
 plot(asdf)
 

### 2016-05-08 R/preprocessing.R

@@ -24,6 +24,10 @@ MOTOR.SYMPTOMS <- c("tremor",
                    "bradykin",
                    "rigidity",
                    "axial")
+MOTOR.PUB <- c("Tremor",
+                   "Bradykinesia",
+                   "Rigidity",
+                   "Axial")
 NMS.30.NAMES <- c(
   'd1-1-lightheaded',
   'd1-2-fainting',
@@ -84,7 +88,7 @@ NMS.30.NAMES.PUB <- c(
   'Sex_drive',
   'Sex_dysfunction',
   'Unexplained_pain',
-  'Gustation/olfacttion',
+  'Gustation_olfaction',
   'Weight_change',
   'Sweating'
 )
@@ -105,6 +109,7 @@ NMS.D.NAMES <- c(
 NMS.D.MAP.PUB <- setNames(NMS.D.NAMES, NMS.D)
 NMS.D.MAP.PUB.N <- setNames(sapply(1:length(NMS.D.NAMES), function(i) paste(i, NMS.D.NAMES[i], sep = "-")), NMS.D)
 NMS.30.MAP.PUB <- setNames(NMS.30.NAMES.PUB, NMS.30)
+NMS.30.LONG.SHORT.MAP <- setNames(NMS.30.NAMES.PUB, NMS.30.NAMES)
 MISC.MAP <- c(
   'age' = 'Age',
   'sex' = 'Sex',
@@ -116,7 +121,8 @@ MISC.MAP <- c(
   'rigidity' = 'Rigidity',
   'axial' = 'Axial',
   'ldopa' = 'ldopa',
-  'surgery' = 'Surgery'
+  'surgery' = 'Surgery',
+  'cluster' = 'Cluster'
 )
 PUB.MAP <- c(NMS.D.MAP.PUB, NMS.30.MAP.PUB, MISC.MAP)
 PUB.MAP.N <- c(NMS.D.MAP.PUB.N, NMS.30.MAP.PUB, MISC.MAP)  # This one just has n

### 2016-05-09 writeup/manuscript/manuscript.tex
Updated analyses and draft of materials/results

@@
 % PREAMBLE ====
-\documentclass[preprint,3p,twocolumn]{elsarticle} % TODO: What model?
+\documentclass[preprint,5p]{elsarticle} % TODO: What model?
 
@@
 \title{Parkinson's disease cluster analysis}
-\author[c3]{Pablo Martinez-Martin}
-\ead{pmartinez@isciii.es}
-\author[bc]{Jesse Mu}
-\ead{muj@bc.edu}
-\author[cig]{Concha Bielza}
-\ead{mcbielza@fi.upm.es}
-\author[cig]{Pedro Larra\~naga}
-\ead{plarranaga@fi.upm.es}
-\address[c3]{Area of Applied Epidemiology, National Centre of Epidemiology and CIBERNED, Carlos III Institute of Health, Madrid, Spain}
-\address[bc]{Department of Computer Science, Boston College, Chestnut Hill, Massachusetts, USA}
-\address[cig]{Computational Intelligence Group, Polytechnic University of Madrid, Madrid, Spain}
+% \author[c3]{Pablo Martinez-Martin}
+% \ead{pmartinez@isciii.es}
+% \author[bc]{Jesse Mu}
+% \ead{muj@bc.edu}
+% \author[cig]{Concha Bielza}
+% \ead{mcbielza@fi.upm.es}
+% \author[cig]{Pedro Larra\~naga}
+% \ead{plarranaga@fi.upm.es}
+% \address[c3]{Area of Applied Epidemiology, National Centre of Epidemiology and CIBERNED, Carlos III Institute of Health, Madrid, Spain}
+% \address[bc]{Department of Computer Science, Boston College, Chestnut Hill, Massachusetts, USA}
+% \address[cig]{Computational Intelligence Group, Polytechnic University of Madrid, Madrid, Spain}
 
@@
 symptoms were scaled to $z$-scores ($\mu = 0$, $\sigma = 1$) before clustering, and unscaled
-afterwards. Analysis was conducted in R 3.2.4 (www.r-project.org).
+afterwards for interpretation. Analyses were conducted in R 3.2.4 (www.r-project.org).
 
@@
 patients in the dataset: the first on the nine aggregate nonmotor symptom
-domains and the four motor symptoms, and the second on the 30 individual
-nonmotor symptoms collected with the four motor symptoms. Complete-linkage
+domains and the four motor symptoms (``domains clustering''), and the second on the 30 individual
+nonmotor symptoms collected with the four motor symptoms (``symptoms clustering''). Complete-linkage
 hierarchical clustering on the 30 nonmotor symptoms was also performed to observe the
@@
 the globally-optimal clusters reported from stability measures of the clValid R package
-\cite{brock11clvalid} stability are 2, 6, and 8 (Figure~\ref{fig:stability}), there appeared to be
+\cite{brock11clvalid} stability were 2, 6, and 8 (Figure~\ref{fig:stability}), there appeared to be
 consensus that $k = 4$ is a locally-optimal number of clusters which supported the gap statistic
@@
   \caption{Plot of gap statistic versus number of clusters with $k$-means on 100 bootstrapped
-    samples. Error bars are standard error. Per the method described in\cite{tibshirani01gap}, 4 is the smallest $k$ such
+    samples. Error bars are standard error. Per the method described in Tibshirani\cite{tibshirani01gap}, 4 is the smallest $k$ such
     that $\text{Gap}(k) \geq \text{Gap}(k + 1) - \text{se}_{k + 1}$.}
@@
 
-For the clustering on nonmotor domains, we displayed the distribution of each symptom for the four
+For the domains clustering, we displayed the distribution of each symptom for the four
 clusters using boxplots, which allowed us to visualize the center and spread of each cluster. Since
-the number of variables was larger for the clustering on the individual symptoms, we visualized
-results for the second analysis with a heatmap. Finally, for the hierarchical clustering on the
-symptoms themselves, we displayed results in a dendrogram and clustered the symptoms into four
-interpretable clusters.
+the number of variables was larger for the symptoms clustering, we visualized results for the
+second analysis with a heatmap. Finally, for the hierarchical clustering on the symptoms
+themselves, we displayed results in a dendrogram and clustered the symptoms into four interpretable
+clusters.
 
@@
 
+To compare the clusterings, we computed the adjusted rand index\cite{hubert1985} (ARI) and
+$v$-measure\cite{rosenberg07vmeasure} to evaluate similarity between the two clusterings, and used
+stacked barplots to visualize the alignment of the clusters.
+
+Lastly, to explore the relationship between symptom severity and disease duration, we computed the
+linear correlation for each symptom with PD duration and fitted smoothed loess curves to the
+data both globally and for each cluster found in the previous analyses.
+
 \section{Results}
 
-\subsection{Clustering with nonmotor domains}
+\subsection{Domains clustering}
 
-$k$-means clustering on the nine nonmotor domains and the four motor symptoms found four clusters,
-available in Table~\ref{tab:nmd} with boxplots in Figure~\ref{fig:box}. Cluster means for all
+$k$-means clustering on the nine nonmotor domains and the four motor symptoms found four clusters
+(Table~\ref{tab:nmd}; boxplots in Figure~\ref{fig:box}). Cluster means for all
 symptoms were found to be statistically different $(p < 0.05 / 13 \approx 0.003)$, with specific
@@
 nonmotor symptoms than Clusters 1 and 3, especially in the sleep/fatigue, mood/cognition and
-miscellaneous domains.
+miscellaneous domains. This group also had a statistically higher percentage of females than
+Cluster 3.
 % But also high variability
@@
   \centering
-  \caption{Summaries of clustering on nonmotor domains. Unless otherwise specified, statistics are
+  \caption{Summaries of domains clustering. Unless otherwise specified, statistics are
     reported as mean (sd). \textbf{Bold} indicates the highest score for that symptom out of the
@@
     \item[4] Significant difference with cluster 4 (adjusted $p < 0.05$)
-    \item[\textdagger] Another footnote.
   \end{tablenotes}
@@
   }
-  \caption{Boxplot summaries for clusters found \\ when clustering with nonmotor domains.}
+  \caption{Boxplot cluster summaries \\ for domains clustering.}
   \label{fig:box}
@@
 
-\subsection{Clustering with nonmotor symptoms}
+\subsection{Symptoms clustering}
 
 $k$-means clustering performed on the 30 individual nonmotor symptoms found similar patterns to the
-first clustering (Table~\ref{tab:nms30}, heatmap in Figure~\ref{fig:nms30-hm}). Means of all symptoms were found to
-differ across clusters except for PD\_Onset and tremor.
+first clustering (Table~\ref{tab:nms30}, heatmap in Figure~\ref{fig:nms30-hm}). Means of all
+symptoms were found to differ across clusters except for PD\_Onset and tremor.
 
 Clusters 1 ($n = 509$) and 4 ($n = 49$), the mild and severe subtypes of the previous clustering,
-retain their characteristics here, while clusters 2 ($n = 97$) and 3 ($n = 249$) are different. 2
-is a mood/cognition-dominant cluster, with relatively higher proportion of female patients and
-additionally a relatively more severe incidence of insomnia. 3 is not characterized by any
-especially high symptoms and is thus an ``average'' group of patients.
+retained their characteristics here, while clusters 2 ($n = 97$) and 3 ($n = 249$) were different.
+2 was a mood/cognition-dominant cluster, with a relatively higher proportion of female patients and
+additionally a more severe incidence of insomnia. 3 was not characterized by any especially high
+symptoms and was thus classified as a group of patients with average severity.
 % Once again, CISI TOTAL...
+
+\subsection{Comparison between clusterings}
+
+The two clusterings are somewhat similar ($\text{ARI} = 0.46$; $\text{$v$-measure} = 0.43$).
+Alignment of the clusters is visualized in Figure~\ref{fig:align}. While clusters 1 and 4 of both
+clusterings align well with each other, main differences in the clusters occur in the mixing of
+clusters 2 and 3 of both clusterings; for example, domains cluster 3 (motor-dominant) is somewhat
+evenly split among symptoms clusters 1 (mild) and 3 (average).
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{cluster-alignment.pdf}
+  \caption{Bar chart indicating the distribution of opposite cluster membership for the patients of
+  a given cluster. For example, the bar furthest left indicates that close to 100\% of patients in
+domains cluster 1 are in symptoms cluster 1, with the rest in symptoms cluster 3.}
+  \label{fig:align}
+\end{figure}
 
@@
   % TODO: Variables not used in analysis? Label color - black nms, blue nonmotor, grey ?
-  \caption{Heatmap of clusters found when clustering with nonmotor symptoms. Color indicates
-    cluster mean as a $z$-score for the given symptom; red is higher, green is lower.}
+  \caption{Heatmap of symptoms clustering. Color indicates cluster mean as a $z$-score for the
+  given symptom.}
   \label{fig:nms30-hm}
@@
 
+\subsection{Longitudinal analysis}
+
+Most variables had little to no correlation with PD duration (Figure~\ref{fig:corr}). Variables
+with the highest correlation include symptoms urinary frequency, swallowing, and drooling; domains
+miscellaneous, urinary, and gastrointestinal; and total CISI score. Scatterplots for CISI Total,
+Tremor, Anxiety, and Depression appear in Figure~\ref{fig:long4}.
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=\linewidth]{cor-unbinned.pdf}
+  \caption{Linear correlation of each symptom with PD duration.}
+  \label{fig:corr}
+\end{figure*}
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=0.8\linewidth]{long4-d.pdf}
+  \caption{Selected symptoms for 901 patients plotted against PD duration. Smoothed loess curves for each cluster
+  are drawn in their respective colors; the black curve is the curve for the entire population.}
+  \label{fig:long4}
+\end{figure*}
+
 \section{Discussion}
 
-$k$-means clustering on this dataset reveals clusters that confirm previous
+Domains clustering reveals clusters that confirm previous
 findings in the field, mainly van Rooden et al.\cite{vanrooden10} and the identification of four
-subtypes of Parkinson's disease: mild, nonmotor-predominant, motor-predominant, and severe.  van
-Rooden's work was done with a separate dataset using a different modeling method
+subtypes of Parkinson's disease: mild, nonmotor-dominant, motor-dominant, and severe. van
+Rooden's work was conducted with a separate dataset using a different modeling method
 (expectation-maximization), and this investigation independently confirms these subtype
@@
 between 2 and 3 (nonmotor/motor predominated) subtypes are insignificant
-(Table~\ref{tab:tukeyhsd}), suggesting different developmental paths of the disease.
+(Table~\ref{tab:nmd}), importantly suggesting different developmental paths of the disease.
 
 Overall, little information was found in pdonset,
-durat\_pd, or current age, according to Tables~\ref{tab:tukeyhsd}
-and~\ref{tab:info_gain}.  Mean ages were similar for subgroups 1, 2, and 3 ($p >
-0.05$), but different for the severe subtype 4, which makes sense given that
-patients in 4 also have longer disease durations. Specifically, clusters 1 and
-4 seem to be phenotypically quite similar, except at different stages of
-disease progression, given cluster 4's higher age and durat\_pd scores.
+durat\_pd, or current age. Mean ages were similar for clusters 1, 2, and 3 ($p >
+0.05$), but different for the severe cluster 4, intuitively since cluster 4 represents a more
+advanced stage of the disease.  Specifically, clusters 1 and 4 seem to be phenotypically quite
+similar, except at different stages of disease progression, given cluster 4's higher age and
+durat\_pd scores.
 
@@
 the motor direction, and one in the nonmotor. Both groups have similar age,
-pdonset, and durat\_pd scores, but differ wildly in symptomatic expression.
-Cluster 2 is dominated by a high prevalence of nonmotor symptoms, such as
-nms\_d2, nms\_d3, nms\_d7, and nms\_d9. Cluster 3, however, is dominated by a
-high prevalence of motor symptoms, while most motor symptoms are similar to the
-mild cluster 1. Of note is that the tremor population mean
-is the highest cluster mean, even higher than the severe subtype 4. This
-motor-dominant cluster may thus overlap with Ma's tremor dominant/slow
-progression cluster\cite{ma15}.
+PD onset, and PD duration scores, but differ significantly in symptomatic expression.
+Cluster 2 is dominated by a high severity of nonmotor domains, especially Sleep/Fatigue,
+Mood/Cognition, Urinary, and Miscellaneous.  Cluster 3, however, is dominated by a high prevalence
+of motor symptoms, where most motor symptoms are similar to the mild cluster 1. Of note is that the
+tremor population mean is the highest cluster mean, even higher than the severe subtype 4. This
+motor-dominant cluster may thus overlap with Ma's tremor dominant/slow progression
+cluster\cite{ma15}.
 
@@
 
-The most important nonmotor symptoms in determining these clusters were nms\_d2
-(sleep) and nms\_d3 (mood/cognition), which echo findings of Fereshtehnejad's
-longitudinal study\cite{fereshtehnejad15} and are similar to Sauerbier's
-identification of sleep dominant and cognitive dominant clinical NMS subtypes\cite{sauerbier15}.
-Compared to Erro et al.\cite{erro13}, nonmotor/motor
-dominant subtypes were indeed found, but an additional subgroup with relatively
-severe levels of both motor and nonmotor symptoms were found. Erro's benign
-subtype groups possibly overlap with the mild cluster 1 found in this
-investigation.
+% The most important nonmotor symptoms in determining these clusters were nms\_d2
+% (sleep) and nms\_d3 (mood/cognition), which echo findings of Fereshtehnejad's
+% longitudinal study\cite{fereshtehnejad15} and are similar to Sauerbier's
+% identification of sleep dominant and cognitive dominant clinical NMS subtypes\cite{sauerbier15}.
+% Compared to Erro et al.\cite{erro13}, nonmotor/motor
+% dominant subtypes were indeed found, but an additional subgroup with relatively
+% severe levels of both motor and nonmotor symptoms were found. Erro's benign
+% subtype groups possibly overlap with the mild cluster 1 found in this
+% investigation.
 
-\subsection{Nonmotor subtype: clustering and modeling}
-Nonmotor symptoms nms\_d2 and nms\_d3 became critical not only in
-classification trees distinguishing between the various symptoms but in the
-nonmotor-predominant subgroup itself. In $k$-means subdivision of the
-nonmotor-dominant subtype where $k = 2$ and $k = 3$, opposite trends were
-confirmed with nms\_d2 and nms\_d3 symptoms. Similarly, in the 2 and 4 vs rest
-decision tree (Figure~\ref{fig:dtree-2and4va-pruned}), nms\_d2 and nms\_d3 nodes were
-used to differentiate various categories of nomotor-dominant patients.
-When $k = 3$, the subtype with the highest nms\_d2 scores and lowest
-nms\_d3 scores had by far the highest axial scores, nms\_d6 (gastrointestinal)
-scores, and nms\_d7 (urinary) scores. Thus subtype 3 of the nonmotor-dominated
-group could include patients falling into the cognitive/depression-dominant or
-autonomic dominant subtypes.
+% \subsection{Nonmotor subtype: clustering and modeling}
 
-Despite the variety in symptomatic expression in this nonmotor group, what
-seems most consistent is the presence of nms\_d9 (miscellaneous) nonmotor
-symptoms, as it is used as the root node of the 2 vs all decision tree
-(Figure~\ref{fig:2va}) and the 2 and 4 vs rest decision tree
-(Figure~\ref{fig:dtree-2and4va-pruned}).
+% Nonmotor symptoms nms\_d2 and nms\_d3 became critical not only in
+% classification trees distinguishing between the various symptoms but in the
+% nonmotor-predominant subgroup itself. In $k$-means subdivision of the
+% nonmotor-dominant subtype where $k = 2$ and $k = 3$, opposite trends were
+% confirmed with nms\_d2 and nms\_d3 symptoms. Similarly, in the 2 and 4 vs rest
+% decision tree (Figure~\ref{fig:dtree-2and4va-pruned}), nms\_d2 and nms\_d3 nodes were
+% used to differentiate various categories of nomotor-dominant patients.
+% When $k = 3$, the subtype with the highest nms\_d2 scores and lowest
+% nms\_d3 scores had by far the highest axial scores, nms\_d6 (gastrointestinal)
+% scores, and nms\_d7 (urinary) scores. Thus subtype 3 of the nonmotor-dominated
+% group could include patients falling into the cognitive/depression-dominant or
+% autonomic dominant subtypes.
 
-First, the longitudinal analysis gives more insight into the clusters found when clustering on
-nms\_d\{1-9\}. According to Figure~\ref{fig:longcorr}, most symptoms are highly correlated, with PD
-duration, but notably, mood/cognition symptoms (nms9, nms10, nms12) and tremor are not correlated
-highly with PD duration. The differences in disease progression can be seen by the corresponding
-graphs, Figures~\ref{fig:nms9-multi} and~\ref{fig:nms10-multi}. In both graphs, what is interesting
-is that Subtype 2 (Nonmotor-Dominant) starts at higher scores for nms\_9 and nms\_10, thus
-indicating that these patients' subtype is can be determined early in PD duration from depressive
-symptom score. Similarly, when examining Subtype 3 (Motor-dominant) in
-Figure~\ref{fig:tremor-multi}, the mean tremor score is substantially higher from PD onset.
-Interestingly, Subtype 4 (Severe) generally starts at lower tremor and motor scores during disease
-onset (Figure~\ref{fig:cisitot-multi}), but then rises sharply, exceeding other Subtypes. More evidence
-that tremor is a unique motor symptom is located in Figure~\ref{fig:nms30m-colhc}, where it is the most
-distant symptom from all other symptoms.
+% Despite the variety in symptomatic expression in this nonmotor group, what
+% seems most consistent is the presence of nms\_d9 (miscellaneous) nonmotor
+% symptoms, as it is used as the root node of the 2 vs all decision tree
+% (Figure~\ref{fig:2va}) and the 2 and 4 vs rest decision tree
+% (Figure~\ref{fig:dtree-2and4va-pruned}).
 
-When examining all 30 symptoms, more evidence is given that the previously-discovered Subtype
-2 (Nonmotor-Dominant) may be primarily characterized by high depressive symptoms. By graphing the
-30 subtypes attached to the original clustering, as in Figure~\ref{fig:d9to30}, the mean scores of
-nms8, nms9, and nms10 for Subtype 2 are substantially higher than Subtypes 1 and 3.
+Our longitudinal analysis gives more insight into the clusters found in the previous analyses.
+According to Figure~\ref{fig:corr}, most symptoms are uncorrelated with PD duration, especially
+mood/cognition symptoms (anxiety, depression) and tremor.
+The differences in disease progression for each cluster can be seen by the corresponding graphs in
+Figure~\ref{fig:long4}. Cluster 2 (Nonmotor-Dominant) starts at higher
+scores for anxiety and depression, and actually decreases with increasing PD duration, thus indicating that these patients' subtype can be determined
+early after PD onset from the depressive symptom score. Similarly, when examining cluster 3
+(Motor-dominant), the mean tremor score is substantially higher from PD onset.
+Interestingly, cluster 4 (Severe) generally starts at lower tremor and motor scores during disease,
+but then rises sharply, exceeding other clusters. More evidence that tremor is a unique motor
+symptom is located in Figure~\ref{fig:hc}, where it is the most distant symptom from all other
+symptoms.
 
-Indeed, PCA on the 30 nonmotor symptoms identifies the second-most prominent component as a general
-mood/cognition component, and $k$-means clustering on the 30 symptoms only
-(Figure~\ref{fig:nms30m-k4-heatmap-improved}) divides the 1000 patients into four slightly
-different groups, a mild, average, depression-dominant, and severe group.
+% Indeed, PCA on the 30 nonmotor symptoms identifies the second-most prominent component as a general
+% mood/cognition component, and $k$-means clustering on the 30 symptoms only
+% (Figure~\ref{fig:nms30m-k4-heatmap-improved}) divides the 1000 patients into four slightly
+% different groups, a mild, average, depression-dominant, and severe group.
 
-The Gaussian mixture model identified in Figure~\ref{fig:nms30m-mclust-heatmap-improved} fragments
-the previously-discovered clusters into more groups. Here, a wide variety of specialized subtypes
-of PD are displayed, including insomnia, urinary, motor, nonmotor, and depression-dominant groups,
-as well as the expected mild, average, and severe subtypes. It is likely that the previous analysis
-with only nms\_d\{1-9\} combined most of those specialized groups into the more general
-nonmotor-dominant Subtype 2.
+% The Gaussian mixture model identified in Figure~\ref{fig:nms30m-mclust-heatmap-improved} fragments
+% the previously-discovered clusters into more groups. Here, a wide variety of specialized subtypes
+% of PD are displayed, including insomnia, urinary, motor, nonmotor, and depression-dominant groups,
+% as well as the expected mild, average, and severe subtypes. It is likely that the previous analysis
+% with only nms\_d\{1-9\} combined most of those specialized groups into the more general
+% nonmotor-dominant Subtype 2.
 
@@
 domain 3 consists of 5 separate measures. Thus, any high expression of depressive symptoms is
-magnified in clustering, since the symptoms are highly similar (Figure~\ref{fig:nms30m-colhc}) and
+magnified in clustering, since the symptoms are highly similar (Figure~\ref{fig:hc}) and
 treated with equal weight. Once again reinforcing what was discovered previously, depressive
@@
 \section{Supplementary Material}
+
+Probably the large table goes here?
+
 \section{References}

### 2016-05-09 R/cleanup.R

@@ -5,6 +5,9 @@ library(xtable)
 library(reshape)
 library(plyr)
 library(infotheo)
+library(ggplot2)
+library(reshape2)
+library(scales)
 
 nmsd.symptoms <- c(
   NMS.D.NAMES,
@@ -404,7 +407,219 @@ if (TRUE) {
   dev.off()
 }
 
-# Compute homogeneity/completeness/v-measure ====
-v.measure(nms30.present$Cluster, present$Cluster)
-# It's actually quite poor!
-c(0.3988046,0.4661785,0.4298677)
+# Compute homogeneity/completeness/v-measure and alignment ====
+# cl.s$cluster, cl$cluster
+redux.30 <- cl.s$cluster
+redux.30 <- factor(redux.30, levels = c("3", "2", "1", "4"))
+redux.30 <- revalue(redux.30, c("3" = "1", "1" = "3"))
+
+redux.d <- cl$cluster
+redux.d <- factor(redux.d, levels = c("4", "1", "3", "2"))
+redux.d <- revalue(redux.d, c("4" = "1", "1" = "2", "2" = "4"))
+
+cat("Homogeneity/completeness/V-measure: ",
+    v.measure(nms30.present$Cluster, present$Cluster), "\n")
+cat("Adjusted rand index: ",
+    mclust::adjustedRandIndex(nms30.present$Cluster, present$Cluster), "\n")
+
+# Bar plots
+# Distribution of nmsd cluster assignments for those who are in nms30
+nmsd.per.nms30 <- data.frame(t(rbind(sapply(1:4, function(i) {
+    cat("Cluster ", i, "\n")
+    present.dist <- table(redux.d[redux.30 == i])
+    round(sapply(present.dist, function(n) n / sum(present.dist)), 2)
+  }))),
+  compcluster = c("\n1", "\n2", "\n3", "\n4"),
+  comparison = rep("\nSymptoms cluster\n", 4)
+)
+names(nmsd.per.nms30) <- c("1", "2", "3", "4", "compcluster", "comparison")
+ndpn30 <- melt(nmsd.per.nms30, id = c("compcluster", "comparison"))
+
+# Distribution of nms30 cluster assignments for those who are in nmsd
+nms30.per.nmsd <- data.frame(t(rbind(sapply(1:4, function(i) {
+    cat("Cluster ", i, "\n")
+    present.dist <- table(redux.30[redux.d == i])
+    round(sapply(present.dist, function(n) n / sum(present.dist)), 2)
+  }))),
[…]

### 2016-05-09 R/longitudinal.R

@@ -362,6 +362,8 @@ if (SAVE.LONG.PLOTS) {
   dev.copy(pdf, '../figures/longitudinal/corr-binned.pdf', width=10, height=10)
   dev.off()
 }
+
+# Correlation with PD Duration BINNING ====
 durat.cor <- function(arr) cor(1:nrow(binned), arr)
 correlations <- sapply(binned.no.sd, durat.cor)
 correlations <- correlations[which(names(correlations) != "counts")]  # Lose counts
@@ -385,24 +387,32 @@ ggplot(correlations.df, aes(x=names, y=r, fill=variable)) +
   geom_text(aes(label=round(r, 2)), position=position_dodge(width=0.9), vjust=2 * (correlations.df$r < 0) - .5) +
   scale_y_continuous(limits = c(0, 1)) +
   ylab("r") +
-  xlab("Variable") +
+  xlab("")+ 
   guides(guides(fill=guide_legend(title="Variable Type"))) +
   theme_pub() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
-  ggtitle("Correlation with PD duration")
+  ggtitle("Correlation with PD duration\n")
 # Only save with different names!
 if (SAVE.LONG.PLOTS) {
   ggsave('../figures/longitudinal/pd-durat-cor.pdf', width=14, height=7)
 }
 
 # Durat cor correlation NO BINNING ====
-durat.cor.everything <- function(symptom) cor(everything.wide$durat_pd, everything.wide[[symptom]])
-correlations.everything <- sapply(c(ALL.SYMPTOMS, NMS.30), durat.cor.everything)
+durat.cor.everything <- function(symptom) cor(everything.wide$durat_pd, as.numeric(everything.wide[[symptom]]))
+durat.cor.test <- function(symptom) cor.test(everything.wide$durat_pd, everything.wide[[symptom]])
+correlations.everything <- sapply(names(everything.wide),
+                                  durat.cor.everything)
+# Get rid of cluster, sex, durat_pd
+to.remove <- c("cluster", "sex", "durat_pd", "pdonset")
+correlations.everything <- correlations.everything[!names(correlations.everything) %in% to.remove]
+names(correlations.everything) <- sapply(names(correlations.everything), function(v) c(NMS.D.MAP.PUB.N, NMS.NUM.TO.PUB, MISC.MAP)[[v]])
[…]

### 2016-05-09 R/preprocessing.R

@@ -106,8 +106,12 @@ NMS.D.NAMES <- c(
   "Cardiovascular", "Sleep_fatigue", "Mood_cognition", "Perception_hallucination", "Attention_memory",
   "Gastrointestinal", "Urinary", "Sexual", "Miscellaneous"
 )
+NMS.NUM.TO.PUB <- setNames(sapply(NMS.30.NAMES, function(s) paste(strsplit(s, "-")[[1]][c(1, 3)], collapse = "-")),
+                           NMS.30)
+NMS.D.TO.NUM <- setNames(sapply(1:length(NMS.D), function(i) paste(i, "-", NMS.D[i]), sep = ""), NMS.D)
 NMS.D.MAP.PUB <- setNames(NMS.D.NAMES, NMS.D)
-NMS.D.MAP.PUB.N <- setNames(sapply(1:length(NMS.D.NAMES), function(i) paste(i, NMS.D.NAMES[i], sep = "-")), NMS.D)
+NMS.D.MAP.PUB.N <- setNames(sapply(1:length(NMS.D),
+                                   function(i) paste(i, NMS.D.NAMES[i], sep = "-")), NMS.D)
 NMS.30.MAP.PUB <- setNames(NMS.30.NAMES.PUB, NMS.30)
 NMS.30.LONG.SHORT.MAP <- setNames(NMS.30.NAMES.PUB, NMS.30.NAMES)
 MISC.MAP <- c(
@@ -194,3 +198,4 @@ if (SAVE.PREPROCESSING.PLOTS) {
 
 # Elbow appears around 3 factors
 # Reset par for the rest of the script
+

### 2016-05-21 writeup/manuscript/manuscript.tex
First round of manuscript corrections

@@
 \usepackage{lmodern}
+\usepackage{times}
 
@@
 
-% Tables, Figures
+% Tables
 \usepackage{booktabs}
@@
   \begin{tabular}[#1]{@{}l@{}}#2\end{tabular}}
+% Reduce table spacing
+\setlength{\tabcolsep}{10pt}
+
+% Figures
 \usepackage{float}
+% Reset numbering for supplementary figures
+\usepackage{newfloat}
+\DeclareFloatingEnvironment[name={Supplementary Figure}]{suppfigure}
 \usepackage{caption}
-\setlength{\tabcolsep}{10pt}
 
@@
 \begin{frontmatter}
-\title{Parkinson's disease cluster analysis}
+\title{Parkinson's Disease cluster analysis}
 % \author[c3]{Pablo Martinez-Martin}
@@
 missing measurements in the variables of interest, resulting in 904 remaining patients. All
-symptoms were scaled to $z$-scores ($\mu = 0$, $\sigma = 1$) before clustering, and unscaled
-afterwards for interpretation. Analyses were conducted in R 3.2.4 (www.r-project.org).
+symptoms were standardized before clustering, and unstandardized afterwards for interpretation.
+Analyses were conducted in R version 3.2.4 (www.r-project.org).
 
@@
 $k$-means was used for cluster analysis. We performed two analyses on the
-patients in the dataset: the first on the nine aggregate nonmotor symptom
-domains and the four motor symptoms (``domains clustering''), and the second on the 30 individual
-nonmotor symptoms collected with the four motor symptoms (``symptoms clustering''). Complete-linkage
-hierarchical clustering on the 30 nonmotor symptoms was also performed to observe the
-categorization of symptoms.
+patients in the dataset: the first ``domains clustering'' on the nine aggregate nonmotor symptom
+domains and the four motor symptoms, and the second ``symptoms clustering'' on the 30 individual
+nonmotor symptoms collected with the four motor symptoms. Complete-linkage agglomerative
+hierarchical clustering on the 30 nonmotor symptoms was also performed to how the symptoms group
+together.
 
@@
 
-Various formal measures were used to determine the optimal number of
-clusters for the dataset. First, the optimal $k$ according to the Gap Statistic and the
-1-standard-error method\cite{tibshirani01gap} was $k = 4$ (Figure~\ref{fig:gap}). Second, although
-the globally-optimal clusters reported from stability measures of the clValid R package
-\cite{brock11clvalid} stability were 2, 6, and 8 (Figure~\ref{fig:stability}), there appeared to be
-consensus that $k = 4$ is a locally-optimal number of clusters which supported the gap statistic
-result.
-
-Other cluster validation methods (SSE scree plot, minimum average silhouette width) generally
-suggested 2 clusters. As a consequence, $k = 2, 3, 4$ was tried, where $k = 2, 3$ simply divided
-the data uninformatively into groups with varying levels of overall PD severity. Thus $k = 4$ was
-selected to offer a good blend of model fit, informativeness, and parsimony.
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{gap-statistic.pdf}
-  \caption{Plot of gap statistic versus number of clusters with $k$-means on 100 bootstrapped
-    samples. Error bars are standard error. Per the method described in Tibshirani\cite{tibshirani01gap}, 4 is the smallest $k$ such
-    that $\text{Gap}(k) \geq \text{Gap}(k + 1) - \text{se}_{k + 1}$.}
-  \label{fig:gap}
-\end{figure}
-
-\begin{figure}[h]
-  \centering
-  \includegraphics[width=\linewidth]{stability-measures.pdf}
-  \caption{Average proportion of non-overlap (APN), average distance (AD), average distance
-    between means (ADM), and figure of merit (FOM) stability measures\cite{datta2006methods,
-    yeung2001validating}. Statistics should be minimized; notice the local optima at $k = 4$ for
-  APN, AD, and ADM.}
-  \label{fig:stability}
-\end{figure}
+Various formal measures were used to determine the optimal number of clusters for the dataset. The
+optimal $k$ according to the Gap Statistic and the 1-standard-error method suggested by
+Tibshirani\cite{tibshirani01gap} was $k = 4$ (Supplementary Material Figure~\ref{fig:gap}). Other cluster validation
+methods (within sum of squares (WSS) scree plot, minimum average silhouette width) suggested $k =
+2, 3, 4$, where $k = 2, 3$ simply divided the data uninformatively into groups with varying levels
+of overall PD severity. Thus $k = 4$ was selected to offer a good blend of model fit,
+informativeness, and parsimony.
 
@@
 
-For each symptom in both clusterings, we used one-way ANOVA and $\chi^2$ tests to test the equality
-of symptom means and proportions across the clusters found, using Bonferroni correction for
-multiple testing with corrected $p < 0.05$ considered significant. Differences among pairwise
-clusters were tested post-hoc using Tukey's range test for continuous means, or pairwise $\chi^2$
-tests with Bonferroni correction for proportions.
+For each symptom in both clusterings, we used one-way ANOVA and $\chi^2$ tests to respectively
+check the equality of symptom means and proportions across the clusters found, using Bonferroni
+correction for multiple testing with corrected $p < 0.05$ considered significant. Differences among
+pairwise clusters were tested post-hoc using Tukey's range test for continuous means, or pairwise
+$\chi^2$ tests with Bonferroni correction for proportions.
 
-To compare the clusterings, we computed the adjusted rand index\cite{hubert1985} (ARI) and
-$v$-measure\cite{rosenberg07vmeasure} to evaluate similarity between the two clusterings, and used
-stacked barplots to visualize the alignment of the clusters.
+To compare the clusterings, we visualized cluster alignment with stacked barplots, and computed the adjusted rand index\cite{hubert1985} (ARI) and
+Rosenberg's $v$-measure.\cite{rosenberg07vmeasure} to evaluate similarity between the two
+clusterings. Both measures range from 0 (no similarity) to 1 (identical), and respectively take a
+contingency table and information-theoretic approach to measuring clustering similarity.
 
 Lastly, to explore the relationship between symptom severity and disease duration, we computed the
-linear correlation for each symptom with PD duration and fitted smoothed loess curves to the
+correlation coefficient for each symptom with PD duration and fitted smoothed loess curves to the
 data both globally and for each cluster found in the previous analyses.
@@
 
-$k$-means clustering on the nine nonmotor domains and the four motor symptoms found four clusters
-(Table~\ref{tab:nmd}; boxplots in Figure~\ref{fig:box}). Cluster means for all
-symptoms were found to be statistically different $(p < 0.05 / 13 \approx 0.003)$, with specific
-pairwise differences noted in the table. Additionally, statistics for variables not used in the
-clustering were explored similarly in Table~\ref{tab:nmd_extra}; cluster means for all variables
-were found to be statistically different, except for PD onset.
+Results from the clustering on the nine nonmotor domains and the four motor symptoms, as well as
+additional variables not used in the analysis, are reported in Table~\ref{tab:nmd}, with boxplots
+in Figure~\ref{fig:box}.
+Cluster means for all variables were found to be statistically different, except for PD onset ($p <
+0.05 / 18 \approx 0.003$ after correcting for the comparisons of the 18 variables). Specific
+pairwise differences are noted in the table.
 
@@
 (mild) subtype especially in tremor, but generally expressed significantly higher scores for
-nonmotor symptoms than Clusters 1 and 3, especially in the sleep/fatigue, mood/cognition and
+nonmotor symptoms than Clusters 1 and 3, especially in the sleep/fatigue, mood/apathy and
 miscellaneous domains. This group also had a statistically higher percentage of females than
@@
   \centering
-  \caption{Summaries of domains clustering. Unless otherwise specified, statistics are
-    reported as mean (sd). \textbf{Bold} indicates the highest score for that symptom out of the
-  four clusters.}
+  \caption{Domains clustering summary. Unless otherwise specified, statistics are
+  reported as mean (sd).}
   \label{tab:nmd}
   \begin{threeparttable}
-  \begin{tabular}{l r r r r}
+    \small
+    \begin{tabular}{lrllll}
     \toprule
-    Cluster & 1 & 2 & 3 & 4 \\
-    $n$ & 406 (45\%) & 189 (21\%) & 221 (24\%) & 88 (10\%) \\
+    & Cluster & 1 & 2 & 3 & 4 \\
+    & $n$ & 406 (45\%) & 189 (21\%) & 221 (24\%) & 88 (10\%) \\
     \midrule
-    \textbf{Nonmotor} (Domains) & & & & \\
-    Cardiovascular (0--24) & 0.7 (1.6)\tnote{24} & 2.3 (2.9)\tnote{134} & 1.1 (2.0)\tnote{24} &
-    \textbf{7.0} (6.0)\tnote{234} \\
-    Sleep/fatigue (0--48) & 4.3 (4.8)\tnote{234} & 14.9 (8.2)\tnote{134} & 6.9 (6.3)\tnote{124}
-    & \textbf{21.0} (10.0)\tnote{123} \\
-    Mood/cognition (0--60) & 3.1 (4.5)\tnote{234} & 16.1 (14.0)\tnote{134} & 6.6
-    (8.1)\tnote{124} & \textbf{23.4} (14.3)\tnote{123} \\
-    Perception/hallucination (0--36) & 0.5 (1.7)\tnote{24} & 1.9 (3.5)\tnote{134} & 0.8
-    (2.0)\tnote{24} & \textbf{8.6} (6.9)\tnote{123} \\
-    Attention/memory (0--36) & 2.8 (4.2)\tnote{24} & 8.9 (8.1)\tnote{134} & 3.2 (4.4)\tnote{24} &
-    \textbf{15.6} (11.1)\tnote{123} \\
-    Gastrointestinal (0--36) & 2.8 (3.9)\tnote{234} & 8.8 (7.1)\tnote{134} & 4.1
-    (4.7)\tnote{124} & \textbf{14.6} (9.6)\tnote{123} \\
-    Urinary (0--36) & 4.5 (5.9)\tnote{234} & 12.4 (9.7)\tnote{134} & 6.1 (6.4)\tnote{124} &
-    \textbf{20.3} (9.9)\tnote{123} \\
-    Sexual function (0--24) & 1.5 (3.1)\tnote{24} & 6.4 (7.4)\tnote{134} & 2.5 (4.1)\tnote{24} &
-    \textbf{9.0} (9.7)\tnote{123} \\
-    Miscellaneous (0--48) & 3.9 (4.7)\tnote{24} & 13.2 (8.6)\tnote{13} & 5.2 (5.8)\tnote{24} &
-    \textbf{14.0} (9.5)\tnote{13} \\
+    Nonmotor domains & Cardiovascular & 0.7 (1.6)\tnote{24} & 2.3 (2.9)\tnote{134} & 1.1 (2.0)\tnote{24} &
+    7.0 (6.0)\tnote{123} \\
+    & Sleep/fatigue & 4.3 (4.8)\tnote{234} & 14.9 (8.2)\tnote{134} & 6.9 (6.3)\tnote{124}
+    & 21.0 (10.0)\tnote{123} \\
+  & Mood/apathy & 3.1 (4.5)\tnote{234} & 16.1 (14.0)\tnote{134} & 6.6
+    (8.1)\tnote{124} & 23.4 (14.3)\tnote{123} \\
+    & Perception/hallucination & 0.5 (1.7)\tnote{24} & 1.9 (3.5)\tnote{134} & 0.8
+    (2.0)\tnote{24} & 8.6 (6.9)\tnote{123} \\
+    & Attention/memory & 2.8 (4.2)\tnote{24} & 8.9 (8.1)\tnote{134} & 3.2 (4.4)\tnote{24} &
+    15.6 (11.1)\tnote{123} \\
+    & Gastrointestinal & 2.8 (3.9)\tnote{234} & 8.8 (7.1)\tnote{134} & 4.1
+    (4.7)\tnote{124} & 14.6 (9.6)\tnote{123} \\
+    & Urinary & 4.5 (5.9)\tnote{234} & 12.4 (9.7)\tnote{134} & 6.1 (6.4)\tnote{124} &
+    20.3 (9.9)\tnote{123} \\
+    & Sexual function & 1.5 (3.1)\tnote{24} & 6.4 (7.4)\tnote{134} & 2.5 (4.1)\tnote{24} &
+    9.0 (9.7)\tnote{123} \\
+    & Miscellaneous & 3.9 (4.7)\tnote{24} & 13.2 (8.6)\tnote{13} & 5.2 (5.8)\tnote{24} &
+    14.0 (9.5)\tnote{13} \\
     \midrule
-    \textbf{Motor} & & & & \\
-    Tremor & 1.9 (1.8)\tnote{34} & 1.6 (1.9)\tnote{34} & \textbf{4.3} (2.8)\tnote{124} & 3.5
+    Motor symptoms & Tremor & 1.9 (1.8)\tnote{34} & 1.6 (1.9)\tnote{34} & 4.3 (2.8)\tnote{124} & 3.5
     (3.8)\tnote{123} \\
-    Bradykinesia & 1.5 (0.9)\tnote{234} & 2.1 (1.1)\tnote{134} & 3.6 (0.9)\tnote{124} &
-    \textbf{4.0} (1.5)\tnote{123} \\
-    Rigidity & 1.5 (0.9)\tnote{234} & 1.8 (1.1)\tnote{134} & 3.3 (0.9)\tnote{124} &
-    \textbf{3.8} (1.4)\tnote{123} \\
-    Axial & 1.8 (1.6)\tnote{234} & 3.3 (2.1)\tnote{134} & 4.3 (2.4)\tnote{124} &
-    \textbf{7.4} (2.9)\tnote{123} \\
+    & Bradykinesia & 1.5 (0.9)\tnote{234} & 2.1 (1.1)\tnote{134} & 3.6 (0.9)\tnote{124} &
+    4.0 (1.5)\tnote{123} \\
+    & Rigidity & 1.5 (0.9)\tnote{234} & 1.8 (1.1)\tnote{134} & 3.3 (0.9)\tnote{124} &
+    3.8 (1.4)\tnote{123} \\
+    & Axial & 1.8 (1.6)\tnote{234} & 3.3 (2.1)\tnote{134} & 4.3 (2.4)\tnote{124} &
+    7.4 (2.9)\tnote{123} \\
+    \midrule
+    Variables not & Sex (\% male) & 63 & 53\tnote{3} & 71\tnote{2} & 58 \\
+    used in analysis & CISI total & 5.7 (3.2)\tnote{234} & 9.2 (3.9)\tnote{14} & 9.7
+    (3.5)\tnote{14} & 14.8 (5.1)\tnote{123} \\
+    & Age (34--89) & 62.7 (9.7)\tnote{34} & 64 (9.0)\tnote{4} & 65 (10.2)\tnote{14} & 70.5 (8.9)\tnote{123} \\
+    & PD onset & 56 (10.6) & 55.2 (10.1) & 56.9 (11.1) & 58.2 (11.4) \\
+    & PD duration & 6.7 (4.8)\tnote{234} & 8.7 (5.7)\tnote{14} & 8 (5.5)\tnote{14} &
+    12.3
+    (8.1)\tnote{123} \\
     \bottomrule
@@
     \small
-    \item[1] Significant difference with cluster 1 (adjusted $p < 0.05$)
-    \item[2] Significant difference with cluster 2 (adjusted $p < 0.05$)
-    \item[3] Significant difference with cluster 3 (adjusted $p < 0.05$)
-    \item[4] Significant difference with cluster 4 (adjusted $p < 0.05$)
+    \item[1] Significant difference with cluster 1 (corrected $p < 0.05$)
+    \item[2] Significant difference with cluster 2 (corrected $p < 0.05$)
+    \item[3] Significant difference with cluster 3 (corrected $p < 0.05$)
+    \item[4] Significant difference with cluster 4 (corrected $p < 0.05$)
   \end{tablenotes}
@@
   }
-  \caption{Boxplot cluster summaries \\ for domains clustering.}
+  \caption{Boxplots for domains clustering.}
   \label{fig:box}
 \end{figure*}
-
-\begin{table*}[t]
-  \centering
-  \caption{Cluster statistics for variables not used in the analysis}
-  \label{tab:nmd_extra}
-  \begin{threeparttable}
-  \begin{tabular}{l r r r r}
-    \toprule
-    Cluster & 1 & 2 & 3 & 4 \\
-    \midrule
-      Sex (\% male) & 63 & 53\tnote{3} & 71\tnote{2} & 58 \\
-      Age & 62.7 (9.7)\tnote{34} & 64 (9.0)\tnote{4} & 65 (10.2)\tnote{14} & 70.5 (8.9)\tnote{123} \\
-      PD\_Onset & 56 (10.6) & 55.2 (10.1) & 56.9 (11.1) & 58.2 (11.4) \\
-      PD\_Duration & 6.7 (4.8)\tnote{234} & 8.7 (5.7)\tnote{14} & 8 (5.5)\tnote{14} & 12.3
-      (8.1)\tnote{123} \\
-      CISI\_Total & 5.7 (3.2)\tnote{234} & 9.2 (3.9)\tnote{14} & 9.7 (3.5)\tnote{14} & 14.8
-      (5.1)\tnote{123} \\
-    \bottomrule
-  \end{tabular}
-  \begin{tablenotes}
-    \small
-    \item[1] Significant difference with cluster 1 (adjusted $p < 0.05$)
-    \item[2] Significant difference with cluster 2 (adjusted $p < 0.05$)
-    \item[3] Significant difference with cluster 3 (adjusted $p < 0.05$)
-    \item[4] Significant difference with cluster 4 (adjusted $p < 0.05$)
-  \end{tablenotes}
-  \end{threeparttable}
-\end{table*}
 
@@
 first clustering (Table~\ref{tab:nms30}, heatmap in Figure~\ref{fig:nms30-hm}). Means of all
-symptoms were found to differ across clusters except for PD\_Onset and tremor.
+symptoms were found to differ across clusters except for PD onset and tremor.
 
 Clusters 1 ($n = 509$) and 4 ($n = 49$), the mild and severe subtypes of the previous clustering,
-retained their characteristics here, while clusters 2 ($n = 97$) and 3 ($n = 249$) were different.
-2 was a mood/cognition-dominant cluster, with a relatively higher proportion of female patients and
-additionally a more severe incidence of insomnia. 3 was not characterized by any especially high
+retained their characteristics here, while clusters 2 ($n = 97$) and 3 ($n = 249$) were
+characterized differently.
+Cluster 2 was a mood/apathy-dominant cluster, with a relatively higher proportion of female patients and
+additionally a more severe incidence of insomnia. Cluster 3 was not characterized by any especially high
 symptoms and was thus classified as a group of patients with average severity.
@@
 
-The two clusterings are somewhat similar ($\text{ARI} = 0.46$; $\text{$v$-measure} = 0.43$).
-Alignment of the clusters is visualized in Figure~\ref{fig:align}. While clusters 1 and 4 of both
-clusterings align well with each other, main differences in the clusters occur in the mixing of
-clusters 2 and 3 of both clusterings; for example, domains cluster 3 (motor-dominant) is somewhat
-evenly split among symptoms clusters 1 (mild) and 3 (average).
+ The two clusterings were somewhat similar ($\text{ARI} = 0.46$; $\text{$v$-measure} = 0.43$).  To
+ help compare the alignment of the two clusterings, we denoted the four clusters in the domains
+ clustering as $D_1, D_2, D_3, D_4$, and the clusters in the symptoms clustering as $S_1, S_2, S_3,
+ S_4$.
+Alignment of the clusters is visualized in Figure~\ref{fig:align}. While clusters $D_1/S_1$ and
+$D_4/S_4$
+aligned well with each other, main differences in the clusterings occurred in the mixing of
+clusters 2 and 3; for example, $D_3$ (motor-dominant) was somewhat
+evenly split among $S_1$ (mild) and $S_3$ (average).
 
@@
   \includegraphics[width=\linewidth]{cluster-alignment.pdf}
-  \caption{Bar chart indicating the distribution of opposite cluster membership for the patients of
+  \caption{Stacked barplots indicating the distribution of opposite cluster membership for the patients of
   a given cluster. For example, the bar furthest left indicates that close to 100\% of patients in
@@
   \caption{Summaries of clustering on individual nonmotor domains. Unless otherwise specified,
-    statistics are reported as mean (sd). \textbf{Bold} indicates the highest score for that
-  symptom out of the four clusters.}
+  statistics are reported as mean (sd).}
   \label{tab:nms30}
@@
   1.\ Cardiovascular &
-  Lightheadedness & 0.6 (1.3)\tnote{234} & 2.4 (3.0)\tnote{14} & 2.1 (2.9)\tnote{14} & \textbf{5.2}
+  Lightheadedness & 0.6 (1.3)\tnote{234} & 2.4 (3.0)\tnote{14} & 2.1 (2.9)\tnote{14} & 5.2
   (4.2)\tnote{123} \\
-  &Fainting & 0.1 (0.7)\tnote{24} & 0.5 (1.4)\tnote{134} & 0.2 (0.6)\tnote{24} & \textbf{2.4}
+  &Fainting & 0.1 (0.7)\tnote{24} & 0.5 (1.4)\tnote{134} & 0.2 (0.6)\tnote{24} & 2.4
   (3.5)\tnote{123} \\
@@
   2.\ Sleep/fatigue &
-  Drowsiness & 1.0 (1.9)\tnote{234} & 2.8 (3.3)\tnote{14} & 2.5 (3.0)\tnote{14} & \textbf{7.0}
+  Drowsiness & 1.0 (1.9)\tnote{234} & 2.8 (3.3)\tnote{14} & 2.5 (3.0)\tnote{14} & 7.0
   (3.8)\tnote{123} \\
-  &Fatigue & 1.3 (2.0)\tnote{234} & 6.3 (4.2)\tnote{13} & 4.8 (3.9)\tnote{124} & \textbf{7.3} (3.5)\tnote{13} \\
-  &Insomnia & 1.2 (2.6)\tnote{234} & \textbf{5.5} (4.8)\tnote{13} & 3.2 (3.8)\tnote{12} & 4.2
+  &Fatigue & 1.3 (2.0)\tnote{234} & 6.3 (4.2)\tnote{13} & 4.8 (3.9)\tnote{124} & 7.3 (3.5)\tnote{13} \\
+  &Insomnia & 1.2 (2.6)\tnote{234} & 5.5 (4.8)\tnote{13} & 3.2 (3.8)\tnote{12} & 4.2
   (3.9)\tnote{1} \\
-  &RLS & 0.7 (1.6)\tnote{234} & 2.3 (3.5)\tnote{14} & 1.9 (3.2)\tnote{14} & \textbf{3.9}
+  &RLS & 0.7 (1.6)\tnote{234} & 2.3 (3.5)\tnote{14} & 1.9 (3.2)\tnote{14} & 3.9
   (3.9)\tnote{123} \\
   \midrule
-  3.\ Mood/cognition &
-  Loss\_interest & 0.4 (1.1)\tnote{234} & \textbf{6.1} (3.8)\tnote{134} & 1.0 (1.8)\tnote{124} & 4.6 (3.7)\tnote{123} \\
-  &Loss\_activities & 0.7 (1.6)\tnote{234} & \textbf{7.1} (4.1)\tnote{134} & 1.7 (2.4)\tnote{124} &
+  3.\ Mood/apathy &
+  Loss\_interest & 0.4 (1.1)\tnote{234} & 6.1 (3.8)\tnote{134} & 1.0 (1.8)\tnote{124} & 4.6 (3.7)\tnote{123} \\
+  &Loss\_activities & 0.7 (1.6)\tnote{234} & 7.1 (4.1)\tnote{134} & 1.7 (2.4)\tnote{124} &
   5.4 (4.0)\tnote{123} \\
-  &Anxiety & 0.9 (1.8)\tnote{234} & \textbf{6.6} (4.3)\tnote{134} & 2.0 (2.8)\tnote{124} & 4.6 (3.9)\tnote{123} \\
-  &Depression & 0.8 (1.6)\tnote{234} & \textbf{7.8} (3.7)\tnote{134} & 2.3 (3.0)\tnote{124} & 4.3 (3.7)\tnote{123} \\
-  &Flat\_affect & 0.4 (1.1)\tnote{234} & \textbf{4.9} (4.3)\tnote{134} & 0.9 (1.9)\tnote{124} & 2.7 (3.0)\tnote{123} \\
-  &Loss\_pleasure & 0.4 (1.2)\tnote{234} & \textbf{6.8} (4.1)\tnote{134} & 1.1 (2.1)\tnote{124} & 3.4 (3.3)\tnote{123} \\
+  &Anxiety & 0.9 (1.8)\tnote{234} & 6.6 (4.3)\tnote{134} & 2.0 (2.8)\tnote{124} & 4.6 (3.9)\tnote{123} \\
+  &Depression & 0.8 (1.6)\tnote{234} & 7.8 (3.7)\tnote{134} & 2.3 (3.0)\tnote{124} & 4.3 (3.7)\tnote{123} \\
+  &Flat\_affect & 0.4 (1.1)\tnote{234} & 4.9 (4.3)\tnote{134} & 0.9 (1.9)\tnote{124} & 2.7 (3.0)\tnote{123} \\
+  &Loss\_pleasure & 0.4 (1.2)\tnote{234} & 6.8 (4.1)\tnote{134} & 1.1 (2.1)\tnote{124} & 3.4 (3.3)\tnote{123} \\
   \midrule
   % This one is spread across multiple cells
-  4.\ Perception/ & Hallucination & 0.2 (1.0)\tnote{234} & 0.8 (1.9)\tnote{14} & 0.6 (1.6)\tnote{14} & \textbf{4.6} (3.6)\tnote{123} \\
+  4.\ Perception/ & Hallucination & 0.2 (1.0)\tnote{234} & 0.8 (1.9)\tnote{14} & 0.6 (1.6)\tnote{14} & 4.6 (3.6)\tnote{123} \\
   hallucination &Delusion & 0.1 (0.7)\tnote{24} & 1.5 (3.1)\tnote{124} & 0.2 (1.0)\tnote{24} &
-  \textbf{3.1} (3.8)\tnote{123} \\
-  &Diplopia & 0.2 (0.7)\tnote{234} & 1.0 (2.5)\tnote{14} & 0.6 (1.4)\tnote{14} & \textbf{4.4}
+  3.1 (3.8)\tnote{123} \\
+  &Diplopia & 0.2 (0.7)\tnote{234} & 1.0 (2.5)\tnote{14} & 0.6 (1.4)\tnote{14} & 4.4
   (4.3)\tnote{123} \\
   \midrule
-  5.\ Attention/ & Loss\_concentration & 0.9 (1.8)\tnote{234} & 4.0 (3.6)\tnote{134} & 2.5 (3.1)\tnote{124} & \textbf{6.6} (3.9)\tnote{123} \\
-  memory &Forget\_explicit & 0.9 (1.6)\tnote{234} & 3.8 (3.9)\tnote{134} & 2.2 (2.9)\tnote{124} & \textbf{6.7} (3.6)\tnote{123} \\
-  &Forget\_implicit & 0.7 (1.6)\tnote{234} & 3.1 (3.7)\tnote{134} & 1.8 (2.6)\tnote{124} & \textbf{6.7} (4.0)\tnote{123} \\
+  5.\ Attention/ & Loss\_concentration & 0.9 (1.8)\tnote{234} & 4.0 (3.6)\tnote{134} & 2.5 (3.1)\tnote{124} & 6.6 (3.9)\tnote{123} \\
+  memory &Forget\_explicit & 0.9 (1.6)\tnote{234} & 3.8 (3.9)\tnote{134} & 2.2 (2.9)\tnote{124} & 6.7 (3.6)\tnote{123} \\
+  &Forget\_implicit & 0.7 (1.6)\tnote{234} & 3.1 (3.7)\tnote{134} & 1.8 (2.6)\tnote{124} & 6.7 (4.0)\tnote{123} \\
   \midrule
   6.\ Gastrointestinal &
-  Drooling & 0.7 (1.6)\tnote{234} & 3.0 (4.1)\tnote{14} & 3.0 (3.8)\tnote{14} & \textbf{6.1} (4.4)\tnote{123} \\
-  &Swallowing & 0.4 (1.2)\tnote{234} & 1.2 (2.1)\tnote{14} & 1.7 (2.8)\tnote{14} & \textbf{4.5} (3.9)\tnote{123} \\
-  &Constipation & 1.6 (2.9)\tnote{234} & 3.6 (4.3)\tnote{14} & 3.3 (4.1)\tnote{14} & \textbf{7.0} (5.1)\tnote{123} \\
+  Drooling & 0.7 (1.6)\tnote{234} & 3.0 (4.1)\tnote{14} & 3.0 (3.8)\tnote{14} & 6.1 (4.4)\tnote{123} \\
+  &Swallowing & 0.4 (1.2)\tnote{234} & 1.2 (2.1)\tnote{14} & 1.7 (2.8)\tnote{14} & 4.5 (3.9)\tnote{123} \\
+  &Constipation & 1.6 (2.9)\tnote{234} & 3.6 (4.3)\tnote{14} & 3.3 (4.1)\tnote{14} & 7.0 (5.1)\tnote{123} \\
   \midrule
   7.\ Urinary &
-  Urinary\_urgency & 1.0 (1.8)\tnote{234} & 3.5 (4.1)\tnote{14} & 3.8 (3.8)\tnote{14} & \textbf{7.8} (4.0)\tnote{123} \\
-  &Urinary\_frequency & 0.9 (1.9)\tnote{234} & 4.0 (4.1)\tnote{14} & 3.8 (4.0)\tnote{14} & \textbf{6.5} (4.0)\tnote{123} \\
-  &Nocturia & 1.7 (2.3)\tnote{234} & 5.0 (4.3)\tnote{14} & 5.1 (4.1)\tnote{14} & \textbf{7.7} (3.9)\tnote{123} \\
+  Urinary\_urgency & 1.0 (1.8)\tnote{234} & 3.5 (4.1)\tnote{14} & 3.8 (3.8)\tnote{14} & 7.8 (4.0)\tnote{123} \\
+  &Urinary\_frequency & 0.9 (1.9)\tnote{234} & 4.0 (4.1)\tnote{14} & 3.8 (4.0)\tnote{14} & 6.5 (4.0)\tnote{123} \\
+  &Nocturia & 1.7 (2.3)\tnote{234} & 5.0 (4.3)\tnote{14} & 5.1 (4.1)\tnote{14} & 7.7 (3.9)\tnote{123} \\
   \midrule
   8.\ Sexual &
-  Sex\_drive & 0.8 (1.9)\tnote{234} & 4.5 (4.9)\tnote{13} & 2.3 (3.6)\tnote{124} & \textbf{5.2} (5.4)\tnote{13} \\
+  Sex\_drive & 0.8 (1.9)\tnote{234} & 4.5 (4.9)\tnote{13} & 2.3 (3.6)\tnote{124} & 5.2 (5.4)\tnote{13} \\
   &Sex\_dysfunction & 0.8 (2.2)\tnote{234} & 3.2 (4.7)\tnote{1} & 2.4 (3.8)\tnote{14} &
-  \textbf{4.6} (5.4)\tnote{13} \\
+  4.6 (5.4)\tnote{13} \\
   \midrule
   9.\ Miscellaneous &
-  Unexplained\_pain & 0.7 (1.9)\tnote{234} & 2.5 (3.7)\tnote{14} & 3.0 (4.1)\tnote{14} & \textbf{4.2} (4.3)\tnote{123} \\
+  Unexplained\_pain & 0.7 (1.9)\tnote{234} & 2.5 (3.7)\tnote{14} & 3.0 (4.1)\tnote{14} & 4.2 (4.3)\tnote{123} \\
   &Gustation\_olfaction & 1.5 (2.9)\tnote{234} & 3.6 (4.3)\tnote{1} & 3.2 (3.9)\tnote{14} &
-  \textbf{4.7} (4.8)\tnote{13} \\
-  &Weight\_change & 0.9 (1.7)\tnote{234} & 2.5 (3.4)\tnote{14} & 1.9 (3.0)\tnote{14} & \textbf{3.9} (4.2)\tnote{123} \\
-  &Sweating & 0.7 (1.8)\tnote{234} & 2.2 (3.5)\tnote{1} & 2.8 (4.1)\tnote{1} & \textbf{3.2}
+  4.7 (4.8)\tnote{13} \\
+  &Weight\_change & 0.9 (1.7)\tnote{234} & 2.5 (3.4)\tnote{14} & 1.9 (3.0)\tnote{14} & 3.9 (4.2)\tnote{123} \\
+  &Sweating & 0.7 (1.8)\tnote{234} & 2.2 (3.5)\tnote{1} & 2.8 (4.1)\tnote{1} & 3.2
   (3.8)\tnote{1} \\
   \midrule
-  Motor &
-  Tremor & 2.5 (2.3) & 2.1 (2.4) & 2.6 (2.6) & \textbf{4.1} (4.3) \\
-  &Bradykinesia & 2.0 (1.2)\tnote{234} & 2.8 (1.4)\tnote{14} & 2.8 (1.4)\tnote{14} & \textbf{3.9} (1.8)\tnote{123} \\
-  &Rigidity & 1.9 (1.2)\tnote{234} & 2.4 (1.3)\tnote{14} & 2.6 (1.4)\tnote{14} & \textbf{3.6} (1.6)\tnote{123} \\
-  &Axial & 2.2 (1.8)\tnote{234} & 4.4 (2.8)\tnote{14} & 4.2 (2.7)\tnote{14} & \textbf{7.2} (3.2)\tnote{123} \\
+  Motor symptoms &
+  Tremor & 2.5 (2.3) & 2.1 (2.4) & 2.6 (2.6) & 4.1 (4.3) \\
+  &Bradykinesia & 2.0 (1.2)\tnote{234} & 2.8 (1.4)\tnote{14} & 2.8 (1.4)\tnote{14} & 3.9 (1.8)\tnote{123} \\
+  &Rigidity & 1.9 (1.2)\tnote{234} & 2.4 (1.3)\tnote{14} & 2.6 (1.4)\tnote{14} & 3.6 (1.6)\tnote{123} \\
+  &Axial & 2.2 (1.8)\tnote{234} & 4.4 (2.8)\tnote{14} & 4.2 (2.7)\tnote{14} & 7.2 (3.2)\tnote{123} \\
   \midrule
@@
   used in analysis & Sex (\% Male) & 65\tnote{2} & 51\tnote{1} & 60 & 63 \\
-  & PD\_Onset & 55.9 (10.8) & 56.2 (10.4) & 56.7 (10.5) & 57.8 (11.0) \\
-  & PD\_Duration & 6.8 (4.8)\tnote{34} & 8.3 (5.3)\tnote{4} & 9.3 (6.3)\tnote{14} & 13.8
+  & PD onset & 55.9 (10.8) & 56.2 (10.4) & 56.7 (10.5) & 57.8 (11.0) \\
+  & PD duration & 6.8 (4.8)\tnote{34} & 8.3 (5.3)\tnote{4} & 9.3 (6.3)\tnote{14} & 13.8
   (8.2)\tnote{123} \\
-  & CISI\_Total & 6.3 (3.4)\tnote{234} & 10.8 (4.7)\tnote{14} & 10 (4.0)\tnote{14} & 15.1 (5.7)\tnote{123} \\
+  & CISI total & 6.3 (3.4)\tnote{234} & 10.8 (4.7)\tnote{14} & 10 (4.0)\tnote{14} & 15.1 (5.7)\tnote{123} \\
    \bottomrule
@@
   \caption{Hierarchical clustering of motor (blue) and nonmotor (black) symptoms. Symptoms are
-  labeled with their name and corresponding domain number. Dendrogram colored with 4 clusters.}
+  labeled with their name and corresponding domain number. Dendrogram colored with 5 clusters.}
   \label{fig:hc}
@@
 dendrogram in Figure~\ref{fig:hc}. Predictably, symptoms in the same nonmotor domains tended to
-cluster together, with the exceptions of diplopia (domain 4), drowsiness (domain 2), and the
-symptoms in domain 9 (miscellaneous). Notably, tremor was the most dissimilar symptom, occupying a
-single branch at the top of the tree.
+cluster together, with the exceptions of diplopia (domain 4) and drowsiness (domain 2), which were
+grouped with the attention/memory symptoms of domain 5.  Notably, tremor was the most dissimilar
+symptom, occupying a single branch at the top of the tree.
 
@@
 
-Most variables had little to no correlation with PD duration (Figure~\ref{fig:corr}). Variables
+Most variables had little to no correlation with PD duration (Supplementary Material Figure~\ref{fig:corr}). Variables
 with the highest correlation include symptoms urinary frequency, swallowing, and drooling; domains
 miscellaneous, urinary, and gastrointestinal; and total CISI score. Scatterplots for CISI Total,
-Tremor, Anxiety, and Depression appear in Figure~\ref{fig:long4}.
-
-\begin{figure*}[p]
-  \centering
-  \includegraphics[width=\linewidth]{cor-unbinned.pdf}
-  \caption{Linear correlation of each symptom with PD duration.}
-  \label{fig:corr}
-\end{figure*}
-
-\begin{figure*}[p]
-  \centering
-  \includegraphics[width=0.8\linewidth]{long4-d.pdf}
-  \caption{Selected symptoms for 901 patients plotted against PD duration. Smoothed loess curves for each cluster
-  are drawn in their respective colors; the black curve is the curve for the entire population.}
-  \label{fig:long4}
-\end{figure*}
+Tremor, Anxiety, and Depression appear in Supplementary Material Figure~\ref{fig:long4}.
 
@@
 Domains clustering reveals clusters that confirm previous
-findings in the field, mainly van Rooden et al.\cite{vanrooden10} and the identification of four
+findings in the field, mainly van Rooden et al.\cite{vanrooden11} and the identification of four
 subtypes of Parkinson's disease: mild, nonmotor-dominant, motor-dominant, and severe. van
@@
 
-Overall, little information was found in pdonset,
-durat\_pd, or current age. Mean ages were similar for clusters 1, 2, and 3 ($p >
-0.05$), but different for the severe cluster 4, intuitively since cluster 4 represents a more
-advanced stage of the disease.  Specifically, clusters 1 and 4 seem to be phenotypically quite
-similar, except at different stages of disease progression, given cluster 4's higher age and
-durat\_pd scores.
+Overall, little information was found in PD onset, PD duration, or current age. Mean ages were
+similar for clusters 1, 2, and 3 ($p > 0.05$), but different for the severe cluster 4, intuitively
+since cluster 4 represents a more advanced stage of the disease.  Specifically, clusters 1 and 4
+seem to be phenotypically quite similar, except at different stages of disease progression, given
+cluster 4's higher age and PD duration scores.
 
@@
 Cluster 2 is dominated by a high severity of nonmotor domains, especially Sleep/Fatigue,
-Mood/Cognition, Urinary, and Miscellaneous.  Cluster 3, however, is dominated by a high prevalence
+Mood/apathy, Urinary, and Miscellaneous.  Cluster 3, however, is dominated by a high prevalence
 of motor symptoms, where most motor symptoms are similar to the mild cluster 1. Of note is that the
 tremor population mean is the highest cluster mean, even higher than the severe subtype 4. This
-motor-dominant cluster may thus overlap with Ma's tremor dominant/slow progression
-cluster\cite{ma15}.
+motor-dominant cluster may thus overlap with Ma et al.'s tremor dominant/slow progression
+cluster.\cite{ma15}
 
-Generally, given stable pdonset scores and predictably increasing durat\_pd
-scores for clusters 1 and 4, Ma et al's rapid disease
-progression/late onset and tremor dominant/slow progression clusters
-\cite{ma15} were mostly not found in this dataset, save for the tremor-dominant
-motor cluster.
+Generally, given stable PD onset scores and predictably increasing PD duration
+scores for clusters 1 and 4, Ma et al.'s rapid disease
+progression/late onset and tremor dominant/slow progression clusters were mostly not found in this
+dataset, save for the tremor-dominant motor cluster.
 
@@
 According to Figure~\ref{fig:corr}, most symptoms are uncorrelated with PD duration, especially
-mood/cognition symptoms (anxiety, depression) and tremor.
+Mood/apathy symptoms (anxiety, depression) and tremor.
 The differences in disease progression for each cluster can be seen by the corresponding graphs in
@@
 
-It's intuitive that a Depression-Dominant group emerges when clustering on nms\{1-30\}, since
-domain 3 consists of 5 separate measures. Thus, any high expression of depressive symptoms is
-magnified in clustering, since the symptoms are highly similar (Figure~\ref{fig:hc}) and
-treated with equal weight. Once again reinforcing what was discovered previously, depressive
+It's intuitive that a Depression-Dominant group emerges from the symptoms clustering, since the
+Mood/apathy domain consists of 5 separate measures. Thus, any high expression of depressive
+symptoms is magnified in clustering, since the symptoms are highly similar (Figure~\ref{fig:hc})
+and treated with equal weight. Once again reinforcing what was discovered previously, depressive
 symptoms have been shown to be very important in determining subtypes of PD.
@@
 
-Probably the large table goes here?
+\begin{suppfigure}[h]
+  \centering
+  \includegraphics[width=\linewidth]{gap-statistic.pdf}
+  \caption{Plot of the gap statistic $\text{Gap}(k)$ versus number of clusters with $k$-means on
+  500 bootstrapped samples of the domains clustering. Error bars represent $\pm$ 1 standard error
+  (se). Per the method described in Tibshirani,\cite{tibshirani01gap} the optimal number of
+  clusters is the smallest $k$ such that $\text{Gap}(k) \geq \text{Gap}(k + 1) - \text{se}_{k +
+  1}$. In this case, $k = 4$.}
+  \label{fig:gap}
+\end{suppfigure}
+
+\begin{suppfigure*}[p]
+  \centering
+  \includegraphics[width=\linewidth]{cor-unbinned.pdf}
+  \caption{Correlation of each symptom with PD duration.}
+  \label{fig:corr}
+\end{suppfigure*}
+
+\begin{suppfigure*}[p]
+  \centering
+  \includegraphics[width=0.8\linewidth]{long4-d.pdf}
+  \caption{Selected symptoms for 901 patients plotted against PD duration. Smoothed loess curves for each cluster
+  are drawn in their respective colors. The black curve is the curve for the entire population, and
+the global mean score is marked with a dotted line.}
+  \label{fig:long4}
+\end{suppfigure*}
 

### 2016-05-21 R/cleanup.R

@@ -3,9 +3,10 @@
 library(doBy)
 library(xtable)
 library(reshape)
-library(plyr)
-library(infotheo)
-library(ggplot2)
+library(grid)
+library(gridExtra)
+library(plyr) library(infotheo) library(ggplot2)
+
 library(reshape2)
 library(scales)
 
@@ -126,7 +127,7 @@ hm.m$colDendrogram %>%
 dev.copy(pdf, "../figures/nms30m-colhc-pub.pdf", width = 15, height = 7)
 dev.off()
 
-# Pub-ready boxplot graph ====
+# Publication-ready boxplots ====
 dev.off()
 clus <- clusters.raw.long[["4"]]
 # Get rid of extra
@@ -140,7 +141,8 @@ clus.pub$Type <- ""
 clus.pub[clus.pub$variable %in%
            sapply(NMS.D, function(s) paste("\n", PUB.MAP.N[as.character(s)][[1]], "\n", sep = "")), ]$Type <- "Nonmotor (analyzed)"
 clus.pub[clus.pub$variable %in% factor(c("\nAxial\n", "\nRigidity\n", "\nBradykinesia\n", "\nTremor\n")), ]$Type <- "Motor (analyzed)"
-clus.pub[!(clus.pub$Type %in% c("Nonmotor (analyzed)", "Motor (analyzed)")), ]$Type <- "Other"
+clus.pub[!(clus.pub$Type %in% c("Nonmotor (analyzed)", "Motor (analyzed)")), ]$Type <- "Other (not analyzed)"
+clus.pub$Type <- factor(clus.pub$Type, levels = c('Nonmotor (analyzed)', 'Motor (analyzed)', 'Other (not analyzed)'))
 p <- ggplot(clus.pub, aes(x = factor(cluster), y = measurement, fill = factor(cluster))) +
   geom_boxplot() +
   guides(fill = FALSE) +
@@ -159,7 +161,7 @@ dummy <- ggplot(clus.pub, aes(x = factor(cluster), y = measurement)) +
         legend.text = element_text(size = 14),
         legend.title = element_text(size = 16)) +
   labs(fill = "Variable Type") +
-  scale_fill_manual(values = c("#d7f6ff", "#d2c5ff", "#fed4f4")) +
+  scale_fill_manual(values = brewer.pal(4, "Pastel2")[c(1, 3, 4)]) +
   theme(legend.position = c(0.5, 0.1))
 dummy
[…]

### 2016-05-21 R/kmeans-dtree.R

@@ -115,14 +115,26 @@ plot(pam(raw.filtered, pamk.best$nc))
 
 # GAP STATISTIC ESTIMATION ====
 set.seed(0)
-gaps <- clusGap(raw.filtered, kmeans, 14, B = 100)
-par(mar = c(4.5, 4.5, 4.5, 4.5), ps = 18)
-plot(gaps)
+# gaps <- clusGap(raw.filtered, kmeans, 14, B = 500)
+par(mar = c(4.3, 4.7, 0.5, 0.5), ps = 18)
+plot(gaps, xlab=expression(k), ylab=expression(Gap(k)), xaxt="n")
+# Embellishments
+bestK <- 4 # Figure this out with method Tibs2001SEmax
+gapk <- as.numeric(gaps$Tab[bestK, "gap"])
+gapk1.se <- as.numeric(gaps$Tab[bestK + 1, "gap"] - gaps$Tab[bestK + 1, "SE.sim"])
+segments(x0 = 4, y0 = 0, x1 = 4, y1 = gapk, lty = 2)
+segments(x0 = 0, y0 = gapk, x1 = 4, y1 = gapk, lty = 2)
+segments(x0 = 5, y0 = 0, x1 = 5, y1 = gapk1.se, lty = 2)
+segments(x0 = 0, y0 = gapk1.se, x1 = 5, y1 = gapk1.se, lty = 2)
+axis(1, at = c(1:14))
+# mtext(expression(paste(Gap(4), phantom("|"))), side = 2, at = c(0, gapk), las = 1)
+# text(x = 5, y = gapk1.se, labels = expression(paste(Gap(5) - se[5], phantom("|"))), pos = 4)
+
 if (SAVE.EXPLORE.PLOTS) {
-  dev.copy(pdf, "../figures/gap-statistic.pdf", width = 6, height = 5)
+  dev.copy(pdf, "../figures/gap-statistic.pdf", width = 8, height = 5)
   dev.off()
 }
-print(gaps, method = "firstSEmax", SE.factor = 1)
+print(gaps, method = "Tibs2001SEmax", SE.factor = 1)
 
 
 # Affinity propagation ====

### 2016-05-21 R/longitudinal.R

@@ -5,6 +5,7 @@ library(ggplot2)
 library(gridExtra)
 library(grid)
 library(gtable)
+library(RColorBrewer)
 library(reshape2)
 library(plyr)
 library(corrplot)
@@ -403,16 +404,18 @@ durat.cor.test <- function(symptom) cor.test(everything.wide$durat_pd, everythin
 correlations.everything <- sapply(names(everything.wide),
                                   durat.cor.everything)
 # Get rid of cluster, sex, durat_pd
-to.remove <- c("cluster", "sex", "durat_pd", "pdonset")
+to.remove <- c("cluster", "sex", "durat_pd")
 correlations.everything <- correlations.everything[!names(correlations.everything) %in% to.remove]
 names(correlations.everything) <- sapply(names(correlations.everything), function(v) c(NMS.D.MAP.PUB.N, NMS.NUM.TO.PUB, MISC.MAP)[[v]])
 correlations.everything <- sort(correlations.everything)  # Sort ascending
 # Pretty meaningless - no negative correlations!!
 is_d.e <- grepl("d", names(correlations.everything))
-is_d.e[which(is_d.e == TRUE)] <- "Domain"
-is_d.e[which(is_d.e == FALSE)] <- "Symptom"
-is_d.e[which(names(correlations.everything) %in% c("Axial", "Bradykinesia", "Rigidity", "Tremor"))] <- "Motor"
-is_d.e[which(names(correlations.everything) %in% c("CISI_Total", "Age", "PD_Onset"))] <- "Other"
+is_d.e[which(is_d.e == TRUE)] <- "Nonmotor symptom"
+is_d.e[which(is_d.e == FALSE)] <- "Nonmotor domain"
+is_d.e[which(names(correlations.everything) %in% c("Axial", "Bradykinesia", "Rigidity", "Tremor"))] <- "Motor symptom"
+is_d.e[which(names(correlations.everything) %in% c("CISI total", "Age", "PD onset"))] <- "Other"
+is_d.e <- factor(is_d.e, levels = c("Nonmotor domain", "Nonmotor symptom", "Motor symptom", "Other"))
+is_d.cols <- c(brewer.pal(4, "Set2")[1], brewer.pal(4, "Pastel2")[1], brewer.pal(4, "Set2")[c(3, 4)])
 
 correlations.df.e <- data.frame(
   names=names(correlations.everything),
[…]

### 2016-05-21 R/preprocessing.R

@@ -103,7 +103,7 @@ NMS.D <- c(
   "nms_d1", "nms_d2", "nms_d3", "nms_d4", "nms_d5", "nms_d6", "nms_d7", "nms_d8", "nms_d9"
 )
 NMS.D.NAMES <- c(
-  "Cardiovascular", "Sleep_fatigue", "Mood_cognition", "Perception_hallucination", "Attention_memory",
+  "Cardiovascular", "Sleep/fatigue", "Mood/apathy", "Perception/hallucination", "Attention/memory",
   "Gastrointestinal", "Urinary", "Sexual", "Miscellaneous"
 )
 NMS.NUM.TO.PUB <- setNames(sapply(NMS.30.NAMES, function(s) paste(strsplit(s, "-")[[1]][c(1, 3)], collapse = "-")),
@@ -117,14 +117,14 @@ NMS.30.LONG.SHORT.MAP <- setNames(NMS.30.NAMES.PUB, NMS.30.NAMES)
 MISC.MAP <- c(
   'age' = 'Age',
   'sex' = 'Sex',
-  'pdonset' = 'PD_Onset',
-  'durat_pd' = 'PD_Duration',
-  'cisitot' = 'CISI_Total',
+  'pdonset' = 'PD onset',
+  'durat_pd' = 'PD duration',
+  'cisitot' = 'CISI total',
   'tremor' = 'Tremor',
   'bradykin' = 'Bradykinesia',
   'rigidity' = 'Rigidity',
   'axial' = 'Axial',
-  'ldopa' = 'ldopa',
+  'ldopa' = 'L-Dopa',
   'surgery' = 'Surgery',
   'cluster' = 'Cluster'
 )

### 2016-05-23 R/preprocessing.R
Get info on excluded patients

@@ -154,7 +154,7 @@ splitdf <- function(dataframe, seed=NULL, trainfrac=0.7) {
 # IMPORT DATA, PREPROCESSING ====
 raw <- read.csv(DB_FILE)
 # Study is irrelevant
-raw$study <- NULL
+# raw$study <- NULL
 # Get rid of NAs (later may use some kind of missing value compensation method
 # This only brings it down to 904, rather than ~700 something!
 raw.omitted <- raw[, INTERPRETED]
@@ -164,7 +164,10 @@ raw.omitted.full <- cbind(
   # Now get the rownames from the raw
   raw[rownames(raw.omitted), -which(names(raw) %in% INTERPRETED)]
 )
-raw.omitted[rownames(raw.omitted), ]
+# To get identity of excluded patients: study/nident/country
+raw[!rownames(raw) %in% rownames(raw.omitted), c("study", "nident", "country")]
+# Verify they all have NAs in the things we care about
+all(apply(raw[!rownames(raw) %in% rownames(raw.omitted), INTERPRETED], MARGIN = 1, anyNA))
 
 # raw with only select nms and motor symptoms
 raw.filtered <- raw.omitted[, SYMPTOMS.TO.USE]

### 2016-05-23 writeup/manuscript/manuscript.tex
More minor tweaks to corrected manuscript

@@
 \usepackage{times}
+\usepackage{soul} % For highlighting
 
@@
 missing measurements in the variables of interest, resulting in 904 remaining patients. All
-symptoms were standardized before clustering, and unstandardized afterwards for interpretation.
+symptoms were standardized before clustering, and unstandardized afterwards for
+interpretation.
 Analyses were conducted in R version 3.2.4 (www.r-project.org).
@@
 domains and the four motor symptoms, and the second ``symptoms clustering'' on the 30 individual
-nonmotor symptoms collected with the four motor symptoms. Complete-linkage agglomerative
-hierarchical clustering on the 30 nonmotor symptoms was also performed to how the symptoms group
+nonmotor symptoms collected with the four motor symptoms. Complete-linkage
+hierarchical agglomerative clustering on the 30 nonmotor symptoms was also performed to how the symptoms group
 together.
@@
 optimal $k$ according to the Gap Statistic and the 1-standard-error method suggested by
-Tibshirani\cite{tibshirani01gap} was $k = 4$ (Supplementary Material Figure~\ref{fig:gap}). Other cluster validation
-methods (within sum of squares (WSS) scree plot, minimum average silhouette width) suggested $k =
+Tibshirani\cite{tibshirani01gap} was $k = 4$ (Supplementary Figure~\ref{fig:gap}). Other cluster validation
+methods (within sum of squares scree plot, minimum average silhouette width) suggested $k =
 2, 3, 4$, where $k = 2, 3$ simply divided the data uninformatively into groups with varying levels
@@
 To compare the clusterings, we visualized cluster alignment with stacked barplots, and computed the adjusted rand index\cite{hubert1985} (ARI) and
-Rosenberg's $v$-measure.\cite{rosenberg07vmeasure} to evaluate similarity between the two
+Rosenberg's $v$-measure\cite{rosenberg07vmeasure} to evaluate similarity between the two
 clusterings. Both measures range from 0 (no similarity) to 1 (identical), and respectively take a
@@
 $k$-means clustering performed on the 30 individual nonmotor symptoms found similar patterns to the
-first clustering (Table~\ref{tab:nms30}, heatmap in Figure~\ref{fig:nms30-hm}). Means of all
+first clustering {(Table~\ref{tab:nms30}, heatmap in Figure~\ref{fig:nms30-hm}). Means of all
 symptoms were found to differ across clusters except for PD onset and tremor.
@@
 clusters 2 and 3; for example, $D_3$ (motor-dominant) was somewhat
-evenly split among $S_1$ (mild) and $S_3$ (average).
+evenly split among $S_1$ (mild) and $S_3$ (average).}
 
@@
 
-Most variables had little to no correlation with PD duration (Supplementary Material Figure~\ref{fig:corr}). Variables
+Most variables had little to no correlation with PD duration (Supplementary Figure~\ref{fig:corr}). Variables
 with the highest correlation include symptoms urinary frequency, swallowing, and drooling; domains
 miscellaneous, urinary, and gastrointestinal; and total CISI score. Scatterplots for CISI Total,
-Tremor, Anxiety, and Depression appear in Supplementary Material Figure~\ref{fig:long4}.
+Tremor, Anxiety, and Depression appear in Supplementary Figure~\ref{fig:long4}.
 
@@
 Our longitudinal analysis gives more insight into the clusters found in the previous analyses.
-According to Figure~\ref{fig:corr}, most symptoms are uncorrelated with PD duration, especially
+According to Supplementary Figure~\ref{fig:corr}, most symptoms are uncorrelated with PD duration, especially
 Mood/apathy symptoms (anxiety, depression) and tremor.
@@
   \includegraphics[width=0.8\linewidth]{long4-d.pdf}
-  \caption{Selected symptoms for 901 patients plotted against PD duration. Smoothed loess curves for each cluster
+  % TODO: Get rid of capital T in  CISI Total
+  \caption{Selected symptoms plotted against PD duration. Smoothed loess curves for each cluster
   are drawn in their respective colors. The black curve is the curve for the entire population, and

### 2016-06-29 writeup/manuscript/manuscript.tex
Update...a lot of things

@@
 Tibshirani\cite{tibshirani01gap} was $k = 4$ (Supplementary Figure~\ref{fig:gap}). Other cluster validation
-methods (within sum of squares scree plot, minimum average silhouette width) suggested $k =
+methods (within sum of squares error scree plot, minimum average silhouette width) suggested $k =
 2, 3, 4$, where $k = 2, 3$ simply divided the data uninformatively into groups with varying levels
@@
     \toprule
-    & Cluster & 1 & 2 & 3 & 4 \\
-    & $n$ & 406 (45\%) & 189 (21\%) & 221 (24\%) & 88 (10\%) \\
+    	 Cluster 	 1 	 2 	 3 	 4 \\
+    	 $n$ 	 406 (45\%) 	 189 (21\%) 	 221 (24\%) 	 88 (10\%) \\
     \midrule
-    Nonmotor domains & Cardiovascular & 0.7 (1.6)\tnote{24} & 2.3 (2.9)\tnote{134} & 1.1 (2.0)\tnote{24} &
+    Nonmotor domains 	 Cardiovascular 	 0.7 (1.6)\tnote{24} 	 2.3 (2.9)\tnote{134} 	 1.1 (2.0)\tnote{24} 	
     7.0 (6.0)\tnote{123} \\
-    & Sleep/fatigue & 4.3 (4.8)\tnote{234} & 14.9 (8.2)\tnote{134} & 6.9 (6.3)\tnote{124}
-    & 21.0 (10.0)\tnote{123} \\
-  & Mood/apathy & 3.1 (4.5)\tnote{234} & 16.1 (14.0)\tnote{134} & 6.6
-    (8.1)\tnote{124} & 23.4 (14.3)\tnote{123} \\
-    & Perception/hallucination & 0.5 (1.7)\tnote{24} & 1.9 (3.5)\tnote{134} & 0.8
-    (2.0)\tnote{24} & 8.6 (6.9)\tnote{123} \\
-    & Attention/memory & 2.8 (4.2)\tnote{24} & 8.9 (8.1)\tnote{134} & 3.2 (4.4)\tnote{24} &
-    15.6 (11.1)\tnote{123} \\
-    & Gastrointestinal & 2.8 (3.9)\tnote{234} & 8.8 (7.1)\tnote{134} & 4.1
-    (4.7)\tnote{124} & 14.6 (9.6)\tnote{123} \\
-    & Urinary & 4.5 (5.9)\tnote{234} & 12.4 (9.7)\tnote{134} & 6.1 (6.4)\tnote{124} &
-    20.3 (9.9)\tnote{123} \\
-    & Sexual function & 1.5 (3.1)\tnote{24} & 6.4 (7.4)\tnote{134} & 2.5 (4.1)\tnote{24} &
-    9.0 (9.7)\tnote{123} \\
-    & Miscellaneous & 3.9 (4.7)\tnote{24} & 13.2 (8.6)\tnote{13} & 5.2 (5.8)\tnote{24} &
-    14.0 (9.5)\tnote{13} \\
+    a	 Sleep/fatigue 	 4.3 (4.8)\tnote{234} 	 14.9 (8.2)\tnote{134} 	 6.9 (6.3)\tnote{124} 21.0 (10.0)\tnote{123} \\
+  	 a  Mood/apathy 	 3.1 (4.5)\tnote{234} 	 16.1 (14.0)\tnote{134} 	 6.6 (8.1)\tnote{124} 	 23.4 (14.3)\tnote{123} \\
+    	 Perception/hallucination 	 0.5 (1.7)\tnote{24} 	 1.9 (3.5)\tnote{134} 	 0.8 (2.0)\tnote{24} 	 8.6 (6.9)\tnote{123} \\
+    	 Attention/memory 	 2.8 (4.2)\tnote{24} 	 8.9 (8.1)\tnote{134} 	 3.2 (4.4)\tnote{24} 	15.6 (11.1)\tnote{123} \\
+    	 Gastrointestinal 	 2.8 (3.9)\tnote{234} 	 8.8 (7.1)\tnote{134} 	 4.1 (4.7)\tnote{124} 	 14.6 (9.6)\tnote{123} \\
+    	 Urinary 	 4.5 (5.9)\tnote{234} 	 12.4 (9.7)\tnote{134} 	 6.1 (6.4)\tnote{124} 	20.3 (9.9)\tnote{123} \\
+    	 Sexual function 	 1.5 (3.1)\tnote{24} 	 6.4 (7.4)\tnote{134} 	 2.5 (4.1)\tnote{24} 	9.0 (9.7)\tnote{123} \\
+    	 Miscellaneous 	 3.9 (4.7)\tnote{24} 	 13.2 (8.6)\tnote{13} 	 5.2 (5.8)\tnote{24} 	14.0 (9.5)\tnote{13} \\
     \midrule
-    Motor symptoms & Tremor & 1.9 (1.8)\tnote{34} & 1.6 (1.9)\tnote{34} & 4.3 (2.8)\tnote{124} & 3.5
-    (3.8)\tnote{123} \\
-    & Bradykinesia & 1.5 (0.9)\tnote{234} & 2.1 (1.1)\tnote{134} & 3.6 (0.9)\tnote{124} &
-    4.0 (1.5)\tnote{123} \\
-    & Rigidity & 1.5 (0.9)\tnote{234} & 1.8 (1.1)\tnote{134} & 3.3 (0.9)\tnote{124} &
-    3.8 (1.4)\tnote{123} \\
-    & Axial & 1.8 (1.6)\tnote{234} & 3.3 (2.1)\tnote{134} & 4.3 (2.4)\tnote{124} &
-    7.4 (2.9)\tnote{123} \\
+    Motor symptoms 	 Tremor 	 1.9 (1.8)\tnote{34} 	 1.6 (1.9)\tnote{34} 	 4.3 (2.8)\tnote{124} 	 3.5 (3.8)\tnote{123} \\
+    	 Bradykinesia 	 1.5 (0.9)\tnote{234} 	 2.1 (1.1)\tnote{134} 	 3.6 (0.9)\tnote{124} 	4.0 (1.5)\tnote{123} \\
+    	 Rigidity 	 1.5 (0.9)\tnote{234} 	 1.8 (1.1)\tnote{134} 	 3.3 (0.9)\tnote{124} 	3.8 (1.4)\tnote{123} \\
+    	 Axial 	 1.8 (1.6)\tnote{234} 	 3.3 (2.1)\tnote{134} 	 4.3 (2.4)\tnote{124} 	7.4 (2.9)\tnote{123} \\
     \midrule
-    Variables not & Sex (\% male) & 63 & 53\tnote{3} & 71\tnote{2} & 58 \\
-    used in analysis & CISI total & 5.7 (3.2)\tnote{234} & 9.2 (3.9)\tnote{14} & 9.7
-    (3.5)\tnote{14} & 14.8 (5.1)\tnote{123} \\
-    & Age (34--89) & 62.7 (9.7)\tnote{34} & 64 (9.0)\tnote{4} & 65 (10.2)\tnote{14} & 70.5 (8.9)\tnote{123} \\
-    & PD onset & 56 (10.6) & 55.2 (10.1) & 56.9 (11.1) & 58.2 (11.4) \\
-    & PD duration & 6.7 (4.8)\tnote{234} & 8.7 (5.7)\tnote{14} & 8 (5.5)\tnote{14} &
-    12.3
-    (8.1)\tnote{123} \\
+    Variables not 	 Sex (\% male) 	 63 	 53\tnote{3} 	 71\tnote{2} 	 58 \\
+    used in analysis 	 CISI total 	 5.7 (3.2)\tnote{234} 	 9.2 (3.9)\tnote{14} 	 9.7 (3.5)\tnote{14} 	 14.8 (5.1)\tnote{123} \\
+    	 Age (34--89) 	 62.7 (9.7)\tnote{34} 	 64 (9.0)\tnote{4} 	 65 (10.2)\tnote{14} 	 70.5 (8.9)\tnote{123} \\
+    	 PD onset 	 56 (10.6) 	 55.2 (10.1) 	 56.9 (11.1) 	 58.2 (11.4) \\
+    	 PD duration 	 6.7 (4.8)\tnote{234} 	 8.7 (5.7)\tnote{14} 	 8 (5.5)\tnote{14} 	12.3 (8.1)\tnote{123} \\
     \bottomrule
@@
 
-Most variables had little to no correlation with PD duration (Supplementary Figure~\ref{fig:corr}). Variables
-with the highest correlation include symptoms urinary frequency, swallowing, and drooling; domains
-miscellaneous, urinary, and gastrointestinal; and total CISI score. Scatterplots for CISI Total,
-Tremor, Anxiety, and Depression appear in Supplementary Figure~\ref{fig:long4}.
+Most variables had little to no correlation with PD duration (Supplementary Figure~\ref{fig:corr}).
+Variables with the highest correlation included symptoms urinary frequency, swallowing, and
+drooling; domains miscellaneous, urinary, and gastrointestinal; and total CISI score. Scatterplots
+for CISI Total, Tremor, Anxiety, and Depression appear in Supplementary Figure~\ref{fig:long4}.
+Total CISI score increased predictably for all four clusters with increasing PD duration, but
+tremor, anxiety, and depression had no significant longitudinal progression; indeed, for cluster
+2, mood/apathy symptoms appeared to be highest at PD onset, decreasing as time progressed. Notably,
+for the motor/nonmotor-dominant groups, differences in severity for motor and nonmotor symptoms
+respectively was present early at PD onset, persisting throughout the range of PD durations
+observed.
 

### 2016-06-29 R/cleanup.R

@@ -5,11 +5,16 @@ library(xtable)
 library(reshape)
 library(grid)
 library(gridExtra)
-library(plyr) library(infotheo) library(ggplot2)
+library(plyr)
+library(infotheo)
+library(ggplot2)
+library(rtf) # For outputting to microsoft word
+library(gplots)
 
 library(reshape2)
 library(scales)
 
+# constants ====
 nmsd.symptoms <- c(
   NMS.D.NAMES,
   "Tremor",
@@ -22,9 +27,9 @@ nmsd.symptoms <- c(
 nmsd.extra <- c(
   'Age',
   'Sex',
-  'PD_Onset',
-  'PD_Duration',
-  'CISI_Total',
+  'PD_onset',
+  'PD_duration',
+  'CISI_PD_total',
   'ldopa',
   'Surgery',
   'Cluster'
@@ -63,31 +68,37 @@ v.measure <- function(a, b) {
 
 # ON NONMOTOR DOMAINS ====
 # Somewhat oddly, the correct clustering vector lies in trees$clusters4$clustering$cluster
-present <- reshape::rename(raw.omitted, PUB.MAP)
-present.full <- reshape::rename(raw.omitted.full, PUB.MAP)
-present$Cluster <- trees$clusters4$clustering$cluster
-present.full$Cluster <- trees$clusters4$clustering$cluster
+present <- reshape::rename(raw.omitted, gsub("/", "_", PUB.MAP))
+present.full <- reshape::rename(raw.omitted.full, gsub("/", "_", PUB.MAP))
+present$Cluster <- cl$cluster
+present.full$Cluster <- cl$cluster
 
 # Funcs for latex
 mean.sd <- function(data, sig = 2) {
   paste(round(mean(data), sig), " (", round(sd(data), sig), ")", sep = "")
 }
-to.latex <- function(df, file = NULL) {
+
+get.xtable <- function(df, file = NULL) {
   summary.t <- t(summaryBy(. ~ Cluster, df, FUN = function(x) mean.sd(x, sig = 1), # Only 1 decimal
                            keep.names = TRUE))
   # Get rid of "cluster" row
   summary.t <- summary.t[-which(rownames(summary.t) == 'Cluster'), ]
-  xt <- xtable(summary.t,
-               sanitize.colnames.function = function(x) gsub(pattern = '\\_', replacement = '/', x = x))
+  xtable(summary.t,
[…]

### 2016-06-29 R/kmeans-dtree.R

@@ -61,6 +61,9 @@ rename.clusters <- function(cl) {
 # NOTE: Doesn't work well, there isn't any elbow
 
 # Just in case it's still defined from below
+library(TeachingDemos)
+seed <- char2seed("Mu")
+set.seed(seed)
 var <- NULL
 wss <- (nrow(raw.filtered)-1)*sum(apply(raw.filtered, 2, var))
 for (i in 2:15) {
@@ -114,10 +117,10 @@ cat("number of clusters estimated by optimum average silhouette width:", pamk.be
 plot(pam(raw.filtered, pamk.best$nc))
 
 # GAP STATISTIC ESTIMATION ====
-set.seed(0)
-# gaps <- clusGap(raw.filtered, kmeans, 14, B = 500)
+set.seed(21) # Randomly chosen seed, for reproducibility
+gaps <- clusGap(raw.filtered, kmeans, 14, B = 500)
 par(mar = c(4.3, 4.7, 0.5, 0.5), ps = 18)
-plot(gaps, xlab=expression(k), ylab=expression(Gap(k)), xaxt="n")
+plot(gaps, xlab=expression(k), ylab=expression(Gap(k)))
 # Embellishments
 bestK <- 4 # Figure this out with method Tibs2001SEmax
 gapk <- as.numeric(gaps$Tab[bestK, "gap"])
@@ -161,16 +164,18 @@ splits$trainset = raw.filtered
 set.seed(0)  # Important to let clusters be the same
 cl <- kmeans(splits$trainset, 4, nstart = 25)
 # Added 20151020 - and rename here
+# Changed 20160613 - added scmmotcp, need to rename
+# 3 (sever)
 print (cl$cluster)
   for (i in 1:length(cl$cluster)) {
     curr.cluster <- cl$cluster[[i]]
 
-    if (curr.cluster == 1) { # nms
+    if (curr.cluster == 1) { # nonmotor
       cl$cluster[[i]] <- 2
-    } else if (curr.cluster == 2) { # all severe
-      cl$cluster[[i]] <- 4
-    } else if (curr.cluster == 3) { # motor
+    } else if (curr.cluster == 2) { # motor
       cl$cluster[[i]] <- 3
+    } else if (curr.cluster == 3) { # severe
+      cl$cluster[[i]] <- 4
     } else if (curr.cluster == 4) { # mild
       cl$cluster[[i]] <- 1
     } else {
[…]

### 2016-06-29 R/longitudinal.R

@@ -399,11 +399,13 @@ if (SAVE.LONG.PLOTS) {
 }
 
 # Durat cor correlation NO BINNING ====
+everything.wide <- cbind(raw.omitted, cluster = cl$cluster)
 durat.cor.everything <- function(symptom) cor(everything.wide$durat_pd, as.numeric(everything.wide[[symptom]]))
 durat.cor.test <- function(symptom) cor.test(everything.wide$durat_pd, everything.wide[[symptom]])
 correlations.everything <- sapply(names(everything.wide),
                                   durat.cor.everything)
 # Get rid of cluster, sex, durat_pd
+# If include PD_onset, then need to change up scale
 to.remove <- c("cluster", "sex", "durat_pd")
 correlations.everything <- correlations.everything[!names(correlations.everything) %in% to.remove]
 names(correlations.everything) <- sapply(names(correlations.everything), function(v) c(NMS.D.MAP.PUB.N, NMS.NUM.TO.PUB, MISC.MAP)[[v]])
@@ -412,10 +414,10 @@ correlations.everything <- sort(correlations.everything)  # Sort ascending
 is_d.e <- grepl("d", names(correlations.everything))
 is_d.e[which(is_d.e == TRUE)] <- "Nonmotor symptom"
 is_d.e[which(is_d.e == FALSE)] <- "Nonmotor domain"
-is_d.e[which(names(correlations.everything) %in% c("Axial", "Bradykinesia", "Rigidity", "Tremor"))] <- "Motor symptom"
-is_d.e[which(names(correlations.everything) %in% c("CISI total", "Age", "PD onset"))] <- "Other"
+is_d.e[which(names(correlations.everything) %in% MOTOR.PUB)] <- "Motor symptom"
+is_d.e[which(names(correlations.everything) %in% c("CISI_PD_total", "Age", "PD_onset"))] <- "Other"
 is_d.e <- factor(is_d.e, levels = c("Nonmotor domain", "Nonmotor symptom", "Motor symptom", "Other"))
-is_d.cols <- c(brewer.pal(4, "Set2")[1], brewer.pal(4, "Pastel2")[1], brewer.pal(4, "Set2")[c(3, 4)])
+is_d.cols <- c(brewer.pal(8, "Set2")[5], brewer.pal(8, "Pastel2")[5], brewer.pal(8, "Set2")[c(6, 7)])
 
 correlations.df.e <- data.frame(
   names=names(correlations.everything),
[…]

### 2016-06-29 R/nms30.R

@@ -42,7 +42,7 @@ nms30.s <- as.data.frame(scale(nms30))
 
 # With motor
 nms30m <- nms30
-nms30m[, MOTOR.SYMPTOMS] <- everything.wide[, MOTOR.SYMPTOMS]
+nms30m[, MOTOR.SYMPTOMS] <- raw.omitted[, MOTOR.SYMPTOMS]
 
 nms30m.scale <- scale(nms30m)
 nms30m.s <- as.data.frame(nms30m.scale)
@@ -102,8 +102,8 @@ rowD <- as.ggdend(hm$rowDendrogram)
 
 # motor
 # NOTE: apcluster interferes with this!!
-hm.m <- heatmap.2(as.matrix(nms30m.s),
-                  hclustfun = function(x) hclust(x, method='complete'),
+hm.m <- heatmap.2(as.matrix(nms30m.s[, 1:35]),
+                  hclustfun = function(x) hclust(x, method='average'),
                   trace='none')
 colD.m <- as.ggdend(hm.m$colDendrogram)
 
@@ -192,7 +192,50 @@ plot(d_clust.m)
 
 # library(cluster)
 # useless
-# clusGap(nms30.s, kmeans, 10, B = 100, verbose = interactive())
+set.seed(21)
+# set.seed(15)
+cg <- clusGap(nms30.s, kmeans, 14, B = 500)
+plot.new()
+print(cg, method = "Tibs2001SEmax")
+par(mar = c(4.3, 4.7, 0.5, 0.5), ps = 18)
+plot(cg, xlab=expression(k), ylab=expression(Gap(k)), xaxt = 'n')
+# Embellishments
+bestK <- 6 # Figure this out with method Tibs2001SEmax
+gapk <- as.numeric(cg$Tab[bestK, "gap"])
+gapk1.se <- as.numeric(cg$Tab[bestK + 1, "gap"] - cg$Tab[bestK + 1, "SE.sim"])
+segments(x0 = bestK, y0 = 0, x1 = bestK, y1 = gapk, lty = 2)
+segments(x0 = 0, y0 = gapk, x1 = bestK, y1 = gapk, lty = 2)
+segments(x0 = bestK + 1, y0 = 0, x1 = bestK + 1, y1 = gapk1.se, lty = 2)
+segments(x0 = 0, y0 = gapk1.se, x1 = bestK + 1, y1 = gapk1.se, lty = 2)
+axis(1, at = c(1:14))
+
[…]

### 2016-06-29 R/preprocessing.R

@@ -7,6 +7,7 @@ library(psych)
 # GLOBAL CONSTANTS ====
 SAVE.PREPROCESSING.PLOTS <- FALSE
 DB_FILE <- "../data/DATABASE_NMS Burden levels_15-4-2012.csv"
+MCP_FILE <- "../data/Database 951_Subtyping_30-5-2016.csv"
 ALL.SYMPTOMS <- c("nms_d1",
                  "nms_d2",
                  "nms_d3",
@@ -19,15 +20,18 @@ ALL.SYMPTOMS <- c("nms_d1",
                  "tremor",
                  "bradykin",
                  "rigidity",
-                 "axial")
+                 "axial",
+                 "scmmotcp")
 MOTOR.SYMPTOMS <- c("tremor",
                    "bradykin",
                    "rigidity",
-                   "axial")
+                   "axial",
+                   "scmmotcp")
 MOTOR.PUB <- c("Tremor",
                    "Bradykinesia",
                    "Rigidity",
-                   "Axial")
+                   "Axial",
+                  "Motor_comp")
 NMS.30.NAMES <- c(
   'd1-1-lightheaded',
   'd1-2-fainting',
@@ -108,7 +112,7 @@ NMS.D.NAMES <- c(
 )
 NMS.NUM.TO.PUB <- setNames(sapply(NMS.30.NAMES, function(s) paste(strsplit(s, "-")[[1]][c(1, 3)], collapse = "-")),
                            NMS.30)
-NMS.D.TO.NUM <- setNames(sapply(1:length(NMS.D), function(i) paste(i, "-", NMS.D[i]), sep = ""), NMS.D)
+NMS.D.TO.NUM <- setNames(sapply(1:length(NMS.D), function(i) paste(i, "-", NMS.D[i], sep = "")), NMS.D)
 NMS.D.MAP.PUB <- setNames(NMS.D.NAMES, NMS.D)
 NMS.D.MAP.PUB.N <- setNames(sapply(1:length(NMS.D),
                                    function(i) paste(i, NMS.D.NAMES[i], sep = "-")), NMS.D)
@@ -117,9 +121,12 @@ NMS.30.LONG.SHORT.MAP <- setNames(NMS.30.NAMES.PUB, NMS.30.NAMES)
 MISC.MAP <- c(
   'age' = 'Age',
   'sex' = 'Sex',
-  'pdonset' = 'PD onset',
[…]

### 2016-06-29 writeup/manuscript/include/help.py

@@ -1,0 +1,29 @@
+from collections import defaultdict
+NCLUS = 6
+
+
+def init():
+    init = [range(1, NCLUS + 1) for x in xrange(NCLUS)]
+    for index, val in enumerate(init):
+        val.pop(index)
+    return init
+
+
+with open('./diffs.txt', 'r') as fin:
+    lines = fin.readlines()
+
+for line in lines:
+    variable, rest = line.split(':')
+    others = [map(int, pair.split('-')) for pair in rest.split()]
+    nonsig = defaultdict(list)
+    for a, b in others:
+        nonsig[a].append(b)
+        nonsig[b].append(a)
+    lst = init()
+    for a, bs in nonsig.iteritems():
+        # Since 0 index
+        for b in bs:
+            lst[a - 1].remove(b)
+    print '{}\t{}'.format(
+        variable, '\t'.join([''.join(map(str, l)) for l in lst])
+    )

### 2017-02-28 writeup/manuscript/include/help.py
Remove .tex manuscript

@@ -1,29 +1,0 @@
-from collections import defaultdict
-NCLUS = 6
-
-
-def init():
-    init = [range(1, NCLUS + 1) for x in xrange(NCLUS)]
-    for index, val in enumerate(init):
-        val.pop(index)
-    return init
-
-
-with open('./diffs.txt', 'r') as fin:
-    lines = fin.readlines()
-
-for line in lines:
-    variable, rest = line.split(':')
-    others = [map(int, pair.split('-')) for pair in rest.split()]
-    nonsig = defaultdict(list)
-    for a, b in others:
-        nonsig[a].append(b)
-        nonsig[b].append(a)
-    lst = init()
-    for a, bs in nonsig.iteritems():
-        # Since 0 index
-        for b in bs:
-            lst[a - 1].remove(b)
-    print '{}\t{}'.format(
-        variable, '\t'.join([''.join(map(str, l)) for l in lst])
-    )

### 2017-02-28 narrative/abbreviated.tex
Make abbreviated narrative the actual narrative

--- 2016-03-03 writeup/abbreviated/abbreviated.tex
@@
 
-Section~\ref{sec:longitudinal} on longitudinal analysis; section~\ref{sec:nms30} on
+Section~\ref{sec:longitudinal} on longitudinal analysis; section~\ref{sec:nms30} on Extended
+Nonmotor Symptoms; Section~\ref{ssec:newconc} on new conclusions from these sections.
 
@@
 symptoms, and they form the broader nms\_d\{1-9\} symptoms. The results are in
-Figure~\ref{fig:d9to30}
+Figure~\ref{fig:d9to30}.
 
@@
 determining the optimal number of eigenvectors is in Figure~\ref{fig:nms30only-eigs}. Most metrics agree
-Nthat the optimal number of eigenvalues is around 6, but notice how one component has by far the
+that the optimal number of eigenvalues is around 6, but notice how one component has by far the
 highest eigenvalue; this component can be interpreted as a general ``nonmotor severity'' measure,
@@
 
-\begin{figure*}[p]
+Using one-way ANOVA for multiple means, only one variable had no significant interaction with the
+cluster ($p > 0.05$): pdonset.
+
+% Post-hoc analysis using Tukey's HSD is available in
+% Table~\ref{tab:tukeyhsd30}. For brevity, only insignificant differences are included.
+
+ibegin{figure*}[p]
   \centering
@@
 
-\begin{table*}[b]
-  \centering
-  \caption{Mean symptom scores for the centers of the $k$-means clustering where $k = 4$. In
-  this case, a Depression-Dominant subtype has been identified.}
+\begin{table*}[p]
+  \centering
+  \caption{Mean symptom scores for the centers of the $k$-means clustering on nms\{1-30\} where $k
+= 4$. In this case, a Depression-Dominant subtype has been identified.}
   \label{tab:nms30_k4}
@@
 
+% \begin{table*}[p]
+  % \centering
+  % \caption{Tukey's HSD for $k$-means clustering on nms\{1-30\}, $k = 4$. For brevity, only insignificant differences ($p > 0.05$) are included.}
+  % \label{tab:tukeyhsd30}
+  % \begin{tabular}{l|l|l}
+    % Variable & Cluster Relation & p \\
+    % \hline
+% d1-1-lightheaded & 3-2 & 0.572 \\
+% \hline
+% d1-2-fainting & 3-1 & 0.974 \\
+% \hline
+% d2-3-drowsiness & 3-2 & 0.814 \\
+% \hline
+% d2-4-fatigue & 4-2 & 0.213 \\
+% \hline
+% d2-5-insomnia & 4-2 & 0.143 \\
+% & 4-3 & 0.168 \\
+% \hline
+% d2-6-rls & 3-2 & 0.504 \\
+% \hline
+% d4-13-hallucination & 3-2 & 0.553 \\
+% \hline
+% d4-14-delusion & 3-1 & 0.977 \\
+% \hline
+% d4-15-diplopia & 3-2 & 0.0843 \\
+% \hline
+% d6-19-drooling & 3-2 & 1.00 \\
+% \hline
+% d6-20-swallowing & 3-2 & 0.167 \\
+% \hline
+% d6-21-constipation & 3-2 & 0.877 \\
+% \hline
+% d7-22-urinary\_urgency & 3-2 & 0.758 \\
+% \hline
+% d7-23-urinary\_frequency & 3-2 & 0.952 \\
+% \hline
+% d7-24-nocturia & 3-2  & 0.983 \\
+% \hline
+% d8-25-sex\_drive & 4-2 & 0.575 \\
+% \hline
+% d8-26-sex\_dysfunction & 3-2 & 0.147 \\
+% & 4-2 & 0.0635 \\
+% \hline
+% d9-27-unexplained\_pain & 3-2 & 0.513 \\
+% \hline
+% d9-28-gust\_olfact & 3-2 & 0.794 \\
+% & 4-2 & 0.220 \\
+% \hline
+% d9-29-weight\_change & 3-2 & 0.105 \\
+% \hline
+% d9-30-sweating & 3-2 & 0.359 \\
+% & 4-2 & 0.215 \\
+% & 4-3 & 0.797 \\
+% \hline
+% age & 2-1 & 0.352 \\
+% & 3-2 & 0.570 \\
+% \hline
+% sex & 3-1 & 0.443 \\
+% & 4-1 & 0.991 \\
+% & 3-2 & 0.373 \\
+% & 4-2 & 0.436 \\
+% & 4-3 & 0.969 \\
+% \hline
+% durat\_pd & 2-1 & 0.067 \\
+ % & 3-2 & 0.436 \\
+% \hline
+% cisitot & 3-2 & 0.428 \\
+% \hline
+% tremor & 2-1 & 0.484 \\
+% & 3-1 & 0.990 \\
+% & 3-2 & 0.429 \\
+% \hline
+% bradykin & 3-2 & 0.972 \\
+% \hline
+% rigidity & 3-2 & 0.610  \\
+% \hline
+% axial & 3-2 & 0.912
+  % \end{tabular}
+% \end{table*}
+
 \subsubsection{Model-based expectation-maximization}
@@
 I created decision trees for the $k$-means $k = 4$ clustering, located in
-Figure~\ref{fig:nms30m-dtree-all}
+Figure~\ref{fig:nms30m-dtree-all}.
 
@@
   \includegraphics[width=\linewidth]{nms30m-dtree-all.pdf}
-  \caption{Pruned decision tree for one vs all, clustered on nms\{1-30\}}
+  \caption{Pruned decision tree, clustered on nms\{1-30\}}
   \label{fig:nms30m-dtree-all}
@@
 
+One vs all decision trees are located in
+Figures~\ref{fig:nms30m-dtree-1va}, \ref{fig:nms30m-dtree-2va}, \ref{fig:nms30m-dtree-3va},
+and \ref{fig:nms30m-dtree-4va}.
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=0.6\linewidth]{nms30m-dtree-1va.pdf}
+  \caption{Pruned 1 vs all decision tree.}
+  \label{fig:nms30m-dtree-1va}
+\end{figure*}
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=0.6\linewidth]{nms30m-dtree-2va.pdf}
+  \caption{Unpruned 2 vs all decision tree. (Pruned was too simple).}
+  \label{fig:nms30m-dtree-2va}
+\end{figure*}
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=\linewidth]{nms30m-dtree-3va.pdf}
+  \caption{Pruned 3 vs all decision tree.}
+  \label{fig:nms30m-dtree-3va}
+\end{figure*}
+
+\begin{figure*}[p]
+  \centering
+  \includegraphics[width=0.6\linewidth]{nms30m-dtree-4va.pdf}
+  \caption{Unpruned 4 vs all decision tree. (Pruned was too simple).}
+  \label{fig:nms30m-dtree-4va}
+\end{figure*}
 \section{Preliminary Conclusions}
@@
 
+\subsection{New conclusions}
+\label{ssec:newconc}
+
+First, the longitudinal analysis gives more insight into the clusters found when clustering on
+nms\_d\{1-9\}. According to Figure~\ref{fig:longcorr}, most symptoms are highly correlated, with PD
+duration, but notably, mood/cognition symptoms (nms9, nms10, nms12) and tremor are not correlated
+highly with PD duration. The differences in disease progression can be seen by the corresponding
+graphs, Figures~\ref{fig:nms9-multi} and~\ref{fig:nms10-multi}. In both graphs, what is interesting
+is that Subtype 2 (Nonmotor-Dominant) starts at higher scores for nms\_9 and nms\_10, thus
+indicating that these patients' subtype is can be determined early in PD duration from depressive
+symptom score. Similarly, when examining Subtype 3 (Motor-dominant) in
+Figure~\ref{fig:tremor-multi}, the mean tremor score is substantially higher from PD onset.
+Interestingly, Subtype 4 (Severe) generally starts at lower tremor and motor scores during disease
+onset (Figure~\ref{fig:cisitot-multi}), but then rises sharply, exceeding other Subtypes. More evidence
+that tremor is a unique motor symptom is located in Figure~\ref{fig:nms30m-colhc}, where it is the most
+distant symptom from all other symptoms.
+
+When examining all 30 symptoms, more evidence is given that the previously-discovered Subtype
+2 (Nonmotor-Dominant) may be primarily characterized by high depressive symptoms. By graphing the
+30 subtypes attached to the original clustering, as in Figure~\ref{fig:d9to30}, the mean scores of
+nms8, nms9, and nms10 for Subtype 2 are substantially higher than Subtypes 1 and 3.
+
+Indeed, PCA on the 30 nonmotor symptoms identifies the second-most prominent component as a general
+mood/cognition component, and $k$-means clustering on the 30 symptoms only
+(Figure~\ref{fig:nms30m-k4-heatmap-improved}) divides the 1000 patients into four slightly
+different groups, a mild, average, depression-dominant, and severe group.
+
+The Gaussian mixture model identified in Figure~\ref{fig:nms30m-mclust-heatmap-improved} fragments
+the previously-discovered clusters into more groups. Here, a wide variety of specialized subtypes
+of PD are displayed, including insomnia, urinary, motor, nonmotor, and depression-dominant groups,
+as well as the expected mild, average, and severe subtypes. It is likely that the previous analysis
+with only nms\_d\{1-9\} combined most of those specialized groups into the more general
+nonmotor-dominant Subtype 2.
+
+It's intuitive that a Depression-Dominant group emerges when clustering on nms\{1-30\}, since
+domain 3 consists of 5 separate measures. Thus, any high expression of depressive symptoms is
+magnified in clustering, since the symptoms are highly similar (Figure~\ref{fig:nms30m-colhc}) and
+treated with equal weight. Once again reinforcing what was discovered previously, depressive
+symptoms have been shown to be very important in determining subtypes of PD.
+
 \bibliographystyle{Science}
